---
title: "Weighted GRIP ND Parity Diagnostics With gflow Fixtures"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S %Z')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 2
params:
  mode: "smoke"
  output_dir: "/Users/pgajer/current_projects/grip/output/weighted-grip-nd-parity"
  gflow_repo: "/Users/pgajer/current_projects/gflow"
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10.5,
  fig.height = 6.2,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = "ragg_png"
)

library(ggplot2)
library(knitr)

repo_dir <- "/Users/pgajer/current_projects/grip"
source(file.path(repo_dir, "tests/testthat/helper-weighted-nd-gflow-parity.R"))
source(file.path(repo_dir, "tests/testthat/helper-weighted-nd-trace-parity.R"))

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(repo_dir, quiet = TRUE)
} else {
  library(grip)
}

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
metrics <- grip_weighted_nd_gflow_parity_run(
  mode = params$mode,
  dims = c(2L, 3L),
  gflow_repo = params$gflow_repo
)
trace_metrics <- grip_weighted_nd_trace_parity_run(dims = c(2L, 3L))
trace_first <- grip_weighted_nd_trace_parity_first_divergence(trace_metrics)
gflow_trace_metrics <- grip_weighted_nd_gflow_trace_run(
  mode = params$mode,
  dims = "intrinsic",
  gflow_repo = params$gflow_repo
)
gflow_trace_first <- grip_weighted_nd_gflow_trace_first_divergence(gflow_trace_metrics)
gflow_trace_all_metrics <- grip_weighted_nd_gflow_trace_run(
  mode = params$mode,
  dims = "all",
  gflow_repo = params$gflow_repo
)
gflow_trace_all_first <- grip_weighted_nd_gflow_trace_first_divergence(gflow_trace_all_metrics)
gflow_trace_cross_metrics <- gflow_trace_all_metrics[
  gflow_trace_all_metrics$layout_dim != gflow_trace_all_metrics$intrinsic_dim,
  ,
  drop = FALSE
]
gflow_trace_cross_first <- gflow_trace_all_first[
  gflow_trace_all_first$case_id %in% unique(gflow_trace_cross_metrics$case_id) &
    mapply(
      function(case_id, layout_dim) {
        any(
          gflow_trace_cross_metrics$case_id == case_id &
            gflow_trace_cross_metrics$layout_dim == layout_dim
        )
      },
      gflow_trace_all_first$case_id,
      gflow_trace_all_first$layout_dim
    ),
  ,
  drop = FALSE
]
refinement_step_metrics <- grip_weighted_nd_gflow_refinement_step_trace_run(
  mode = params$mode,
  gflow_repo = params$gflow_repo
)
refinement_step_target <- attr(refinement_step_metrics, "target")
refinement_step_term_metrics <- attr(refinement_step_metrics, "attraction_terms")
refinement_step_summary <- grip_weighted_nd_gflow_refinement_step_summary(
  refinement_step_metrics
)
refinement_step_term_summary <- grip_weighted_nd_gflow_refinement_step_term_summary(
  refinement_step_metrics
)
refinement_source <- grip_weighted_nd_gflow_refinement_source_run(
  mode = params$mode,
  gflow_repo = params$gflow_repo
)
metrics$layout_dim_label <- paste0(metrics$layout_dim, "D layout")
metrics$case_label <- paste(metrics$family, metrics$intrinsic_dim, "D", sep = " / ")
metrics$direct_rmse_plot <- pmax(metrics$direct_rmse, 1e-15)
metrics$procrustes_rmse_plot <- pmax(metrics$procrustes_rmse, 1e-15)
trace_metrics$frame_label <- paste0(
  trace_metrics$fixture_id,
  " / ",
  trace_metrics$dim,
  "D"
)
trace_metrics$procrustes_rmse_plot <- pmax(trace_metrics$procrustes_rmse, 1e-15)
gflow_trace_metrics$frame_label <- paste0(
  gflow_trace_metrics$family,
  " / ",
  gflow_trace_metrics$layout_dim,
  "D"
)
gflow_trace_metrics$procrustes_rmse_plot <- pmax(gflow_trace_metrics$procrustes_rmse, 1e-15)
gflow_trace_cross_metrics$frame_label <- paste0(
  gflow_trace_cross_metrics$family,
  " / ",
  gflow_trace_cross_metrics$intrinsic_dim,
  "D fixture in ",
  gflow_trace_cross_metrics$layout_dim,
  "D"
)
gflow_trace_cross_metrics$procrustes_rmse_plot <- pmax(
  gflow_trace_cross_metrics$procrustes_rmse,
  1e-15
)
thresholds <- grip_weighted_nd_gflow_parity_thresholds()

csv_path <- file.path(params$output_dir, "weighted_grip_nd_gflow_parity_metrics.csv")
rds_path <- file.path(params$output_dir, "weighted_grip_nd_gflow_parity_metrics.rds")
trace_csv_path <- file.path(params$output_dir, "weighted_grip_nd_trace_parity_metrics.csv")
trace_rds_path <- file.path(params$output_dir, "weighted_grip_nd_trace_parity_metrics.rds")
gflow_trace_csv_path <- file.path(params$output_dir, "weighted_grip_nd_gflow_trace_parity_metrics.csv")
gflow_trace_rds_path <- file.path(params$output_dir, "weighted_grip_nd_gflow_trace_parity_metrics.rds")
gflow_trace_all_csv_path <- file.path(params$output_dir, "weighted_grip_nd_gflow_trace_all_parity_metrics.csv")
gflow_trace_all_rds_path <- file.path(params$output_dir, "weighted_grip_nd_gflow_trace_all_parity_metrics.rds")
refinement_step_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_step_trace_metrics.csv")
refinement_step_summary_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_step_trace_summary.csv")
refinement_step_rds_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_step_trace_metrics.rds")
refinement_step_term_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_step_attraction_terms.csv")
refinement_step_term_summary_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_step_attraction_term_summary.csv")
refinement_source_target_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_source_target.csv")
refinement_source_summary_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_source_summary.csv")
refinement_source_vertex_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_source_vertex_trace.csv")
refinement_source_term_csv_path <- file.path(params$output_dir, "weighted_grip_nd_refinement_source_term_trace.csv")
write.csv(metrics, csv_path, row.names = FALSE)
saveRDS(metrics, rds_path)
write.csv(trace_metrics, trace_csv_path, row.names = FALSE)
saveRDS(trace_metrics, trace_rds_path)
write.csv(gflow_trace_metrics, gflow_trace_csv_path, row.names = FALSE)
saveRDS(gflow_trace_metrics, gflow_trace_rds_path)
write.csv(gflow_trace_all_metrics, gflow_trace_all_csv_path, row.names = FALSE)
saveRDS(gflow_trace_all_metrics, gflow_trace_all_rds_path)
write.csv(refinement_step_metrics, refinement_step_csv_path, row.names = FALSE)
write.csv(refinement_step_summary, refinement_step_summary_csv_path, row.names = FALSE)
saveRDS(refinement_step_metrics, refinement_step_rds_path)
write.csv(refinement_step_term_metrics, refinement_step_term_csv_path, row.names = FALSE)
write.csv(refinement_step_term_summary, refinement_step_term_summary_csv_path, row.names = FALSE)
write.csv(refinement_source$target, refinement_source_target_csv_path, row.names = FALSE)
write.csv(refinement_source$summary, refinement_source_summary_csv_path, row.names = FALSE)
write.csv(refinement_source$vertex_trace, refinement_source_vertex_csv_path, row.names = FALSE)
write.csv(refinement_source$term_trace, refinement_source_term_csv_path, row.names = FALSE)

run_metadata <- data.frame(
  field = c(
    "grip version",
    "gflow version",
    "mode",
    "gflow repository",
    "metrics CSV",
    "metrics RDS",
    "trace metrics CSV",
    "trace metrics RDS",
    "gflow trace metrics CSV",
    "gflow trace metrics RDS",
    "gflow all-dimension trace metrics CSV",
    "gflow all-dimension trace metrics RDS",
    "refinement-step trace metrics CSV",
    "refinement-step trace summary CSV",
    "refinement-step trace metrics RDS",
    "refinement-step attraction terms CSV",
    "refinement-step attraction term summary CSV",
    "refinement source target CSV",
    "refinement source summary CSV",
    "refinement source vertex trace CSV",
    "refinement source term trace CSV"
  ),
  value = c(
    as.character(utils::packageVersion("grip")),
    as.character(utils::packageVersion("gflow")),
    params$mode,
    normalizePath(params$gflow_repo, mustWork = FALSE),
    normalizePath(csv_path, mustWork = FALSE),
    normalizePath(rds_path, mustWork = FALSE),
    normalizePath(trace_csv_path, mustWork = FALSE),
    normalizePath(trace_rds_path, mustWork = FALSE),
    normalizePath(gflow_trace_csv_path, mustWork = FALSE),
    normalizePath(gflow_trace_rds_path, mustWork = FALSE),
    normalizePath(gflow_trace_all_csv_path, mustWork = FALSE),
    normalizePath(gflow_trace_all_rds_path, mustWork = FALSE),
    normalizePath(refinement_step_csv_path, mustWork = FALSE),
    normalizePath(refinement_step_summary_csv_path, mustWork = FALSE),
    normalizePath(refinement_step_rds_path, mustWork = FALSE),
    normalizePath(refinement_step_term_csv_path, mustWork = FALSE),
    normalizePath(refinement_step_term_summary_csv_path, mustWork = FALSE),
    normalizePath(refinement_source_target_csv_path, mustWork = FALSE),
    normalizePath(refinement_source_summary_csv_path, mustWork = FALSE),
    normalizePath(refinement_source_vertex_csv_path, mustWork = FALSE),
    normalizePath(refinement_source_term_csv_path, mustWork = FALSE)
  ),
  stringsAsFactors = FALSE
)
```

## Purpose

This report evaluates whether the opt-in, side-by-side
`grip.layout.weighted.nd()` backend currently reproduces the legacy
`grip.layout.weighted()` embedding in the two
dimensions where the legacy implementation is defined. The comparison uses
`gflow`-generated fixtures rather than reimplementing their geometry inside
`grip`: flat and nonuniform 2D grids, flat and nonuniform 3D grids, and 2D/3D
quadform graphs whose edge lengths are produced by `gflow` utilities.

The central question is strict parity: when the graph, weights, dimension,
seed, and shared tuning parameters are held fixed, does the ND entry point
return the same embedding as the legacy weighted-GRIP backend in 2D and 3D?
This is a release gate for the ND API: higher-dimensional support should not
come at the cost of changing the authoritative 2D/3D weighted-GRIP behavior.
The current ND implementation is intentionally evaluated on its own path; it
does not route `dim = 2` or `dim = 3` through the legacy core.

## Trace-Level Sanity Check

The smallest connected fixture is used first as a phase-level smoke test. With
the same 3-by-3 mesh, weights, seed, rounds, final rounds, and `num_init`, the
legacy weighted-GRIP trace initializes only its coarse MISF level and then emits
a `level_start` transition when the full graph is inserted. The ND backend now
emits the same active-level skeleton from its side-by-side weighted MISF
builder and legacy-style refinement loop.

```{r trace-sanity}
trace_edges <- edges.mesh(3, 3)
trace_weights <- rep(1, nrow(trace_edges))
legacy_trace <- grip.layout.trace.weighted(
  edges = trace_edges,
  edge_weights = trace_weights,
  n = 9,
  dim = 2,
  placement = "barycenter",
  rounds = 3,
  final_rounds = 2,
  num_init = 5,
  num_nbrs = 6,
  r = 0.03,
  s = 6.0,
  repulsion_factor = 1.5,
  coarse_repulsion_factor = 1.5,
  coarse_repulsion_sample = 100000L,
  coarse_repulsion_exact_below = 100000L,
  tinit_factor = 6,
  trace = "round",
  trace.every = 1,
  seed = 81
)
nd_trace_fn <- get("grip.layout.weighted.nd.trace", asNamespace("grip"))
nd_trace <- nd_trace_fn(
  edges = trace_edges,
  edge_weights = trace_weights,
  n = 9,
  dim = 2,
  rounds = 3,
  final_rounds = 2,
  num_init = 5,
  num_nbrs = 6,
  trace.every = 1,
  seed = 81
)
trace_summary <- rbind(
  data.frame(
    backend = "legacy weighted-GRIP",
    frames = length(legacy_trace$frames),
    phases = paste(legacy_trace$meta$phase, collapse = ", "),
    finite_rows_at_init = sum(stats::complete.cases(legacy_trace$frames[[1]])),
    stringsAsFactors = FALSE
  ),
  data.frame(
    backend = "current ND backend",
    frames = length(nd_trace$frames),
    phases = paste(nd_trace$meta$phase, collapse = ", "),
    finite_rows_at_init = sum(stats::complete.cases(nd_trace$frames[[1]])),
    stringsAsFactors = FALSE
  )
)
kable(trace_summary)
```

This trace does not prove final layout quality by itself. Its value is narrower
and more useful: it verifies that the ND backend and the legacy weighted backend
are now comparable frame by frame. The next diagnostic uses those matching
frames to locate the first coordinate-level divergence.

## Trace Parity Ratchet

The following diagnostic compares each traced frame after centering,
orthogonal Procrustes alignment, and scaling. It is intentionally small: a
3-by-3 mesh and an 8-cycle with nonuniform edge lengths, each in 2D and 3D.
The purpose is not to certify all scientific fixtures; it is to ratchet the
port one trace frame at a time and identify the first phase where the ND
trajectory stops matching the legacy weighted trajectory.

```{r trace-first-divergence-table}
kable(trace_first, digits = 6)
```

```{r trace-first-divergence-plot, fig.height=6.2}
ggplot(trace_metrics, aes(x = frame, y = procrustes_rmse_plot, color = phase)) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  geom_point(size = 2.1) +
  facet_wrap(~ frame_label, scales = "free_x", ncol = 2) +
  scale_y_log10() +
  labs(
    x = "Trace frame",
    y = "Frame Procrustes RMSE, log scale",
    color = "Phase",
    title = "First divergence is now localized to traced phases"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Frames whose Procrustes RMSE is near numerical noise indicate that the ported
mechanics are effectively matching at that point. In the current run, all
frames in these tiny fixtures are within the strict trace tolerance and the
metadata matches across all frames. This does not prove full gflow-fixture
shape parity, but it does mean the first coarse-to-fine schedule, initialization,
insertion, local-temperature, and refinement ratchet is now passing on the
small diagnostic fixtures.

## Canonical gflow Trace Ratchet

The same frame-by-frame diagnostic is also run on the smallest canonical
gflow-backed grid and lattice fixtures: flat 2D grid, nonuniform 2D grid, flat
3D lattice, and nonuniform 3D grid. These runs use each fixture's intrinsic
dimension and the smoke-report tuning, so they are larger than the tiny
mesh/cycle ratchet but still small enough to inspect frame by frame.

```{r gflow-trace-first-divergence-table}
kable(gflow_trace_first, digits = 6)
```

```{r gflow-trace-ratchet-plot, fig.height=7.2}
ggplot(gflow_trace_metrics, aes(x = frame, y = procrustes_rmse_plot, color = phase)) +
  geom_line(linewidth = 0.55, alpha = 0.75) +
  geom_point(size = 1.8) +
  facet_wrap(~ frame_label, scales = "free_x", ncol = 2) +
  scale_y_log10() +
  labs(
    x = "Trace frame",
    y = "Frame Procrustes RMSE, log scale",
    color = "Phase",
    title = "Canonical gflow trace ratchet stays within tolerance"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

In the current smoke run, the canonical grid and lattice trace ratchet remains
within tolerance for intrinsic 2D/3D embeddings. This narrows the remaining
smoke-report disagreement to broader final-layout cases, especially embeddings
requested in a dimension different from the fixture's intrinsic dimension.

## Cross-Dimensional gflow Trace Ratchet

The intrinsic-dimensional trace ratchet answers whether the ND path can follow
legacy weighted-GRIP on each fixture's natural dimension. The larger remaining
residual in the smoke report is different: it occurs when a graph generated
from a 3D fixture is embedded in 2D. The diagnostic below reruns the same
canonical gflow grid and lattice fixtures in both supported legacy dimensions,
then isolates the cross-dimensional requests.

```{r gflow-cross-trace-first-divergence-table}
kable(gflow_trace_cross_first, digits = 6)
```

```{r gflow-cross-trace-ratchet-plot, fig.height=6.6}
ggplot(gflow_trace_cross_metrics, aes(x = frame, y = procrustes_rmse_plot, color = phase)) +
  geom_line(linewidth = 0.55, alpha = 0.75) +
  geom_point(size = 1.8) +
  facet_wrap(~ frame_label, scales = "free_x", ncol = 2) +
  scale_y_log10() +
  labs(
    x = "Trace frame",
    y = "Frame Procrustes RMSE, log scale",
    color = "Phase",
    title = "Cross-dimensional traces localize the remaining smoke residual"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The current run localizes the first strict-threshold failure to the nonuniform
3D grid requested in 2D. Metadata still matches, so this is not a schedule or
active-set mismatch. The first measured break occurs in the full-graph
refinement phase after insertion has completed, which makes the next debugging
target much narrower than the original package-wide parity problem.

## Refinement-Step Trace

The step-level trace drills into the first failing cross-dimensional case. It
captures each active vertex for the two rounds before the first divergent
frame and the divergent round itself. For each backend it records coordinates
before the round, the pre-temperature displacement, local temperature and
cosine state before and after the temperature update, the displacement actually
applied, and coordinates after application.

```{r refinement-step-target-table}
kable(refinement_step_target, digits = 6)
```

```{r refinement-step-summary-table}
kable(head(refinement_step_summary, 16L), digits = 6)
```

```{r refinement-step-summary-plot, fig.height=5.8}
plot_summary <- head(refinement_step_summary, 16L)
plot_summary$metric <- factor(plot_summary$metric, levels = rev(plot_summary$metric))
ggplot(plot_summary, aes(x = max_abs, y = metric)) +
  geom_col(width = 0.68, fill = "#356D8C") +
  scale_x_log10() +
  labs(
    x = "Maximum absolute legacy-ND difference, log scale",
    y = NULL,
    title = "First failing trace is dominated by pre-temperature displacement"
  ) +
  theme_minimal(base_size = 11)
```

The largest discrepancy in the current run is already present in the
pre-temperature displacement at round 11. The applied displacement and
coordinate-after differences follow from that earlier force-level difference.
The component split shows that the largest contribution to that row-level
difference is the weighted edge-attraction term, while the edge and
metric-neighbor provenance strings still match. This points the next
implementation pass toward the arithmetic used inside full-graph weighted-FR
force accumulation, rather than toward input ordering, local-temperature update,
or coordinate application order.

```{r refinement-step-attraction-term-summary-table}
kable(head(refinement_step_term_summary, 16L), digits = 6)
```

The attraction-term table expands the weighted-attraction component one
adjacent edge at a time. For each term it records the adjacent vertex, edge
weight, coordinate delta, squared norm, desired weighted length, scale after the
legacy-style float cast, per-edge step vector, and cumulative attraction vector.
This is the diagnostic to use for the next code fix: if a term's delta already
differs, the coordinate entering that edge contribution is the issue; if delta
matches but norm or scale differs, the arithmetic sequence or cast is the
issue; if individual steps match but cumulative attraction differs, the
accumulation order is the issue.

## Refinement Source Trace

The source trace reruns the same failing level from round 1 through the first
divergent round and follows the two vertices involved in the largest
attraction-term discrepancy. In the current smoke run this is the edge from
vertex 34 to vertex 9.

```{r refinement-source-target-table}
kable(refinement_source$target, digits = 6)
```

```{r refinement-source-summary-table}
kable(refinement_source$summary, digits = 6)
```

```{r refinement-source-vertex-table}
kable(refinement_source$vertex_trace, digits = 6)
```

```{r refinement-source-term-table}
kable(refinement_source$term_trace, digits = 6)
```

The source trace shows tiny coordinate and force differences from the first
full-graph round, with a clear escalation at rounds 7-10. At round 11, the
coordinate delta entering the edge 34-to-9 attraction term is large enough that
the legacy-style float-cast scale differs, producing the much larger per-edge
attraction-step gap. That makes the next implementation target earlier than
round 11: the arithmetic that creates the growing coordinate drift through
rounds 7-10.

## Coordinate Parity

The direct coordinate diagnostic is the root-mean-square difference between the
raw legacy and ND coordinate matrices,

\[
\operatorname{RMSE}_{direct} =
\sqrt{\frac{1}{nd}\sum_{i=1}^{n}\sum_{j=1}^{d}(X_{ij}^{legacy}-X_{ij}^{ND})^2}.
\]

If the ND entry point recovers the legacy path with the same initialization and
update order, this quantity should be near machine precision. Large values
would indicate that the two algorithms are following different trajectories
before considering translation, rotation, reflection, or scale.

```{r direct-coordinate-parity}
ggplot(metrics, aes(x = reorder(case_id, direct_rmse_plot), y = direct_rmse_plot, fill = layout_dim_label)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  coord_flip() +
  scale_y_log10() +
  labs(
    x = NULL,
    y = "Direct coordinate RMSE, log scale",
    fill = NULL,
    title = "Raw legacy and ND coordinates are not numerically identical"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Large direct coordinate differences are expected for independent graph-layout
trajectories, but they are still useful: if the ND implementation eventually
ports the legacy initialization and update order phase by phase, this diagnostic
should fall toward numerical noise before alignment is even considered. The
figure uses a small plotting floor only to make near-zero values visible on the
log scale.

## Shape Parity After Alignment

Some layout differences are geometrically irrelevant. To separate coordinate
frame changes from shape changes, the ND layout is centered, optimally rotated
or reflected, and scaled to the legacy layout. The plotted quantity is

\[
\operatorname{RMSE}_{Proc} =
\sqrt{\frac{1}{nd}\|X_c^{legacy} - sX_c^{ND}R\|_F^2}.
\]

This diagnostic is stricter than a visual comparison but more meaningful than
raw coordinate equality for graph layouts.

```{r procrustes-shape-parity}
ggplot(metrics, aes(x = reorder(case_id, procrustes_rmse_plot), y = procrustes_rmse_plot, fill = layout_dim_label)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  coord_flip() +
  scale_y_log10() +
  labs(
    x = NULL,
    y = "Procrustes-aligned RMSE, log scale",
    fill = NULL,
    title = "Alignment reduces frame effects but does not yet recover parity"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

After alignment, the residuals measure shape disagreement rather than arbitrary
translation, rotation, reflection, or scale. Non-negligible residuals indicate
that the current ND backend remains algorithmically different from the legacy
weighted-GRIP engine. This is the diagnostic the ND port must drive down before
the function should be treated as release-ready for weighted-GRIP parity.

## Weighted Edge-Length Fidelity

Parity with the legacy implementation is distinct from weighted-layout
quality. This diagnostic compares the normalized layout edge lengths with the
normalized input weights,

\[
E =
\sqrt{\frac{1}{m}\sum_{(i,j)\in E}
\left(
\frac{\|x_i-x_j\|}{\operatorname{median}(\|x_i-x_j\|)}
-
\frac{w_{ij}}{\operatorname{median}(w)}
\right)^2}.
\]

Lower values indicate that the embedding respects the supplied local edge
lengths more closely.

```{r edge-fidelity, fig.height=7.2}
edge_long <- rbind(
  data.frame(
    case_id = metrics$case_id,
    layout_dim_label = metrics$layout_dim_label,
    method = "legacy weighted",
    edge_rmse = metrics$legacy_edge_rmse
  ),
  data.frame(
    case_id = metrics$case_id,
    layout_dim_label = metrics$layout_dim_label,
    method = "ND weighted",
    edge_rmse = metrics$nd_edge_rmse
  )
)
edge_long$case_id <- factor(edge_long$case_id, levels = rev(unique(metrics$case_id)))

ggplot(edge_long, aes(x = edge_rmse, y = case_id, color = method, shape = method)) +
  geom_point(size = 3.0, alpha = 0.9) +
  facet_wrap(~ layout_dim_label, ncol = 1) +
  scale_x_log10() +
  labs(
    x = "Normalized edge-length RMSE, log scale",
    y = NULL,
    color = NULL,
    shape = NULL,
    title = "Edge-length fidelity is a quality diagnostic, not a parity guarantee"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10)
  )
```

The edge-length diagnostic shows whether each layout preserves the supplied
weights on each fixture. It is not a substitute for parity: two algorithms can
have similar edge-length fidelity and still produce different global shapes.
Here it helps distinguish weighted-layout quality from the stricter question of
whether the ND path has inherited the legacy weighted-GRIP mechanics.

## Results Summary And Discussion

The current evidence does not yet support strict 2D/3D embedding parity between
`grip.layout.weighted.nd()` and `grip.layout.weighted()` on all canonical
gflow-backed fixtures. The ND backend is finite and deterministic, and it is
almost shape-identical on the smoke fixtures, but strict thresholds still flag
small residuals in the cross-dimensional nonuniform 3D grid embedded in 2D. The
trace-level diagnostics show a much more actionable picture: the active-level
metadata now matches exactly on tiny weighted fixtures, coarsest initialization
is numerically identical, every traced frame in the small 2D/3D ratchet fixtures
is within strict Procrustes tolerance, the intrinsic-dimension gflow grid and
lattice traces are within tolerance, and the remaining cross-dimensional
failure first appears during full-graph refinement. The refinement-step trace
then narrows that failure further: the first large row-level discrepancy is in
the pre-temperature displacement, before heat scaling or coordinate application,
and the component split shows the largest term is weighted edge attraction. The
term-level attraction trace now provides the per-adjacent-edge arithmetic needed
to choose the next fix. The source trace further shows that the large round-11
edge term is an amplification of coordinate drift that starts much earlier in
the full-graph refinement loop.

The gflow-backed harness exercises the graph families we care about without
copying quadform formulas into `grip`. The fixtures are generated from the
gflow sources that already own the grid, lattice, and quadform constructions.
This gives us a development test bed for future weighted-GRIP ND work while
keeping ordinary package checks stable.

The implication for implementation is precise: this remains a
compatibility-port project, not a tuning project. The next work should inspect
the full-graph refinement arithmetic for the nonuniform 3D grid embedded in 2D,
starting with vertex-level force accumulation at the first divergent round
reported by the refinement-step trace. The current evidence suggests that the
remaining gap is no longer in the smallest traced mechanics,
intrinsic-dimensional grid/lattice traces, the multiscale active-set schedule,
edge/metric-neighbor input identity, local-temperature scaling, or coordinate
application order.

## Appendix

### Run Metadata

```{r metadata-table}
kable(run_metadata)
```

### Strict Parity Thresholds

The gated test harness can enforce strict thresholds by setting
`GRIP_RUN_GFLOW_PARITY_TESTS=true` and `GRIP_ENFORCE_GFLOW_PARITY=true`.
For the current side-by-side ND backend, those strict checks are expected to
fail until the legacy phases have been ported to ND.

```{r threshold-table}
kable(data.frame(
  metric = names(thresholds),
  threshold = unlist(thresholds),
  row.names = NULL
))
```

### Summary Metrics

```{r metrics-table}
kable(metrics[, c(
  "case_id",
  "layout_dim",
  "n",
  "m",
  "direct_rmse",
  "direct_max_abs",
  "procrustes_rmse",
  "procrustes_max_abs",
  "legacy_edge_rmse",
  "nd_edge_rmse",
  "legacy_elapsed_sec",
  "nd_elapsed_sec"
)], digits = 5)
```

### Trace Metrics

```{r trace-metrics-table}
kable(trace_metrics[, c(
  "fixture_id",
  "dim",
  "frame",
  "phase",
  "active_vertices",
  "direct_rmse",
  "direct_max_abs",
  "procrustes_rmse",
  "procrustes_max_abs",
  "metadata_match"
)], digits = 6)
```

### gflow Trace Metrics

```{r gflow-trace-metrics-table}
kable(gflow_trace_metrics[, c(
  "case_id",
  "layout_dim",
  "frame",
  "phase",
  "misf_level",
  "round_in_level",
  "active_vertices",
  "direct_rmse",
  "direct_max_abs",
  "procrustes_rmse",
  "procrustes_max_abs",
  "metadata_match"
)], digits = 6)
```

### gflow All-Dimension Trace Metrics

```{r gflow-all-trace-metrics-table}
kable(gflow_trace_all_metrics[, c(
  "case_id",
  "intrinsic_dim",
  "layout_dim",
  "frame",
  "phase",
  "misf_level",
  "round_in_level",
  "active_vertices",
  "direct_rmse",
  "direct_max_abs",
  "procrustes_rmse",
  "procrustes_max_abs",
  "metadata_match"
)], digits = 6)
```

### Refinement-Step Trace Metrics

```{r refinement-step-metrics-table}
kable(refinement_step_metrics[, c(
  "case_id",
  "layout_dim",
  "level_index",
  "misf_level",
  "round_in_level",
  "order_index",
  "vertex",
  "heat_before_delta",
  "heat_after_delta",
  "old_cos_before_delta",
  "old_cos_after_delta",
  "pre_temp_disp_norm_delta",
  "coords_before_max_abs",
  "pre_temp_disp_max_abs",
  "attraction_disp_max_abs",
  "repulsion_disp_max_abs",
  "applied_disp_max_abs",
  "coords_after_max_abs",
  "attraction_edges_match",
  "repulsion_neighbors_match"
)], digits = 6)
```

### Refinement-Step Attraction Terms

```{r refinement-step-attraction-terms-table}
kable(refinement_step_term_metrics[, c(
  "case_id",
  "layout_dim",
  "round_in_level",
  "order_index",
  "vertex",
  "term_index",
  "neighbor",
  "weight_delta",
  "norm2_delta",
  "desired_delta",
  "desired2_delta",
  "scale_delta",
  "delta_max_abs",
  "step_max_abs",
  "cumulative_max_abs"
)], digits = 6)
```

### Refinement Source Trace

```{r refinement-source-tables}
kable(refinement_source$summary, digits = 6)
kable(refinement_source$term_trace, digits = 6)
```

### Reproducibility

Render this report from the `grip` repository root with:

```{r reproducibility-command, eval=FALSE, echo=TRUE}
rmarkdown::render(
  "dev/weighted-grip-nd-parity/weighted_grip_nd_gflow_parity_report.Rmd",
  output_dir = "output/weighted-grip-nd-parity",
  output_file = "weighted_grip_nd_gflow_parity_report.html",
  params = list(
    mode = "smoke",
    output_dir = "/Users/pgajer/current_projects/grip/output/weighted-grip-nd-parity",
    gflow_repo = "/Users/pgajer/current_projects/gflow"
  )
)
```

Run the gated package test with:

```{bash eval=FALSE}
GRIP_RUN_GFLOW_PARITY_TESTS=true Rscript -e 'devtools::test(filter = "gflow-parity")'
```

Run the trace-level first-divergence diagnostic with:

```{bash eval=FALSE}
GRIP_RUN_TRACE_PARITY_TESTS=true Rscript -e 'devtools::test(filter = "trace-parity")'
```

Run the gflow-backed trace ratchet with:

```{bash eval=FALSE}
GRIP_RUN_GFLOW_TRACE_PARITY_TESTS=true Rscript -e 'devtools::test(filter = "gflow-parity")'
```

Run the cross-dimensional gflow-backed trace ratchet with:

```{bash eval=FALSE}
GRIP_RUN_GFLOW_TRACE_ALL_TESTS=true Rscript -e 'devtools::test(filter = "gflow-parity")'
```

Run the refinement-step gflow-backed trace with:

```{bash eval=FALSE}
GRIP_RUN_GFLOW_REFINEMENT_STEP_TRACE_TESTS=true Rscript -e 'devtools::test(filter = "gflow-parity")'
```

Run strict parity enforcement with:

```{bash eval=FALSE}
GRIP_RUN_GFLOW_PARITY_TESTS=true GRIP_ENFORCE_GFLOW_PARITY=true \
  Rscript -e 'devtools::test(filter = "gflow-parity")'
```

Run strict trace parity enforcement with:

```{bash eval=FALSE}
GRIP_RUN_TRACE_PARITY_TESTS=true GRIP_ENFORCE_TRACE_PARITY=true \
  Rscript -e 'devtools::test(filter = "trace-parity")'
```

Run strict gflow trace parity enforcement with:

```{bash eval=FALSE}
GRIP_RUN_GFLOW_TRACE_PARITY_TESTS=true GRIP_ENFORCE_GFLOW_TRACE_PARITY=true \
  Rscript -e 'devtools::test(filter = "gflow-parity")'
```

Run strict cross-dimensional gflow trace parity enforcement with:

```{bash eval=FALSE}
GRIP_RUN_GFLOW_TRACE_ALL_TESTS=true GRIP_ENFORCE_GFLOW_TRACE_ALL_PARITY=true \
  Rscript -e 'devtools::test(filter = "gflow-parity")'
```
