---
title: "grip: Multiscale Graph Layout and Geodesic Embedding Tools in R"
author:
  - name: Pawel Gajer
    affiliation: Center for Advanced Microbiome Research and Innovation (CAMRI), Institute for Genome Sciences, Department of Microbiology and Immunology, University of Maryland School of Medicine
    address: Baltimore, Maryland, USA
    orcid: ~
    email: pgajer@gmail.com
  - name: Jacques Ravel
    affiliation: Center for Advanced Microbiome Research and Innovation (CAMRI), Institute for Genome Sciences, Department of Microbiology and Immunology, University of Maryland School of Medicine
    address: Baltimore, Maryland, USA
    orcid: ~
    email: ~
abstract: >
  grip is an R package for multiscale graph layout and experimental
  graph-geodesic embedding tools in 2D and 3D. Its stable interface brings the
  GRIP algorithm to R through a C++ core, with an explicit choice between hop
  distance and edge length as the graph metric and trace diagnostics for the
  coarse-to-fine solve. The package also provides full, landmark, and
  multiscale geodesic-KK functions that assess or refine fixed input shortest
  paths using their lengths after embedding. These research interfaces expose
  current objectives and approximations rather than a definitive geodesic
  multidimensional scaling method. Presets, graph generators, and common
  scoring functions support reproducible comparison of candidate embeddings.
  This article describes grip 0.1.3 and illustrates its stable and experimental
  interfaces on synthetic weighted graphs and a microbial network.
preamble: |
  \usepackage{amsmath}
  \usepackage{xcolor}
  \usepackage{etoolbox}
  \usepackage{float}
  \AtBeginEnvironment{table}{\small}
  \AtBeginEnvironment{thebibliography}{\small}
  \providecommand{\manuscriptversion}{unversioned}
  \providecommand{\manuscriptbuildnumber}{NA}
  \providecommand{\manuscriptcommit}{unknown}
  \providecommand{\manuscriptbuilddatetime}{unknown}
  \InputIfFileExists{build/manuscript_build_info.tex}{}{}
  \newcommand{\grippaperbuildstamp}{%
    \par
    \vspace{-0.35em}%
    \begin{center}%
    \scriptsize\color{gray!75}%
    Version \texttt{\manuscriptversion}\enspace | \enspace
    Build \texttt{\manuscriptbuildnumber}\enspace | \enspace
    Commit \texttt{\manuscriptcommit}\\
    Built \manuscriptbuilddatetime
    \end{center}%
    \vspace{0.35em}%
  }
  \makeatletter
  \apptocmd{\maketitle}{\grippaperbuildstamp}{}{}
  \makeatother
date: "2026-08-20"
date_received: ~
journal:
  firstpage: 1
  lastpage: ~
slug: gajer-grip
bibliography: grip-software-paper.bib
output:
  rjtools::rjournal_article:
    self_contained: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)

paper.table.colnames <- c(
  graph = "Graph",
  candidate = "Cand",
  method = "Meth",
  dim = "Dim",
  `n.vertices` = "nV",
  `n.edges` = "nE",
  `n.runs` = "Runs",
  `elapsed.sec.mean` = "Sec",
  `score.composite` = "Comp",
  `sampled.stress` = "Stress",
  `sampled.stress.mean` = "Stress",
  `edge.length.cv` = "EdgeCV",
  `edge.length.cv.mean` = "EdgeCV",
  `sampled.nonedge.sep.ratio` = "NESR",
  `sampled.nonedge.sep.ratio.mean` = "NESR",
  `cluster.separation` = "ClSep",
  `cluster.separation.mean` = "ClSep",
  `stability.procrustes.mean` = "PStab",
  `edge.crossings` = "Cross",
  `gkk.weighted.rel.rmse` = "GKK Rel",
  `gkk.weighted.rmse` = "GKK RMSE",
  `gkk.mean.rel.path.error` = "Path Rel"
)

paper_table_caption <- function(title, abbreviations) {
  paste0(title, " Abbrev.: ", abbreviations)
}

paper_kable <- function(x, caption, digits = 3) {
  col.names <- unname(ifelse(
    names(x) %in% names(paper.table.colnames),
    paper.table.colnames[names(x)],
    names(x)
  ))
  align <- c("l", rep("r", max(0, ncol(x) - 1L)))
  knitr::kable(
    x,
    digits = digits,
    col.names = col.names,
    row.names = FALSE,
    align = align,
    caption = caption,
    position = "H"
  )
}

find.extdata.file <- function(...) {
  rel.path <- file.path(...)
  candidates <- c(
    system.file("extdata", rel.path, package = "grip"),
    file.path("..", "..", "..", "..", "inst", "extdata", rel.path),
    file.path("..", "..", "..", "inst", "extdata", rel.path),
    file.path("inst", "extdata", rel.path)
  )
  path <- candidates[file.exists(candidates)][1L]
  if (!length(path) || is.na(path) || !nzchar(path)) {
    stop("could not locate ", rel.path)
  }
  path
}

read.extdata.csv <- function(file.name) {
  utils::read.csv(find.extdata.file(file.name), stringsAsFactors = FALSE)
}

read.extdata.rds <- function(...) {
  readRDS(find.extdata.file(...))
}

plot_layout_panel <- function(coords,
                              edges,
                              main,
                              projection = NULL,
                              vertex.col = "black",
                              edge.col = "gray82",
                              cex = 0.5) {
  plot.layout(
    coords, edges,
    projection = projection,
    vertex.col = vertex.col,
    edge.col = edge.col,
    pch = 16,
    cex = cex,
    main = main
  )
}

edge.matrix.from.adj <- function(adj.list) {
  edges <- list()
  idx <- 0L
  for (u in seq_along(adj.list)) {
    nbrs <- adj.list[[u]][adj.list[[u]] > u]
    for (v in nbrs) {
      idx <- idx + 1L
      edges[[idx]] <- c(u, v)
    }
  }
  do.call(rbind, edges)
}
```

# Introduction

Graph embeddings are commonly used to produce readable drawings, but their
coordinates can also be intended to represent graph geometry. For weighted
graphs, these goals need not coincide: a visually clear arrangement may
nevertheless distort edge lengths and shortest-path relationships. grip
provides a stable multiscale interface for 2D and 3D graph layout, together
with experimental tools for assessing and refining graph-geodesic fidelity.

The original GRIP algorithm [@gajer2002; @gajer2004] addresses this problem by
constructing a hierarchy of maximal independent sets, placing vertices from
coarse to fine resolution, and refining locally at each level. The grip package
exposes this multiscale strategy in R through a C++ core accessed with Rcpp
[@eddelbuettel2011]. In practice, that hierarchy is what makes it feasible to
produce layouts for graphs with hundreds to thousands of vertices within an R
workflow, including both 2D and 3D embeddings.

The graph metric is explicit in the main interface. Hop distance supports
topology-oriented layouts, whereas edge length supports weighted graphs whose
edges encode geometry. The experimental geodesic-KK functions ask a related
but distinct question: how closely do shortest paths chosen from the input
graph retain their lengths after the graph is embedded? This distinction makes
grip both a layout package and a research platform for studying when a graph
drawing can also serve as a metric representation.

This article describes the stable layout workflow, including family-specific
presets, trace diagnostics, graph generators, and comparison across random
seeds with common quality metrics. It also defines the narrower role of the
experimental geodesic-KK interfaces and illustrates how their path-based scores
can distinguish candidate embeddings. The methodological development of a
general geodesic multidimensional scaling framework remains future work.

The package is complementary to existing R tools rather than a replacement for
them. igraph [@csardi2006] supplies several strong general-purpose layouts;
ggraph [@pedersen2024] provides a plotting grammar for network graphics; and
graphlayouts [@schoch2023] offers additional layout algorithms, including
stress-based methods. The contribution of grip is the combination of fast
multiscale layout generation, geometry-aware weighted layouts, common scoring
tools that can assess layouts computed elsewhere, and experimental fixed-path
geodesic diagnostics and refinement.

# Package overview

At a high level, grip keeps the original GRIP idea of working from a coarse
representation of the graph back to the full graph. The paper does not attempt
to re-derive the algorithm in detail; instead, it focuses on the package
interfaces that expose this strategy in a form suitable for routine R use. The
practical consequence of that hierarchy is speed: the package can produce 2D
and 3D layouts quickly enough to support comparison workflows rather than only
single final figures.

## Layout engines and interfaces

The main layout functions are summarized in Table \@ref(tab:api-summary). All
return an $n \times d$ numeric matrix of coordinates, where $d$ is 2 or 3.

```{r api-summary, echo=FALSE}
api.summary <- data.frame(
  Function = c(
    "grip(metric = \"hop\")",
    "grip(metric = \"edge_length\")",
    "trace.grip(metric = \"hop\")",
    "trace.grip(metric = \"edge_length\")"
  ),
  Purpose = c(
    "Combinatorial multiscale layout",
    "Weighted multiscale layout",
    "Combinatorial layout with per-level snapshots",
    "Weighted layout with per-level snapshots"
  ),
  `Graph metric` = c("Hop", "Edge length", "Hop", "Edge length"),
  check.names = FALSE
)

knitr::kable(
  api.summary,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "Core layout interfaces in the grip package.",
  position = "H"
)
```

The `metric` argument states which graph geometry the layout should use.
`metric = "hop"`, the default, treats every edge as one step. With
`metric = "edge_length"`, supplied edge lengths affect the optimization. The
traced interface uses the same metric choice and records intermediate states
and diagnostics for inspection or animation.

The package accepts graphs either as two-column integer edge matrices or as
adjacency lists with optional parallel weight lists. A minimal example of both
calling patterns is shown below.

```{r api-examples, echo=TRUE}
library(grip)
if (utils::packageVersion("grip") < "0.1.3") {
  stop("This article requires grip 0.1.3 or later.")
}

edges <- edges.mesh(8, 8)
coords.mesh <- grip(edges, n = 64, dim = 2, preset = "mesh", seed = 1)

adj.list <- list(c(2L), c(1L, 3L), c(2L, 4L), c(3L))
weight.list <- list(c(1.0), c(1.0, 2.0), c(2.0, 1.5), c(1.5))
coords.weighted.small <- grip(
  adj_list = adj.list,
  weight_list = weight.list,
  n = 4,
  dim = 2,
  metric = "edge_length",
  seed = 12
)
```

The most important user-facing parameters are layout dimension, initialization
placement, the number of multiscale and final refinement rounds, local
neighborhood size, repulsion strength, and the random seed. The package also
packs disconnected components by default, so disconnected graphs can still be
laid out with a single call.

## From graph layout to graph-geodesic fidelity

Let $G=(V,E,\ell)$ be a connected weighted graph and let
$Z=(z_1,\ldots,z_n)$ be an embedding of its vertices. Stress-based graph
drawing commonly compares a target graph distance $d_G(i,j)$ with the ambient
Euclidean distance $\lVert z_i-z_j\rVert_2$ between two embedded vertices, as
in Kamada--Kawai graph drawing [@kamada1989]. This criterion concerns the
straight chord between the coordinates, even when the target distance was
defined by a path through the graph.

The experimental geodesic-KK functions in grip instead prepare an input
shortest path $P_{ij}$ and measure its length after embedding,

$$
L_Z(P_{ij}) = \sum_{(u,v)\in P_{ij}} \lVert z_u-z_v\rVert_2.
$$

They compare $L_Z(P_{ij})$ with the corresponding input graph distance, up to
a fitted or specified global scale. By default, preparation retains one
deterministic shortest path for each evaluated pair; an alternative mode
averages exactly over tied input shortest paths. The paths are fixed during
scoring and optimization rather than recomputed after every coordinate update.

This path-based criterion gives a precise meaning to graph-geodesic fidelity
within the package. Exact preservation of every edge length would preserve the
length of every graph path and hence the induced shortest-path metric. In a
low-dimensional embedding, however, exact preservation may be impossible, and
a low geodesic-KK score establishes fidelity only to the prepared graph metric
and path family. It does not by itself establish recovery of an unknown
underlying data geometry.

## Stable and experimental interfaces

The package contains several related interface families, but they do not all
have the same maturity or computational role. Table \@ref(tab:family-scope)
summarizes the distinction used throughout this article.

```{r family-scope, echo=FALSE}
family.scope <- data.frame(
  Family = c(
    "Multiscale GRIP",
    "Full geodesic KK",
    "Landmark geodesic KK",
    "MISF geodesic KK"
  ),
  Status = c(
    "Stable",
    "Experimental",
    "Experimental",
    "Experimental"
  ),
  Purpose = c(
    "Primary layout and trace interface.",
    "All-pairs fixed-path scoring and gradient polish.",
    "Sparse fixed-path scoring and warm-started polish.",
    "Coarse-to-fine geodesic optimization and scoring."
  ),
  check.names = FALSE
)

knitr::kable(
  family.scope,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "Maturity and purpose of the main layout and geodesic interface families in grip.",
  position = "H"
)
```

The stable entry points are `grip()` and `trace.grip()`. The full experimental
family consists of `prepare.geodesic.kk()`, `score.geodesic.kk()`, and
`geodesic.kk()`; corresponding landmark and MISF families use the same
prepare--score--optimize pattern. The landmark family retains a sparse set of
pairs for scalable evaluation and warm-started polish. The MISF family adds
coarsening, a top-level geodesic solve, coarse-to-fine insertion, anchored
active-level refinement, and final geodesic polish.

These experimental interfaces expose related objectives and approximations,
not a single definitive geodesic multidimensional scaling algorithm. They are
useful now for evaluating and refining weighted layouts, while questions of
general formulation, automatic strategy selection, and validation across data
geometries remain open research problems.

## Presets, tracing, and graph families

The package includes family-specific presets for common graph classes. Hop
metric layouts provide presets for meshes, carpets, trees, and tori.
Edge-length layouts add presets for cylinders, spheres, irregular manifolds,
and other families where graph geometry matters. These presets are
starting points, not guarantees of optimality, but they are useful because they
make the package easier to use on recurring graph families.

The trace functions expose the coarse-to-fine solve explicitly. Figure
\@ref(fig:trace) shows four trace frames for a small mesh layout.

```{r trace, fig.cap="Multiscale trace of a 6x6 mesh. The four panels show the coarsest initialization, an early intermediate state, a late intermediate state, and the final layout.", fig.alt="Four panels showing a mesh layout evolving from a sparse coarse configuration to a regular grid-like final arrangement.", fig.width=10, fig.height=2.8, fig.pos="H"}
trace.edges <- edges.mesh(6, 6)
trace.solve <- trace.grip(
  trace.edges, n = 36, dim = 2,
  preset = "mesh", seed = 31,
  metric = "hop",
  trace = "level"
)

frame.idx <- unique(c(
  1L,
  max(2L, floor(length(trace.solve$frames) / 3)),
  max(3L, floor(2 * length(trace.solve$frames) / 3)),
  length(trace.solve$frames)
))

labels <- c("Coarsest start", "Early refinement", "Late refinement", "Final layout")
op <- par(mfrow = c(1, 4), mar = c(1, 1, 2.5, 1))
for (i in seq_along(frame.idx)) {
  coords <- trace.solve$frames[[frame.idx[i]]]
  active <- complete.cases(coords[, 1:2])
  xy <- coords[active, 1:2, drop = FALSE]
  active.edges <- trace.edges[active[trace.edges[, 1]] & active[trace.edges[, 2]], , drop = FALSE]

  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- max(0.08 * diff(xlim), 0.2)
  ypad <- max(0.08 * diff(ylim), 0.2)

  plot(xy[, 1], xy[, 2], type = "n", asp = 1, axes = FALSE,
       xlab = "", ylab = "",
       xlim = xlim + c(-xpad, xpad),
       ylim = ylim + c(-ypad, ypad),
       main = labels[i])
  if (nrow(active.edges) > 0) {
    segments(
      coords[active.edges[, 1], 1], coords[active.edges[, 1], 2],
      coords[active.edges[, 2], 1], coords[active.edges[, 2], 2],
      col = "gray82"
    )
  }
  points(xy[, 1], xy[, 2], pch = 16, cex = 0.6)
}
par(op)
```

The package also ships graph generators for regular grids, trees, recursive
families such as Sierpinski carpets, and weighted geometric families generated
from surfaces. These are helpful for examples, regression tests, and
family-specific benchmarking.

## Representative runtime scale

The main software claim of grip is practical runtime scale rather than a claim
to dominate every other layout method on every graph. Table
\@ref(tab:runtime-scale) therefore summarizes three representative layouts that
cover both 2D and 3D settings. The torus row is evaluated directly in the paper
build, while the larger carpet and HMP rows use bundled package artifacts so
that the manuscript remains quick to render.

```{r runtime-scale}
benchmark.results <- read.extdata.rds("vs_alternatives", "benchmark_results.rds")
hmp.results <- read.extdata.rds("hmp_u01_gc_coarse", "vignette_results.rds")

torus.edges <- edges.torus(20, 20)
torus.elapsed <- system.time(
  torus.coords <- grip(
    torus.edges,
    n = max(torus.edges),
    dim = 3,
    preset = "torus",
    seed = 1
  )
)[["elapsed"]]
torus.score <- score.layout(
  torus.coords,
  edges = torus.edges,
  n = max(torus.edges),
  sample.size.stress = 2000L,
  sample.size.nonedge = 2000L,
  edge.crossings = "never"
)

carpet.runtime <- merge(
  benchmark.results$carpet$scores[
    benchmark.results$carpet$scores$method == "grip carpet preset",
    c("method", "n.vertices", "n.edges", "dim", "sampled.stress")
  ],
  benchmark.results$carpet$timing[
    benchmark.results$carpet$timing$method == "grip carpet preset",
    c("method", "elapsed.sec")
  ],
  by = "method"
)

hmp.default <- hmp.results$preset_summary[
  hmp.results$preset_summary$candidate == "default",
  ,
  drop = FALSE
]

runtime.scale <- data.frame(
  graph = c("20x20 torus", "Carpet level 4", "HMP/U01 coarse"),
  dim = c(3L, carpet.runtime$dim, 3L),
  `n.vertices` = c(max(torus.edges), carpet.runtime$n.vertices, 1828L),
  `n.edges` = c(nrow(torus.edges), carpet.runtime$n.edges, 4656L),
  context = c(
    "Single 3D preset solve",
    "Single 2D preset solve",
    "Bundled weighted 3D default candidate"
  ),
  elapsed.sec = c(
    unname(torus.elapsed),
    carpet.runtime$elapsed.sec,
    hmp.default$elapsed.sec.mean
  ),
  sampled.stress = c(
    torus.score$sampled.stress[[1L]],
    carpet.runtime$sampled.stress,
    hmp.default$sampled.stress.mean
  ),
  check.names = FALSE
)

paper_kable(
  runtime.scale,
  caption = paper_table_caption(
    "Representative 2D and 3D runtime scale for grip.",
    "nV = number of vertices; nE = number of edges; Sec = elapsed seconds; Stress = sampled stress."
  ),
  digits = 3
)
```

These rows are not intended as a formal benchmark against all alternatives.
They show the practical regime the package is designed for: graphs with
hundreds to thousands of vertices, including 3D embeddings and a weighted
real-data example, can be laid out in seconds rather than forcing users into a
slow manual tuning loop.

# Layout assessment workflow

The strongest package-level feature of grip is not that it always produces the
best layout on every graph. Rather, it gives users a reproducible way to score
candidate layouts, compare them across seeds, and inspect the trade-offs
between graph families and layout objectives.

## Quality metrics

Table \@ref(tab:metrics-table) summarizes the metrics returned by
`score.layout()`. The package uses them directly for single-layout scoring
and for aggregate comparisons across seeds in `compare.layouts()`.

```{r metrics-table, echo=FALSE}
metrics.table <- data.frame(
  Metric = c(
    "Sampled stress",
    "Edge-length CV",
    "Sampled non-edge separation",
    "Edge crossings",
    "Cluster separation",
    "Procrustes stability"
  ),
  Meaning = c(
    "Agreement between layout distances and graph distances",
    "Uniformity of edge lengths",
    "Breathing room between non-neighbors",
    "Readability in 2D",
    "Separation of known external groups",
    "Seed-to-seed reproducibility"
  ),
  Better = c("Lower", "Lower", "Higher", "Lower", "Higher", "Lower"),
  check.names = FALSE
)

knitr::kable(
  metrics.table,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "Quality metrics used by score.layout().",
  position = "H"
)
```

`compare.layouts()` runs multiple candidates across seeds, computes these
metrics for each run, and returns both run-level and summary-level results. The
summary includes mean quality metrics, a rank-based composite score, and a
Procrustes stability estimate when multiple seeds are used.

## Evidence for family-specific presets

To make the preset claims concrete, Table \@ref(tab:preset-benchmark) compares
the default layout settings with package-supplied presets on two structured
families, each run across three seeds. The results are generated directly in
the paper build.

```{r preset-benchmark}
mesh.edges <- edges.mesh(12, 12)
mesh.compare <- compare.layouts(
  edges = mesh.edges,
  n = 144,
  dim = 2,
  candidates = c("default", "mesh"),
  seeds = 1:3,
  edge.crossings = "always"
)

carpet.edges <- edges.sierpinski.carpet(3)
carpet.n <- max(carpet.edges)
carpet.compare <- compare.layouts(
  edges = carpet.edges,
  n = carpet.n,
  dim = 2,
  candidates = c("default", "carpet"),
  seeds = 1:3,
  edge.crossings = "never"
)

preset.benchmark <- rbind(
  transform(
    mesh.compare$summary[, c(
      "candidate", "n.runs", "elapsed.sec.mean", "sampled.stress.mean",
      "edge.length.cv.mean", "sampled.nonedge.sep.ratio.mean",
      "stability.procrustes.mean"
    )],
    graph = "12x12 mesh"
  ),
  transform(
    carpet.compare$summary[, c(
      "candidate", "n.runs", "elapsed.sec.mean", "sampled.stress.mean",
      "edge.length.cv.mean", "sampled.nonedge.sep.ratio.mean",
      "stability.procrustes.mean"
    )],
    graph = "Carpet level 3"
  )
)

preset.benchmark <- preset.benchmark[, c(
  "graph", "candidate", "n.runs", "elapsed.sec.mean",
  "sampled.stress.mean", "edge.length.cv.mean",
  "sampled.nonedge.sep.ratio.mean", "stability.procrustes.mean"
)]
preset.benchmark$candidate <- factor(
  preset.benchmark$candidate,
  levels = c("default", "mesh", "carpet")
)
preset.benchmark <- preset.benchmark[order(preset.benchmark$graph, preset.benchmark$candidate), ]

paper_kable(
  preset.benchmark,
  caption = paper_table_caption(
    "Preset benchmark on two structured graph families, each averaged over three seeds.",
    "Runs = number of seeds; Sec = mean elapsed seconds; Stress = sampled stress; EdgeCV = edge-length coefficient of variation; NESR = sampled non-edge separation ratio; PStab = Procrustes stability."
  ),
  digits = 3
)
```

These results show why the presets matter. On both graph families, the
package-supplied preset reduces sampled stress, improves edge-length
uniformity, increases non-edge separation, and is substantially more stable
across seeds. The runtime cost is small relative to the improvement in layout
quality on these families.

# Worked examples

## Family-specific improvement on a recursive graph

Figure \@ref(fig:carpet-figure) visualizes the Sierpinski-carpet benchmark from
the previous section. The figure is paired with the benchmark table so that the
claim is visual and quantitative rather than purely visual.

```{r carpet-figure, fig.cap="Level-3 Sierpinski carpet laid out with the default settings (left) and the carpet preset (right). The preset yields the lower-stress and more stable solution reported in Table \\@ref(tab:preset-benchmark).", fig.alt="Two panels showing a Sierpinski carpet graph. The default layout is more distorted, while the preset layout better preserves the recursive holes.", fig.width=10, fig.height=4, fig.pos="H"}
carpet.default.coords <- grip(carpet.edges, n = carpet.n, dim = 2, seed = 12)
carpet.preset.coords <- grip(carpet.edges, n = carpet.n, dim = 2, preset = "carpet", seed = 12)

op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
plot_layout_panel(carpet.default.coords, carpet.edges, "Default", cex = 0.3)
plot_layout_panel(carpet.preset.coords, carpet.edges, "Carpet preset", cex = 0.3)
par(op)
```

## Weighted graph with geometric edge lengths

The edge-length metric is easiest to motivate on a graph whose topology is
simple but whose geometry is not. The example below uses a $5 \times 5$ mesh
whose edge lengths are induced by a curved saddle surface. The graph topology
is unchanged, but the intended metric is no longer the flat grid metric.

The comparison includes three candidates: the combinatorial layout, the
weighted layout, and the weighted layout followed by a small amount of LGKK
polish. The scoring uses `score.geodesic.kk()`, which evaluates the
embedded lengths of fixed input shortest paths against the weighted graph
metric.

```{r weighted-summary}
weighted.mesh <- mesh.surface.graph(5, 5, surface = "saddle", amplitude = 0.8)

coords.combinatorial <- grip(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  dim = 3,
  preset = "mesh",
  seed = 1
)

coords.weighted <- grip(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  metric = "edge_length",
  seed = 1
)

coords.weighted.lgkk <- grip(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  metric = "edge_length",
  lgkk_polish_rounds = 6L,
  seed = 1
)

gkk.prepared <- prepare.geodesic.kk(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights
)

weighted.summary <- do.call(
  rbind,
  list(
    cbind(
      method = "Combinatorial GRIP",
      score.geodesic.kk(
        coords.combinatorial,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rel.rmse",
        "gkk.weighted.rmse",
        "gkk.mean.rel.path.error"
      )]
    ),
    cbind(
      method = "Weighted GRIP",
      score.geodesic.kk(
        coords.weighted,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rel.rmse",
        "gkk.weighted.rmse",
        "gkk.mean.rel.path.error"
      )]
    ),
    cbind(
      method = "Weighted GRIP + LGKK polish",
      score.geodesic.kk(
        coords.weighted.lgkk,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rel.rmse",
        "gkk.weighted.rmse",
        "gkk.mean.rel.path.error"
      )]
    )
  )
)

paper_kable(
  weighted.summary,
  caption = paper_table_caption(
    "Weighted saddle-surface example scored with geodesic-KK criteria.",
    "GKK Rel = weighted relative RMSE; GKK RMSE = weighted RMSE; Path Rel = mean relative path error."
  ),
  digits = 3
)
```

On this example, the weighted solve lowers the weighted relative RMSE from
`r round(weighted.summary$gkk.weighted.rel.rmse[1], 3)` to
`r round(weighted.summary$gkk.weighted.rel.rmse[2], 3)`, and a short LGKK
polish reduces it further to
`r round(weighted.summary$gkk.weighted.rel.rmse[3], 3)`. This is the package
argument for making the metric explicit: the same interface solves a different
layout problem when edge lengths, rather than hop counts, define graph
geometry. It also demonstrates the narrower role of the experimental layer:
the path-based score provides a stated fidelity criterion for comparing the
three candidates, without claiming that this small example establishes general
recovery of curved data geometry.

```{r weighted-figure, fig.cap="Weighted mesh example. Top left: target saddle geometry. Top right: combinatorial GRIP. Bottom left: weighted GRIP. Bottom right: weighted GRIP followed by LGKK polish.", fig.alt="A four-panel comparison of a saddle-surface mesh showing the target geometry, the combinatorial layout, the weighted layout, and the weighted layout with LGKK polish.", fig.width=11.2, fig.height=7.2, fig.pos="H"}
op <- par(mfrow = c(2, 2), mar = c(1, 1, 3, 1))
plot_layout_panel(
  weighted.mesh$coords_surface,
  weighted.mesh$edges,
  "Target geometry",
  projection = "ortho",
  vertex.col = "#666666"
)
plot_layout_panel(
  coords.combinatorial,
  weighted.mesh$edges,
  "Combinatorial GRIP",
  projection = "ortho"
)
plot_layout_panel(
  coords.weighted,
  weighted.mesh$edges,
  "Weighted GRIP",
  projection = "ortho",
  vertex.col = "#1F3B73"
)
plot_layout_panel(
  coords.weighted.lgkk,
  weighted.mesh$edges,
  "Weighted GRIP + LGKK",
  projection = "ortho",
  vertex.col = "#7A2E2E"
)
par(op)
```

## Large real-data example: HMP microbial network

The package bundles a coarsened graph derived from Human Microbiome Project
data [@hmp2012]. The original giant connected component had 6,474 vertices; the
bundled coarsened graph has 1,828 vertices and 4,656 edges, which is large
enough that ad hoc manual layout selection becomes impractical. This example is
included not only for biological realism but also because it shows that a
weighted 3D workflow at this graph size is still practical inside R.

```{r hmp-data}
data(hmp.u01.gc.coarse)

cat("Vertices:", length(hmp.u01.gc.coarse$adj_list), "\n")
cat("Edges:", nrow(edge.matrix.from.adj(hmp.u01.gc.coarse$adj_list)), "\n")
cat(
  "CST levels:",
  paste(sort(unique(hmp.u01.gc.coarse$vertex_data$cst)), collapse = ", "),
  "\n"
)
```

The full HMP workflow uses repeated scoring and candidate comparison, so the
package ships precomputed artifacts for the vignette and the paper. Even in
that broader workflow, the bundled default weighted candidate runs in about
`r round(hmp.results$preset_summary$elapsed.sec.mean[hmp.results$preset_summary$candidate == "default"], 1)`
seconds. Table \@ref(tab:hmp-summary) reports three candidates: the default
preset, the tree preset, and the strongest local-search candidate by
cluster-separation score among the bundled precomputed results.

```{r hmp-summary}
hmp.results <- read.extdata.rds("hmp_u01_gc_coarse", "vignette_results.rds")

hmp.local.best.idx <- which.max(hmp.results$local_search_summary$cluster.separation.mean)
hmp.local.best <- hmp.results$local_search_summary[hmp.local.best.idx, , drop = FALSE]
hmp.display <- rbind(
  transform(
    hmp.results$preset_summary[hmp.results$preset_summary$candidate == "default", , drop = FALSE],
    candidate = "default"
  ),
  transform(
    hmp.results$preset_summary[hmp.results$preset_summary$candidate == "tree", , drop = FALSE],
    candidate = "tree"
  ),
  transform(
    hmp.local.best,
    candidate = "local search"
  )
)

hmp.display <- hmp.display[, c(
  "candidate", "elapsed.sec.mean", "sampled.stress.mean",
  "sampled.nonedge.sep.ratio.mean", "cluster.separation.mean",
  "score.composite"
)]

paper_kable(
  hmp.display,
  caption = paper_table_caption(
    "Precomputed HMP candidate summary for the bundled 1,828-vertex graph.",
    "Sec = elapsed seconds; Stress = sampled stress; NESR = sampled non-edge separation ratio; ClSep = cluster separation; Comp = composite score."
  ),
  digits = 3
)
```

The table shows a real trade-off. The default candidate is the fastest of the
three and has the lowest sampled stress, while the tree and local-search
candidates produce substantially larger cluster-separation scores. That is
exactly the kind of choice the comparison workflow is meant to surface: speed
makes it practical to inspect several plausible 3D layouts, but the package
still leaves the final selection explicit and evidence-based.

```{r hmp-figure, fig.cap="Precomputed 3D layouts for the HMP microbial network, colored by community state type. Left: default candidate. Middle: tree preset. Right: best precomputed local-search candidate by cluster-separation score.", fig.alt="Three orthographic 3D network layouts of the HMP graph colored by community state type, comparing default, tree, and local-search candidates.", fig.width=11, fig.height=3.5, fig.pos="H"}
hmp.edges <- edge.matrix.from.adj(hmp.u01.gc.coarse$adj_list)
hmp.cst <- hmp.u01.gc.coarse$vertex_data$cst
cst.levels <- sort(unique(hmp.cst))
cst.cols <- setNames(grDevices::hcl.colors(length(cst.levels), "Dark 3"), cst.levels)
hmp.vertex.cols <- cst.cols[hmp.cst]
hmp.local.name <- as.character(hmp.results$local_search_summary$candidate[hmp.local.best.idx])

op <- par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))
plot_layout_panel(
  hmp.results$layouts$preset$default,
  hmp.edges,
  "Default",
  projection = "ortho",
  vertex.col = hmp.vertex.cols,
  cex = 0.45
)
plot_layout_panel(
  hmp.results$layouts$preset$tree,
  hmp.edges,
  "Tree preset",
  projection = "ortho",
  vertex.col = hmp.vertex.cols,
  cex = 0.45
)
plot_layout_panel(
  hmp.results$layouts$local[[hmp.local.name]],
  hmp.edges,
  "Local search",
  projection = "ortho",
  vertex.col = hmp.vertex.cols,
  cex = 0.45
)
par(op)
```

# Comparison with other R tools

For package evaluation, the most useful comparison is not a checklist of
features but an evidence-based benchmark. Table \@ref(tab:karate-benchmark)
therefore uses bundled precomputed results on the Zachary karate club graph
[@zachary1977], scored with the same layout-quality metrics for all methods.

```{r karate-benchmark}
benchmark.results <- read.extdata.rds("vs_alternatives", "benchmark_results.rds")

karate.display <- benchmark.results$karate$scores[, c(
  "method",
  "sampled.stress",
  "edge.length.cv",
  "sampled.nonedge.sep.ratio",
  "cluster.separation",
  "edge.crossings"
)]

paper_kable(
  karate.display,
  caption = paper_table_caption(
    "Small-graph comparison on the Zachary karate club network using a shared scoring function.",
    "Stress = sampled stress; EdgeCV = edge-length coefficient of variation; NESR = sampled non-edge separation ratio; ClSep = cluster separation; Cross = edge crossings."
  ),
  digits = 3
)
```

This comparison is intentionally modest in tone. On a very small 2D social
graph, single-scale methods from igraph and graphlayouts perform very well, and
grip does not dominate them. That is an informative result. It supports the
claim that grip is most useful as a fast multiscale and evaluation-oriented
package,
especially for structured graph families, weighted geometric graphs, and
workflow-driven layout selection. It also illustrates why the scoring
functions are valuable on their own: they can be used to assess layouts
computed outside the package.

# Reproducibility

This article is written in R Markdown using the rjtools output format currently
recommended by the journal. The tables and the small-graph figures are produced
from evaluated code in the manuscript. The larger HMP case study and the
external-package comparison both use bundled precomputed `.rds` artifacts so
that the paper can be rendered quickly during review while still pointing to
the scripts and vignettes that generate those artifacts.

In the package source tree, the external comparison artifacts are created by
`inst/scripts/precompute-vs-alternatives.R`, and the HMP workflow is described
in the real-data vignette and the HMP case-study article. This split keeps the
paper build practical without hiding how the heavier results were produced.

# Software and data availability

grip is available from CRAN at
<https://CRAN.R-project.org/package=grip>. The development repository, issue
tracker, manuscript source, and scripts used to generate the bundled benchmark
artifacts are available at <https://github.com/pgajer/grip>. The HMP example
data and the precomputed results used here are distributed with the package.

# Discussion

grip brings multiscale 2D and 3D layout generation, graph-geodesic diagnostics,
and a reproducible assessment workflow into R. The package is most useful in
three settings:

- graphs large enough that single-scale manual tuning becomes slow, especially
  when 3D embeddings are useful;
- weighted graphs whose edge lengths encode geometry and should therefore be
  used as the layout metric rather than treated as ordinary metadata;
- analysis workflows where several plausible layouts must be compared
  reproducibly rather than selected by eye.

The main interface remains `grip()`, with `trace.grip()` for inspecting the
multiscale solve. Full GKK, sparse LGKK, and MISF-GKK are explicitly
experimental research interfaces. They make fixed-path geodesic objectives,
diagnostics, and several refinement strategies available for studying when a
layout can also represent graph path geometry.

There are also clear limits. The preset library is finite, so some real graphs
will still require manual tuning. Layout-quality metrics are proxies for
readability, while geodesic-KK scores quantify fidelity only to the prepared
graph metric and fixed input path family. The experimental functions do not yet
constitute a general GMDS method, and the package does not provide an automatic
rule for selecting among full, landmark, and multiscale strategies. Exact
edge-crossing counts are expensive on dense 2D graphs, and the current
implementation does not use GPU acceleration for very large problems.

Future methodological work will examine broader GMDS formulations, automatic
strategy selection without external geometric truth, and empirical conditions
under which graph-geodesic fidelity corresponds to scientifically meaningful
data geometry. The present package supplies a tested layout core and a concrete
experimental platform for that work without conflating the two levels of
maturity.

# Acknowledgments

The GRIP algorithm was originally developed in collaboration with Stephen G.
Kobourov and Michael T. Goodrich. The HMP microbial network data were derived
from publicly available Human Microbiome Project resources. Research reported
in this publication was supported by grant number INV-048956 from the Gates
Foundation.
