---
title: "grip: Multiscale Graph Layout and Geodesic Embedding Tools in R"
author:
  - name: Pawel Gajer
    affiliation: Center for Advanced Microbiome Research and Innovation (CAMRI), Institute for Genome Sciences, Department of Microbiology and Immunology, University of Maryland School of Medicine
    address: Baltimore, Maryland, USA
    orcid: ~
    email: pgajer@gmail.com
  - name: Jacques Ravel
    affiliation: Center for Advanced Microbiome Research and Innovation (CAMRI), Institute for Genome Sciences, Department of Microbiology and Immunology, University of Maryland School of Medicine
    address: Baltimore, Maryland, USA
    orcid: ~
    email: ~
abstract: >
  grip is an R package for multiscale graph layout and experimental
  graph-geodesic embedding tools in 2D and 3D. Its stable interface brings the
  GRIP algorithm to R through a C++ core, with an explicit choice between hop
  distance and edge length as the graph metric and trace diagnostics for the
  coarse-to-fine solve. The package also provides full, landmark, and
  multiscale geodesic variants of the Kamada--Kawai (KK) algorithm that assess
  or refine fixed input shortest paths using their lengths after embedding.
  These research interfaces expose
  current objectives and approximations rather than a definitive geodesic
  multidimensional scaling method. Presets, graph generators, and common
  scoring functions support reproducible comparison of candidate embeddings.
  In a broader data-to-embedding workflow, grip addresses the
  graph-to-embedding stage, downstream of graph construction tools such as the
  companion dgraphs package.
  A shared benchmark illustrates interoperability with igraph and
  graphlayouts. This article describes grip 0.2.0 and illustrates
  its stable and experimental interfaces on structured graphs, a synthetic
  weighted graph, and an HMP-only Illumina 16S microbial network.
preamble: |
  \usepackage{amsmath}
  \usepackage{xcolor}
  \usepackage{etoolbox}
  \usepackage{float}
  \AtBeginEnvironment{table}{\small}
  \AtBeginEnvironment{thebibliography}{\small}
  \providecommand{\manuscriptversion}{unversioned}
  \providecommand{\manuscriptbuildnumber}{NA}
  \providecommand{\manuscriptcommit}{unknown}
  \providecommand{\manuscriptbuilddatetime}{unknown}
  \newif\ifgrippapershowbuildstamp
  \grippapershowbuildstampfalse
  \InputIfFileExists{build/manuscript_build_info.tex}{}{}
  \newcommand{\grippaperbuildstamp}{%
    \par
    \vspace{-0.35em}%
    \begin{center}%
    \scriptsize\color{gray!75}%
    Version \texttt{\manuscriptversion}\enspace | \enspace
    Build \texttt{\manuscriptbuildnumber}\enspace | \enspace
    Commit \texttt{\manuscriptcommit}\\
    Built \manuscriptbuilddatetime
    \end{center}%
    \vspace{0.35em}%
  }
  \makeatletter
  \ifgrippapershowbuildstamp
    \apptocmd{\maketitle}{\grippaperbuildstamp}{}{}
  \fi
  \makeatother
date: "2026-08-23"
date_received: ~
journal:
  firstpage: 1
  lastpage: ~
slug: gajer-grip
bibliography: grip-software-paper.bib
output:
  rjtools::rjournal_article:
    self_contained: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)

if (!requireNamespace("grip", quietly = TRUE) ||
    utils::packageVersion("grip") < "0.2.0") {
  stop("This article requires grip 0.2.0 or later.")
}

paper.table.colnames <- c(
  graph = "Graph",
  candidate = "Cand",
  method = "Meth",
  dim = "Dim",
  `n.vertices` = "nV",
  `n.edges` = "nE",
  `n.runs` = "Runs",
  `elapsed.sec.mean` = "Sec",
  `elapsed.sec.median` = "MedSec",
  `elapsed.sec.iqr` = "IQR",
  `score.composite` = "Comp",
  `sampled.stress` = "Stress",
  `sampled.stress.mean` = "Stress",
  `edge.length.cv` = "EdgeCV",
  `edge.length.cv.mean` = "EdgeCV",
  `sampled.nonedge.sep.ratio` = "NESR",
  `sampled.nonedge.sep.ratio.mean` = "NESR",
  `stability.procrustes.mean` = "PStab",
  `edge.crossings` = "Cross",
  `gkk.weighted.rel.rmse` = "GKK Rel",
  `gkk.mean.rel.path.error` = "Path Rel"
)

paper_table_caption <- function(title, abbreviations) {
  paste0(title, " Abbrev.: ", abbreviations)
}

html_attribute_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}

accessible_kable <- function(x,
                             caption,
                             alt,
                             row.names = FALSE,
                             align = NULL,
                             col.names = NULL,
                             digits = getOption("digits"),
                             position = "H") {
  args <- list(
    x = x,
    digits = digits,
    col.names = col.names,
    row.names = row.names,
    align = align,
    caption = caption,
    position = position
  )
  args <- Filter(Negate(is.null), args)
  if (knitr::is_html_output()) {
    args$table.attr <- sprintf(
      'aria-label="%s"',
      html_attribute_escape(alt)
    )
  }
  do.call(knitr::kable, args)
}

paper_kable <- function(x, caption, alt, digits = 3) {
  col.names <- unname(ifelse(
    names(x) %in% names(paper.table.colnames),
    paper.table.colnames[names(x)],
    names(x)
  ))
  align <- c("l", rep("r", max(0, ncol(x) - 1L)))
  accessible_kable(
    x,
    digits = digits,
    col.names = col.names,
    row.names = FALSE,
    align = align,
    caption = caption,
    alt = alt,
    position = "H"
  )
}

find.extdata.file <- function(...) {
  rel.path <- file.path(...)
  candidates <- c(
    file.path("reproducibility", "precomputed", rel.path),
    system.file("extdata", rel.path, package = "grip"),
    file.path("..", "..", "..", "..", "inst", "extdata", rel.path),
    file.path("..", "..", "..", "inst", "extdata", rel.path),
    file.path("inst", "extdata", rel.path)
  )
  path <- candidates[file.exists(candidates)][1L]
  if (!length(path) || is.na(path) || !nzchar(path)) {
    stop("could not locate ", rel.path)
  }
  path
}

read.extdata.csv <- function(file.name) {
  utils::read.csv(find.extdata.file(file.name), stringsAsFactors = FALSE)
}

read.extdata.rds <- function(...) {
  readRDS(find.extdata.file(...))
}

plot_layout_panel <- function(coords,
                              edges,
                              main,
                              projection = NULL,
                              vertex.col = "black",
                              edge.col = "gray82",
                              cex = 0.5,
                              ...) {
  plot.layout(
    coords, edges,
    projection = projection,
    vertex.col = vertex.col,
    edge.col = edge.col,
    pch = 16,
    cex = cex,
    main = main,
    ...
  )
}

align_to_reference <- function(coords, reference) {
  centered <- sweep(coords, 2, colMeans(coords), "-", check.margin = FALSE)
  target <- sweep(reference, 2, colMeans(reference), "-", check.margin = FALSE)
  rotation.svd <- svd(crossprod(centered, target))
  rotation <- rotation.svd$u %*% t(rotation.svd$v)
  scale <- sum(rotation.svd$d) / sum(centered^2)
  sweep(scale * centered %*% rotation, 2, colMeans(reference), "+", check.margin = FALSE)
}

plot_common_projection <- function(coords,
                                   edges,
                                   main,
                                   xlim,
                                   ylim,
                                   vertex.col = "black") {
  plot(
    coords[, 1], coords[, 2],
    type = "n", asp = 1, axes = FALSE,
    xlab = "", ylab = "", xlim = xlim, ylim = ylim,
    main = main
  )
  segments(
    coords[edges[, 1], 1], coords[edges[, 1], 2],
    coords[edges[, 2], 1], coords[edges[, 2], 2],
    col = "gray82"
  )
  points(coords[, 1], coords[, 2], pch = 16, cex = 0.55, col = vertex.col)
}

edge.matrix.from.adj <- function(adj.list) {
  edges <- list()
  idx <- 0L
  for (u in seq_along(adj.list)) {
    nbrs <- adj.list[[u]][adj.list[[u]] > u]
    for (v in nbrs) {
      idx <- idx + 1L
      edges[[idx]] <- c(u, v)
    }
  }
  do.call(rbind, edges)
}
```

# Introduction

Many data-analysis workflows represent observations by a graph and then
represent that graph in low-dimensional coordinates. Either step can alter the
geometry: graph construction specifies the edges and weights that induce the
graph's intrinsic shortest-path geometry, whereas an embedding places the
vertices in an ambient Euclidean space and thereby induces a second geometry
through Euclidean chord lengths and embedded graph-path lengths. The question
of fidelity therefore has two parts: how closely the graph-geodesic structure
reflects the geometry in the data that the analysis is intended to preserve,
and how closely the geometry induced by the embedding represents the graph's
intrinsic geometry. The \CRANpkg{dgraphs} package addresses the first question,
whereas \CRANpkg{grip} focuses on the second [@gajer2026dgraphs]. The companion
packages separate these stages:

$$
X \xrightarrow{\texttt{dgraphs}} G
  \xrightarrow{\texttt{grip}} Z.
$$

Within this division of responsibility, \CRANpkg{grip} begins with a supplied
graph, whether produced by \CRANpkg{dgraphs} or another source. It generates,
traces, scores, compares, and experimentally refines embeddings. These
operations support two drawing aims that can conflict in weighted graphs:
readability and faithful representation of edge lengths and shortest-path
relationships. This article focuses on that downstream package.

The stable interface implements the original GRIP algorithm [@gajer2002;
@gajer2004] in a C++ core accessed from R through Rcpp [@eddelbuettel2011].
GRIP constructs a hierarchy of maximal independent sets, places vertices
from coarse to fine resolution, and refines the layout locally at each
level. This hierarchy makes it practical to generate and compare multiple 2D
and 3D candidate layouts for graphs with hundreds to thousands of vertices.
The original GRIP refinement is computationally localized rather than global:
intermediate filtration levels use Kamada--Kawai (KK)-style forces over
restricted neighborhoods, and the final level uses Fruchterman--Reingold
refinement [@gajer2002; @kamada1989; @fruchterman1991]. GRIP therefore draws on
KK geometry without optimizing the classical global all-pairs KK objective.
The interface also makes the graph metric explicit:
hop distance supports topology-oriented layouts, whereas edge length allows
weighted graph geometry to guide the layout.

GRIP belongs to a broader multilevel graph-drawing tradition. Related methods
construct coarse graph hierarchies by matching, clustering, or graph-distance
sampling and then refine layouts from coarse to fine resolution
[@harel2002; @walshaw2003; @hachul2005; @hu2005]. The methods differ in their
coarsening rules, force approximations, and refinement objectives; the package
contribution here is an R interface to the GRIP maximal-independent-set
hierarchy rather than a claim that multilevel layout is unique to GRIP.

A separate, explicitly experimental layer provides geodesic variants of KK.
These methods assess or refine fixed input shortest paths by comparing their
embedded lengths with the corresponding graph distances. They provide concrete
objectives for studying graph-geodesic fidelity, but they do not yet constitute
a general geodesic multidimensional scaling method.

Within the R ecosystem, \CRANpkg{igraph} provides general-purpose layouts
[@csardi2006; @igraph2026], \CRANpkg{ggraph} supplies a plotting grammar
[@pedersen2025], and \CRANpkg{graphlayouts} provides additional layout
algorithms [@schoch2023]. The contribution of \CRANpkg{grip} is the integration
of multiscale candidate generation, weighted graph geometry, reproducible layout
assessment that also applies to layouts computed elsewhere, and experimental
fixed-path geodesic diagnostics. This article describes the layout and trace
interfaces, family-specific presets and graph generators, runtime
characteristics, and a common workflow for scoring and comparing candidate
embeddings. The examples cover structured graphs, a weighted geometric
graph, and a microbial network.

# Package overview

This section describes the interfaces that expose the multiscale strategy: the
layout engines, the graph-geodesic fidelity criterion, the split between stable
and experimental families, and the preset and tracing facilities.

## Layout engines and interfaces

The main layout functions are summarized in Table \@ref(tab:api-summary). All
return an $n \times d$ numeric matrix of coordinates, where $d$ is 2 or 3.

```{r api-summary, echo=FALSE}
api.summary <- data.frame(
  Function = c(
    "grip(): hop",
    "grip(): edge length",
    "trace.grip(): hop",
    "trace.grip(): edge length"
  ),
  Purpose = c(
    "Multiscale layout",
    "Weighted multiscale layout",
    "Layout with per-level snapshots",
    "Weighted layout with snapshots"
  ),
  `Graph metric` = c("Hop", "Edge length", "Hop", "Edge length"),
  check.names = FALSE
)

accessible_kable(
  api.summary,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "Core layout interfaces in the grip package.",
  alt = paste(
    "Four rows list the grip and trace.grip interfaces for hop and edge-length",
    "metrics, together with each interface's purpose and graph metric."
  ),
  position = "H"
)
```

The `metric` argument states which graph geometry the layout should use.
`metric = "hop"`, the default, treats every edge as one step. With
`metric = "edge_length"`, supplied edge lengths affect the optimization. The
traced interface uses the same metric choice and records intermediate states
and diagnostics for inspection or animation.

The package accepts graphs either as two-column integer edge matrices or as
adjacency lists with optional parallel weight lists. A minimal example of both
calling patterns is shown below.

```{r api-examples, echo=TRUE}
library(grip)

edges <- edges.mesh(8, 8)
coords.mesh <- grip(edges, n = 64, dim = 2, preset = "mesh", seed = 1)

adj.list <- list(c(2L), c(1L, 3L), c(2L, 4L), c(3L))
weight.list <- list(c(1.0), c(1.0, 2.0), c(2.0, 1.5), c(1.5))
coords.weighted.small <- grip(
  adj_list = adj.list,
  weight_list = weight.list,
  n = 4,
  dim = 2,
  metric = "edge_length",
  seed = 12
)
```

The most important user-facing parameters are layout dimension, initialization
placement, the number of multiscale and final refinement rounds, local
neighborhood size, repulsion strength, and the random seed. The package also
packs disconnected components by default, so disconnected graphs can still be
laid out with a single call.

The stable default solve does not invoke the geodesic-KK research layer. The
`lgkk_polish_rounds` argument and other `lgkk_*` multiscale controls explicitly
opt a `grip()` or `trace.grip()` call into experimental landmark-geodesic
refinement when nonzero. These controls retain the stable input and output types,
but their numerical semantics are not part of the stable contract.
`plot.layout()` is a registered S3 method; `trace.grip()` and `score.layout()`
are standalone functions whose dot-delimited names predate the current
snake-case argument convention.

## From graph layout to graph-geodesic fidelity

Let $G=(V,E,\ell)$ be a connected weighted graph and let
$Z=(z_1,\ldots,z_n)$ be an embedding of its vertices. Stress-based graph
drawing commonly compares a target graph distance $d_G(i,j)$ with the ambient
Euclidean distance $\lVert z_i-z_j\rVert_2$ between two embedded vertices, as
in Kamada--Kawai graph drawing [@kamada1989]. This criterion concerns the
straight chord between the coordinates, even when the target distance was
defined by a path through the graph.

The experimental geodesic-KK functions in \CRANpkg{grip} instead prepare an input
shortest path $P_{ij}$ and measure its length after embedding,

$$
L_Z(P_{ij}) = \sum_{(u,v)\in P_{ij}} \lVert z_u-z_v\rVert_2.
$$

They compare $L_Z(P_{ij})$ with the corresponding input graph distance, up to
a fitted or specified global scale. With graph distances $g_p$, embedded path
lengths $h_p$, and Kamada--Kawai weights $k_p=1/g_p^2$, the default profiled
scale is the weighted least-squares solution

$$
L_0 = \frac{\sum_p k_p g_p h_p}{\sum_p k_p g_p^2}.
$$

All examples in this article use this profiled mode. By default, preparation retains one
deterministic shortest path for each evaluated pair; an alternative mode
averages exactly over tied input shortest paths. The paths are fixed during
scoring and optimization rather than recomputed after every coordinate update.

This path-based criterion gives a precise meaning to graph-geodesic fidelity
within the package. Exact preservation of every edge length would preserve the
length of every graph path and hence the induced shortest-path metric. In a
low-dimensional embedding, however, exact preservation may be impossible, and
a low geodesic-KK score establishes fidelity only to the prepared graph metric
and path family. It does not by itself establish recovery of an unknown
underlying data geometry.

This objective is related to, but distinct from, geodesic multidimensional
scaling. Isomap applies classical MDS to shortest-path distance estimates
[@tenenbaum2000], and its landmark variants reduce the cost by using a subset
of reference points [@desilva2002]. Stress majorization and PivotMDS likewise
optimize or approximate chord-distance representations of graph distances
[@gansner2005; @brandes2007]. The criterion above instead compares each fixed
input path's length *after embedding* with its graph length. Its landmark
approximation inherits the scalability motivation of landmark MDS, but not its
chord-distance objective.

## Stable and experimental interfaces

The package contains several related interface families, but they do not all
have the same maturity or computational role. Table \@ref(tab:family-scope)
summarizes the distinction used throughout this article.

```{r family-scope, echo=FALSE}
family.scope <- data.frame(
  Family = c(
    "Multiscale GRIP",
    "Full geodesic KK",
    "Landmark geodesic KK",
    "MISF geodesic KK"
  ),
  Status = c(
    "Stable",
    "Experimental",
    "Experimental",
    "Experimental"
  ),
  Purpose = c(
    "Primary layout and trace interface.",
    "All-pairs path scoring and polish.",
    "Landmark path scoring and polish.",
    "Multiscale path scoring and polish."
  ),
  check.names = FALSE
)

accessible_kable(
  family.scope,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "Maturity and purpose of the main layout and geodesic interface families in grip.",
  alt = paste(
    "Four rows distinguish the stable multiscale GRIP interface from the",
    "experimental full, landmark, and MISF geodesic-KK interface families."
  ),
  position = "H"
)
```

The stable entry points are `grip()` and `trace.grip()`. The full experimental
family has separate preparation, scoring, and optimization entry points:
`prepare.geodesic.kk()`, `score.geodesic.kk()`, and `geodesic.kk()`.
Corresponding landmark geodesic Kamada--Kawai (LGKK) and
maximal-independent-set filtration (MISF) families use the same
prepare--score--optimize pattern. LGKK retains, by default, the 20 nearest
graph-metric neighbors and eight deterministic farthest-point landmarks per
vertex. The MISF family adds
coarsening, a top-level geodesic solve, coarse-to-fine insertion, anchored
active-level refinement, and final geodesic polish.

Experimental functions and opt-in `lgkk_*` controls may change as the research
interface develops. Such changes are recorded in `NEWS`; deprecation periods
are provided where feasible, but only the stable entry points and their default
behavior carry the package's compatibility commitment.

## Presets, tracing, and graph families

The package includes family-specific presets for common graph classes. Hop
metric layouts provide presets for meshes, carpets, trees, and tori.
Edge-length layouts add presets for cylinders, spheres, irregular manifolds,
and other families where graph geometry matters.

The trace functions expose the coarse-to-fine solve explicitly. Figure
\@ref(fig:trace) shows four trace frames for a small mesh layout.

```{r trace, fig.cap="Multiscale trace of a 6x6 mesh. The four panels show the coarsest initialization, an early intermediate state, a late intermediate state, and the final layout.", fig.alt="Four panels showing a mesh layout evolving from a sparse coarse configuration to a regular grid-like final arrangement.", fig.width=10, fig.height=2.8, fig.pos="H"}
trace.edges <- edges.mesh(6, 6)
trace.solve <- trace.grip(
  trace.edges, n = 36, dim = 2,
  preset = "mesh", seed = 31,
  metric = "hop",
  trace = "level"
)

frame.idx <- unique(c(
  1L,
  max(2L, floor(length(trace.solve$frames) / 3)),
  max(3L, floor(2 * length(trace.solve$frames) / 3)),
  length(trace.solve$frames)
))

labels <- c("Coarsest start", "Early refinement", "Late refinement", "Final layout")
op <- par(mfrow = c(1, 4), mar = c(1, 1, 2.5, 1))
for (i in seq_along(frame.idx)) {
  coords <- trace.solve$frames[[frame.idx[i]]]
  active <- complete.cases(coords[, 1:2])
  xy <- coords[active, 1:2, drop = FALSE]
  active.edges <- trace.edges[active[trace.edges[, 1]] & active[trace.edges[, 2]], , drop = FALSE]

  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- max(0.08 * diff(xlim), 0.2)
  ypad <- max(0.08 * diff(ylim), 0.2)

  plot(xy[, 1], xy[, 2], type = "n", asp = 1, axes = FALSE,
       xlab = "", ylab = "",
       xlim = xlim + c(-xpad, xpad),
       ylim = ylim + c(-ypad, ypad),
       main = labels[i])
  if (nrow(active.edges) > 0) {
    segments(
      coords[active.edges[, 1], 1], coords[active.edges[, 1], 2],
      coords[active.edges[, 2], 1], coords[active.edges[, 2], 2],
      col = "gray82"
    )
  }
  points(xy[, 1], xy[, 2], pch = 16, cex = 0.6)
}
par(op)
```

The package also ships graph generators for regular grids, trees, recursive
families such as Sierpinski carpets, and weighted geometric families generated
from surfaces. They provide examples, regression fixtures, and controlled
benchmarks for the graph-to-embedding interfaces. They are not substitutes for
data-driven graph construction, which is upstream of the problem addressed by
\CRANpkg{grip}.

```{r precomputed-artifacts, include=FALSE}
benchmark.results <- read.extdata.rds("vs_alternatives", "benchmark_results.rds")
```

The original GRIP studies considered substantially larger graphs than the
examples in this article [@gajer2002; @gajer2004]. In the R package, the layout
solve remains practical beyond the low-thousands regime, but all-pairs graph
distances, experimental geodesic preparation, repeated scoring, and exact
crossing counts can become the limiting steps before the multiscale solve
itself. The scaled benchmark below therefore reports layout time separately
from scoring time and keeps exact crossings disabled.

# Assessing and refining candidate embeddings

\CRANpkg{grip} separates candidate generation from assessment. A candidate may be
produced by `grip()`, by one of the package's experimental embedding functions,
or by another R package. It can then be evaluated with the same diagnostics,
compared across random seeds or parameter choices, and, when appropriate,
passed to an experimental refinement step. This separation makes the package
useful as an evaluation framework rather than only as a collection of layout
engines.

## General layout diagnostics

Table \@ref(tab:metrics-table) lists the diagnostics used in this article from
`score.layout()`. `compare.layouts()` applies the same scoring function to
multiple candidates and summarizes repeated seeded runs.

```{r metrics-table, echo=FALSE}
metrics.table <- data.frame(
  Metric = c(
    "Sampled stress",
    "Edge-length CV",
    "Sampled non-edge separation",
    "Edge crossings",
    "Procrustes stability"
  ),
  Meaning = c(
    "Chord agreement with graph distances",
    "Variation among embedded edges",
    "Spacing of sampled non-neighbors",
    "2D segment intersections",
    "Across-seed reproducibility"
  ),
  Better = c("Lower", "Lower", "Higher", "Lower", "Lower"),
  check.names = FALSE
)

accessible_kable(
  metrics.table,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "General layout diagnostics used in the examples.",
  alt = paste(
    "Five rows define sampled stress, edge-length variation, non-edge",
    "separation, edge crossings, and Procrustes stability, and indicate",
    "whether lower or higher values are preferred."
  ),
  position = "H"
)
```

For sampled stress, let $S$ be the unique unordered vertex pairs obtained from
uniform pair sampling, $g_p$ the graph distance, and $e_p$ the Euclidean layout
distance for $p\in S$. The implementation fits
$a=\sum_{p\in S}e_pg_p/\sum_{p\in S}g_p^2$ and reports the scale-invariant
quantity

$$
\operatorname{Stress}_S =
\sqrt{\frac{\sum_{p\in S}(e_p-a g_p)^2}{\sum_{p\in S}e_p^2}}.
$$

The examples use 2,000 sampled pairs unless stated otherwise. Edge-length CV is
the standard deviation divided by the mean of embedded edge lengths. Sampled
non-edge separation is the minimum distance among up to 5,000 sampled
non-edges divided by the median embedded edge length. Edge crossings count
strict intersections between nonincident segments in 2D; `"auto"` computes
them for at most 1,000 edges, `"always"` overrides that threshold, and
`"never"` skips the calculation. Procrustes stability is the mean pairwise
vertex RMSE after centering each seeded layout, scaling it to unit maximum
radius, and applying the optimal orthogonal rotation or reflection. The paper
reports the component diagnostics rather than a composite rank because their
relative importance depends on the analysis.
Because sampled non-edge separation depends on graph size, the number of
sampled non-edges, and the sampled pairs, it should be compared only for the
same graph under the same sampling protocol; it is not a graph-independent
quality scale.

## Graph-geodesic diagnostics and refinement

The general diagnostics above concern layout quality and chord-distance
agreement. The experimental graph-geodesic interface asks a different
question: how well an embedding preserves the metric of a fixed input graph
when graph paths are measured after embedding. Table
\@ref(tab:assessment-functions) shows how the principal functions fit into a
generate--assess--refine workflow.

```{r assessment-functions, echo=FALSE}
assessment.functions <- data.frame(
  Task = c(
    "Generate candidate",
    "Assess layout",
    "Prepare graph paths",
    "Assess geodesic fidelity",
    "Prepare edge targets",
    "Refine candidate"
  ),
  Function = c(
    "grip()",
    "score.layout(), compare.layouts()",
    "prepare.geodesic.kk()",
    "score.gmds(), score.geodesic.kk()",
    "prepare.edge.kk()",
    "geodesic.kk(), edge.kk()"
  ),
  check.names = FALSE
)

accessible_kable(
  assessment.functions,
  row.names = FALSE,
  align = c("l", "l"),
  caption = "Package functions in the candidate generation, assessment, and refinement workflow.",
  alt = paste(
    "Six rows map multiscale generation, general scoring, all-pairs path",
    "preparation, graph-geodesic scoring, lightweight edge preparation, and",
    "experimental refinement to grip functions."
  ),
  position = "H"
)
```

`score.gmds()` supplies a common diagnostic panel for any coordinate matrix on
a prepared graph. With an all-pairs preparation it reports edge fidelity,
fixed-path geodesic stress, distance-band summaries, and ordinary
metric-MDS chord stress. With the lightweight object returned by
`prepare.edge.kk()`, only edge diagnostics are available; the all-pairs
quantities are reported as unavailable. This distinction matters in practice:
edge-only assessment and repair avoid a dense shortest-path matrix, whereas
the full path-based diagnostics retain substantially more information and have
correspondingly higher memory and runtime costs.

The following executable example shows the lightweight
prepare--score--refine pattern. Here `prepare.edge.kk()` avoids all-pairs path
construction; `edge.kk()` returns refined coordinates and a diagnostic panel.

```{r geodesic-workflow, echo=TRUE, results="hide"}
cycle.edges <- edges.cycle(8)
theta <- seq(0, 2 * pi, length.out = 9)[-9]
initial <- cbind(1.4 * cos(theta), 0.7 * sin(theta))

prepared <- prepare.edge.kk(cycle.edges, n = 8)
before <- score.gmds(initial, prepared = prepared)
fit <- edge.kk(
  coords = initial,
  prepared = prepared,
  density_mix_schedule = 1,
  max_iter = 4,
  return_trace = FALSE
)
after <- score.gmds(fit$coords, prepared = prepared)
```

# Worked package workflows

## Structured graph presets and tracing

To make the preset behavior concrete, Table \@ref(tab:preset-benchmark)
compares the default settings with package-supplied presets on two structured
families, each run across ten seeds. The results are generated directly in the
paper build.

```{r preset-benchmark}
mesh.edges <- edges.mesh(12, 12)
mesh.compare <- compare.layouts(
  edges = mesh.edges,
  n = 144,
  dim = 2,
  candidates = c("default", "mesh"),
  seeds = 1:10,
  edge.crossings = "always"
)

carpet.edges <- edges.sierpinski.carpet(3)
carpet.n <- max(carpet.edges)
carpet.compare <- compare.layouts(
  edges = carpet.edges,
  n = carpet.n,
  dim = 2,
  candidates = c("default", "carpet"),
  seeds = 1:10,
  edge.crossings = "never"
)

summary.cols <- c(
  "candidate", "n.runs", "elapsed.sec.mean", "sampled.stress.mean",
  "edge.length.cv.mean", "sampled.nonedge.sep.ratio.mean",
  "stability.procrustes.mean"
)

preset.benchmark <- rbind(
  transform(mesh.compare$summary[, summary.cols], graph = "12x12 mesh"),
  transform(carpet.compare$summary[, summary.cols], graph = "Carpet level 3")
)

preset.benchmark <- preset.benchmark[, c(
  "graph", "candidate", "n.runs", "elapsed.sec.mean",
  "sampled.stress.mean", "edge.length.cv.mean",
  "sampled.nonedge.sep.ratio.mean", "stability.procrustes.mean"
)]
preset.benchmark$candidate <- factor(
  preset.benchmark$candidate,
  levels = c("default", "mesh", "carpet")
)
preset.benchmark <- preset.benchmark[order(preset.benchmark$graph, preset.benchmark$candidate), ]

paper_kable(
  preset.benchmark,
  caption = paper_table_caption(
    "Preset benchmark on two structured graph families, each averaged over ten seeds.",
    "Runs = number of seeds; Sec = mean elapsed seconds; Stress = sampled stress; EdgeCV = edge-length coefficient of variation; NESR = sampled non-edge separation ratio; PStab = Procrustes stability."
  ),
  alt = paste(
    "Four rows compare default and family-specific presets on a mesh and",
    "Sierpinski carpet using runtime, stress, edge variation, non-edge",
    "separation, and stability metrics."
  ),
  digits = 3
)
```

For these two controlled families, the package-supplied preset reduces sampled
stress and edge-length variation and produces more reproducible layouts across
seeds. PStab is itself the across-seed dispersion summary used for this
comparison. The result is family-specific evidence for the supplied defaults, not a
claim that one parameter profile is preferable for arbitrary graphs. Figure
\@ref(fig:carpet-figure) shows the corresponding carpet layouts.

```{r carpet-figure, fig.cap="Level-3 Sierpinski carpet laid out with the default settings (left) and the carpet preset (right). The preset yields the lower-stress and more stable solution reported in the preceding benchmark table.", fig.alt="Two panels showing a Sierpinski carpet graph. The default layout is more distorted, while the preset layout better preserves the recursive holes.", fig.width=10, fig.height=4, fig.pos="H"}
carpet.default.coords <- grip(carpet.edges, n = carpet.n, dim = 2, seed = 12)
carpet.preset.coords <- grip(carpet.edges, n = carpet.n, dim = 2, preset = "carpet", seed = 12)

op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
plot_layout_panel(
  carpet.default.coords, carpet.edges, "Default", cex = 0.3,
  axes = FALSE, xlab = "", ylab = ""
)
plot_layout_panel(
  carpet.preset.coords, carpet.edges, "Carpet preset", cex = 0.3,
  axes = FALSE, xlab = "", ylab = ""
)
par(op)
```

## Fixed weighted graph: graph-geodesic assessment

The edge-length metric is easiest to motivate on a graph whose topology is
simple but whose geometry is not. The example below uses a $5 \times 5$ mesh
whose edge lengths are induced by a curved saddle surface. The graph topology
is unchanged, but the intended metric is no longer the flat grid metric.

The comparison includes combinatorial and weighted GRIP layouts, the weighted
layout followed by six experimental LGKK polish rounds, and the weighted
three-dimensional Kamada--Kawai implementation in \CRANpkg{igraph}. All candidates
use seed 1 and otherwise retain their package defaults. The graph is fixed
before the candidates are generated, and `score.geodesic.kk()` evaluates their
embedded fixed-path lengths against the same weighted graph metric. The example
therefore illustrates the graph-to-embedding workflow; it does not evaluate
how well the graph itself represents an underlying data set.

```{r weighted-summary}
weighted.saddle <- benchmark.results$weighted_saddle
weighted.mesh <- list(
  edges = weighted.saddle$edges,
  n = weighted.saddle$n,
  edge_weights = weighted.saddle$edge_weights,
  coords_surface = weighted.saddle$target_coords
)

coords.combinatorial <- weighted.saddle$layouts[["Combinatorial GRIP"]]
coords.weighted <- weighted.saddle$layouts[["Weighted GRIP"]]
coords.weighted.lgkk <- weighted.saddle$layouts[["Weighted GRIP + LGKK polish"]]
coords.weighted.kk <- weighted.saddle$layouts[["Weighted KK (igraph)"]]

gkk.cols <- c(
  "gkk.weighted.rel.rmse",
  "gkk.mean.rel.path.error"
)

weighted.summary <- weighted.saddle$scores[, c("method", gkk.cols)]

paper_kable(
  weighted.summary,
  caption = paper_table_caption(
    "Weighted saddle-surface example scored with geodesic-KK criteria.",
    "GKK Rel = weighted relative RMSE; Path Rel = mean relative path error."
  ),
  alt = paste(
    "Four rows compare combinatorial GRIP, weighted GRIP, weighted GRIP with",
    "LGKK polish, and weighted igraph Kamada-Kawai using geodesic-KK scores."
  ),
  digits = 3
)
```

The scale-normalized columns are comparable across solvers whose coordinate
scales differ. The LGKK-polished candidate is optimized against the same
fixed-path objective used for scoring, so its low value is an in-sample
optimization result rather than independent validation. The comparison with
weighted KK shows how the package can apply that criterion to a layout that
was not optimized for it.

Figure \@ref(fig:weighted-figure) also exposes a limitation of the criterion.
Preserving the weighted graph paths does not uniquely determine the target's
extrinsic curvature: a comparatively flat realization can have low path loss.
The panels are similarity-aligned to the target and use one projection and one
coordinate scale, so this difference is not a viewing-axis artifact.

```{r weighted-figure, fig.cap="Weighted mesh example after similarity alignment to the target and display with a common orthographic projection and scale. Panels show the target geometry, combinatorial GRIP, weighted GRIP, weighted GRIP with LGKK polish, and weighted igraph KK.", fig.alt="Five consistently projected panels compare the target saddle mesh with combinatorial GRIP, weighted GRIP, LGKK-polished weighted GRIP, and weighted igraph Kamada-Kawai. The low-loss polished layout remains visibly flatter than the target.", fig.width=11.2, fig.height=6.8, fig.pos="H"}
weighted.layouts.aligned <- list(
  "Target geometry" = weighted.mesh$coords_surface,
  "Combinatorial GRIP" = align_to_reference(
    coords.combinatorial, weighted.mesh$coords_surface
  ),
  "Weighted GRIP" = align_to_reference(
    coords.weighted, weighted.mesh$coords_surface
  ),
  "Weighted GRIP + LGKK" = align_to_reference(
    coords.weighted.lgkk, weighted.mesh$coords_surface
  ),
  "Weighted KK (igraph)" = align_to_reference(
    coords.weighted.kk, weighted.mesh$coords_surface
  )
)
weighted.projected <- lapply(
  weighted.layouts.aligned,
  project.3d,
  azimuth = 35,
  elevation = 22
)
all.projected <- do.call(rbind, weighted.projected)
xpad <- 0.05 * diff(range(all.projected[, 1]))
ypad <- 0.05 * diff(range(all.projected[, 2]))
common.xlim <- range(all.projected[, 1]) + c(-xpad, xpad)
common.ylim <- range(all.projected[, 2]) + c(-ypad, ypad)
weighted.colors <- c("#666666", "black", "#1F3B73", "#7A2E2E", "#3A6B35")

op <- par(mfrow = c(2, 3), mar = c(1, 1, 3, 1))
for (i in seq_along(weighted.projected)) {
  plot_common_projection(
    weighted.projected[[i]], weighted.mesh$edges,
    names(weighted.projected)[[i]],
    xlim = common.xlim, ylim = common.ylim,
    vertex.col = weighted.colors[[i]]
  )
}
plot.new()
par(op)
```

## Real-data network at package scale

```{r hmp-data, include=FALSE}
data(hmp.gc)
hmp.edges <- edge.matrix.from.adj(hmp.gc$adj_list)
hmp.info <- hmp.gc$graph_info
hmp.data.summary <- data.frame(
  Quantity = c(
    "Explicitly labeled HMP Illumina samples",
    "Samples in the giant component",
    "Graph edges",
    "Retained taxonomic features",
    "Symmetric kNN parameter"
  ),
  Value = c(
    format(hmp.info$hmp_illumina_rows, big.mark = ","),
    format(hmp.info$original_vertices, big.mark = ","),
    format(nrow(hmp.edges), big.mark = ","),
    format(hmp.info$retained_features, big.mark = ","),
    hmp.info$selected_k
  ),
  check.names = FALSE
)
```

The package bundles a graph derived only from vaginal microbiome samples
explicitly labeled as Human Microbiome Project data [@hmp2012] and generated
by Illumina 16S rRNA amplicon sequencing. Before preprocessing, eligible
samples must have a `Project` value beginning with `HMP` and a `16S_Platform`
value equal to `Illumina`.

Taxonomic features detected in at least 1% of the retained HMP samples are
converted to relative abundance, followed by a ten-component PCA. A symmetric
$k=3$ nearest-neighbor graph is constructed in that representation: an
undirected edge is present when either endpoint occurs among the other's three
nearest neighbors. Edge weights are Euclidean distances in the PCA
representation. The largest connected component contains 4,391 of the 4,411
eligible samples and is used without coarsening.

```{r hmp-data-table}
accessible_kable(
  hmp.data.summary,
  row.names = FALSE,
  align = c("l", "l"),
  caption = "Construction summary for the bundled HMP-only Illumina 16S graph.",
  alt = paste(
    "Five rows report the eligible HMP Illumina sample count, giant-component",
    "sample and edge counts, retained feature count, and symmetric kNN k."
  ),
  position = "H"
)
```

# Interoperability and computational scope

\CRANpkg{grip} is intended for R users who need reproducible 2D or 3D graph
layout, candidate comparison, and metric-aware embedding diagnostics. The
stable GRIP layout core is single-threaded; selected experimental GMDS
refinements can use up to two compiled-engine threads.

The assessment functions accept coordinate matrices produced outside
\CRANpkg{grip}. The bundled benchmark applies a common scoring function to
Fruchterman--Reingold [@fruchterman1991] and DrL layouts from
\CRANpkg{igraph} [@igraph2026], stress majorization [@gansner2005] from
\CRANpkg{graphlayouts}, and a
default hop-metric GRIP layout on the same HMP-only graph. All four methods are
run without edge weights, and `score.layout()` uses hop distance. This isolates
topology-oriented layout and avoids treating PCA distances as attraction
strengths in packages whose weight conventions differ. The weighted workflow
in the preceding section evaluates edge-length geometry separately.

## Runtime and shared scoring at package scale

The same script benchmarks four methods on the bundled HMP graph. Each
two-dimensional layout call is repeated five times with its random-number seed reset to 1;
Table \@ref(tab:hmp-runtime-benchmark) reports the median elapsed wall time and
interquartile range. Garbage collection is performed before, rather than
during, each timed call, and layout scoring is excluded from the reported
time. The coordinates from the first repeat are evaluated by the common
scoring function.

```{r hmp-runtime-benchmark}
hmp.scaled <- merge(
  benchmark.results$hmp$timing,
  benchmark.results$hmp$scores,
  by = "method",
  sort = FALSE
)
hmp.scaled <- hmp.scaled[, c(
  "method", "n.runs", "elapsed.sec.median", "elapsed.sec.iqr",
  "sampled.stress", "edge.length.cv",
  "sampled.nonedge.sep.ratio"
)]

paper_kable(
  hmp.scaled,
  caption = paper_table_caption(
    paste0(
      "Repeated two-dimensional topology-only layout benchmark on the bundled HMP graph (n = ",
      format(benchmark.results$hmp$n, big.mark = ","), ", m = ",
      format(benchmark.results$hmp$m, big.mark = ","), ")."
    ),
    "Runs = repeated layout calls; MedSec = median elapsed seconds; IQR = interquartile range; Stress = sampled stress; EdgeCV = edge-length coefficient of variation; NESR = sampled non-edge separation ratio."
  ),
  alt = paste(
    "Four rows compare repeated runtime and common quality scores for",
    "Fruchterman-Reingold, DrL, stress majorization, and default grip on the",
    "4,391-vertex HMP graph. All methods and scores use the unweighted topology."
  ),
  digits = 3
)
```

The timing column measures candidate generation, while the adjacent columns
show what the common scoring interface reports for every candidate. The
methods optimize different objectives, so the table is read as a runtime and
diagnostic profile rather than a single ranking.
The run was performed on `r benchmark.results$benchmark_metadata$hardware_model` with
`r format(as.numeric(benchmark.results$benchmark_metadata$memory_bytes) / 2^30, digits = 3)`
GiB of memory and CPU `r benchmark.results$benchmark_metadata$cpu`, using
`r benchmark.results$benchmark_metadata$r_version`,
`r benchmark.results$benchmark_metadata$platform`, and the R framework
reference BLAS. The R-relative BLAS path is retained in the provenance record.

## Computational boundaries

The bundled example contains the full 4,391-vertex giant component of the
HMP-only symmetric kNN graph; no graph coarsening is used. The stable multiscale
layout can be applied to larger sparse graphs, and `prepare.edge.kk()` supports
edge-only diagnostics and repair without constructing an all-pairs path cache.
By contrast, metric MDS, full geodesic-KK preparation, and complete
graph-geodesic diagnostic panels require shortest-path information whose time
and memory costs can dominate the layout solve. Larger quadform and
data-derived graph studies therefore belong in a separate, systematically
designed GMDS evaluation rather than as unsupported scaling claims in this
package article.

# Reproducibility and availability

This article is written with \CRANpkg{rmarkdown} [@allaire2026rmarkdown],
evaluated with \CRANpkg{knitr} [@xie2025knitr], and rendered using the
\CRANpkg{rjtools} output format [@oharawild2026rjtools] currently recommended
by the journal. The tables and the small-graph figures are produced from
evaluated code in the manuscript. The larger HMP case study, weighted
external comparison, and repeated scaled benchmark use bundled precomputed
`.rds` artifacts so that the paper renders quickly during review.

The submission supplement contains the exact precomputed artifacts used here,
their generation scripts, the HMP-only edge and sample tables, the feature
screening manifest, and a provenance record with file checksums.
`BENCHMARK_PROVENANCE.md` records the
timing protocol, hardware, software versions, and boundary between layout and
scoring time. `build-hmp-only-graph.R` applies the HMP/Illumina cohort filter
before feature screening and graph construction, while
`precompute-vs-alternatives.R` recreates the weighted and cross-package
benchmarks. The final HMP graph and all paper results derived from it are
independently inspectable and reproducible from the supplied files. The two
combined upstream abundance and metadata tables are not redistributed and no
public acquisition procedure for those exact tables is currently supplied;
therefore, the supplement does not independently regenerate the HMP graph from
primary abundance data. The graph-construction script documents and executes
that upstream transformation only when those tables are supplied separately.

The version evaluated in this article is \CRANpkg{grip} 0.2.0, available from CRAN at
<https://CRAN.R-project.org/package=grip>. The development repository, issue
tracker, manuscript source, and artifact-generation scripts are available at
<https://github.com/pgajer/grip>. The HMP example data and the precomputed
results used here are distributed with the package.

# Discussion

In the data-to-graph-to-embedding picture introduced at the beginning of the
article, \CRANpkg{grip} operationalizes the second transformation. Its stable
interface generates and traces multiscale layouts, while its scoring functions
allow candidates from \CRANpkg{grip} or other packages to be evaluated on common
terms. The structured, weighted, and HMP-only workflows illustrate this package
role without turning the article into a general comparison of embedding
methods.

The graph-geodesic layer extends that workflow beyond conventional layout
diagnostics. Full geodesic Kamada--Kawai, LGKK, MISF-GKK, metric MDS, and
edge-KK provide experimental baselines, fixed-path objectives, and refinement
strategies for investigating how a supplied graph metric is represented after
embedding. They are a research sandbox within the package, not a completed
general GMDS method.

The distinction between the two fidelity questions is also a limitation on
interpretation. A low graph-geodesic loss shows agreement with the supplied
graph and prepared path family; it does not establish that the graph itself
faithfully represents the geometry of the original data. That upstream
question must be assessed during graph construction, for example with the
companion \CRANpkg{dgraphs} infrastructure. Within \CRANpkg{grip}, all-pairs path
preparation can become more costly than the multiscale layout, the preset
library cannot cover every graph family, exact crossing counts are expensive
on dense 2D graphs, and the current experimental interface provides no
automatic rule for selecting among full, landmark, edge-only, and multiscale
strategies.

Future work will connect the two transformations more systematically: choosing
graphs by explicit data-to-graph losses, comparing graph-to-embedding losses
over controlled graph families and sizes, and determining when improvements in
graph-geodesic fidelity preserve scientifically relevant data geometry. The
current package provides the software infrastructure needed to formulate and
reproduce those investigations while keeping methodological claims separate
from the stable package contribution.

# Acknowledgments

The GRIP algorithm was originally developed in collaboration with Stephen G.
Kobourov and Michael T. Goodrich. The HMP microbial network data were derived
from publicly available Human Microbiome Project resources. Research reported
in this publication was supported by grant number INV-048956 from the Gates
Foundation.
