---
title: "grip: Multiscale Graph Layout and Geodesic Embedding Tools in R"
author:
  - name: Pawel Gajer
    affiliation: Center for Advanced Microbiome Research and Innovation (CAMRI), Institute for Genome Sciences, Department of Microbiology and Immunology, University of Maryland School of Medicine
    address: Baltimore, Maryland, USA
    orcid: 0000-0002-0974-7798
    email: pgajer@gmail.com
  - name: Jacques Ravel
    affiliation: Center for Advanced Microbiome Research and Innovation (CAMRI), Institute for Genome Sciences, Department of Microbiology and Immunology, University of Maryland School of Medicine
    address: Baltimore, Maryland, USA
    orcid: 0000-0002-0851-2233
    email: jravel@som.umaryland.edu
abstract: >
  A graph embedding places the vertices of a graph in Euclidean space. However,
  the resulting geometry need not faithfully represent the graph's intrinsic
  shortest-path geometry. grip is an R package for generating,
  evaluating, comparing, and refining graph embeddings. Its multiscale layout
  functions generate 2D and 3D candidate coordinates and can record the
  intermediate stages of a coarse-to-fine layout. A common scoring framework uses
  conventional
  layout diagnostics and graph-geodesic losses to evaluate graph embeddings
  produced by grip or other software. These losses compare embedded graph-path
  lengths with shortest-path distances in the input graph and support
  investigation of geodesic multidimensional scaling (GMDS), which seeks
  embeddings that preserve graph-geodesic geometry. The package also provides
  experimental GMDS tools, including a refinement workflow that initializes
  coordinates using metric MDS and then applies edge-restricted
  Kamada--Kawai (edge-KK) optimization to edge-length discrepancies. Within the
  two-stage transformation from data to graph to embedding, the supported
  grip() and trace.grip() interfaces serve general multiscale-layout use;
  the GMDS methods are experimental tools for studying graph-to-embedding
  fidelity.
preamble: |
  \usepackage{amsmath}
  \usepackage{graphicx}
  \usepackage{xcolor}
  \usepackage{etoolbox}
  \usepackage{float}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \graphicspath{{build/}{}}
  \AtBeginEnvironment{table}{\small}
  \AtBeginEnvironment{thebibliography}{\small}
  \providecommand{\manuscriptversion}{unversioned}
  \providecommand{\manuscriptbuildnumber}{NA}
  \providecommand{\manuscriptcommit}{unknown}
  \providecommand{\manuscriptbuilddatetime}{unknown}
  \newif\ifgrippapershowbuildstamp
  \grippapershowbuildstampfalse
  \InputIfFileExists{manuscript_build_info.tex}{}{%
    \InputIfFileExists{build/manuscript_build_info.tex}{}{}%
  }
  \newcommand{\grippaperbuildstamp}{%
    \par
    \vspace{-0.35em}%
    \begin{center}%
    \scriptsize\color{gray!75}%
    Version \texttt{\manuscriptversion}\enspace | \enspace
    Build \texttt{\manuscriptbuildnumber}\enspace | \enspace
    Commit \texttt{\manuscriptcommit}\\
    Built \manuscriptbuilddatetime
    \end{center}%
    \vspace{0.35em}%
  }
  \makeatletter
  \ifgrippapershowbuildstamp
    \apptocmd{\maketitle}{\grippaperbuildstamp}{}{}
  \fi
  \makeatother
date: "2026-08-28"
journal:
  firstpage: 1
  lastpage: ~
slug: gajer-grip
bibliography: grip-software-paper.bib
output:
  rjtools::rjournal_article:
    self_contained: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)

if (!requireNamespace("grip", quietly = TRUE) ||
    utils::packageVersion("grip") < "0.2.0") {
  stop("This article requires grip 0.2.0 or later.")
}
if (!requireNamespace("dgraphs", quietly = TRUE) ||
    utils::packageVersion("dgraphs") < "0.2.0") {
  stop("This article requires dgraphs 0.2.0 or later.")
}
if (!requireNamespace("igraph", quietly = TRUE)) {
  stop("This article requires igraph to render graph-path examples.")
}
suppressPackageStartupMessages(library(grip))

paper.table.colnames <- c(
  graph = "Graph",
  candidate = "Cand",
  method = "Meth",
  dim = "Dim",
  `n.vertices` = "nV",
  `n.edges` = "nE",
  `n.runs` = "Runs",
  `elapsed.sec.mean` = "Sec",
  `elapsed.sec.median` = "MedSec",
  `elapsed.sec.iqr` = "IQR",
  `score.composite` = "Comp",
  `sampled.stress` = "Stress",
  `sampled.stress.mean` = "Stress",
  `edge.length.cv` = "EdgeCV",
  `edge.length.cv.mean` = "EdgeCV",
  `sampled.nonedge.sep.ratio` = "NESR",
  `sampled.nonedge.sep.ratio.mean` = "NESR",
  `stability.procrustes.mean` = "PStab",
  `edge.crossings` = "Cross",
  `gkk.weighted.rel.rmse` = "GKK Rel",
  `gkk.mean.rel.path.error` = "Path Rel"
)

paper_table_caption <- function(title, abbreviations) {
  paste0(title, " Abbrev.: ", abbreviations)
}

html_attribute_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}

accessible_kable <- function(x,
                             caption,
                             alt,
                             row.names = FALSE,
                             align = NULL,
                             col.names = NULL,
                             digits = getOption("digits"),
                             position = "H") {
  args <- list(
    x = x,
    digits = digits,
    col.names = col.names,
    row.names = row.names,
    align = align,
    caption = caption,
    position = position
  )
  args <- Filter(Negate(is.null), args)
  if (knitr::is_html_output()) {
    args$table.attr <- sprintf(
      'aria-label="%s"',
      html_attribute_escape(alt)
    )
  }
  do.call(knitr::kable, args)
}

paper_kable <- function(x, caption, alt, digits = 3) {
  col.names <- unname(ifelse(
    names(x) %in% names(paper.table.colnames),
    paper.table.colnames[names(x)],
    names(x)
  ))
  align <- c("l", rep("r", max(0, ncol(x) - 1L)))
  accessible_kable(
    x,
    digits = digits,
    col.names = col.names,
    row.names = FALSE,
    align = align,
    caption = caption,
    alt = alt,
    position = "H"
  )
}

find.extdata.file <- function(...) {
  rel.path <- file.path(...)
  candidates <- c(
    file.path("reproducibility", "precomputed", rel.path),
    system.file("extdata", rel.path, package = "grip"),
    file.path("..", "..", "..", "..", "inst", "extdata", rel.path),
    file.path("..", "..", "..", "inst", "extdata", rel.path),
    file.path("inst", "extdata", rel.path)
  )
  path <- candidates[file.exists(candidates)][1L]
  if (!length(path) || is.na(path) || !nzchar(path)) {
    stop("could not locate ", rel.path)
  }
  path
}

read.extdata.csv <- function(file.name) {
  utils::read.csv(find.extdata.file(file.name), stringsAsFactors = FALSE)
}

read.extdata.rds <- function(...) {
  readRDS(find.extdata.file(...))
}

plot_layout_panel <- function(coords,
                              edges,
                              main,
                              projection = NULL,
                              vertex.col = "black",
                              edge.col = "gray82",
                              cex = 0.5,
                              ...) {
  plot.layout(
    coords, edges,
    projection = projection,
    vertex.col = vertex.col,
    edge.col = edge.col,
    pch = 16,
    cex = cex,
    main = main,
    ...
  )
}

align_to_reference <- function(coords, reference) {
  centered <- sweep(coords, 2, colMeans(coords), "-", check.margin = FALSE)
  target <- sweep(reference, 2, colMeans(reference), "-", check.margin = FALSE)
  rotation.svd <- svd(crossprod(centered, target))
  rotation <- rotation.svd$u %*% t(rotation.svd$v)
  scale <- sum(rotation.svd$d) / sum(centered^2)
  sweep(scale * centered %*% rotation, 2, colMeans(reference), "+", check.margin = FALSE)
}

plot_common_projection <- function(coords,
                                   edges,
                                   main,
                                   xlim,
                                   ylim,
                                   vertex.col = "black",
                                   cex = 0.55) {
  plot(
    coords[, 1], coords[, 2],
    type = "n", asp = 1, axes = FALSE,
    xlab = "", ylab = "", xlim = xlim, ylim = ylim,
    main = main
  )
  segments(
    coords[edges[, 1], 1], coords[edges[, 1], 2],
    coords[edges[, 2], 1], coords[edges[, 2], 2],
    col = "gray82"
  )
  points(coords[, 1], coords[, 2], pch = 16, cex = cex, col = vertex.col)
}

edge.matrix.from.adj <- function(adj.list) {
  edges <- list()
  idx <- 0L
  for (u in seq_along(adj.list)) {
    nbrs <- adj.list[[u]][adj.list[[u]] > u]
    for (v in nbrs) {
      idx <- idx + 1L
      edges[[idx]] <- c(u, v)
    }
  }
  do.call(rbind, edges)
}
```

# Introduction

Multidimensional scaling (MDS) begins with pairwise dissimilarities
$\Delta=(\delta_{ij})$ and seeks coordinates
$Z=\{z_1,\ldots,z_n\}\subset\mathbb{R}^m$ whose Euclidean endpoint distances
approximate them. Stress-based formulations quantify the discrepancy between
the target dissimilarities and the configuration distances
[@kruskal1964multidimensional]. The choice of target distances matters when
$n$ observations $X=\{x_1,\ldots,x_n\}\subset\mathbb{R}^p$ lie on a curved,
branching, or otherwise non-Euclidean structure: ambient straight-line
distances may then fail to represent the intrinsic relationships that an
analysis is intended to preserve. Those unobserved intrinsic relationships have
distance matrix $D_X=(d_{ij}^X)$. Isomap constructs a weighted neighborhood
graph $G=(V,E,\ell)$ whose vertices correspond to the observations, uses its
shortest-path distance matrix $D_G=(d_{ij}^G)$ as an estimate of $D_X$, and
applies classical MDS to $D_G$ [@tenenbaum2000].

For real data, $D_X$ is generally unknown, so the first fidelity question is
whether $D_G$ is an adequate approximation to it. In controlled synthetic
examples, the generating model can make $D_X$ known and thereby allow the
discrepancy between $D_X$ and $D_G$ to be evaluated directly. Examining that
discrepancy across geometries, sampling densities, noise levels, and sample
sizes supports calibration of graph constructors for corresponding classes of
real data.

The graph creates a second fidelity question. An embedding assigns each vertex
of $G$ a coordinate $z_i\in\mathbb{R}^m$; in the common
dimensionality-reduction setting illustrated below, $p\gg m$. The matrix
$D_{G,Z}=(d_{ij}^{G,Z})$ records the lengths, after embedding, of selected
geodesics from the input graph. Chord-based MDS criteria compare $D_G$ with
the endpoint chords $d_{ij}^{Z}=\lVert z_i-z_j\rVert_2$. Thus, although
Isomap uses graph geodesics to construct its input targets, this view of
output fidelity remains Euclidean: it does not measure the embedded length of
the corresponding input-side geodesic. The two transformations and the roles
of the companion packages are

```{r two-stage-fidelity-diagram, echo=FALSE, results="asis"}
if (knitr::is_latex_output()) {
  cat(r"(
\begin{center}
\resizebox{0.92\linewidth}{!}{%
\begin{tikzpicture}[>=stealth]
  \node (Rp) at (0,1.20) {$\mathbb{R}^{p}$};
  \node at (7.4,1.20) {$p\gg m$};
  \node (Rm) at (14.8,1.20) {$\mathbb{R}^{m}$};

  \node (X) at (0,0) {$X$};
  \node (G) at (7.4,0) {$G$};
  \node (Z) at (14.8,0) {$Z$};

  \draw[right hook->] (X) -- (Rp);
  \draw[right hook->] (Z) -- (Rm);

  \draw[->] (X) -- (G)
    node[midway, above=0.06cm, align=center, font=\scriptsize,
         inner sep=0pt, text depth=0.22em]
      {\texttt{dgraphs}\\[-0.16em] construction and assessment}
    node[midway, below=0.10cm, font=\scriptsize, inner sep=0pt]
      {$D_X\ \text{versus}\ D_G$};

  \draw[->] (G) -- (Z)
    node[midway, above=0.06cm, align=center, font=\scriptsize,
         inner sep=0pt, text depth=0.22em]
      {\texttt{grip}\\[-0.16em] layout, scoring, and refinement}
    node[midway, below=0.10cm, font=\scriptsize, inner sep=0pt]
      {$D_G\ \text{versus}\ D_{G,Z}$};
\end{tikzpicture}%
}
\end{center}
)"
  )
} else {
  cat(r"{
<div style="margin: 1.2em auto; max-width: 920px;">
<svg viewBox="0 0 1200 170" role="img"
     aria-label="Data-to-graph and graph-to-embedding fidelity diagram"
     style="display: block; width: 100%; height: auto;">
  <defs>
    <marker id="grip-arrowhead" markerWidth="9" markerHeight="7"
            refX="8" refY="3.5" orient="auto" markerUnits="strokeWidth">
      <path d="M0,0 L9,3.5 L0,7" fill="none" stroke="currentColor"
            stroke-width="1.2"/>
    </marker>
  </defs>

  <g fill="currentColor" stroke="currentColor" stroke-width="1.3"
     font-family="serif">
    <path d="M48,111 L579,111" fill="none"
          marker-end="url(#grip-arrowhead)"/>
    <path d="M621,111 L1152,111" fill="none"
          marker-end="url(#grip-arrowhead)"/>
    <path d="M27,95 C27,102 20,102 20,95 L20,34" fill="none"
          marker-end="url(#grip-arrowhead)"/>
    <path d="M1179,95 C1179,102 1172,102 1172,95 L1172,34" fill="none"
          marker-end="url(#grip-arrowhead)"/>

    <g stroke="none" text-anchor="middle">
      <text x="20" y="22" font-size="22">ℝ<tspan dy="-7" font-size="14">p</tspan></text>
      <text x="600" y="22" font-size="22" font-style="italic">p ≫ m</text>
      <text x="1172" y="22" font-size="22">ℝ<tspan dy="-7" font-size="14">m</tspan></text>

      <text x="20" y="119" font-size="24" font-style="italic">X</text>
      <text x="600" y="119" font-size="24" font-style="italic">G</text>
      <text x="1172" y="119" font-size="24" font-style="italic">Z</text>

      <text x="310" y="74" font-size="15" font-family="monospace">dgraphs</text>
      <text x="310" y="95" font-size="16">construction and assessment</text>
      <text x="890" y="73" font-size="15" font-family="monospace">grip</text>
      <text x="890" y="94" font-size="16">layout, scoring, and refinement</text>

      <text x="310" y="140" font-size="17" font-style="italic">D<tspan baseline-shift="sub" font-size="12">X</tspan><tspan baseline-shift="baseline" font-style="normal"> versus </tspan>D<tspan baseline-shift="sub" font-size="12">G</tspan></text>
      <text x="890" y="140" font-size="17" font-style="italic">D<tspan baseline-shift="sub" font-size="12">G</tspan><tspan baseline-shift="baseline" font-style="normal"> versus </tspan>D<tspan baseline-shift="sub" font-size="12">G,Z</tspan></text>
    </g>
  </g>
</svg>
</div>
}"
  )
}
```

The first comparison asks how well a data-derived graph recovers the intended
data geometry. The companion \pkg{dgraphs} package supports that
data-to-graph question through graph construction and calibration tools
[@gajer2026dgraphs]. Once a graph construction has been calibrated for a class
of data---or a set of plausible constructions has been retained---the second
comparison asks which coordinate representations preserve the resulting graph
geometry with the least distortion. The \pkg{grip} package addresses this
graph-to-embedding question. A low graph-to-embedding loss cannot recover
information already lost during $X\to G$, so the two losses remain separate
and the interpretation of the second is conditional on the first.

The package supports two complementary views of the graph-to-embedding stage.
Chord-based MDS criteria compare $D_G$ with Euclidean distances between
embedded vertices. Geodesic multidimensional scaling (GMDS) instead compares
$D_G$ with the lengths of corresponding input-graph paths after embedding.
\pkg{grip} provides a common interface for scoring
coordinates produced by its own layout functions or by other software under
both views and supplements them with conventional layout diagnostics. It also
provides experimental GMDS-facing objectives and refinement methods.

The core layout functions implement the original GRIP algorithm [@gajer2002;
@gajer2004] in a C++ core accessed from R through Rcpp [@eddelbuettel2011].
GRIP constructs a hierarchy of maximal independent sets, places vertices from
coarse to fine resolution, and refines the layout locally at each level. This
hierarchy makes it practical to generate and compare multiple 2D and 3D
candidate layouts for graphs with hundreds to thousands of vertices. The
original GRIP refinement is computationally localized rather than global:
intermediate filtration levels use Kamada--Kawai (KK)-style forces over
restricted neighborhoods, and the final level uses Fruchterman--Reingold
refinement [@gajer2002; @kamada1989; @fruchterman1991]. GRIP therefore draws on
both KK and Fruchterman--Reingold, but it does not optimize the classical global
all-pairs KK objective.

Figure \@ref(fig:trace) illustrates the coarse-to-fine construction using four
states returned by `trace.grip()`.

```{r trace, fig.cap="Coarse-to-fine construction in the GRIP algorithm, exposed by trace.grip() for a 6x6 mesh. The panels show the coarsest initialization, an early intermediate state, a late intermediate state, and the final layout.", fig.alt="Four panels showing the GRIP algorithm placing and refining a mesh from a sparse coarse configuration to a regular grid-like final arrangement.", fig.width=10, fig.height=2.8, fig.pos="H"}
trace.edges <- edges.mesh(6, 6)
trace.solve <- trace.grip(
  trace.edges, n = 36, dim = 2,
  preset = "mesh", seed = 31,
  metric = "hop",
  trace = "level"
)

frame.idx <- unique(c(
  1L,
  max(2L, floor(length(trace.solve$frames) / 3)),
  max(3L, floor(2 * length(trace.solve$frames) / 3)),
  length(trace.solve$frames)
))

labels <- c("Coarsest start", "Early refinement", "Late refinement", "Final layout")
op <- par(mfrow = c(1, 4), mar = c(1, 1, 2.5, 1))
for (i in seq_along(frame.idx)) {
  coords <- trace.solve$frames[[frame.idx[i]]]
  active <- complete.cases(coords[, 1:2])
  xy <- coords[active, 1:2, drop = FALSE]
  active.edges <- trace.edges[active[trace.edges[, 1]] & active[trace.edges[, 2]], , drop = FALSE]

  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- max(0.08 * diff(xlim), 0.2)
  ypad <- max(0.08 * diff(ylim), 0.2)

  plot(xy[, 1], xy[, 2], type = "n", asp = 1, axes = FALSE,
       xlab = "", ylab = "",
       xlim = xlim + c(-xpad, xpad),
       ylim = ylim + c(-ypad, ypad),
       main = labels[i])
  if (nrow(active.edges) > 0) {
    segments(
      coords[active.edges[, 1], 1], coords[active.edges[, 1], 2],
      coords[active.edges[, 2], 1], coords[active.edges[, 2], 2],
      col = "gray82"
    )
  }
  points(xy[, 1], xy[, 2], pch = 16, cex = 0.6)
}
par(op)
```

The original formulation can also construct layouts in $\mathbb{R}^m$ and
project them to two or three dimensions; Gajer et al. reported that these
projections were often smoother and more symmetric than layouts computed
directly in the display dimension [@gajer2004]. The main layout and trace
interfaces use $m\in\{2,3\}$, whereas `weighted.grip.nd()` and selected
GMDS-facing functions support $m>3$. Retaining higher-dimensional coordinates
allows the package diagnostics to be used in searches for low-distortion
Euclidean representations of $D_G$. Such coordinates can support downstream
methods defined on Euclidean vectors and, when $m$ is smaller than the original
feature dimension, provide a graph-mediated form of dimensionality reduction.
This fidelity remains conditional on the supplied graph and does not establish
recovery of $D_X$.

GRIP belongs to a broader multilevel graph-drawing tradition. Related methods
construct coarse graph hierarchies by matching, clustering, or graph-distance
sampling and then refine layouts from coarse to fine resolution
[@harel2002; @walshaw2003; @hachul2005; @hu2005]. Within the R ecosystem,
\CRANpkg{igraph} and \CRANpkg{graphlayouts} implement broad collections of
graph-layout algorithms [@csardi2006; @igraph2026; @schoch2023], while
\CRANpkg{ggraph} provides additional layouts and a common visualization
interface to algorithms from both packages [@pedersen2025].

The contribution of \pkg{grip} is an R implementation of the original
multiscale GRIP layout extended to allow supplied edge lengths, rather than a
fixed unit edge length, to influence candidate generation [@gajer2002]. The
package also provides reproducible assessment of coordinates generated by
\pkg{grip} or other software. Its experimental GMDS layer adds fixed-path
losses and refinement methods for studying graph-geodesic fidelity.

# Package overview

This article evaluates \pkg{grip} 0.2.0 and requires
\pkg{dgraphs} 0.2.0. Matching source releases are available at
<https://github.com/pgajer/grip/tree/v0.2.0> and
<https://github.com/pgajer/dgraphs/tree/v0.2.0> while their CRAN
submissions are pending CRAN review; once accepted, the stable releases will be
available from <https://CRAN.R-project.org/package=grip> and
<https://CRAN.R-project.org/package=dgraphs>. The companion
\pkg{dgraphs} package constructs and calibrates data-derived graphs,
whereas \pkg{grip} generates, compares, and assesses their coordinate
representations. Extended worked examples are provided in the package
vignettes and at <https://pgajer.github.io/grip/>.

## Core layout API and supporting utilities

The package accepts graphs as two-column integer edge matrices, with an
optional edge-length vector, or as adjacency lists with optional parallel
weight lists.

The supported core layout API consists of two functions. `grip()` returns an
$n \times m$ numeric coordinate matrix, whereas `trace.grip()` returns the
final coordinates together with intermediate coarse-to-fine states and
diagnostics. The core layout and trace functions use $m\in\{2,3\}$;
`weighted.grip.nd()` and selected GMDS-facing methods support any $m\geq 2$.

The `metric` argument states which graph geometry the layout should use.
`metric = "hop"`, the default, treats every edge as one step. With
`metric = "edge_length"`, supplied edge lengths affect the optimization. The
traced interface uses the same metric choice and records intermediate states
and diagnostics for inspection or animation.

The most important user-facing parameters are layout dimension, initialization
placement, the number of multiscale and final refinement rounds, local
neighborhood size, repulsion strength, and the random seed. For a disconnected
input, `grip()` lays out each connected component separately and then places the
component layouts together in the returned coordinate matrix; it does not add
edges between components. `trace.grip()` requires a connected graph. When a
connected data-derived graph is required, several \pkg{dgraphs} constructors
can optionally add MST bridge edges between components with
`connect.components = TRUE` [@gajer2026dgraphs]. Those bridges change the graph
geometry and should be treated as an explicit construction choice.

`trace.grip()` exposes the per-level states of the coarse-to-fine layout process
illustrated in Figure \@ref(fig:trace). The package also supplies graph
generators for regular grids, trees, recursive families such as Sierpinski
carpets, and weighted geometric surfaces. These generators provide examples,
regression fixtures, and controlled benchmarks; they are not substitutes for
data-driven graph construction.

Family-specific presets provide reproducible starting values for several of
these controlled graph classes. Hop-metric presets cover meshes, carpets,
trees, and tori, while edge-length presets cover cylinders, spheres, and other
weighted geometric families. Their purpose is to make packaged examples and
targeted comparisons reproducible, not to prescribe settings for arbitrary
real-data graphs.

## Reproducibility of long-running benchmarks

All small examples, tables, and figures in this article are evaluated while the
document is rendered. The weighted saddle comparison and the repeated UMB-HMP
runtime benchmark are exceptions: recomputing them exceeds the journal's
ordinary review-time budget. The article reads the exact checked cache below;
the supplementary archive supplies the cache, its SHA-256 checksum, the
minimal inputs, and reproducibility/scripts/precompute-vs-alternatives.R for
a deferred computational review.

```{r precomputed-artifacts, echo=TRUE}
# Exact cache used by the weighted-saddle and UMB-HMP benchmark tables.
benchmark.results <- read.extdata.rds("vs_alternatives", "benchmark_results.rds")
```

The setup helpers accessible_kable(), paper_kable(), find.extdata.file(),
plot_layout_panel(), plot_common_projection(), and align_to_reference() are
manuscript-only scaffolding. In particular, align_to_reference() is used only
to place the weighted-saddle panels in a common display frame; it is not a
\pkg{grip} function and is unnecessary for using or reproducing the
package layout and scoring interfaces.

## From graph layout to graph-geodesic fidelity

Graph-geodesic scoring asks whether a selected input-graph shortest path retains
its length after embedding, so that path must remain fixed while the coordinates
change. For each evaluated pair $i<j$, the retained input-graph geodesic
$\gamma_{ij}$ realizes $d_{ij}^G$; the collection of retained routes is
$\Gamma=\{\gamma_{ij}\}$. By default, the geodesic-KK preparation functions
retain one deterministic route per pair, whereas an alternative mode averages
exactly over tied input geodesics. The embedded length of $\gamma_{ij}$ is

$$
d_{ij}^{G,Z}
= L_Z(\gamma_{ij})
= \sum_{(u,v)\in\gamma_{ij}}\lVert z_u-z_v\rVert_2.
$$

The paths are fixed by the input graph rather than recomputed as the
coordinates change. The matrix $D_{G,Z}=(d_{ij}^{G,Z})$ contains their embedded
path lengths. The corresponding endpoint chord is
$d_{ij}^Z=\lVert z_i-z_j\rVert_2$.

Kruskal's Stress-1, shown here in its metric-MDS form
[@kruskal1964multidimensional], compares target dissimilarities $\delta_{ij}$
with ambient chord distances:

$$
\operatorname{Stress}_1(\Delta, D_Z)
=
\sqrt{
  \frac{\sum_{i<j}(d_{ij}^Z-\delta_{ij})^2}
       {\sum_{i<j}(d_{ij}^Z)^2}
}.
$$

The fixed-path GMDS criterion implemented in \pkg{grip} instead compares
graph distances with the embedded lengths of the corresponding fixed
geodesics. At identity scale, its target-normalized fixed-path relative RMSE is

$$
\operatorname{RelRMSE}^{\mathrm{path}}(D_G, D_{G,Z}\mid\Gamma)
=
\sqrt{
  \frac{\sum_{i<j}(d_{ij}^{G,Z}-d_{ij}^G)^2}
       {\sum_{i<j}(d_{ij}^G)^2}
}.
$$

The difference between chord and fixed-path criteria is visible in a
three-leaf star (Figure \@ref(fig:star-path-geometry)). Placing two leaves on
opposite sides of their common center makes their endpoint chord equal the
two-edge graph distance. A noncollinear drawing can preserve the length of the
same two-edge route even when its endpoint chord is shorter. The central and
right panels therefore test different geometric quantities.

```{r star-path-geometry, fig.cap="Endpoint and fixed-path views of a three-leaf star. The blue route between vertices 1 and 2 has graph length two. A collinear placement reproduces that value as an endpoint chord (middle), whereas a branched placement preserves the two-edge route length while its endpoint chord is shorter (right).", fig.alt="Three panels show a three-leaf star, a collinear placement where the chord between two leaves equals the graph distance, and a branched placement where the highlighted two-edge route retains its length but its endpoint chord is shorter.", fig.width=10, fig.height=3.1, fig.pos="H"}
star.edges <- rbind(c(4L, 1L), c(4L, 2L), c(4L, 3L))
star.input <- rbind(c(-0.85, 0.65), c(0.85, 0.65), c(0, -0.9), c(0, 0))
star.chord <- rbind(c(-1, 0), c(1, 0), c(0, 0.9), c(0, 0))
star.branch <- rbind(
  c(0, 1),
  c(cos(7 * pi / 6), sin(7 * pi / 6)),
  c(cos(11 * pi / 6), sin(11 * pi / 6)),
  c(0, 0)
)

draw.star.panel <- function(coords, title, chord = FALSE, annotation = NULL) {
  plot(
    coords[, 1], coords[, 2], type = "n", asp = 1, axes = FALSE,
    xlab = "", ylab = "", xlim = c(-1.35, 1.35), ylim = c(-1.2, 1.25),
    main = title
  )
  segments(
    coords[star.edges[, 1], 1], coords[star.edges[, 1], 2],
    coords[star.edges[, 2], 1], coords[star.edges[, 2], 2],
    col = "gray78", lwd = 1.5
  )
  segments(
    coords[c(1L, 4L), 1], coords[c(1L, 4L), 2],
    coords[c(4L, 2L), 1], coords[c(4L, 2L), 2],
    col = "#1F77B4", lwd = 3
  )
  if (isTRUE(chord)) {
    segments(
      coords[1L, 1], coords[1L, 2], coords[2L, 1], coords[2L, 2],
      col = "#C44E52", lwd = 2, lty = 2
    )
  }
  points(coords[, 1], coords[, 2], pch = 21, bg = "white", cex = 1.35)
  text(coords[, 1], coords[, 2], labels = c("1", "2", "3", "c"), cex = 0.75)
  if (!is.null(annotation)) {
    mtext(annotation, side = 1, line = -0.2, cex = 0.78)
  }
}

op <- par(mfrow = c(1, 3), mar = c(2.1, 0.7, 2.4, 0.7))
draw.star.panel(star.input, "Input graph path", annotation = "dG(1,2) = 2")
draw.star.panel(
  star.chord, "Endpoint chord", chord = TRUE,
  annotation = "||z1-z2|| = 2"
)
draw.star.panel(
  star.branch, "Embedded fixed path", chord = TRUE,
  annotation = "LZ(gamma12) = 2"
)
par(op)
```

One family of regularized GMDS objectives addresses this indeterminacy by
adding a Euclidean dispersion term,

$$
F_\lambda(Z;G,\Gamma)
=
\sum_{i<j}(d_{ij}^{G,Z}-d_{ij}^G)^2
+\lambda R(Z),
\qquad
R(Z)=\sum_{i<j}f_r(d_{ij}^Z),
\quad \lambda>0,
$$

\noindent where $f_r$ is decreasing. Separately, the full geodesic KK (gKK) objective
applies Kamada--Kawai weights to fixed-path discrepancies:

$$
E_{\Gamma,L_0}^{\mathrm{gKK}}(Z)
=
\frac{1}{2}\sum_{i<j} k_{ij}
\bigl(d_{ij}^{G,Z}-L_0d_{ij}^G\bigr)^2,
\qquad
k_{ij}=\frac{K_0}{\max\{d_{ij}^G,\varepsilon_d\}^2}.
$$

Landmark geodesic KK (LGKK) evaluates selected local and landmark pairs rather
than all pairs. The edge-KK relaxation retains only single-edge paths:

$$
E_{K}^{\mathrm{edge}}(Z)
=
\sum_{(i,j)\in E} k_{ij}
\bigl(d_{ij}^{Z}-\ell_{ij}\bigr)^2.
$$

`score.gmds()` evaluates chord, edge, and fixed-path discrepancies for
coordinates produced by any embedding method; the full gKK, LGKK, and edge-KK
interfaces prepare, score, or refine their respective objectives.
`repulsive.state()`, `repulsive.stage()`, `edge.repulsive.state()`, and
`edge.repulsive.stage()` expose dispersion and edge-regularized components.

An arbitrary common coordinate scale should not determine a comparison of
configuration shape. For each evaluated pair $p$, the graph distance $d_p^G$,
embedded path length $d_p^{G,Z}$, and Kamada--Kawai weight
$k_p=1/(d_p^G)^2$ define the profiled gKK scale $L_0$ as the minimizer of the
weighted squared discrepancy

$$
L_0
=\underset{L\geq 0}{\operatorname{argmin}}
\sum_p k_p\bigl(d_p^{G,Z}-L d_p^G\bigr)^2.
$$

\noindent Its closed-form value is

$$
L_0 = \frac{\sum_p k_p d_p^G d_p^{G,Z}}
           {\sum_p k_p (d_p^G)^2}.
$$

## Which criterion answers which question

Chord and fixed-path criteria are not competing approximations to one common
notion of fidelity. Their difference begins with an asymmetry between the two
sides of the embedding. Isomap uses $D_G$ because ambient straight-line
distances can misrepresent the intrinsic geometry of the observations. The
output space $\mathbb{R}^m$, however, is flat and complete by construction, so
its geodesic between $z_i$ and $z_j$ is the chord
$d_{ij}^Z=\lVert z_i-z_j\rVert_2$. Chord-based stress therefore does not
assume that the input geometry is chord-like. It asks how nearly $D_G$ admits
an isometric Euclidean representation in dimension $m$; a large value is an
informative answer about the cost of flattening that graph metric.

This distinction mirrors two uses of isometry. A Riemannian isometric embedding
$f:(M,g)\rightarrow\mathbb{R}^m$ preserves the metric $g$, and thus
the lengths of curves and the intrinsic path metric of its image; it need not
preserve the ambient Euclidean chord between two image points. Chord stress
asks for the stronger metric-space representation
$d_M(x,y)=\lVert f(x)-f(y)\rVert_2$. Fixed-path relative RMSE instead evaluates
the discrete analogue of the intrinsic question: whether the embedded
$1$-skeleton, regarded as a metric graph in $\mathbb{R}^m$, retains the
lengths of selected input paths.

Figure \@ref(fig:noisy-circle-path-geometry) gives a data-derived version of
this distinction. The displayed observations themselves give the identity
embedding $z_i=x_i$ into $\mathbb{R}^2$. In that embedding, the highlighted
graph shortest path has exactly its input-graph length. More generally, because
the graph uses Euclidean edge lengths from these coordinates,
$D_{G,Z}=D_G$ and fixed-path relative RMSE is zero for this identity embedding.
The chord between the highlighted endpoints, shown in the legend, is still
shorter than their graph distance. Thus the two criteria already disagree on
one fixed graph and one embedding; neither length is defective, because they
answer different questions.
The appropriate diagnostic consequently depends on what a downstream use
consumes. Proximity, neighborhood, clustering, and centroid-distance readings
in the displayed coordinates consume chords. Trajectory and pseudotime
analyses instead use graph or lineage paths to order cells and follow changes
along those paths [@haghverdi2016; @street2018; @wolf2019]. Chord fidelity
governs the first class of readings; fixed-path fidelity governs the second.

```{r noisy-circle-path-geometry, fig.cap="A graph-to-embedding contrast for the identity embedding of a data-derived graph. Left: 120 observations sampled around a unit circle with truncated Gaussian radial offsets. Right: their native connected symmetric 6NN graph drawn at the same coordinates, so $z_i=x_i$. The white endpoints are joined by one fixed graph shortest path (blue), whose embedded length exactly equals its input-graph length, and by their Euclidean chord (dashed red); the legend gives their lengths. The display compares two metrics on the same graph and embedding; it does not establish that a chosen neighborhood graph recovers latent geometry.", fig.alt="Two panels show a noisy circular point cloud and its symmetric six-nearest-neighbor graph. In the graph panel, which uses the identity embedding, two highlighted endpoints are joined by a thick blue graph shortest path around the circle and a dashed red Euclidean chord across the center. The graph path retains its input-graph length and is longer than the chord.", fig.width=10, fig.height=4.9, fig.pos="H"}
RNGversion("4.0.0") # Preserve this illustrative realization's RNG stream.
set.seed(20260830)
noisy.n <- 120L
noisy.theta <- sort(stats::runif(noisy.n, 0, 2 * pi))
noisy.radial.offset <- pmax(
  pmin(stats::rnorm(noisy.n, sd = 0.045), 0.126),
  -0.126
)
noisy.circle <- cbind(
  x = (1 + noisy.radial.offset) * cos(noisy.theta),
  y = (1 + noisy.radial.offset) * sin(noisy.theta)
)
noisy.graph <- dgraphs::create.sknn.graph(
  noisy.circle,
  k = 6,
  neighbor.method = "exact",
  connect.components = FALSE
)
stopifnot(noisy.graph$n_components_before == 1L)
noisy.edges <- dgraphs::convert.adjacency.to.edge.matrix(
  noisy.graph$adj_list
)$edge.matrix
noisy.from <- which.min(abs(noisy.theta))
noisy.to <- which.min(abs(noisy.theta - pi))
noisy.igraph <- dgraphs::as_igraph(noisy.graph)
noisy.path.vertices <- as.integer(igraph::shortest_paths(
  noisy.igraph,
  from = noisy.from,
  to = noisy.to,
  weights = igraph::E(noisy.igraph)$weight
)$vpath[[1L]])
noisy.path.edges <- cbind(
  noisy.path.vertices[-length(noisy.path.vertices)],
  noisy.path.vertices[-1L]
)
noisy.graph.distance <- dgraphs::graph.geodesic.distances(noisy.graph)[
  noisy.from,
  noisy.to
]
noisy.chord.distance <- sqrt(sum(
  (noisy.circle[noisy.from, ] - noisy.circle[noisy.to, ])^2
))

noisy.x.limits <- c(-1.25, 1.25)
noisy.y.limits <- c(-1.40, 1.20)
op <- par(mfrow = c(1, 2), mar = c(1.0, 1.2, 1.8, 1.0))
plot(
  noisy.circle,
  asp = 1,
  xlim = noisy.x.limits,
  ylim = noisy.y.limits,
  axes = FALSE,
  xlab = "",
  ylab = "",
  pch = 16,
  cex = 0.55,
  col = "#335C81"
)
title(main = "Noisy circular observations", line = 0.2)
symbols(
  0,
  0,
  circles = 1,
  inches = FALSE,
  add = TRUE,
  fg = grDevices::adjustcolor("#9AA7B1", alpha.f = 0.75),
  lwd = 1.2
)
text(
  0,
  -1.30,
  expression(paste("truncated radial Gaussian; ", sigma, " = 0.045")),
  cex = 0.9
)

plot(
  noisy.circle,
  asp = 1,
  xlim = noisy.x.limits,
  ylim = noisy.y.limits,
  axes = FALSE,
  xlab = "",
  ylab = "",
  type = "n"
)
title(main = "Symmetric 6NN graph: path versus chord", line = 0.2)
segments(
  noisy.circle[noisy.edges[, 1L], 1L],
  noisy.circle[noisy.edges[, 1L], 2L],
  noisy.circle[noisy.edges[, 2L], 1L],
  noisy.circle[noisy.edges[, 2L], 2L],
  col = grDevices::adjustcolor("#7E8790", alpha.f = 0.36),
  lwd = 0.7
)
segments(
  noisy.circle[noisy.from, 1L],
  noisy.circle[noisy.from, 2L],
  noisy.circle[noisy.to, 1L],
  noisy.circle[noisy.to, 2L],
  col = "#C33C54",
  lwd = 2,
  lty = 2
)
segments(
  noisy.circle[noisy.path.edges[, 1L], 1L],
  noisy.circle[noisy.path.edges[, 1L], 2L],
  noisy.circle[noisy.path.edges[, 2L], 1L],
  noisy.circle[noisy.path.edges[, 2L], 2L],
  col = "#1F5A94",
  lwd = 3.4
)
points(noisy.circle, pch = 16, cex = 0.48, col = "#335C81")
points(
  noisy.circle[c(noisy.from, noisy.to), , drop = FALSE],
  pch = 21,
  cex = 1.35,
  lwd = 1.1,
  bg = "#F7F7F7",
  col = "#1A1A1A"
)
legend(
  "bottom",
  inset = 0.03,
  legend = c(
    sprintf("fixed graph shortest path = %.2f", noisy.graph.distance),
    sprintf("Euclidean chord = %.2f", noisy.chord.distance)
  ),
  col = c("#1F5A94", "#C33C54"),
  lwd = c(3.4, 2),
  lty = c(1, 2),
  bty = "n",
  cex = 0.85
)
par(op)
```

Neither criterion certifies the other. A fixed-path loss constrains sums of
edge lengths along retained routes, leaving endpoint separation and extrinsic
shape underdetermined; the star in Figure \@ref(fig:star-path-geometry) is the
minimal example, and the weighted saddle below makes the limitation visible at
larger scale. The regularized family $F_\lambda(Z;G,\Gamma)$ acknowledges this
directly: $R(Z)=\sum_{i<j}f_r(d_{ij}^Z)$ reintroduces endpoint information.
Accordingly, score.gmds() reports chord and fixed-path diagnostics on the same
prepared graph without combining them into a single score.

## Method families and function entry points

The package combines an established multiscale layout implementation with
GMDS-facing scoring and refinement methods. These components differ in maturity
and computational role. Table \@ref(tab:family-scope) maps each family to its
principal functions.

```{r family-scope, echo=FALSE}
family.scope <- data.frame(
  Family = c(
    "Multiscale GRIP",
    "Full gKK",
    "LGKK",
    "MISF gKK",
    "Edge KK",
    "Repulsive terms"
  ),
  Maturity = c(
    "Core",
    "Exp.",
    "Exp.",
    "Exp.",
    "Exp.",
    "Exp."
  ),
  Purpose = c(
    "Candidate layout and trace.",
    "All-pairs scoring and polish.",
    "Landmark scoring and polish.",
    "Multiscale scoring and polish.",
    "Edge scoring and polish.",
    "Dispersion and edge regularization."
  ),
  `Principal functions` = c(
    "grip(), trace.grip()",
    "prepare.geodesic.kk(), score.geodesic.kk(), geodesic.kk()",
    "prepare.landmark.geodesic.kk(), score.landmark.geodesic.kk(), landmark.geodesic.kk()",
    "prepare.misf.geodesic.kk(), score.misf.geodesic.kk(), misf.geodesic.kk()",
    "prepare.edge.kk(), edge.kk()",
    "repulsive.state(), repulsive.stage(), edge.repulsive.state(), edge.repulsive.stage()"
  ),
  check.names = FALSE
)

accessible_kable(
  family.scope,
  row.names = FALSE,
  align = if (knitr::is_latex_output()) {
    c("l", "l", "p{0.22\\linewidth}", "p{0.42\\linewidth}")
  } else {
    c("l", "l", "l", "l")
  },
  caption = "Maturity, purpose, and principal functions of the main layout and GMDS-facing method families in grip.",
  alt = paste(
    "Six rows distinguish the core multiscale GRIP implementation from the",
    "experimental full, landmark, MISF, edge-KK, and repulsive families and",
    "list the principal functions associated with each family."
  ),
  position = "H"
)
```

`grip()` and `trace.grip()` form the supported core layout API. The remaining
families in Table \@ref(tab:family-scope) are experimental GMDS methods and may
evolve as the research program develops.

# Assessing and refining candidate embeddings

Candidate embeddings require comparable assessment whether they come from
`grip()`, one of the package's GMDS-facing methods, or another R package. The
same diagnostics can then compare candidates across random seeds or parameter
choices and, when appropriate, pass a selected candidate to an experimental
GMDS-facing refinement method. This separation makes the package useful as an
evaluation framework rather than only as a collection of layout engines.

## General layout diagnostics

Table \@ref(tab:metrics-table) lists the general diagnostics used in this
article.

```{r metrics-table, echo=FALSE}
metrics.table <- data.frame(
  Metric = c(
    "Sampled stress",
    "Edge-length CV",
    "Sampled non-edge separation",
    "Edge crossings",
    "Procrustes stability"
  ),
  Meaning = c(
    "Chord agreement with graph distances",
    "Variation among embedded edges",
    "Spacing of sampled non-neighbors",
    "2D segment intersections",
    "Across-seed reproducibility"
  ),
  Better = c("Lower", "Lower", "Higher", "Lower", "Lower"),
  check.names = FALSE
)

accessible_kable(
  metrics.table,
  row.names = FALSE,
  align = c("l", "l", "l"),
  caption = "General layout diagnostics used in the examples.",
  alt = paste(
    "Five rows define sampled stress, edge-length variation, non-edge",
    "separation, edge crossings, and Procrustes stability, and indicate",
    "whether lower or higher values are preferred."
  ),
  position = "H"
)
```

All metrics in Table \@ref(tab:metrics-table) are returned by `score.layout()`;
`compare.layouts()` applies the same diagnostics across repeated candidate
layouts.

All-pairs scoring can be unnecessarily expensive for a general layout
diagnostic, so sampled stress uses the unique unordered pairs $S$ obtained by
uniform sampling. For a sampled pair $p\in S$, $d_p^G$ is its graph distance
and $d_p^Z$ its Euclidean layout distance. The implementation fits the scale
$a_S=\sum_{p\in S}d_p^Zd_p^G/\sum_{p\in S}(d_p^G)^2$ and reports the
scale-invariant discrepancy

$$
\operatorname{Stress}_S =
\sqrt{\frac{\sum_{p\in S}(d_p^Z-a_Sd_p^G)^2}
                 {\sum_{p\in S}(d_p^Z)^2}}.
$$

The examples use 2,000 sampled pairs unless stated otherwise. Edge-length CV is
the standard deviation divided by the mean of embedded edge lengths. Sampled
non-edge separation is the minimum distance among up to 5,000 sampled
non-edges divided by the median embedded edge length. Edge crossings count
strict intersections between nonincident segments in 2D; `"auto"` computes
them for at most 1,000 edges, `"always"` overrides that threshold, and
`"never"` skips the calculation. Procrustes stability is the mean vertex RMSE
over all unordered pairs of seeded layouts (45 pairs for ten seeds), after
centering each layout, scaling it to unit maximum radius, and applying the
optimal orthogonal rotation or reflection. The paper reports the component
diagnostics rather than a composite rank because their relative importance
depends on the analysis.
Because sampled non-edge separation depends on graph size, the number of
sampled non-edges, and the sampled pairs, it should be compared only for the
same graph under the same sampling protocol; it is not a graph-independent
quality scale.

## Graph-geodesic diagnostics and refinement

The general diagnostics above assess layout quality and endpoint-chord
agreement. When the quantity of interest is the length of graph paths after
embedding, the package instead fixes an input graph and evaluates its
graph-geodesic metric. Table \@ref(tab:assessment-functions) places the
principal functions in a generate--assess--refine workflow.

```{r assessment-functions, echo=FALSE}
assessment.functions <- data.frame(
  Task = c(
    "Generate candidate",
    "Assess layout",
    "Prepare graph paths",
    "Assess geodesic fidelity",
    "Prepare edge targets",
    "Evaluate regularization terms",
    "Refine candidate"
  ),
  Function = c(
    "grip()",
    "score.layout(), compare.layouts()",
    "prepare.geodesic.kk()",
    "score.gmds(), score.geodesic.kk()",
    "prepare.edge.kk()",
    "repulsive.state(), edge.repulsive.state()",
    "geodesic.kk(), landmark.geodesic.kk(), edge.kk()"
  ),
  check.names = FALSE
)

accessible_kable(
  assessment.functions,
  row.names = FALSE,
  align = c("l", "l"),
  caption = "Package functions in the candidate generation, assessment, and refinement workflow.",
  alt = paste(
    "Seven rows map multiscale generation, general scoring, all-pairs path",
    "preparation, graph-geodesic scoring, lightweight edge preparation, and",
    "GMDS-facing regularization and refinement to grip functions."
  ),
  position = "H"
)
```

`score.gmds()` supplies a common diagnostic panel for any coordinate matrix on
a prepared graph. With an all-pairs preparation it reports edge fidelity,
target-normalized fixed-path and chord relative RMSE, and distance-band
summaries. With the lightweight object returned by
`prepare.edge.kk()`, only edge diagnostics are available; the all-pairs
quantities are reported as unavailable. This distinction matters in practice:
edge-only assessment and repair avoid a dense shortest-path matrix, whereas
the full path-based diagnostics retain substantially more information and have
correspondingly higher memory and runtime costs.

For an edge-only refinement, `prepare.edge.kk()` avoids all-pairs path
construction before `edge.kk()` returns refined coordinates and a diagnostic
panel. The executable example below follows that prepare--score--refine path.

```{r geodesic-workflow, echo=TRUE, results="hide"}
workflow.edges <- edges.mesh(4, 4)
workflow.grid <- as.matrix(expand.grid(
  x = seq(-1, 1, length.out = 4),
  y = seq(-1, 1, length.out = 4)
))
initial <- cbind(
  1.45 * workflow.grid[, 1] + 0.22 * workflow.grid[, 2]^2,
  0.55 * workflow.grid[, 2] + 0.10 * workflow.grid[, 1]
)

prepared <- prepare.edge.kk(workflow.edges, n = 16)
before <- score.gmds(initial, prepared = prepared)
fit <- edge.kk(
  coords = initial,
  prepared = prepared,
  density_mix_schedule = 1,
  max_iter = 8,
  return_trace = FALSE
)
after <- score.gmds(fit$coords, prepared = prepared)
```

Figure \@ref(fig:edge-kk-workflow) shows the coordinates produced by this
example before and after refinement. Because the edge-only diagnostic profiles
out a common scale, the panels are centered and scaled only for display; the
figure illustrates the resulting configuration and its edge-length fidelity,
rather than selecting a unique polygonal shape.

```{r edge-kk-workflow, fig.cap=sprintf("Illustrative output of the edge-KK workflow. Eight refinement iterations reduce profiled edge relative RMSE from %.3f to %.3f by equalizing the embedded mesh-edge lengths. The edge-only objective does not select a unique shape, so this deterministic example demonstrates the workflow rather than a general performance comparison.", before$edge.rel.rmse, after$edge.rel.rmse), fig.alt="Two panels show the 16-vertex 4-by-4 mesh from the executable edge-KK workflow. The initial coordinates form a compressed, curved grid, while refinement produces a more regular grid with nearly equal embedded edge lengths. Each panel gives the corresponding profiled edge relative RMSE.", fig.width=7.5, fig.height=3.4, fig.pos="H"}
display.coords <- function(coords) {
  centered <- sweep(coords, 2, colMeans(coords), "-", check.margin = FALSE)
  centered / sqrt(mean(rowSums(centered^2)))
}

workflow.layouts <- list(
  "Initial mesh coordinates" = display.coords(initial),
  "After edge-KK refinement" = display.coords(fit$coords)
)
workflow.scores <- c(before$edge.rel.rmse, after$edge.rel.rmse)
workflow.range <- range(do.call(rbind, workflow.layouts))
workflow.pad <- 0.12 * diff(workflow.range)

op <- par(mfrow = c(1, 2), mar = c(1, 1, 3.1, 1))
for (i in seq_along(workflow.layouts)) {
  plot_common_projection(
    workflow.layouts[[i]], workflow.edges,
    sprintf("%s\nEdge relative RMSE = %.3f", names(workflow.layouts)[i], workflow.scores[i]),
    xlim = workflow.range + c(-workflow.pad, workflow.pad),
    ylim = workflow.range + c(-workflow.pad, workflow.pad),
    vertex.col = if (i == 1L) "#666666" else "#1F3B73"
  )
}
par(op)
```

# Worked package workflows

## Linking the two fidelity stages

A variable-density unit circle makes the two fidelity stages observable because
its intrinsic arc-distance matrix $D_X$ is known. The example draws 60
unevenly spaced points, constructs symmetric $k$-nearest-neighbor graphs with
the companion \pkg{dgraphs} package, and compares $D_X$ with each graph
distance matrix $D_G$. For every graph, \pkg{grip} then compares a weighted
GRIP layout and metric MDS using the same graph-to-embedding diagnostics. The
two losses remain separate rather than being combined into a single score.

```{r paired-fidelity-example}
set.seed(271)
n.circle <- 60L
circle.parameter <- sort(stats::runif(n.circle)^1.65)
circle.angle <- 2 * pi * circle.parameter
circle.data <- cbind(cos(circle.angle), sin(circle.angle))
circle.arc <- outer(circle.angle, circle.angle, function(a, b) abs(a - b))
D.circle <- pmin(circle.arc, 2 * pi - circle.arc)

paired.rows <- lapply(c(5L, 7L, 9L), function(k) {
  graph <- dgraphs::create.sknn.graph(
    circle.data,
    k = k,
    neighbor.method = "exact",
    connect.components = TRUE
  )
  stopifnot(graph$n_components_before == 1L, nrow(graph$mst_edge_matrix) == 0L)
  D.graph <- dgraphs::graph.geodesic.distances(graph)
  graph.loss <- dgraphs::isometry.rel.rms.error(D.graph, D.circle)
  prepared <- prepare.graph.geodesic.mds(
    adj_list = graph$adj_list,
    weight_list = graph$weight_list
  )

  layouts <- list(
    "Weighted GRIP" = grip(
      adj_list = graph$adj_list,
      weight_list = graph$weight_list,
      n = n.circle,
      dim = 2,
      metric = "edge_length",
      seed = 19
    ),
    "Metric MDS" = metric.mds(prepared = prepared, dim = 2)$coords
  )

  do.call(rbind, lapply(names(layouts), function(method) {
    score <- score.gmds(layouts[[method]], prepared = prepared)
    data.frame(
      k = k,
      `X-G Rel` = graph.loss,
      Embedding = method,
      `G-Z Path` = score$gmds.stress,
      `G-Z Chord` = score$metric.chord.stress,
      check.names = FALSE
    )
  }))
})
paired.fidelity <- do.call(rbind, paired.rows)

accessible_kable(
  paired.fidelity,
  digits = 3,
  row.names = FALSE,
  align = c("r", "r", "l", "r", "r"),
  caption = paper_table_caption(
    paste(
      "Separate data-to-graph and graph-to-embedding losses for one",
      "controlled variable-density circle realization. All three graphs",
      "are natively connected and require no repair edge."
    ),
    paste(
      "X-G Rel = target-normalized data-to-graph relative RMSE;",
      "G-Z Path = fixed-path relative RMSE; G-Z Chord = chord relative RMSE."
    )
  ),
  alt = paste(
    "Six rows report separate data-to-graph and graph-to-embedding losses for",
    "three symmetric nearest-neighbor graphs, each embedded with weighted GRIP",
    "and metric MDS."
  ),
  position = "H"
)
```

This is a framework demonstration, not a benchmark of either constructor or
embedding method. In this realization, the $k=5$ graph is connected but fails
to close the sparsely sampled part of the circle, producing a large
data-to-graph error. Metric MDS represents that supplied graph with small
downstream loss, but the low value cannot recover the circular relationship
already lost upstream. The $k=7$ and $k=9$ graphs approximate $D_X$ much more
closely, while the graph-to-embedding columns still distinguish how their
geometry is represented. The two losses are therefore retained as separate,
conditional diagnostics rather than added or multiplied. A substantive study
would repeat this calculation over a prespecified calibration family and
multiple realizations, as illustrated in the companion \pkg{dgraphs}
article.

## Supporting graph families and preset behavior

The graph generators and presets support reproducible examples and regression
tests. Table \@ref(tab:preset-benchmark) illustrates their behavior on two
controlled structured families, each run across ten seeds. The results are
generated directly in the paper build; they do not provide a prescription for
selecting parameters on real-data graphs.

```{r preset-benchmark}
mesh.edges <- edges.mesh(12, 12)
mesh.compare <- compare.layouts(
  edges = mesh.edges,
  n = 144,
  dim = 2,
  candidates = c("default", "mesh"),
  seeds = 1:10,
  edge.crossings = "always"
)

carpet.edges <- edges.sierpinski.carpet(3)
carpet.n <- max(carpet.edges)
carpet.compare <- compare.layouts(
  edges = carpet.edges,
  n = carpet.n,
  dim = 2,
  candidates = c("default", "carpet"),
  seeds = 1:10,
  edge.crossings = "never"
)

summary.cols <- c(
  "candidate", "n.runs", "elapsed.sec.mean", "sampled.stress.mean",
  "edge.length.cv.mean", "stability.procrustes.mean"
)

preset.benchmark <- rbind(
  transform(mesh.compare$summary[, summary.cols], graph = "12x12 mesh"),
  transform(carpet.compare$summary[, summary.cols], graph = "Carpet level 3")
)

preset.benchmark <- preset.benchmark[, c(
  "graph", "candidate", "n.runs", "elapsed.sec.mean",
  "sampled.stress.mean", "edge.length.cv.mean",
  "stability.procrustes.mean"
)]
preset.benchmark$candidate <- factor(
  preset.benchmark$candidate,
  levels = c("default", "mesh", "carpet")
)
preset.benchmark <- preset.benchmark[order(preset.benchmark$graph, preset.benchmark$candidate), ]

paper_kable(
  preset.benchmark,
  caption = paper_table_caption(
    "Preset benchmark on two structured graph families, each averaged over ten seeds.",
    "Runs = number of seeds; Sec = mean elapsed seconds; Stress = sampled stress; EdgeCV = edge-length coefficient of variation; PStab = Procrustes stability."
  ),
  alt = paste(
    "Four rows compare default and family-specific presets on a mesh and",
    "Sierpinski carpet using runtime, stress, edge variation, and stability",
    "metrics."
  ),
  digits = 3
)
```

For these two controlled families, the package-supplied preset reduces sampled
stress and edge-length variation and produces more reproducible layouts across
seeds. PStab is itself the across-seed dispersion summary used for this
comparison. The result is family-specific evidence for the supplied defaults, not a
claim that one parameter profile is preferable for arbitrary graphs. Figure
\@ref(fig:carpet-figure) shows the corresponding carpet layouts.

```{r carpet-figure, fig.cap="Level-3 Sierpinski carpet laid out with the default settings (left) and the carpet preset (right). The preset yields the lower-stress and more stable solution reported in the preceding benchmark table.", fig.alt="Two panels showing a Sierpinski carpet graph. The default layout is more distorted, while the preset layout better preserves the recursive holes.", fig.width=10, fig.height=4, fig.pos="H"}
carpet.default.coords <- grip(carpet.edges, n = carpet.n, dim = 2, seed = 12)
carpet.preset.coords <- grip(carpet.edges, n = carpet.n, dim = 2, preset = "carpet", seed = 12)

op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
plot_layout_panel(
  carpet.default.coords, carpet.edges, "Default", cex = 0.3,
  axes = FALSE, xlab = "", ylab = ""
)
plot_layout_panel(
  carpet.preset.coords, carpet.edges, "Carpet preset", cex = 0.3,
  axes = FALSE, xlab = "", ylab = ""
)
par(op)
```

## Fixed weighted graph: graph-geodesic assessment

An edge-length metric matters when a graph's topology is simple but its geometry
is not. A $5 \times 5$ mesh whose edge lengths are induced by a curved saddle
surface provides that contrast: its topology is unchanged, but the intended
metric is no longer the flat-grid metric.

The comparison includes combinatorial and weighted GRIP layouts, the weighted
layout followed by six experimental LGKK polish rounds, and the weighted
three-dimensional Kamada--Kawai implementation in \CRANpkg{igraph}. All candidates
use seed 1 and otherwise retain their package defaults. The graph is fixed
before the candidates are generated, and `score.geodesic.kk()` evaluates their
embedded fixed-path lengths against the same weighted graph metric. The example
therefore illustrates the graph-to-embedding workflow; it does not evaluate
how well the graph itself represents an underlying data set.

The coordinates and scores used in this comparison are read from a
precomputed artifact so that the article can be rendered within the journal's
review-time budget. The supplementary archive provides both the artifact and
the `precompute-vs-alternatives.R` script used to generate it.

```{r weighted-summary}
weighted.saddle <- benchmark.results$weighted_saddle
weighted.mesh <- list(
  edges = weighted.saddle$edges,
  n = weighted.saddle$n,
  edge_weights = weighted.saddle$edge_weights,
  coords_surface = weighted.saddle$target_coords
)

coords.combinatorial <- weighted.saddle$layouts[["Combinatorial GRIP"]]
coords.weighted <- weighted.saddle$layouts[["Weighted GRIP"]]
coords.weighted.lgkk <- weighted.saddle$layouts[["Weighted GRIP + LGKK polish"]]
coords.weighted.kk <- weighted.saddle$layouts[["Weighted KK (igraph)"]]

gkk.cols <- c(
  "gkk.weighted.rel.rmse",
  "gkk.mean.rel.path.error"
)

weighted.summary <- weighted.saddle$scores[, c("method", gkk.cols)]

paper_kable(
  weighted.summary,
  caption = paper_table_caption(
    "Weighted saddle-surface example scored with geodesic-KK criteria.",
    "GKK Rel = weighted relative RMSE; Path Rel = mean relative path error."
  ),
  alt = paste(
    "Four rows compare combinatorial GRIP, weighted GRIP, weighted GRIP with",
    "LGKK polish, and weighted igraph Kamada-Kawai using geodesic-KK scores."
  ),
  digits = 3
)
```

The scale-normalized columns are comparable across methods whose coordinate
scales differ. The LGKK-polished candidate is optimized against the same
fixed-path objective used for scoring, so its low value is an in-sample
optimization result rather than independent validation. The comparison with
weighted KK shows how the package can apply that criterion to a layout that
was not optimized for it.

Figure \@ref(fig:weighted-figure) also exposes a limitation of the criterion.
Preserving the weighted graph paths does not uniquely determine the target's
extrinsic curvature: a comparatively flat realization can have low path loss.
The panels are similarity-aligned to the target and use one projection and one
coordinate scale, so this difference is not a viewing-axis artifact.

```{r weighted-figure, fig.cap="Weighted mesh example after similarity alignment to the target and display with a common orthographic projection and scale. Panels show the target geometry, combinatorial GRIP, weighted GRIP, weighted GRIP with LGKK polish, and weighted igraph KK.", fig.alt="Five consistently projected panels compare the target saddle mesh with combinatorial GRIP, weighted GRIP, LGKK-polished weighted GRIP, and weighted igraph Kamada-Kawai. The low-loss polished layout remains visibly flatter than the target.", fig.width=11.2, fig.height=6.8, fig.pos="H"}
weighted.layouts.aligned <- list(
  "Target geometry" = weighted.mesh$coords_surface,
  "Combinatorial GRIP" = align_to_reference(
    coords.combinatorial, weighted.mesh$coords_surface
  ),
  "Weighted GRIP" = align_to_reference(
    coords.weighted, weighted.mesh$coords_surface
  ),
  "Weighted GRIP + LGKK" = align_to_reference(
    coords.weighted.lgkk, weighted.mesh$coords_surface
  ),
  "Weighted KK (igraph)" = align_to_reference(
    coords.weighted.kk, weighted.mesh$coords_surface
  )
)
weighted.projected <- lapply(
  weighted.layouts.aligned,
  project.3d,
  azimuth = 35,
  elevation = 22
)
all.projected <- do.call(rbind, weighted.projected)
xpad <- 0.05 * diff(range(all.projected[, 1]))
ypad <- 0.05 * diff(range(all.projected[, 2]))
common.xlim <- range(all.projected[, 1]) + c(-xpad, xpad)
common.ylim <- range(all.projected[, 2]) + c(-ypad, ypad)
weighted.colors <- c("#666666", "black", "#1F3B73", "#7A2E2E", "#3A6B35")

op <- par(mfrow = c(2, 3), mar = c(1, 1, 3, 1))
for (i in seq_along(weighted.projected)) {
  plot_common_projection(
    weighted.projected[[i]], weighted.mesh$edges,
    names(weighted.projected)[[i]],
    xlim = common.xlim, ylim = common.ylim,
    vertex.col = weighted.colors[[i]]
  )
}
plot.new()
par(op)
```

## Real-data network at package scale

```{r hmp-data, include=FALSE}
data(hmp.gc)
hmp.edges <- edge.matrix.from.adj(hmp.gc$adj_list)
hmp.info <- hmp.gc$graph_info
hmp.data.summary <- data.frame(
  Quantity = c(
    "Explicitly labeled HMP Illumina samples",
    "Samples in the giant component",
    "Graph edges",
    "Retained taxonomic features",
    "Symmetric kNN parameter"
  ),
  Value = c(
    format(hmp.info$hmp_illumina_rows, big.mark = ","),
    format(hmp.info$original_vertices, big.mark = ","),
    format(nrow(hmp.edges), big.mark = ","),
    format(hmp.info$retained_features, big.mark = ","),
    hmp.info$selected_k
  ),
  check.names = FALSE
)
```

The package bundles a graph derived from the Ravel-led University of Maryland
Baltimore Human Microbiome Project (UMB-HMP), a longitudinal study in which
135 nonpregnant women of reproductive age were enrolled and asked to
self-collect vaginal swabs daily for ten weeks [@ravel2013daily]. This HMP
demonstration-project cohort is distinct from the NIH HMP healthy-reference
cohort. Later work identifies the longitudinal data as UMB-HMP and describes
the available profiles as V3--V4 16S rRNA data [@lee2023equilibrium]. The
processed working tables used here contain later Illumina profiles: before
preprocessing, eligible samples must have a `Project` value beginning with
`HMP` and a `16S_Platform` value equal to `Illumina`. Thus, the cited studies
establish the cohort and assay lineage; the supplementary input tables record
the exact 4,411 profiles used to construct the graph.

Taxonomic features detected in at least 1% of the retained HMP samples are
converted to relative abundance, followed by a ten-component PCA. A symmetric
$k=3$ nearest-neighbor graph is constructed in that representation: an
undirected edge is present when either endpoint occurs among the other's three
nearest neighbors. Edge weights are Euclidean distances in the PCA
representation. The largest connected component contains 4,391 of the 4,411
eligible samples and is used without coarsening.

The supplementary archive contains the minimal UMB-HMP-only feature-count and
technical metadata tables, the graph-construction script, the feature manifest,
checksums, and provenance records needed to reconstruct this graph. It also
contains the precomputed repeated-layout benchmark used below and the script
that generates it.

```{r hmp-data-table}
accessible_kable(
  hmp.data.summary,
  row.names = FALSE,
  align = c("l", "l"),
  caption = "Construction summary for the bundled UMB-HMP-only Illumina 16S graph.",
  alt = paste(
    "Five rows report the eligible HMP Illumina sample count, giant-component",
    "sample and edge counts, retained feature count, and symmetric kNN k."
  ),
  position = "H"
)
```

Figure \@ref(fig:hmp-layouts) compares two topology-only layouts of the same
graph. Vertices are colored by their supplied community-state-type (CST) labels
only to make the displayed graph easier to inspect; the layouts were computed
without those labels.

```{r hmp-layouts, fig.cap="Topology-only layouts of the bundled UMB-HMP graph, colored by supplied community-state-type labels. The left panel shows default hop-metric GRIP and the right panel stress majorization. Each coordinate matrix is centered and scaled independently for display; CST labels are not layout inputs.", fig.alt="Two panels show the 4,391-vertex UMB-HMP graph with vertices colored by seven community-state-type labels. The left panel is a default hop-metric GRIP layout and the right panel is a stress-majorization layout. Both show the same color legend and graph topology.", fig.width=10, fig.height=4.5, fig.pos="H"}
hmp.cst <- as.character(hmp.gc$vertex_data$cst)
hmp.cst.levels <- c("I", "II", "III", "IV-A", "IV-B", "IV-C", "V")
hmp.cst.colors <- c(
  "I" = "#1B9E77", "II" = "#D95F02", "III" = "#7570B3",
  "IV-A" = "#E7298A", "IV-B" = "#66A61E", "IV-C" = "#E6AB02",
  "V" = "#A6761D"
)
hmp.layouts <- list(
  "Default GRIP (hop)" = benchmark.results$hmp$layouts[["grip default (hop)"]],
  "Stress majorization" = benchmark.results$hmp$layouts[["Stress (graphlayouts)"]]
)
hmp.display <- lapply(hmp.layouts, function(coords) {
  centered <- sweep(coords, 2, colMeans(coords), "-", check.margin = FALSE)
  centered / max(sqrt(rowSums(centered^2)))
})
hmp.coordinates <- do.call(rbind, hmp.display)
hmp.xlim <- range(hmp.coordinates[, 1])
hmp.ylim <- range(hmp.coordinates[, 2])
hmp.xpad <- 0.04 * diff(hmp.xlim)
hmp.ypad <- 0.04 * diff(hmp.ylim)

op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
for (i in seq_along(hmp.display)) {
  plot_common_projection(
    hmp.display[[i]], hmp.edges, names(hmp.display)[i],
    xlim = hmp.xlim + c(-hmp.xpad, hmp.xpad),
    ylim = hmp.ylim + c(-hmp.ypad, hmp.ypad),
    vertex.col = unname(hmp.cst.colors[hmp.cst]),
    cex = 0.34
  )
  legend(
    "bottomleft", legend = hmp.cst.levels, col = hmp.cst.colors[hmp.cst.levels],
    pch = 16, cex = 0.72, bty = "n", ncol = 2
  )
}
par(op)
```

# Interoperability and computational scope

R users can apply the same assessment to coordinates generated by
\pkg{grip} or by another layout package. The core GRIP layout
implementation is single-threaded; selected experimental GMDS refinements can
use up to two compiled-engine threads.

The assessment functions accept coordinate matrices produced outside
\pkg{grip}. The bundled benchmark applies a common scoring function to
Fruchterman--Reingold [@fruchterman1991] and DrL layouts from
\CRANpkg{igraph} [@igraph2026], stress majorization [@gansner2005] from
\CRANpkg{graphlayouts}, and a
default hop-metric GRIP layout on the same UMB-HMP-only graph. All four methods are
run without edge weights, and `score.layout()` uses hop distance. This isolates
topology-oriented layout and avoids treating PCA distances as attraction
strengths in packages whose weight conventions differ. The weighted workflow
in the preceding section evaluates edge-length geometry separately.

## Runtime and shared scoring at package scale

Candidate-generation time and shared layout diagnostics answer different
questions, so the bundled HMP script reports both for four methods. Each
two-dimensional layout call is repeated five times with its random-number seed
reset to 1; Table \@ref(tab:hmp-runtime-benchmark) reports the median elapsed
wall time and interquartile range. Garbage collection is performed before,
rather than during, each timed call, and layout scoring is excluded from the
reported time. The common scoring function evaluates the coordinates from the
first repeat.

```{r hmp-runtime-benchmark}
hmp.scaled <- merge(
  benchmark.results$hmp$timing,
  benchmark.results$hmp$scores,
  by = "method",
  sort = FALSE
)
hmp.scaled <- hmp.scaled[, c(
  "method", "n.runs", "elapsed.sec.median", "elapsed.sec.iqr",
  "sampled.stress", "edge.length.cv",
  "sampled.nonedge.sep.ratio"
)]

paper_kable(
  hmp.scaled,
  caption = paper_table_caption(
    paste0(
      "Repeated two-dimensional topology-only layout benchmark on the bundled HMP graph (n = ",
      format(benchmark.results$hmp$n, big.mark = ","), ", m = ",
      format(benchmark.results$hmp$m, big.mark = ","), ")."
    ),
    "Runs = repeated layout calls; MedSec = median elapsed seconds; IQR = interquartile range; Stress = sampled stress; EdgeCV = edge-length coefficient of variation; NESR = sampled non-edge separation ratio, comparable here because every row uses the same graph and sampling protocol."
  ),
  alt = paste(
    "Four rows compare repeated runtime and common quality scores for",
    "Fruchterman-Reingold, DrL, stress majorization, and default grip on the",
    "4,391-vertex HMP graph. All methods and scores use the unweighted topology."
  ),
  digits = 3
)
```

The timing column measures candidate generation, while the adjacent columns
show what the common scoring interface reports for every candidate. The
methods optimize different objectives, so the table is read as a runtime and
diagnostic profile rather than a single ranking.
The run was performed on `r benchmark.results$benchmark_metadata$hardware_model` with
`r format(as.numeric(benchmark.results$benchmark_metadata$memory_bytes) / 2^30, digits = 3)`
GiB of memory and CPU `r benchmark.results$benchmark_metadata$cpu`, using
`r benchmark.results$benchmark_metadata$r_version`,
`r benchmark.results$benchmark_metadata$platform`, and the R framework
reference BLAS. The R-relative BLAS path is retained in the provenance record.

## Computational boundaries

The bundled example contains the full 4,391-vertex giant component of the
UMB-HMP-only symmetric kNN graph; no graph coarsening is used. This article
demonstrates the current R implementation at that scale and does not provide a
larger within-package scalability benchmark. The original GRIP studies
considered substantially larger graphs than the examples in this article
[@gajer2002; @gajer2004]. Within the R package, however, all-pairs graph
distances, geodesic path preparation, repeated scoring, and exact crossing
counts can become limiting steps before the multiscale layout itself.
`prepare.edge.kk()` therefore supports edge-only diagnostics and repair without
constructing an all-pairs path cache.
By contrast, metric MDS, full geodesic-KK preparation, and complete
graph-geodesic diagnostic panels require shortest-path information whose time
and memory costs can dominate the layout computation. Larger quadform and
data-derived graph studies therefore belong in a separate, systematically
designed GMDS evaluation rather than as unsupported scaling claims in this
package article.

# Discussion

A low graph-geodesic loss establishes agreement with the supplied graph and
prepared path family; it does not establish that the graph faithfully represents
the geometry of the original data. That upstream fidelity question belongs to
graph construction, for example through the companion \pkg{dgraphs}
infrastructure. When a synthetic calibration family leaves several graph
constructors or parameter settings plausible, embedding assessment should be
repeated over that set rather than conditioned on a single nominal winner. Such
transfer remains conditional on how well the synthetic family represents the
topology, sampling process, and noise structure of the application.

Within that boundary, \pkg{grip} operationalizes the graph-to-embedding
stage. Its core layout functions generate and trace multiscale layouts, while
its scoring functions evaluate coordinates from \pkg{grip} or other
packages on common terms. Full geodesic Kamada--Kawai, LGKK, MISF-GKK, metric
MDS, and edge-KK provide experimental baselines, fixed-path objectives, and
refinement strategies for investigating how a supplied graph metric is
represented after embedding. They are a research sandbox within the package,
not a completed general GMDS method.

The scope also has computational and methodological limits. All-pairs path
preparation can become more costly than the multiscale layout, graph-family
presets are illustrative rather than exhaustive, and exact crossing counts are
expensive on dense 2D graphs. The current GMDS-facing methods provide no
automatic rule for selecting among full, landmark, edge-only, and multiscale
strategies. The structured, weighted, and UMB-HMP-only workflows illustrate
these roles and limits without turning the article into a general comparison of
embedding methods.

The tables and small-graph figures are evaluated from code in the manuscript.
The weighted comparison and repeated HMP benchmark use precomputed artifacts
to keep the article within the journal's review-time budget. The supplementary
archive supplies those artifacts and their generation scripts, together with
the minimal UMB-HMP-only inputs, graph builder, manifest, checksums, and provenance
records needed to reproduce the reported results.

Future work will connect the two transformations more systematically through a
two-dimensional experimental design: graph constructors and settings along one
axis, and embedding objectives and algorithms along the other. Evaluating this
design over controlled calibration families and multiple realizations can show
which conclusions are stable to uncertainty in the graph rather than merely
optimal for one selected construction. The current package provides the
software infrastructure needed to formulate and reproduce those investigations
while keeping methodological claims separate from the established package
contribution.

# Acknowledgments

The GRIP algorithm was originally developed in collaboration with Stephen G.
Kobourov and Michael T. Goodrich. The bundled microbial-network example was
derived from the UMB-HMP longitudinal vaginal microbiome cohort
[@ravel2013daily; @lee2023equilibrium]. Research reported in this publication
was supported by grant number INV-048956 from the Gates Foundation.
