---
title: "Embedding Fermat distances on a paraboloid, saddle and helix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Embedding Fermat distances on a paraboloid, saddle and helix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment='#>',fig.width=8,fig.height=5,warning=FALSE,message=FALSE)
available <- all(vapply(c('dgraphs','igraph','plotly'),requireNamespace,logical(1),quietly=TRUE))
available <- available && file.exists(system.file('examples','fermat-surfaces.R',package='dgraphs')) &&
  'return.graph' %in% names(formals(dgraphs::fermat.distances)) &&
  'distance.matrix' %in% names(formals(grip::metric.mds))
knitr::opts_chunk$set(eval=available)
```

```{r dependencies,echo=FALSE,eval=TRUE,results='asis'}
if(!available) cat('This executable example requires development versions of dgraphs and grip with the direct distance and supporting-graph APIs, plus igraph and plotly. Install these dependencies to render the calculated examples.\n')
```

This example fits three-dimensional metric MDS directly to **dense complete-graph
Fermat distance matrices**. No kNN graph is required. We then deliberately restrict
paths to symmetric Fermat kNN graphs with k from 3 through 10 and compare their
embeddings with the dense-target fits. The companion dgraphs vignette explains the
distance estimator and supporting path graphs.

## Sample surfaces and supply a distance matrix

We use the paraboloid $(u,v,u^2+v^2)$ and saddle $(u,v,u^2-v^2)$ over the unit
disk. The existing dgraphs sampler draws uniformly with respect to **surface
area**, using the area element of the graph surface; we do not replace it with
uniform sampling of the two parameter coordinates.

```{r dense-example}
library(grip)
s <- dgraphs::sample.synthetic.geometry(
  dgraphs::synthetic.quadform(2,3,forms=list(diag(c(1,-1)))),
  dgraphs::synthetic.sampling.quadform.lab('disk',mode='area'),
  n=150,seed=4101)
X <- s$predictors
D <- dgraphs::fermat.distances(points=X,p=2,backend='implicit',rooted=FALSE)
fit <- metric.mds(distance.matrix=D,dim=3,backend='sgd',init='random',
  seed=3051,max.iter=300,diagnostics=FALSE,
  sgd.control=list(final.rate=1e-5))
Y <- fit$coords
```

For observations $X_i$, the supplied target is
$D_p(i,j)=\min_{\pi:i\leadsto j}\sum_{(a,b)\in\pi}\|X_a-X_b\|^p$.
The minimum allows every observation as an intermediate point. The exponent is
applied to individual ambient steps, and the returned distance is unrooted.
The MDS coordinates $Y_i\in\mathbb R^3$ seek to minimize an unweighted sum of squared
residuals $(\|Y_i-Y_j\|-D_p(i,j))^2$ over distinct pairs. The fixed 300-epoch SGD
budget is an optimization budget, not a convergence certificate.

## A matched comparison

The comparison uses powers 1, 2, 3 and 4. Every fit uses the same seeded random starting configuration for retained
samples, with the initial global scale fitted separately to its target matrix, uses uniform pair weights and the
same optimization controls. For a disconnected kNN graph only its largest
component is embedded. All within-component graph shortest-path distances are
supplied as a dense matrix; this isolates **graph restriction**, without adding
a landmark approximation to this small example.

```{r shared, include=FALSE}
source(Sys.getenv('FERMAT_SURFACE_HELPER',
  unset=system.file('examples','fermat-surfaces.R',package='dgraphs')))
cache.root <- Sys.getenv('FERMAT_SURFACE_RESULTS')
results <- lapply(c('paraboloid','saddle','helix'),function(shape) {
  path <- file.path(cache.root,paste0(shape,'_4101.rds'))
  if(nzchar(cache.root) && file.exists(path)) {
    cached<-readRDS(path)
    stopifnot(nrow(cached$reference$pairs)==1000L,
      shape=='helix' || cached$reference$method=='best_of_six')
    return(cached)
  }
  s <- fermat.surface.sample(shape,150L,4101L)
  fermat.surface.experiment(s,fermat.surface.reference(s))
})
metrics <- do.call(rbind,lapply(results,`[[`,'records'))
```

We keep three questions separate. **Distance error** compares the input distances
with reference surface geodesics. **Graph restriction error** compares graph
shortest paths with complete-graph Fermat distances. **Embedding error** compares
Euclidean distances in the fitted coordinates with their supplied targets.
For vectors of estimated distances $d$ and reference distances $g$, we report
\[
E(d,g;a)=\sqrt{\frac{\sum_r(ad_r-g_r)^2}{\sum_r g_r^2}}.
\]
Here $r$ indexes valid comparison pairs. For distance-versus-geodesic comparisons,
$a=\sum_r d_rg_r/\sum_r d_r^2$ fits one global scale because different powers have
different units. For embedding and graph restriction error, $a=1$: the targets
and their estimates are in the same units. Smaller errors are better; zero means
agreement on those pairs. Reference geodesics use existing continuous solvers
on 1,000 sampled pairs using `best_of_six` for both surfaces, with
`paraboloid_clairaut` on the same paraboloid pairs. Each of the six refinement
searches receives 64 rounds; no global-optimality certificate is implied.

```{r comparison-table, echo=FALSE}
knitr::kable(metrics[metrics$k==0,c('shape','p','distance_nrmse','embedding_nrmse',
 'embedded_geodesic_nrmse')],digits=3,col.names=c('Example','p',
 'Input vs geodesic','Embedding vs input','Embedding vs geodesic'))
```

Table 1. Dense-target fits. The two geodesic columns allow a global scale change;
the embedding-versus-input column does not. A good MDS fit can faithfully reproduce
an inaccurate geodesic estimator, and these columns distinguish those cases.

```{r k-comparison, echo=FALSE}
old<-par(mfrow=c(1,3),mar=c(4,4,2,1))
for(shape in unique(metrics$shape)) {
 z<-metrics[metrics$shape==shape,]
 plot(c(0,10),c(0,1.3*max(c(0,z$embedded_geodesic_nrmse),na.rm=TRUE)),type='n',
 xlab='k (0 = complete targets)',ylab='Embedding vs geodesic error',main=shape)
 for(p in 1:4){q<-z[z$p==p,];lines(q$k,q$embedded_geodesic_nrmse,col=p,pch=p,type='b')}
 legend('topright',legend=paste('p =',1:4),col=1:4,lty=1,pch=1:4,bty='n',cex=.7)
};par(old)
```

**Figure 1.** Scaled agreement between embedded Euclidean distances and the
intrinsic reference. The point at k=0 is the complete-target baseline; points at
k=3--10 are restricted graphs. Connecting segments help follow powers and do not
represent fits at untested k values. Disconnected results use fewer pairs:

```{r coverage,echo=FALSE}
z<-metrics[metrics$components>1,]
knitr::kable(aggregate(cbind(retained,components,reference_pairs)~shape+p,z,range))
```

The table summarizes ranges across disconnected k values. Complete coverage is
150 samples. Small within-component errors do not establish whole-sample recovery.

## Overlay embeddings and their supporting paths

The reference surface is translucent gray, original observations are gray points,
and fitted coordinates are colored by their first latent coordinate. Thin blue
lines show the selected shortest-path edges for the complete graph, or Fermat
neighbor edges for the restricted graph. Views are aligned to the observations
using least-squares similarity Procrustes alignment: translation, rotation or
reflection, and one uniform scale, matching each embedded point to its original
observation. Plot titles report the root mean squared Euclidean alignment
residual per retained sample, in target-coordinate units (zero is exact overlap).
Alignment
changes only display coordinates; numerical errors above use the original fits.

```{r paraboloid,echo=FALSE}
fermat.surface.overlay(results[[1]],p=2,k=0)
```

**Figure 2.** Dense p=2 Fermat embedding on the paraboloid. Rotate the interactive
view to inspect displacements from the gray surface.

```{r saddle,echo=FALSE}
fermat.surface.overlay(results[[2]],p=2,k=0)
```

**Figure 3.** Dense p=2 Fermat embedding on the saddle, with its supporting paths.
An intrinsic metric need not have an exact three-dimensional Euclidean realization.

```{r restricted,echo=FALSE}
fermat.surface.overlay(results[[2]],p=2,k=3)
```

**Figure 4.** The same saddle samples after restriction to a symmetric k=3 Fermat
graph. Its edges summarize potentially multistep Fermat paths; they are not the
individual powered ambient edges shown in Figure 3. If disconnected, only the
largest component appears. Edge legends report any display thinning beyond 1,800
edges; underlying graph data remain complete.

## Why the helix should unfold

The helix $(\cos t,\sin t,0.2t)$ has arc-length distance
$\sqrt{1.04}|t_i-t_j|$. Therefore its intrinsic distance matrix has an exact
one-dimensional realization. Fitting those distances in three coordinates should
produce a nearly straight line, not reconstruct the coil. Three-dimensional
metric MDS compares straight distances, not travel constrained to a curve.

```{r helix,echo=FALSE}
fermat.surface.overlay(results[[3]],p=2,k=0)
```

**Figure 5.** Dense p=2 Fermat embedding over the original helix. Agreement with the
coil is not the success criterion for intrinsic-distance reconstruction.

```{r helix-baseline,echo=FALSE}
h<-results[[3]]; b<-h$geodesic_baseline
knitr::kable(data.frame(target='Exact helix arc length',
 embedding_error=fermat.surface.score(as.matrix(dist(b$coords))[lower.tri(b$target)],
 b$target[lower.tri(b$target)],rescale=FALSE)['nrmse']),digits=6,row.names=FALSE)
plotly::plot_ly(x=b$coords[,1],y=b$coords[,2],z=b$coords[,3],type='scatter3d',
 mode='markers',marker=list(size=3,color=h$sample$latent[,1],colorscale='Viridis'))
```

**Figure 6.** Exact arc-length target baseline. Its coordinates should approach a
line; remaining transverse spread measures finite optimization error. The surface
examples do not have corresponding all-pairs exact-reference baselines: their
numerical reference checks cover selected pairs only.

## Larger datasets

The direct dense API above is suitable for small comparisons. For larger datasets,
dgraphs can compute only selected source rows with `sources = landmarks`.
`landmark.mds.constraints()` converts those distances into explicit weighted pair
fitting targets for sparse `metric.mds()`. This is a separate approximation in the
embedding objective, not a restriction of the paths defining each Fermat distance.


## Chord lengths before and after scaling

At p=1, complete-graph Fermat distances equal ambient Euclidean chord lengths.
A 3D MDS fit can therefore reproduce the sampled shape without reproducing its
intrinsic distances. The earlier distance score fits a global scale; it is not
the discrepancy of the original chord lengths.

For each of the same evaluated pairs, write $c_r$ for the chord and $g_r$ for
the intrinsic reference. The following table reports $E(c,g;1)$ and
$E(c,g;a)$ with $a=\sum c_rg_r/\sum c_r^2$, together with the median and maximum
ratio $g_r/c_r$. Errors are dimensionless and zero means agreement. The ratio
is one for equal lengths; 1.3 means the geodesic is 30% longer than its chord.
These are the 1,000 evaluated pairs for seed 4101, not all sample pairs.

```{r chord-audit,echo=FALSE}
chord.audit <- do.call(rbind,lapply(results,function(r) {
 ij<-r$reference$pairs; g<-r$reference$distances
 c<-sqrt(rowSums((r$sample$predictors[ij[,1],]-r$sample$predictors[ij[,2],])^2))
 ok<-is.finite(g)&g>0&c>0; g<-g[ok];c<-c[ok];a<-sum(c*g)/sum(c^2)
 data.frame(Example=r$sample$shape,Pairs=length(g),Scale=a,
 Raw.error=sqrt(sum((c-g)^2)/sum(g^2)),Scaled.error=sqrt(sum((a*c-g)^2)/sum(g^2)),
 Median.ratio=median(g/c),Maximum.ratio=max(g/c))
}))
knitr::kable(chord.audit,digits=4)
```

Scaling reduces chord underestimation on these patches, but does not make chords
intrinsic distances. Larger discrepancies remain for individual endpoint pairs.
On the helix, accurately preserving arc lengths unfolds the curve into a line;
matching the original coil is a different objective.

Independent solver checks are attempts, not guaranteed successes. Candidate
lengths and close agreement where both solvers succeed do not certify global
minimality. The table below makes the successful comparison count explicit.

```{r reference-success,echo=FALSE}
knitr::kable(do.call(rbind,lapply(results,function(r) {
 v<-r$reference$check.distances
 data.frame(Example=r$sample$shape,Attempted=length(v),Finite=sum(is.finite(v)),
 Failed=sum(!is.finite(v)),Check=if(r$sample$shape=='helix')'Analytic identity' else if(!is.null(r$reference$check.method))r$reference$check.method else 'None')
})))
```

## Paraboloid reference-method comparison

Best-of-six and Clairaut evaluate the same 1,000 pairs. Positive signed relative
difference $(b_r-h_r)/h_r$ means the best-of-six candidate $b_r$ is longer than
the Clairaut candidate $h_r$. Neither is assumed globally certified.

```{r clairaut-comparison,echo=FALSE}
r<-results[[1]]; b<-r$reference$distances; h<-r$reference$check.distances
ok<-is.finite(b)&is.finite(h)&h>0; delta<-(b[ok]-h[ok])/h[ok]
knitr::kable(data.frame(Pairs=sum(ok),Median.signed=median(delta),
 Maximum.absolute=max(abs(delta))),digits=6)
knitr::kable(do.call(rbind,lapply(1:4,function(p){
 d<-r$matrices[[as.character(p)]][r$reference$pairs]
 data.frame(p=p,Best.of.six=unname(fermat.surface.score(d,b)['nrmse']),
 Clairaut=unname(fermat.surface.score(d,h)['nrmse']))})),digits=5)
```

The second table compares scaled Fermat-distance errors under each reference.
It uses identical targets and endpoint pairs, isolating reference-method effects.

## Follow-up: sampling density, intermediate powers and stronger curvature

A separate distance-only experiment varied total sample size from 150 to 2,400,
with five repetitions, on paraboloids and saddles with coefficients 1, 2 and 4,
and an area-uniform Swiss roll. It reused the existing quadform area sampler.
For each surface, 64 fixed sampled endpoints supplied 500 calibration pairs and
1,000 different evaluation pairs. Additional path-support points varied across
repetitions and were nested across sample sizes. All observations could be used
by the complete-graph paths; this was not a landmark approximation.

Errors below are percentages: the square root of the ratio of summed squared
distance residuals to summed squared reference distances. Each method's single
multiplicative scale was fitted on calibration pairs only. Values are means over
five repetitions conditional on the fixed endpoints; they are new measurements,
not replacements for the original 150-point fits above. No MDS error enters them.

| Surface | Power p | Error at n=150 (%) | Error at n=2,400 (%) | Ambient error (%) |
|---|---:|---:|---:|---:|
| Paraboloid, coefficient 1 | 1.25 | 3.73 | 1.82 | 8.51 |
| Paraboloid, coefficient 1 | 2 | 11.72 | 5.67 | 8.51 |
| Saddle, coefficient 1 | 1.25 | 4.55 | 1.75 | 8.74 |
| Saddle, coefficient 1 | 2 | 14.27 | 5.58 | 8.74 |
| Saddle, coefficient 4 | 1.5 | 14.82 | 2.55 | 28.32 |
| Swiss roll | 3 | 42.33 | 7.70 | 55.84 |

Ambient error is constant because the endpoints are fixed. These results expose
both a sample-size effect and a limitation of the original integer-only power
grid: p=1.25 already beats ambient chords on the new mild-surface samples at
n=150. Larger powers are needed to suppress shortcuts on the rolled surface.
The displayed powers illustrate these effects; they are not universal tuning
recommendations.

Ordinary Euclidean kNN graph distances were more accurate on these uniform
examples at the larger sample size. For k=20 and n=2,400, errors were 0.62%,
0.49% and 0.35% on the mild paraboloid, mild saddle and Swiss roll. This separates
recovering ordinary geodesics from deliberately imposing a density-weighted
geometry.

References used best-of-six on quadforms and analytic distances on the roll,
with Clairaut sensitivity checks on every paraboloid pair. For the steepest
paraboloid, p=1.25 at n=2,400 had 4.82% error against best-of-six but 2.05% against
Clairaut. Reference-method disagreement therefore matters for this harder
surface. Increasing the quadform coefficient changes its curvature profile and
area; it does not multiply curvature uniformly over the patch. The complete
experiment, pair definitions, coverage and uncertainty are recorded in the
research experiment `fermat_density_curvature_01_oct_2026`.
