---
title: "Finding your way around grip"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Finding your way around grip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
table { width: 100%; }
td, th { vertical-align: top; }
td code { white-space: normal; overflow-wrap: anywhere; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     fig.width = 6, fig.height = 4.5)
# Installed help and pkgdown have different reference-page locations.
help_url <- function(topic) {
  prefix <- if (identical(Sys.getenv("IN_PKGDOWN"), "true")) {
    "../reference/"
  } else "../html/"
  paste0(prefix, topic, ".html")
}
```

What do you want to learn from your graph: its connectivity, its edge-length
geometry, or how a drawing compares with a known reference? Start with that
question. `grip()` computes a drawing; scoring, tracing, and graph-family
helpers support different decisions around that drawing.

## Choose a starting point

| Your question | Entry point | Follow-up |
|---|---|---|
| What does the connectivity look like? | `grip(metric = "hop")` | [Getting Started](grip-examples.html) |
| What if edges have meaningful lengths? | `grip(metric = "edge_length")` | [Weighted Graph Layouts](weighted-grip-intro.html) |
| Which settings and seeds give useful drawings? | `compare.layouts()`, then `params.from.summary()` | [Choosing Layouts for Real Data](grip-real-data.html) |
| Does my drawing preserve graph distances? | `score.layout()` or the more specialized `score.gmds()` | [Choose a score](#choose-a-score) |
| Does it recover a reference configuration? | `score.coordinates()`; `score.surface()` for triangular surfaces | [Graph examples](synthetic-graph-families.html) |
| Where did a fold or collapse appear? | `trace.grip()` | [Tracing and Diagnosing Layouts](grip-trace-and-diagnostics.html) |
| How do I construct a reproducible example? | An `edges.*()` generator or a complete `*.graph()` bundle | [Graph examples](synthetic-graph-families.html) |
| How do I browse many saved results? | `gripui_project()` and `run_gripui()` in an interactive session | [Website explorer article](https://pgajer.github.io/grip/articles/gripui-explorer.html) |

The seven linked package vignettes are installed with grip. The
[gallery](https://pgajer.github.io/grip/articles/gallery.html),
[comparison article](https://pgajer.github.io/grip/articles/grip-vs-alternatives.html),
and interactive explorer are website-only articles.

## A first drawing and diagnostic

Here 20 vertices form a rectangular grid. Sampled stress measures how well
Euclidean separation in the drawing matches shortest-path hop counts after
fitting a common scale. Zero means perfect agreement on the evaluated pairs;
it does not certify that every geometric feature or unsampled pair is correct.

```{r first-layout, fig.alt="A 20-vertex rectangular mesh drawn with the mesh preset, showing its rows and columns."}
library(grip)
edges <- edges.mesh(4, 5)
coords <- grip(edges, n = 20, dim = 2, preset = "mesh", seed = 11)
stopifnot(identical(dim(coords), c(20L, 2L)), all(is.finite(coords)))
plot.layout(coords, edges = edges, pch = 16, cex = 0.7,
            main = "A 4 by 5 mesh")
quality <- score.layout(coords, edges = edges, n = 20,
                        sample.size.stress = 200, stress.seed = 11,
                        edge.crossings = "never")
knitr::kable(quality[, c("sampled.stress", "edge.length.cv")], digits = 3)
```

The edge-length coefficient of variation (`edge.length.cv`) is the standard
deviation divided by the mean of drawn edge lengths. It describes their
uniformity, which is useful here because the graph is unweighted. Unequal
lengths can be intentional on other graphs.

### Input and output conventions

Supply an undirected edge list as a two-column matrix of **1-based integer
vertex indices**, one edge per row. Supply `n` explicitly when the highest
numbered vertex is isolated, when the edge list is empty, or when a generator
has a known vertex count. Inferring `n = max(edges)` cannot recover isolated
vertices beyond the largest endpoint. Avoid duplicate edges and self-loops;
the edge-list layout converter ignores self-loops, and duplicate treatment is
not a common contract across all helpers.

Alternatively use `adj_list`, a list of length `n`: element `i` lists the
neighbors of vertex `i`. Include both directions of each undirected edge,
with matching lengths in a parallel `weight_list`. An isolated vertex has an
empty neighbor vector (and empty length vector if weighted). Do not supply
both graph representations: this is an error, as are fractional or nonfinite
vertex indices and counts. Supply lengths with their matching representation
(`edge_weights` with `edges`, or `weight_list` with `adj_list`). Preserve the same vertex ordering in graph,
coordinates, labels, and references; row names do not establish correspondence.

`grip()` defaults to `dim = 3`; request `dim = 2` explicitly for base 2D plots.
It returns an `n` by `dim` numeric matrix. By default disconnected components
are laid out separately and packed into that matrix, including isolated
vertices. Their relative positions are display choices, not finite distances
between components. Set `disconnected = "error"` to reject such inputs.
Sampled graph-distance stress uses reachable pairs; full all-pairs geodesic
preparations require a connected graph. Split components before using them.

```{r isolated-vertex}
path.with.isolate <- grip(edges.path(5), n = 6, dim = 2, seed = 11)
stopifnot(nrow(path.with.isolate) == 6L, all(is.finite(path.with.isolate)))
```

### Hop counts or edge lengths?

The historical names `edge_weights` and `weight_list` mean **finite positive
lengths or traversal costs**, not connection strengths. With an edge list,
`edge_weights[j]` belongs to row `j`. Larger lengths request greater separation;
convert strengths to scientifically meaningful lengths before fitting.

- `metric = "hop"` uses hop counts for the standard hierarchy, neighborhoods,
  and insertion anchors. If lengths are supplied, they still set adjacent-edge
  attractive-force targets, without global normalization. Omit lengths for a
  purely unweighted baseline. Optional landmark-geodesic refinement stages
  use supplied lengths even when the standard stages use hops.
- `metric = "edge_length"` requires lengths and uses sums of those lengths
  along shortest paths throughout the multiscale engine. Its default
  `length_normalization = "median"` divides by the median length; `"mean"`
  divides by the mean and `"none"` retains their numerical scale. These
  normalization controls and `metric_neighbor_cap` belong only to this mode.
  A finite neighbor cap enables an approximate weighted search.

A graph-family constructor may already normalize lengths: inspect its
`weight_scale` and `normalize` fields before choosing solver normalization.
`score.layout()` has no `metric` argument: it uses weighted shortest paths
when lengths are supplied and hops when they are omitted, independently of
how the drawing was fitted.

## Dimensions, result objects, and methods

| Operation | Coordinate dimensions and result |
|---|---|
| `grip()`, global-repulsion variants, legacy layouts | 2 or 3 columns; coordinate matrix. |
| `weighted.grip.nd()` | At least 2 columns, including dimensions above 3; weighted multiscale layout matrix. Opt in explicitly. |
| `classical.mds()`, `metric.mds()`, `edge.kk()` | Support higher dimensions (MDS dimension must be smaller than vertex count); list of class `grip_gmds_layout`, with `$coords`, `$diagnostics`, `$metadata`, and method information. |
| `score.layout()`, full/landmark/MISF geodesic-KK, `kernel.gram.gkk()` and local-star tools | 2D or 3D workflows; do not infer higher-dimensional support from the MDS interfaces. |
| `score.gmds()`, `score.coordinates()` | Higher-dimensional matrices supported; corresponding-coordinate inputs must have identical shapes. |
| `score.surface()` | Three-column coordinates plus explicit triangular faces for each surface. |
| `trace.grip()` | List with `$final`, `$frames`, `$meta`, and `$diagnostics`; frames have 2 or 3 columns. |
| `plot.layout()`, `project.3d()` | 2D drawing or a view of the first three coordinate columns; additional columns are ignored, not fitted into a new layout. |

Use `plot.layout(coords, edges = edges)` on the ordinary matrix returned by
`grip()`. The registered [`plot.layout` method](`r help_url("plot.layout")`)
is also dispatched by `plot()` on an object with class `"layout"`; grip's
ordinary matrix does not acquire that class. `plot()` alone on a matrix is
not the graph plotting workflow.

Call `print(fit)` on a `grip_gmds_layout` to see its
[method summary](`r help_url("print.grip_gmds_layout")`), then plot
`fit$coords`. Call `print(hierarchy)` on the `grip_misf` returned by
`build.misf()` for its [hierarchy summary](`r help_url("print.grip_misf")`).
These are the other two S3 registrations; they are not additional exports.
`build.weighted.misf()` returns its documented hierarchy list and does not
share that print method. Full and landmark geodesic-KK refinements return
lists with `$coords`, `$score`, and optional traces, not ordinary matrices.

For static 3D figures, always request `projection = "ortho"` in
`plot.layout()`. Its default 3D route uses optional rgl interactively.
`project.3d()` rotates the first three columns and returns two projected
columns. Score the original coordinates to assess the fitted layout; scoring
the projection answers a different question.

### A common extraction step

[`layout.coords()`](`r help_url("layout.coords")`) returns the original coordinate
matrix from a matrix, trace, MDS fit, or supported refinement result. It preserves
vertex order, dimensions, names, and values; it does not align or project results.
Use the same graph when scoring the extracted matrices:

```{r common-extraction}
small.edges <- edges.path(6)
fits <- list(
  ordinary = grip(small.edges, n = 6, dim = 2, seed = 1),
  trace = trace.grip(small.edges, n = 6, dim = 2, seed = 1),
  classical = classical.mds(edges = small.edges, n = 6)
)
matrices <- lapply(fits, layout.coords)
vapply(matrices, function(z) score.layout(z, edges = small.edges, n = 6,
  sample.size.stress = 15, stress.seed = 1,
  edge.crossings = "never")$sampled.stress, numeric(1))
# Plot any extracted matrix with the same graph:
# plot.layout(matrices[[1]], edges = small.edges)
```

Direct `$final` and `$coords` extraction remains supported. In MDS results,
`method`, `coords`, `diagnostics`, and `metadata` remain available; stopping
metadata depends on the method. Prepared caches and coordinate-dependent state
are internal/version-dependent and should be rebuilt when their contract changes.
The accessor rejects unknown lists and ambiguous results instead of guessing a field.

## Classical scaling and metric stress MDS

`classical.mds()` runs `stats::cmdscale()` on graph shortest-path distances.
Classical scaling approximates the double-centered squared-distance (Gram)
matrix by a low-rank Euclidean representation, usually called minimizing
*strain*. It does not directly minimize the sum of squared distance residuals.
Non-Euclidean graph distances can yield negative eigenvalues; `add` and `eig`
are controls of this classical method.

`metric.mds()` minimizes unweighted raw distance stress,
\(\sum_{i<j}(\|z_i-z_j\|-d_{ij})^2\), through native stochastic gradient descent (SGD) by default,
with `backend = "sgd"`, or optional **smacof** selected with
`backend = "smacof"`. The SGD pair updates follow Zheng, Pawar and Goodman
(2018); the default hybrid learning-rate schedule follows Hangan, Kobourov and
Miller (2026). Both papers are listed in the [SGD references](#sgd-references).
Edge lengths determine the shortest-path targets, not pairwise stiffnesses.
Returned coordinates are rescaled to the input distance units. Multiple starts
can help with local optima; iteration limits do not certify convergence or a
global optimum. `init` accepts `"classical"`, `"random"`, or a coordinate
matrix; `n_init` includes the first start. A non-NULL `seed` preserves the
caller's R random-number state. `scale_mode` changes diagnostics only.

In versions through 0.2.0, `metric.mds()` performed classical scaling. Use
`classical.mds()` to retain that behavior, including `add` and `eig`. See the
[MDS migration help](`r help_url("grip-mds-migration")`).

```{r mds-baselines}
prepared <- prepare.graph.geodesic.mds(edges.cycle(8), n = 8)
classical <- classical.mds(prepared = prepared, dim = 2)
print(classical)
if (requireNamespace("smacof", quietly = TRUE)) {
  metric <- metric.mds(prepared = prepared, dim = 2, backend = "smacof", n_init = 2,
                       max_iter = 100, seed = 11)
  metric$metadata$raw_stress
} else {
  message("Install smacof for this optional-backend comparison; the SGD example below needs no extra backend.")
}
```

The default SGD backend requires no additional runtime package. Its learning-rate and
checkpoint defaults are provisional calibration choices. A pass visits every
unordered pair. The function's `max_iter` default is 1000; the 30 passes
requested below are an initial trial budget, not a convergence rule.
Schedule completion emits an iteration-limit warning; the example suppresses
that expected warning and displays the termination status instead.
The `eps` argument belongs to SMACOF and must be omitted for SGD. Inspect the
checkpoint history and compare repeated starts before drawing conclusions.

```{r mds-sgd}
sgd <- suppressWarnings(metric.mds(prepared = prepared, dim = 3,
  init = "random", max_iter = 30, seed = 11))
sgd$metadata$engine  # Omitting backend selects "sgd".
sgd$metadata$termination  # Schedule exhausted, not a convergence certificate.
sgd$metadata$target_normalized_rmse
sgd$metadata$sgd[[1]]$trace[c(1, 31), ]
# Use edge.kk(prepared = prepared, coords = sgd$coords, dim = 3) to refine it.
```

For matched-start stress and timing comparisons, quadratic surfaces, and
interactive 3D overlays, see [Comparing SGD and SMACOF](metric-mds-backends.html).

## Choose a score

A *chord* is the Euclidean separation between two drawn vertices. A *path
length* adds the drawn lengths of consecutive graph edges along a retained
route. A folded chain can preserve its path length while bringing its
endpoints close together. These measurements answer different questions.

| Question | Score and targets | Alignment, scale, and units |
|---|---|---|
| Is the drawing useful without a reference embedding? | `score.layout()`: sampled chords versus shortest graph distances, edge-length variation, non-neighbor separation, and optional 2D crossings or clusters. | Sampled stress fits one scalar and is dimensionless. Edge CV and separation ratios are dimensionless; crossing counts are counts. Sampling seeds and sizes affect the result. |
| Are local lengths, retained paths, and endpoint separations faithful? | `score.gmds()`: separate edge, fixed-path, and chord residual panels on a prepared graph. Path targets follow its retained-route convention; they may differ slightly from strict distances near ties. | `"profiled"` fits a separate scale per panel; `"identity"` fixes one; `"user"` supplies panel scales. Relative RMSE is dimensionless; absolute RMSE is in drawn-coordinate units against scaled targets. No coordinate-reference alignment. |
| Did corresponding vertices recover particular coordinates? | `score.coordinates(coords, reference)`: root mean squared Euclidean vertex displacement. Rows must correspond. | Default rigid alignment allows translation, rotation and reflection. `allow_reflection = FALSE` forbids reflection. Similarity also fits uniform scale; `"none"` uses the supplied alignment. RMSE is in reference-coordinate units; relative RMSE divides by the reference RMS radius. |
| Do two already aligned triangular surfaces occupy the same region? | `score.surface()`: area-weighted, symmetric RMS closest-point distance to triangles, estimated by sampling each surface. Different meshes are allowed. | No alignment, scaling, or triangulation is performed. Output has coordinate-distance units. Its Monte Carlo standard error measures sampling error, not mesh or alignment uncertainty. It is not a maximum-distance measure. |

The geodesic-KK scorers use their own path objectives, stiffnesses, and scale
policies. Their values are not interchangeable with sampled chord stress,
metric-MDS raw stress, or reference errors. Compare layouts using the same
graph, lengths, vertex order, samples, preparation, scale policy, and score
definition. Candidate composite scores are relative to their search and
weighting choices, not universal quality grades.

## Reuse preparations and control cost

| Preparation | What can be reused | When to rebuild |
|---|---|---|
| `prepare.edge.kk()` | Edges and targets for repeated edge repair and edge diagnostics; no all-pairs cache. | Changed topology, vertex order/count, or edge lengths. |
| `prepare.graph.geodesic.mds()` / `prepare.geodesic.kk()` | Connected-graph distances and retained routes for multiple coordinate candidates. | Graph changes, a new tie policy, or older caches predating the shortest-path symmetry fix. |
| `prepare.landmark.geodesic.kk()` | A sparse set of local and landmark routes. | Graph changes or new local-neighbor/landmark counts. |
| `prepare.misf.geodesic.kk()` | Multiscale independent-set filtration (MISF), level routes, and initializers. | Graph changes or changed hierarchy, tie, or initialization settings. |
| `build.misf()` / `build.weighted.misf()` | Inspectable multiscale hierarchy, with the latter also storing weighted neighborhoods and anchors. | Changed graph, metric, seed, or construction controls. They are not a generic `prepared` argument for `grip()`. |

Graph inputs must be undirected: adjacency entries have reciprocal entries with
matching lengths and multiplicities. Self-loops are rejected; remove them before
layout or preparation. Ordinary GRIP preserves parallel-edge multiplicity.
Full and landmark geodesic preparations collapse parallel pairs in their canonical
edge table; edge-only preparation rejects duplicate undirected edges. Simplify
parallel edges explicitly before comparing these different workflows, particularly
when their lengths differ. The validators do not silently simplify your input.

Treat preparations as immutable graph-specific objects. Do not combine a
`prepared` object with raw graph inputs (`edges`, `adj_list`, `edge_weights`,
or `weight_list`); these calls fail rather than silently choosing a graph.
An explicit `n` must be an integer matching the prepared vertex count.
Rebuild the preparation to change the graph. Coordinate
candidates on the same graph may change without rebuilding the graph cache;
a state containing coordinate-dependent forces or a local-star geometry must
be recomputed when those coordinates change.

Use [`estimate.preparation()`](`r help_url("estimate.preparation")`) before a
large graph-distance preparation:

```{r preparation-estimate}
estimate.preparation(10000, n.edges = 20000)
```

This is a lower bound for one matrix, not a peak-memory prediction. Dense
preparations warn before graph searches when that bound exceeds 512 MiB.
Set `options(grip.preparation.warn.bytes = ...)` to a different positive byte
count, or `Inf` to acknowledge and suppress the advisory warning. Edge-only
preparation has no dense distance matrix. **Landmark preparation still computes
dense distances**; it reduces retained paths, not that storage requirement.

Native construction/refinement checks for interrupts at main-thread boundaries.
Threaded geodesic-MDS work joins its workers before raising an interrupt;
cancellation may wait for the current evaluation and does not return a partial fit.

A dense distance matrix takes roughly `8 * n^2` bytes before workspace and
copies. Full route caches can cost much more, depending on path lengths. Use
`diagnostics = FALSE` with raw inputs in the two MDS methods to avoid the full
route cache; the dense distance matrix is still required. Classical scaling
also uses a dense eigendecomposition. Landmark and multiscale preparations
reduce retained pairs or work on smaller levels, but their cost still depends
on graph searches and route lengths. Begin with GRIP, small candidate sets,
and level traces; retaining every round of a large solve can itself be costly.

## Experimental refinement and current initializers

Try the appropriate GRIP metric, a plausible preset, several seeds, and the
scoring/tracing workflows first. Edge-KK, full geodesic-KK, landmark geodesic-KK,
MISF geodesic-KK, repulsive stages, and kernel/Gram local-star tools are
**public experimental interfaces**. Improving their objective need not improve
a reference embedding or the readability of a drawing.

`edge.kk()` accepts supplied `coords`, or `init = "classical_mds"` (default),
`"metric_mds"`, `"weighted_grip"`, or `"random"`. The MDS choices need
all-pairs distances; supplying coordinates or using a non-MDS start permits
edge-only preparation. The weighted-GRIP initializer uses the 2D/3D `grip()`
interface; supply higher-dimensional coordinates directly for edge repair.
`kernel.gram.gkk()` accepts `"classical_mds"` (default), `"metric_mds"`, or
`"random"`, or supplied coordinates. Both methods now interpret
`"metric_mds"` as stress minimization through native SGD. To use SMACOF,
fit `metric.mds(backend = "smacof", ...)` first and supply its coordinates.

`geodesic.kk()` and `landmark.geodesic.kk()` require starting coordinates.
MISF preparation uses `top_level_init = "geometric"`, `"cmdscale"`, or
`"random"`; `misf.geodesic.kk(top_level_init = NULL)` inherits the prepared
choice or defaults to `"geometric"`. Its `"cmdscale"` spelling still means
classical scaling. Initializer names are not a uniform package-wide vocabulary.

## Function catalog

This guide covers **106 explicit public exports and three registered S3
methods**. `plot.layout()` is both exported and registered, so there are 108
unique exported or registered function names, not 107. The catalog gives each
export exactly one row. Methods selected by R's generic functions are explained
through their objects below. Internal helpers and the compatibility aliases
retired in 0.2.0 are not public entry points.

Each row below is one explicit export. **Help** opens the detailed help page;
in an offline R session use `help("function.name", package = "grip")`.
Shared help topics may document several exports, but these rows keep their
purposes distinct. Maintainers can run `make audit-api-guide` to check coverage,
method accounting, and help targets against the current namespace.

### Compute layouts

| Function | Purpose | Details |
|---|---|---|
| `grip()` | Compute a multiscale layout using hop counts or positive edge lengths. | [Help](`r help_url("grip")`) |
| `weighted.grip.nd()` | Compute a weighted multiscale layout in two or more dimensions. | [Help](`r help_url("weighted.grip.nd")`) |
| `classical.mds()` | Fit a classical-scaling baseline from graph shortest-path distances. | [Help](`r help_url("classical.mds")`) |
| `metric.mds()` | Minimize raw distance stress with native SGD (default) or SMACOF and multiple starts. | [Help](`r help_url("metric.mds")`) |
| `globalrep.grip()` | Use the explicit coarse-global-repulsion hop-layout interface. | [Help](`r help_url("globalrep.grip")`) |
| `globalrep.weighted.grip()` | Use the explicit coarse-global-repulsion weighted-layout interface. | [Help](`r help_url("globalrep.weighted.grip")`) |
| `legacy.grip()` | Reproduce the legacy GRIP profile for historical comparisons. | [Help](`r help_url("legacy.grip")`) |

### Select and compare candidates

| Function | Purpose | Details |
|---|---|---|
| `compare.layouts()` | Run candidate settings across seeds and summarize layout quality and stability. | [Help](`r help_url("compare.layouts")`) |
| `params.from.summary()` | Recover reusable layout parameters from a comparison summary row. | [Help](`r help_url("params.from.summary")`) |

### Score graph distances or reference geometry

| Function | Purpose | Details |
|---|---|---|
| `score.layout()` | Evaluate sampled graph-distance fidelity and drawing-quality heuristics. | [Help](`r help_url("score.layout")`) |
| `score.gmds()` | Report separate edge, retained-path, and chord diagnostic panels. | [Help](`r help_url("score.gmds")`) |
| `score.coordinates()` | Measure error against corresponding reference vertices with explicit alignment. | [Help](`r help_url("score.coordinates")`) |
| `score.surface()` | Estimate symmetric area-weighted distance between already aligned triangular surfaces. | [Help](`r help_url("score.surface")`) |
| `geometry.diagnostics()` | Compute reference-aware geometric diagnostics for trace frames. | [Help](`r help_url("geometry.diagnostics")`) |

### Trace multiscale layouts

| Function | Purpose | Details |
|---|---|---|
| `trace.grip()` | Record multiscale frames and diagnostics under either graph metric. | [Help](`r help_url("trace.grip")`) |
| `trace.legacy.grip()` | Record frames from the legacy layout profile. | [Help](`r help_url("trace.legacy.grip")`) |

### Plot and project

| Function | Purpose | Details |
|---|---|---|
| `layout.coords()` | Extract unchanged coordinates from supported result families | [Help](`r help_url("layout.coords")`) |
| `plot.layout()` | Draw a coordinate matrix and graph edges; also the registered plot method for class layout. | [Help](`r help_url("plot.layout")`) |
| `project.3d()` | Rotate and project the first three coordinate columns to a static two-column view. | [Help](`r help_url("project.3d")`) |

### Prepare graphs and inspect multiscale structures

| Function | Purpose | Details |
|---|---|---|
| `build.misf()` | Build the hop-distance maximal independent-set filtration used by GRIP. | [Help](`r help_url("build.misf")`) |
| `build.weighted.misf()` | Build weighted multiscale levels, neighborhoods, and insertion anchors. | [Help](`r help_url("build.weighted.misf")`) |
| `estimate.preparation()` | Estimate dense storage and retained-pair bounds without preparing a graph | [Help](`r help_url("estimate.preparation")`) |
| `prepare.edge.kk()` | Prepare edges and length targets without an all-pairs cache for experimental edge repair. | [Help](`r help_url("prepare.edge.kk")`) |
| `prepare.graph.geodesic.mds()` | Prepare connected-graph all-pairs distances and retained geodesic routes. | [Help](`r help_url("prepare.graph.geodesic.mds")`) |
| `prepare.geodesic.kk()` | Prepare full retained routes for experimental geodesic-KK scoring and refinement. | [Help](`r help_url("prepare.geodesic.kk")`) |
| `prepare.landmark.geodesic.kk()` | Prepare sparse local and landmark routes for experimental refinement. | [Help](`r help_url("prepare.landmark.geodesic.kk")`) |
| `prepare.misf.geodesic.kk()` | Prepare multiscale hierarchies and per-level routes for experimental geodesic-KK. | [Help](`r help_url("prepare.misf.geodesic.kk")`) |

### Experimental refinement and state diagnostics

| Function | Purpose | Details |
|---|---|---|
| `edge.kk()` | Repair adjacent-edge lengths from a starting layout under edge stress. | [Help](`r help_url("edge.kk")`) |
| `edge.length.density.stiffness()` | Convert edge lengths to normalized spring stiffnesses for edge repair. | [Help](`r help_url("edge.length.density.stiffness")`) |
| `geodesic.kk()` | Refine supplied coordinates using full retained-path geodesic-KK energy. | [Help](`r help_url("geodesic.kk")`) |
| `score.geodesic.kk()` | Evaluate full geodesic-KK path error and energy. | [Help](`r help_url("score.geodesic.kk")`) |
| `landmark.geodesic.kk()` | Refine supplied coordinates using sparse local and landmark routes. | [Help](`r help_url("landmark.geodesic.kk")`) |
| `score.landmark.geodesic.kk()` | Evaluate the sparse landmark geodesic-KK objective. | [Help](`r help_url("score.landmark.geodesic.kk")`) |
| `misf.geodesic.kk()` | Compute a layout through multiscale geodesic-KK insertion and refinement. | [Help](`r help_url("misf.geodesic.kk")`) |
| `score.misf.geodesic.kk()` | Evaluate a layout on a selected prepared MISF level. | [Help](`r help_url("score.misf.geodesic.kk")`) |
| `edge.repulsive.state()` | Evaluate edge repair combined with repulsion at fixed coordinates. | [Help](`r help_url("edge.repulsive.state")`) |
| `edge.repulsive.stage()` | Optimize an edge-repair stage with repulsion. | [Help](`r help_url("edge.repulsive.stage")`) |
| `repulsive.state()` | Evaluate the repulsion-only objective and gradient at fixed coordinates. | [Help](`r help_url("repulsive.state")`) |
| `repulsive.stage()` | Optimize a repulsion-only stage from supplied coordinates. | [Help](`r help_url("repulsive.stage")`) |
| `graph.riemannian.star.structure()` | Build local reference-star geometry for kernel/Gram refinement. | [Help](`r help_url("graph.riemannian.star.structure")`) |
| `kernel.gram.gkk()` | Refine a layout using edge stress and local kernel/Gram shape constraints. | [Help](`r help_url("kernel.gram.gkk")`) |

### Synthetic graphs: primitive edge generators

| Function | Purpose | Details |
|---|---|---|
| `edges.path()` | Return consecutive edges of a path. | [Help](`r help_url("graph_generators")`) |
| `edges.cycle()` | Return edges of a closed cycle. | [Help](`r help_url("graph_generators")`) |
| `edges.mesh()` | Return edges of a rectangular grid. | [Help](`r help_url("graph_generators")`) |
| `edges.cube()` | Return edges of a three-dimensional grid. | [Help](`r help_url("graph_generators")`) |
| `edges.cylinder()` | Return edges of a grid wrapped in one direction. | [Help](`r help_url("graph_generators")`) |
| `edges.torus()` | Return edges of a grid wrapped in both directions. | [Help](`r help_url("graph_generators")`) |
| `edges.kary.tree()` | Return a rooted tree with k children per internal vertex. | [Help](`r help_url("graph_generators")`) |
| `edges.sierpinski.carpet()` | Return occupied-cell adjacency for a Sierpinski carpet. | [Help](`r help_url("graph_generators")`) |
| `edges.sierpinski.triangle()` | Return a recursively subdivided Sierpinski triangle graph. | [Help](`r help_url("graph_generators")`) |
| `edges.sierpinski.tetrahedron()` | Return a recursively subdivided Sierpinski tetrahedron graph. | [Help](`r help_url("graph_generators")`) |

### Synthetic graphs: configurable keep masks

| Function | Purpose | Details |
|---|---|---|
| `keep.periodic.holes()` | Create a finite-grid keep matrix with periodic holes. | [Help](`r help_url("perforated_grid_helpers")`) |
| `keep.asymmetric.notches()` | Create a finite-grid keep matrix with asymmetric notches. | [Help](`r help_url("perforated_grid_helpers")`) |
| `keep.slit.channels()` | Create a finite-grid keep matrix with slit channels. | [Help](`r help_url("perforated_grid_helpers")`) |
| `keep.staggered.windows()` | Create a finite-grid keep matrix with staggered windows. | [Help](`r help_url("perforated_grid_helpers")`) |
| `mask.border()` | Create a square recursive mask retaining its border. | [Help](`r help_url("mask_pattern_helpers")`) |
| `mask.corner()` | Create a square recursive corner pattern. | [Help](`r help_url("mask_pattern_helpers")`) |
| `mask.cross()` | Create a square recursive cross pattern. | [Help](`r help_url("mask_pattern_helpers")`) |
| `mask.asymmetric.holes()` | Create a square recursive pattern with asymmetric holes. | [Help](`r help_url("mask_pattern_helpers")`) |
| `mask.cube.periodic.tunnels()` | Create a cubic recursive keep array with periodic tunnels. | [Help](`r help_url("cube_mask_pattern_helpers")`) |
| `mask.cube.asymmetric.cavities()` | Create a cubic recursive keep array with offset cavities. | [Help](`r help_url("cube_mask_pattern_helpers")`) |
| `mask.cube.channel.network()` | Create a cubic recursive keep array with connected channels. | [Help](`r help_url("cube_mask_pattern_helpers")`) |
| `mask.triangle.classic()` | Select the three corner subtriangles of the classic recursive triangle. | [Help](`r help_url("triangle_mask_helpers")`) |
| `mask.triangle.bridge()` | Select a recursive triangle pattern including a bridge. | [Help](`r help_url("triangle_mask_helpers")`) |
| `mask.tetrahedron.classic()` | Select the four corner subtetrahedra of the classic recursive tetrahedron. | [Help](`r help_url("tetrahedron_mask_helpers")`) |
| `mask.tetrahedron.corner.missing()` | Select a recursive tetrahedron pattern with one missing corner. | [Help](`r help_url("tetrahedron_mask_helpers")`) |

### Synthetic graphs: complete surface and solid bundles

| Function | Purpose | Details |
|---|---|---|
| `mesh.surface.graph()` | Lift a rectangular mesh to a surface with induced edge lengths. | [Help](`r help_url("mesh_surface_helpers")`) |
| `occupied.mesh.surface.graph()` | Lift a finite occupied grid specified by a keep matrix. | [Help](`r help_url("occupied_mesh_surface_helpers")`) |
| `cylinder.surface.graph()` | Construct a weighted cylindrical surface bundle. | [Help](`r help_url("cylinder_surface_helpers")`) |
| `sphere.surface.graph()` | Construct a weighted pole-and-latitude-ring sphere bundle. | [Help](`r help_url("sphere_surface_helpers")`) |
| `torus.surface.graph()` | Construct a weighted toroidal surface bundle. | [Help](`r help_url("torus_surface_helpers")`) |
| `irregular.rectangle.surface.graph()` | Construct a deterministically irregular rectangular surface bundle. | [Help](`r help_url("irregular_rectangle_surface_helpers")`) |
| `irregular.annulus.surface.graph()` | Construct an irregular annular surface bundle. | [Help](`r help_url("irregular_annulus_surface_helpers")`) |
| `irregular.sphere.surface.graph()` | Construct an irregular spherical surface bundle. | [Help](`r help_url("irregular_sphere_surface_helpers")`) |
| `irregular.torus.surface.graph()` | Construct an irregular toroidal surface bundle. | [Help](`r help_url("irregular_torus_surface_helpers")`) |
| `irregular.double.torus.surface.graph()` | Construct an irregular double-torus surface bundle. | [Help](`r help_url("irregular_double_torus_surface_helpers")`) |
| `irregular.pair.of.pants.surface.graph()` | Construct an irregular surface with three boundary components. | [Help](`r help_url("irregular_pair_of_pants_surface_helpers")`) |
| `irregular.ball.solid.graph()` | Construct a graph sampling a three-dimensional ball interior. | [Help](`r help_url("irregular_ball_solid_helpers")`) |
| `irregular.shell.solid.graph()` | Construct a graph sampling a three-dimensional shell volume. | [Help](`r help_url("irregular_shell_solid_helpers")`) |
| `triangulated.annulus.surface.graph()` | Construct a lifted triangular-lattice annulus bundle. | [Help](`r help_url("triangulated_annulus_surface_helpers")`) |
| `triangulated.pair.of.pants.surface.graph()` | Construct a lifted triangular-lattice surface with two holes. | [Help](`r help_url("triangulated_pair_of_pants_surface_helpers")`) |
| `triangulated.polyhedron.surface.graph()` | Construct a subdivided polyhedral graph with reference coordinates. | [Help](`r help_url("triangulated_polyhedron_surface_helpers")`) |
| `sampled.rectangle.surface.graph()` | Sample a rectangle and construct one intersection-nearest-neighbor graph bundle. | [Help](`r help_url("sampled_rectangle_surface_helpers")`) |
| `sampled.rectangle.surface.graphs()` | Reuse one rectangular sample across a sequence of neighborhood sizes. | [Help](`r help_url("sampled_rectangle_surface_helpers")`) |
| `kary.tree.weighted.graph()` | Assign intrinsic tree lengths by depth and child slot, with tree metadata. | [Help](`r help_url("kary_tree_weighted_graph_helpers")`) |

### Synthetic graphs: complete recursive bundles

| Function | Purpose | Details |
|---|---|---|
| `recursive.mask.grid.surface.graph()` | Recursively apply a square mask and lift the retained cells. | [Help](`r help_url("recursive_mask_grid_surface_helpers")`) |
| `recursive.cube.mask.surface.graph()` | Recursively apply a cubic mask and embed occupied-cell adjacency. | [Help](`r help_url("recursive_cube_mask_surface_helpers")`) |
| `recursive.triangle.mask.surface.graph()` | Recursively apply a triangle mask and return a weighted bundle. | [Help](`r help_url("recursive_triangle_mask_surface_helpers")`) |
| `recursive.tetrahedron.mask.surface.graph()` | Recursively apply a tetrahedron mask and return a weighted bundle. | [Help](`r help_url("recursive_tetrahedron_mask_surface_helpers")`) |
| `sierpinski.carpet.surface.graph()` | Construct a weighted Sierpinski carpet with a surface lift. | [Help](`r help_url("sierpinski_carpet_surface_helpers")`) |
| `sierpinski.triangle.surface.graph()` | Construct a weighted Sierpinski triangle with reference coordinates. | [Help](`r help_url("sierpinski_triangle_surface_helpers")`) |
| `sierpinski.tetrahedron.surface.graph()` | Construct a weighted Sierpinski tetrahedron with reference coordinates. | [Help](`r help_url("sierpinski_tetrahedron_surface_helpers")`) |
| `menger.sponge.surface.graph()` | Construct a weighted Menger occupied-cube graph bundle. | [Help](`r help_url("menger_sponge_surface_helpers")`) |
| `vicsek.surface.graph()` | Construct a weighted recursive cross-family bundle. | [Help](`r help_url("vicsek_surface_helpers")`) |
| `cube.periodic.tunnels.surface.graph()` | Construct a recursively perforated cube bundle with periodic tunnels. | [Help](`r help_url("porous_cube_surface_helpers")`) |
| `cube.asymmetric.cavities.surface.graph()` | Construct a recursively perforated cube bundle with offset cavities. | [Help](`r help_url("porous_cube_surface_helpers")`) |
| `cube.channel.network.surface.graph()` | Construct a recursively perforated cube bundle with a channel network. | [Help](`r help_url("porous_cube_surface_helpers")`) |

### Interactive exploration

| Function | Purpose | Details |
|---|---|---|
| `gripui_project()` | Assemble graph, layouts, scores, and metadata into an explorer project. | [Help](`r help_url("gripui_project")`) |
| `gripui_project_from_compare()` | Convert a layout-comparison result into an explorer project. | [Help](`r help_url("gripui_project_from_compare")`) |
| `gripui_project_from_dir()` | Load a saved explorer project from a directory. | [Help](`r help_url("gripui_project_from_dir")`) |
| `gripui_validate_project()` | Validate a project's graph and layout structures. | [Help](`r help_url("gripui_validate_project")`) |
| `gripui_app()` | Construct the Shiny layout-explorer app object. | [Help](`r help_url("gripui_app")`) |
| `run_gripui()` | Launch the layout explorer in an interactive session. | [Help](`r help_url("run_gripui")`) |
| `gripui_graph_family_catalog()` | List family identifiers and geometry choices for the family explorer. | [Help](`r help_url("gripui_graph_family_catalog")`) |
| `gripui_family_app()` | Construct the Shiny synthetic-family explorer app object. | [Help](`r help_url("gripui_family_app")`) |
| `run_gripui_family()` | Launch the synthetic-family explorer in an interactive session. | [Help](`r help_url("run_gripui_family")`) |

## Find detailed help

```{r navigation, eval=FALSE}
help(package = "grip")
help("grip-package", package = "grip")
help("grip", package = "grip")
vignette(package = "grip")
vignette("synthetic-graph-families", package = "grip")
```

The [0.2 migration help](`r help_url("grip-0.2-migration")`) documents retired
aliases. Internal builders and standalone embeddings used within graph bundles
are implementation details; use the public bundle constructors above.

This guide adopts the short function-and-purpose listings and connections
between overview, examples, and detailed help found in the
[Hmisc documentation](https://hbiostat.org/r/hmisc/) and its
[package overview](https://raw.githubusercontent.com/harrelfe/Hmisc/master/man/Overview.Rd).
The groups here follow graph-layout tasks specific to grip.

## SGD references

- Zheng, J. X., Pawar, S. and Goodman, D. F. (2018).
  *Graph Drawing by Stochastic Gradient Descent*.
  [doi:10.1109/TVCG.2018.2859997](https://doi.org/10.1109/TVCG.2018.2859997).
- Hangan, D., Kobourov, S. and Miller, J. (2026).
  *Bridging Graph Drawing and Dimensionality Reduction with Stochastic Stress
  Optimization*. [arXiv:2605.00641](https://arxiv.org/abs/2605.00641).
