---
title: "Graph gallery: SGD metric MDS"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Graph gallery: SGD metric MDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This gallery fits the same six graph families as the
[multiscale GRIP gallery](gallery_grip.html), using SGD metric MDS.
Graphs, edge lengths and seeds are retained so their
representations are easy to compare. This is an illustration, not a controlled
benchmark of optimizer quality or runtime. All SGD fits here use three dimensions;
the preserved GRIP tree example uses two.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  fig.width = 11,
  fig.height = 4.2
)
```

Each fit uses weighted graph shortest-path distances, uniform pair weights,
and full metric MDS with the SGD backend. The examples are small enough for
full distances. No edge-KK refinement follows MDS. For larger graphs see
[sparse SGD](sparse-metric-mds.html).

The reference coordinates generate edge lengths; they are not the MDS targets.
Shortest-path distances need not admit an exact Euclidean representation, so
a fitted layout can differ from the reference geometry even with low stress.
The ivue viewers support rotation, zoom and selection of displayed geometries.
Coordinates are centered for display without rotation or scale fitting; these
are not aligned shape-recovery comparisons. The parameter grid is planar and
is displayed at z = 0. Install `ivue` and `rgl` to build the interactive views.

```{r}
library(grip)
```

```{r view-helpers, include=FALSE}
source(system.file('scripts', 'vignette-views.R', package='grip'), local=TRUE)
stopifnot(documentation.interactive())
```

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.rglWebGL { max-width: 100%; }
</style>

## Family classes at a glance

The table below lists a representative subset of the larger family collection.

| Family class | Representative helper | Why it matters |
|:--|:--|:--|
| Lifted lattices | `mesh.surface.graph()` | Plain topology, nontrivial metric geometry |
| Wrapped surfaces | `torus.surface.graph()` | Closed periodic surface geometry |
| Recursive fractals | `vicsek.surface.graph()` | Self-similar bottlenecks and holes |
| Irregular manifolds | `irregular.annulus.surface.graph()` | Non-lattice surface geometry with boundary |
| Porous 3D families | `cube.channel.network.surface.graph()` | Volumetric geodesic detours |
| Intrinsic trees | `kary.tree.weighted.graph()` | Geometry from edge lengths rather than ambient space |

This article shows one representative example from each class. The interactive
family browser described in `Interactive Exploration with gripui` exposes a
much larger catalog.

## Lifted lattices

The simplest geometry-rich family keeps the graph topology of a rectangular
mesh but induces edge lengths from a curved embedding in `R^3`.

```{r mesh-family}
mesh.graph <- mesh.surface.graph(
  5, 5,
  surface = "saddle",
  amplitude = 0.8
)

mesh.fit <- metric.mds(
  edges = mesh.graph$edges,
  n = mesh.graph$n,
  edge.weights = mesh.graph$edge_weights,
  dim = 3,
  seed = 1,
  backend = "sgd", pair.weights = "uniform",
  max.iter = 1000, diagnostics = FALSE
)
mesh.coords <- mesh.fit$coords

mesh.summary <- data.frame(
  family = mesh.graph$family,
  surface = mesh.graph$surface,
  n = mesh.graph$n,
  m = nrow(mesh.graph$edges),
  weight_scale = mesh.graph$weight_scale
)

knitr::kable(mesh.summary, digits = 3)
```

```{r mesh-interactive}
documentation.view(
  list(`SGD metric MDS`=mesh.coords,
       `Reference geometry`=mesh.graph$coords_surface,
       `Parameter grid`=cbind(mesh.graph$coords_param, z=0)),
  edges=mesh.graph$edges, aligned=TRUE,
  description="Mesh: drag to rotate, scroll to zoom; use the selector to compare geometries."
)
```

The useful design pattern here is that the topology stays simple while the
metric geometry becomes nontrivial.

## Wrapped surfaces

Wrapped lattices extend the same idea to periodic topologies such as a torus.

```{r torus-family}
torus.graph <- torus.surface.graph(
  5, 6,
  surface = "pinched",
  minor.radius = 0.33,
  amplitude = 0.18,
  twist = 0.18
)

torus.fit <- metric.mds(
  edges = torus.graph$edges,
  n = torus.graph$n,
  edge.weights = torus.graph$edge_weights,
  dim = 3,
  seed = 2,
  backend = "sgd", pair.weights = "uniform",
  max.iter = 1000, diagnostics = FALSE
)
torus.coords <- torus.fit$coords

knitr::kable(
  data.frame(
    family = torus.graph$family,
    surface = torus.graph$surface,
    n = torus.graph$n,
    m = nrow(torus.graph$edges)
  )
)
```

```{r torus-interactive}
documentation.view(
  list(`SGD metric MDS`=torus.coords,
       `Reference geometry`=torus.graph$coords_surface),
  edges=torus.graph$edges, aligned=TRUE,
  description="Torus: drag to rotate, scroll to zoom; use the selector to compare geometries."
)
```

These wrapped families are especially useful when ordinary planar intuition is
misleading and a 3D layout is the natural target.

## Recursive and fractal families

The package now includes recursive square-mask families such as the Vicsek
fractal and Sierpinski carpet variants, all of which can be lifted into curved
surface geometries.

```{r vicsek-family}
vicsek.graph <- vicsek.surface.graph(
  level = 2,
  surface = "ripple",
  amplitude = 0.55
)

vicsek.fit <- metric.mds(
  edges = vicsek.graph$edges,
  n = vicsek.graph$n,
  edge.weights = vicsek.graph$edge_weights,
  dim = 3,
  seed = 3,
  backend = "sgd", pair.weights = "uniform",
  max.iter = 1000, diagnostics = FALSE
)
vicsek.coords <- vicsek.fit$coords
```

```{r vicsek-interactive}
documentation.view(
  list(`SGD metric MDS`=vicsek.coords,
       `Reference geometry`=vicsek.graph$coords_surface),
  edges=vicsek.graph$edges, aligned=TRUE,
  description="Vicsek: drag to rotate, scroll to zoom; use the selector to compare geometries."
)
```

Recursive families are useful because they create holes, bottlenecks, and
multi-scale detours that are hard to capture with simpler lattices.

## Irregular manifolds

Not all surface families in `grip` are based on regular grids. The package also
includes irregular triangulated manifolds with boundary and closed irregular
surfaces.

```{r annulus-family}
annulus.graph <- irregular.annulus.surface.graph(
  rings = 5,
  outer.count = 24,
  inner.radius = 0.42,
  surface = "folded",
  amplitude = 0.2
)

annulus.fit <- metric.mds(
  edges = annulus.graph$edges,
  n = annulus.graph$n,
  edge.weights = annulus.graph$edge_weights,
  dim = 3,
  seed = 4,
  backend = "sgd", pair.weights = "uniform",
  max.iter = 1000, diagnostics = FALSE
)
annulus.coords <- annulus.fit$coords
```

```{r annulus-interactive}
documentation.view(
  list(`SGD metric MDS`=annulus.coords,
       `Reference geometry`=annulus.graph$coords_surface),
  edges=annulus.graph$edges, aligned=TRUE,
  description="Annulus: drag to rotate, scroll to zoom; use the selector to compare geometries."
)
```

These families are important because they reduce lattice bias and produce more
variable local geometry than mesh-like constructions.

## Porous 3D families

The cubical branch introduces graphs that are genuinely volumetric rather than
surface-like. Channel networks and related porous families create 3D geodesic
detours that are hard to flatten cleanly.

```{r cube-family}
cube.graph <- cube.channel.network.surface.graph(
  side = 5,
  level = 1,
  surface = "twisted",
  twist = 0.2
)

cube.fit <- metric.mds(
  edges = cube.graph$edges,
  n = cube.graph$n,
  edge.weights = cube.graph$edge_weights,
  dim = 3,
  seed = 5,
  backend = "sgd", pair.weights = "uniform",
  max.iter = 1000, diagnostics = FALSE
)
cube.coords <- cube.fit$coords

knitr::kable(
  data.frame(
    family = cube.graph$family,
    n = cube.graph$n,
    m = nrow(cube.graph$edges)
  )
)
```

```{r cube-interactive}
documentation.view(
  list(`SGD metric MDS`=cube.coords,
       `Reference geometry`=cube.graph$coords_surface),
  edges=cube.graph$edges, aligned=TRUE,
  description="Cube: drag to rotate, scroll to zoom; use the selector to compare geometries."
)
```

This family class is especially useful for testing whether a method can
preserve large-scale volumetric structure without collapsing tunnels and
channels.

## Intrinsic weighted trees

Some family geometries are intrinsic rather than ambient. The weighted tree
helpers preserve the exact tree topology while assigning edge lengths directly
from branch-depth and branch-position rules.

```{r tree-family}
tree.graph <- kary.tree.weighted.graph(
  k = 2,
  depth = 4,
  depth.rule = "geometric",
  depth.decay = 0.82,
  branch.rule = "linear",
  branch.spread = 0.25
)

tree.fit <- metric.mds(
  edges = tree.graph$edges,
  n = tree.graph$n,
  edge.weights = tree.graph$edge_weights,
  dim = 3,
  seed = 6,
  backend = "sgd", pair.weights = "uniform",
  max.iter = 1000, diagnostics = FALSE
)
tree.coords <- tree.fit$coords

knitr::kable(
  head(tree.graph$edge_table[, c(
    "parent",
    "child",
    "child_depth",
    "branch_index",
    "edge_weight"
  )]),
  digits = 3
)
```

```{r tree-interactive}
documentation.view(
  list(`SGD metric MDS`=tree.coords),
  edges=tree.graph$edges, aligned=TRUE,
  description="Tree: drag to rotate, scroll to zoom; use the selector to compare geometries."
)
```

These tree families are useful because they separate intrinsic weighted
geometry from any arbitrary 3D embedding choice.

## Stopping status

The table records each fit's starts. An iteration-limit warning means that
coordinates were returned without meeting the stopping criterion; it does not
establish convergence or an optimum. Assess graph-distance error and stability
across starts before interpreting a layout as evidence about geometry.

```{r stopping-status, results='asis'}
fits <- list(Saddle=mesh.fit, Torus=torus.fit, Vicsek=vicsek.fit,
             Annulus=annulus.fit, Porous=cube.fit, Tree=tree.fit)
for (name in names(fits)) {
  cat("\n\n### ", name, "\n\n", sep="")
  print(knitr::kable(fits[[name]]$metadata$starts, digits=5))
}
```

## Learn more

- [Multiscale GRIP versions of these examples](gallery_grip.html).
- [SGD, SMACOF and stress weights](metric-mds-backends.html).
- [Sparse SGD for larger graphs](sparse-metric-mds.html).
- [Interactive synthetic graph comparisons](synthetic-graph-families.html).
