---
title: "Choosing Layouts for Real Data"
output: rmarkdown::html_vignette
bibliography: grip-real-data.bib
vignette: >
  %\VignetteIndexEntry{Choosing Layouts for Real Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5.5
)
```

## Why comparison helps

Synthetic graphs sometimes come with a known target picture, so a layout can be
judged against that target directly. Real-world graphs are different: there is
usually no single correct picture waiting to be recovered. The practical goal is
to choose a layout that is faithful to graph structure, visually uncluttered,
and reasonably stable when the random seed changes.

In practice, the most useful workflow is:

- try the default layout plus one or two structurally plausible presets,
- compare them in 3D across a few seeds, and
- inspect a small set of quality metrics before choosing a final plot.

The `grip` package provides two helpers for this:

- `grip.score.layout()` is the low-level scoring function for one realized layout,
- `grip.compare.layouts()` runs several candidates, including local parameter searches.

For small and medium graphs, seed effects are often smaller than the difference
between obviously weak and obviously strong parameter settings. Even so, a few
repeated runs are useful because they reveal brittle candidates that only look
good for one lucky initialization.

```{r}
library(grip)
```

```{r}
read.extdata.csv <- function(file.name) {
  candidates <- c(
    system.file("extdata", file.name, package = "grip"),
    file.path("inst", "extdata", file.name),
    file.path("..", "inst", "extdata", file.name)
  )
  path <- candidates[file.exists(candidates)][1L]
  if (!length(path) || is.na(path) || !nzchar(path)) {
    stop("could not locate ", file.name)
  }
  utils::read.csv(path, stringsAsFactors = FALSE)
}

read.edge.csv <- function(file.name) {
  as.matrix(read.extdata.csv(file.name))
}

read.karate.club <- function() {
  labels <- read.extdata.csv("karate-club-membership.csv")
  labels <- labels[order(labels$vertex), , drop = FALSE]
  labels$club
}

compact.summary <- function(x) {
  keep <- intersect(c(
    "candidate",
    "preset",
    "rounds",
    "final.rounds",
    "num.nbrs",
    "repulsion.factor",
    "score.composite",
    "sampled.stress.mean",
    "edge.length.cv.mean",
    "sampled.nonedge.sep.ratio.mean",
    "stability.procrustes.mean",
    "cluster.separation.mean"
  ), names(x))
  x[, keep, drop = FALSE]
}

candidate.from.summary <- function(row) {
  row <- row[1, , drop = FALSE]
  out <- list(
    placement = row$placement[[1L]],
    rounds = row$rounds[[1L]],
    final_rounds = row$final.rounds[[1L]],
    num_init = row$num.init[[1L]],
    num_nbrs = row$num.nbrs[[1L]],
    r = row$r[[1L]],
    s = row$s[[1L]],
    repulsion_factor = row$repulsion.factor[[1L]],
    tinit_factor = row$tinit.factor[[1L]]
  )
  if (!is.na(row$preset[[1L]]) && nzchar(row$preset[[1L]])) {
    out$preset <- row$preset[[1L]]
  }
  out
}

project.coords <- function(coords, azimuth = 35, elevation = 22) {
  if (ncol(coords) == 2L) {
    return(coords)
  }
  az <- azimuth * pi / 180
  el <- elevation * pi / 180
  rz <- matrix(c(
    cos(az), -sin(az), 0,
    sin(az),  cos(az), 0,
    0,        0,       1
  ), nrow = 3, byrow = TRUE)
  rx <- matrix(c(
    1, 0, 0,
    0, cos(el), -sin(el),
    0, sin(el),  cos(el)
  ), nrow = 3, byrow = TRUE)
  rotated <- coords %*% rz %*% rx
  rotated[, 1:2, drop = FALSE]
}

plot.projected.graph <- function(coords,
                                 edges,
                                 main = "",
                                 azimuth = 35,
                                 elevation = 22,
                                 vertex.col = "black",
                                 edge.col = "gray80",
                                 pch = 16,
                                 cex = 0.8) {
  xy <- project.coords(coords, azimuth = azimuth, elevation = elevation)
  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- 0.08 * diff(xlim)
  ypad <- 0.08 * diff(ylim)
  if (!is.finite(xpad) || xpad == 0) xpad <- 0.2
  if (!is.finite(ypad) || ypad == 0) ypad <- 0.2

  plot(
    xy[, 1], xy[, 2],
    type = "n",
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    xlim = xlim + c(-xpad, xpad),
    ylim = ylim + c(-ypad, ypad),
    main = main
  )
  if (!is.null(edges) && nrow(edges) > 0) {
    apply(edges, 1, function(e) {
      graphics::segments(
        xy[e[1], 1], xy[e[1], 2],
        xy[e[2], 1], xy[e[2], 2],
        col = edge.col
      )
    })
  }
  points(xy[, 1], xy[, 2], pch = pch, cex = cex, col = vertex.col)
}

plot.layout.triptych <- function(layouts, edges, titles, vertex.cols = NULL) {
  if (is.null(vertex.cols)) {
    vertex.cols <- rep(list("black"), length(layouts))
  }
  op <- par(mfrow = c(1, length(layouts)), mar = c(1.2, 1.2, 3.2, 1.2))
  on.exit(par(op), add = TRUE)
  for (i in seq_along(layouts)) {
    plot.projected.graph(
      layouts[[i]],
      edges = edges,
      main = titles[[i]],
      vertex.col = vertex.cols[[i]]
    )
  }
}

plot.layout.views <- function(coords, edges, title, vertex.col = "black") {
  op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3.2, 1.2))
  on.exit(par(op), add = TRUE)
  plot.projected.graph(coords, edges, main = paste(title, "\nview 1"),
                       azimuth = 35, elevation = 22, vertex.col = vertex.col)
  plot.projected.graph(coords, edges, main = paste(title, "\nview 2"),
                       azimuth = -45, elevation = 28, vertex.col = vertex.col)
}

plot.search.slices <- function(summary.df, title.prefix = "") {
  op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
  on.exit(par(op), add = TRUE)

  palette.cols <- grDevices::hcl.colors(length(unique(summary.df$num.nbrs)), "Dark 3")
  color.idx <- match(summary.df$num.nbrs, sort(unique(summary.df$num.nbrs)))

  plot(
    summary.df$repulsion.factor,
    summary.df$sampled.stress.mean,
    pch = 19,
    col = palette.cols[color.idx],
    xlab = "repulsion.factor",
    ylab = "sampled.stress.mean",
    main = paste(title.prefix, "stress")
  )
  legend(
    "topright",
    legend = paste("num.nbrs =", sort(unique(summary.df$num.nbrs))),
    col = palette.cols,
    pch = 19,
    bty = "n",
    cex = 0.8
  )

  plot(
    summary.df$final.rounds,
    summary.df$sampled.nonedge.sep.ratio.mean,
    pch = 19,
    col = palette.cols[color.idx],
    xlab = "final.rounds",
    ylab = "sampled.nonedge.sep.ratio.mean",
    main = paste(title.prefix, "separation")
  )
}
```

## Benchmark graphs used here

This vignette uses two bundled benchmark networks:

- the Zachary karate club friendship network, a 34-person social network with a
  known split into the `Mr. Hi` and `Officer` factions [@zachary1977],
- the Krackhardt kite network, a 10-person organizational benchmark commonly
  used to discuss centrality and informal structure [@krackhardt1990].

```{r}
karate.edges <- read.edge.csv("karate-club-edges.csv")
karate.n <- max(karate.edges)
karate.club <- read.karate.club()

kite.edges <- read.edge.csv("krackhardt-kite-edges.csv")
kite.n <- max(kite.edges)

benchmark.info <- data.frame(
  graph = c("Zachary karate club", "Krackhardt kite"),
  n.vertices = c(karate.n, kite.n),
  n.edges = c(nrow(karate.edges), nrow(kite.edges)),
  stringsAsFactors = FALSE
)

knitr::kable(benchmark.info)
```

## Shortlisting 3D presets on the karate club network

For a real social network like the karate club graph, there is no obvious
family-specific preset that must be correct ahead of time. A sensible first pass
is to compare the generic default against one or two plausible alternatives and
look for candidates that are both numerically strong and visually coherent.

```{r}
karate.presets <- grip.compare.layouts(
  edges = karate.edges,
  n = karate.n,
  dim = 3,
  candidates = c("default", "tree", "mesh"),
  seeds = 1:3,
  sample.size.stress = 1000L,
  sample.size.nonedge = 2000L,
  edge.crossings = "never",
  return.layouts = TRUE
)

knitr::kable(compact.summary(karate.presets$summary), digits = 3)
```

```{r fig.width=10, fig.height=3.8}
plot.layout.triptych(
  layouts = list(
    karate.presets$layouts$default[["1"]],
    karate.presets$layouts$tree[["1"]],
    karate.presets$layouts$mesh[["1"]]
  ),
  edges = karate.edges,
  titles = c("default", "tree", "mesh")
)
```

The summary table already gives a useful shortlist. The `score.composite` column
is a rank-based aggregate across the available metrics, so it is best treated as
a screening tool rather than as the final decision by itself.

## Local parameter search around a promising candidate

Once a preset looks plausible, the next step is not to stare harder at one
picture. Instead, search a small parameter neighborhood and see how the quality
metrics move. The search API below generates a grid of candidate settings around
the leading preset from the first pass.

```{r}
karate.base <- karate.presets$summary[1, , drop = FALSE]

karate.search <- list(
  candidate.prefix = "karate.local",
  placement = karate.base$placement[[1L]],
  rounds = karate.base$rounds[[1L]],
  final_rounds = sort(unique(as.integer(c(
    max(32L, karate.base$final.rounds[[1L]] - 32L),
    karate.base$final.rounds[[1L]],
    karate.base$final.rounds[[1L]] + 32L
  )))),
  num_init = karate.base$num.init[[1L]],
  num_nbrs = sort(unique(as.integer(c(
    max(4L, karate.base$num.nbrs[[1L]] - 4L),
    karate.base$num.nbrs[[1L]],
    karate.base$num.nbrs[[1L]] + 4L
  )))),
  r = karate.base$r[[1L]],
  s = karate.base$s[[1L]],
  repulsion_factor = sort(unique(c(
    max(0, karate.base$repulsion.factor[[1L]] - 0.5),
    karate.base$repulsion.factor[[1L]],
    karate.base$repulsion.factor[[1L]] + 0.5
  )))
)
if (!is.na(karate.base$preset[[1L]]) && nzchar(karate.base$preset[[1L]])) {
  karate.search$preset <- karate.base$preset[[1L]]
}

karate.local <- grip.compare.layouts(
  edges = karate.edges,
  n = karate.n,
  dim = 3,
  search = karate.search,
  seeds = 1:2,
  sample.size.stress = 1000L,
  sample.size.nonedge = 2000L,
  edge.crossings = "never",
  return.layouts = TRUE
)

knitr::kable(head(compact.summary(karate.local$summary), 8L), digits = 3)
```

```{r}
plot.search.slices(karate.local$summary, title.prefix = "Karate local search")
```

These two plots are only slices through a multidimensional search, but they are
still helpful. They show how the shortlisted settings move as `repulsion.factor`
and `final.rounds` vary, while color indicates `num.nbrs`.

```{r}
karate.local.best <- karate.local$summary[1, , drop = FALSE]
karate.local.best.seed <- karate.local$runs$seed[
  karate.local$runs$candidate == karate.local.best$candidate[[1L]] &
    karate.local$runs$status == "ok"
][[1L]]
karate.local.best.coords <- karate.local$layouts[[karate.local.best$candidate[[1L]]]][[as.character(karate.local.best.seed)]]
```

```{r fig.width=8, fig.height=4.2}
plot.layout.views(
  karate.local.best.coords,
  karate.edges,
  title = paste("Karate local-search winner:", karate.local.best$candidate[[1L]])
)
```

### Reading the quality metrics

The layout-quality metrics are meant to be simple heuristics rather than formal
proofs that one picture is correct:

- lower `sampled.stress` means graph distances and Euclidean distances are more
  consistent, so vertices that are close in the graph tend to stay close in the layout,
- lower `edge.length.cv` means adjacent vertices are drawn with more uniform edge lengths,
- higher `sampled.nonedge.sep.ratio` means non-neighboring vertices have more breathing room,
- lower `stability.procrustes.mean` means repeated runs give a more similar picture after alignment,
- higher `cluster.separation` means known groups are more clearly separated,
- `edge.crossings` matters mainly for 2D layouts, so it is not used in the 3D examples here.

No single metric should dominate the decision. In practice, a good candidate is
one that scores well on several of them at once and still looks interpretable.

## Using external metadata: the observed karate-club split

The karate data come with real external labels: the eventual split between the
`Mr. Hi` and `Officer` factions reported in the original study [@zachary1977].
Those labels can be passed through `clusters` so that `grip.compare.layouts()`
also reports `cluster.separation`.

```{r}
karate.metadata <- grip.compare.layouts(
  edges = karate.edges,
  n = karate.n,
  dim = 3,
  candidates = list(
    default = NULL,
    tuned = candidate.from.summary(karate.local.best)
  ),
  clusters = karate.club,
  seeds = 1:3,
  sample.size.stress = 1000L,
  sample.size.nonedge = 2000L,
  edge.crossings = "never",
  return.layouts = TRUE
)

knitr::kable(
  karate.metadata$summary[, c(
    "candidate",
    "score.composite",
    "sampled.stress.mean",
    "sampled.nonedge.sep.ratio.mean",
    "cluster.separation.mean"
  )],
  digits = 3
)
```

```{r}
karate.metadata.best <- karate.metadata$summary[1, , drop = FALSE]
karate.metadata.best.seed <- karate.metadata$runs$seed[
  karate.metadata$runs$candidate == karate.metadata.best$candidate[[1L]] &
    karate.metadata$runs$status == "ok"
][[1L]]
karate.metadata.best.coords <- karate.metadata$layouts[[karate.metadata.best$candidate[[1L]]]][[as.character(karate.metadata.best.seed)]]
karate.club.cols <- ifelse(karate.club == "Mr. Hi", "#1b9e77", "#d95f02")
```

```{r fig.width=8, fig.height=4.2}
plot.layout.views(
  karate.metadata.best.coords,
  karate.edges,
  title = paste("Karate best layout with metadata:", karate.metadata.best$candidate[[1L]]),
  vertex.col = karate.club.cols
)
legend(
  "bottomleft",
  legend = c("Mr. Hi", "Officer"),
  col = c("#1b9e77", "#d95f02"),
  pch = 16,
  bty = "n"
)
```

This is where metadata become genuinely useful: if a layout already looks good
geometrically, `cluster.separation` gives you one more way to ask whether known
groups are being displayed clearly rather than mixed together.

## A second example: the Krackhardt kite network

The Krackhardt kite graph is small, but it is still a useful example because it
has recognizable organizational structure and is fast enough for a targeted
local search.

```{r}
kite.compare <- grip.compare.layouts(
  edges = kite.edges,
  n = kite.n,
  dim = 3,
  candidates = c("default", "tree"),
  search = list(
    candidate.prefix = "kite.tree.local",
    preset = "tree",
    final_rounds = c(128L, 160L, 192L),
    num_nbrs = c(6L, 8L),
    repulsion_factor = c(0.0, 0.5)
  ),
  seeds = 1:2,
  sample.size.stress = 400L,
  sample.size.nonedge = 800L,
  edge.crossings = "never",
  return.layouts = TRUE
)

knitr::kable(head(compact.summary(kite.compare$summary), 8L), digits = 3)
```

```{r}
kite.best <- kite.compare$summary[1, , drop = FALSE]
kite.best.seed <- kite.compare$runs$seed[
  kite.compare$runs$candidate == kite.best$candidate[[1L]] &
    kite.compare$runs$status == "ok"
][[1L]]
kite.best.coords <- kite.compare$layouts[[kite.best$candidate[[1L]]]][[as.character(kite.best.seed)]]
```

```{r fig.width=8, fig.height=4.2}
plot.layout.views(
  kite.best.coords,
  kite.edges,
  title = paste("Krackhardt kite winner:", kite.best$candidate[[1L]])
)
```

## When `grip.score.layout()` is useful directly

The main workflow for real data is usually `grip.compare.layouts()`, because
layout quality only becomes meaningful once there is something to compare
against. The lower-level `grip.score.layout()` function is still useful when you
already have a chosen layout, perhaps from a cached run or from another package,
and you want to report its metrics explicitly.

```{r}
knitr::kable(
  grip.score.layout(
    coords = karate.metadata.best.coords,
    edges = karate.edges,
    n = karate.n,
    clusters = karate.club,
    edge.crossings = "never"
  ),
  digits = 3
)
```

## Practical guidance

For real-world graphs, a good default workflow is:

1. Start in 3D with the default layout plus one or two structurally plausible presets.
2. Keep only the strongest one or two candidates after a short comparison across a few seeds.
3. Run a small local search around the most promising candidate with `grip.compare.layouts(search = ...)`.
4. Use `sampled.stress`, `edge.length.cv`, `sampled.nonedge.sep.ratio`, and `stability.procrustes.mean` as the main generic metrics.
5. Add `cluster.separation` when you have real metadata such as communities, classes, or known splits.
6. Use `grip.score.layout()` directly only when you already have a layout in hand and want to annotate it.

## References
