---
title: "Choosing Layouts for Real Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Layouts for Real Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

## Why scoring helps

For real-world graphs there is usually no canonical embedding, so layout choice
cannot rely on a direct geometric error like Procrustes RMSE. In practice, the
most useful workflow is:

1. try the default layout plus one or two structurally plausible presets,
2. compare them across several seeds, and
3. inspect a small set of quality metrics before choosing a final plot.

The `grip` package provides two helpers for this:

- `grip.score.layout()` scores one realized layout,
- `grip.compare.layouts()` compares several presets or parameter sets across seeds.

The examples below use two small real-world benchmark networks bundled with the
package under `inst/extdata/`.

```{r}
library(grip)
```

```{r}
read.edge.csv <- function(file.name) {
  candidates <- c(
    system.file("extdata", file.name, package = "grip"),
    file.path("inst", "extdata", file.name),
    file.path("..", "inst", "extdata", file.name)
  )
  path <- candidates[file.exists(candidates)][1L]
  if (!length(path) || is.na(path) || !nzchar(path)) {
    stop("could not locate ", file.name)
  }
  as.matrix(utils::read.csv(path))
}

compact.summary <- function(x) {
  keep <- c(
    "candidate",
    "score.composite",
    "sampled.stress.mean",
    "edge.length.cv.mean",
    "sampled.nonedge.sep.ratio.mean",
    "stability.procrustes.mean",
    "edge.crossings.mean"
  )
  x[, keep]
}
```

## Scoring one layout

As a first example, we score a `preset = "tree"` layout of the Krackhardt kite
social network, a classic small organizational-network benchmark.

```{r}
kite.edges <- read.edge.csv("krackhardt-kite-edges.csv")
kite.n <- max(kite.edges)

kite.coords <- grip.layout(
  kite.edges,
  n = kite.n,
  dim = 2,
  preset = "tree",
  seed = 1
)

kite.score <- grip.score.layout(
  coords = kite.coords,
  edges = kite.edges,
  n = kite.n,
  edge.crossings = "always"
)

knitr::kable(kite.score, digits = 3)
```

```{r}
grip.plot(kite.coords, kite.edges, main = "Krackhardt kite with preset = 'tree'",
          pch = 16, cex = 1)
```

Interpretation:

- lower `sampled.stress` means Euclidean distances better track graph distances,
- lower `edge.length.cv` means adjacent vertices are drawn more uniformly,
- higher `sampled.nonedge.sep.ratio` means less clutter,
- lower `edge.crossings` is usually preferable in 2D.

## Comparing presets across seeds

The Zachary karate club friendship network is a good example of a real graph
where there is no obviously correct preset ahead of time.

```{r}
karate.edges <- read.edge.csv("karate-club-edges.csv")
karate.n <- max(karate.edges)

karate.compare <- grip.compare.layouts(
  edges = karate.edges,
  n = karate.n,
  dim = 2,
  candidates = c("default", "tree", "mesh"),
  seeds = 1:3,
  sample.size.stress = 1000L,
  sample.size.nonedge = 2000L,
  edge.crossings = "always",
  return.layouts = TRUE
)

knitr::kable(compact.summary(karate.compare$summary), digits = 3)
```

The `score.composite` column is a rank-based summary of the available metrics,
so lower values are better. It is meant to help shortlist candidates rather
than replace visual judgment.

```{r}
best.karate <- karate.compare$summary$candidate[[1L]]
best.seed <- karate.compare$runs$seed[
  karate.compare$runs$candidate == best.karate &
    karate.compare$runs$status == "ok"
][[1L]]

best.karate.coords <- karate.compare$layouts[[best.karate]][[as.character(best.seed)]]

grip.plot(best.karate.coords, karate.edges,
          main = paste("Karate club best candidate:", best.karate),
          pch = 16, cex = 0.8)
```

## Comparing presets with a custom candidate

You can mix presets and explicit parameter lists in the same comparison.

```{r}
kite.compare <- grip.compare.layouts(
  edges = kite.edges,
  n = kite.n,
  dim = 2,
  candidates = list(
    default = NULL,
    tree = "tree",
    tree.long = list(preset = "tree", final_rounds = 224)
  ),
  seeds = 1:3,
  sample.size.stress = 800L,
  sample.size.nonedge = 1500L,
  edge.crossings = "always"
)

knitr::kable(compact.summary(kite.compare$summary), digits = 3)
```

## Using external metadata

If you have known communities or other node labels, pass them through the
`clusters` argument. `grip.score.layout()` and `grip.compare.layouts()` will
then report `cluster.separation`, which is higher when groups are more cleanly
separated in the layout.

```r
club.labels <- c(...)  # one label per vertex

grip.score.layout(
  coords = best.karate.coords,
  edges = karate.edges,
  n = karate.n,
  clusters = club.labels
)
```

## Practical guidance

For real-world data, a good default workflow is:

1. Start with the default layout plus one or two structurally plausible presets.
2. Compare them across several seeds with `grip.compare.layouts()`.
3. Use `sampled.stress`, `edge.length.cv`, and `sampled.nonedge.sep.ratio` as
   the main generic metrics.
4. In 2D, include `edge.crossings` when readability matters.
5. If you have node metadata, add `cluster.separation`.
6. Inspect the top one or two candidates visually before making a final choice.
