---
title: "Choosing and evaluating a layout"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Choosing and evaluating a layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE, comment='#>', fig.width=9, fig.height=5,
                     out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

A useful layout reveals graph structure while keeping distance distortion and
sensitivity to initialization in view. This guide compares a small set of
settings on **494_bus**, a 494-vertex power network supplied with grip.
Its source electrical values are not layout distances: we use the bundled
unit-length connectivity graph.

## Choose a method for the question

| Question | Starting point |
|---|---|
| Explore graph connectivity | `grip(metric='hop')` |
| Respect positive traversal lengths | `grip(metric='edge_length')` |
| Fit all shortest-path distances directly | `metric.mds()`; SGD is the default backend |
| Reduce the cost of full MDS | `metric.mds(approximation='sparse')` |
| Adjust edge lengths of an existing drawing | `edge.kk(coords=...)`, followed by independent scoring |

The [function overview](function-guide.html) explains input conventions and
provides the complete categorized catalog. [Metric MDS](metric-mds-backends.html)
compares its two optimizers; [weighted layouts](weighted-grip-intro.html)
explains distance-valued edges.

## Compare settings across seeds

```{r candidates}
data(zheng.graphs)
g <- zheng.graphs[['494_bus']]
cmp <- compare.layouts(edges=g$edges,n=g$n,dim=3,
  candidates=c('default','tree','mesh'),seeds=1:3,
  sample.size.stress=2000L,sample.size.nonedge=3000L,
  edge.crossings='never',return.layouts=TRUE)
knitr::kable(cmp$summary[,c('candidate','sampled.stress.mean',
  'edge.length.cv.mean','sampled.nonedge.sep.ratio.mean',
  'stability.procrustes.mean')],digits=3)
```

These are three GRIP presets with three seeds each. They are a teaching
comparison, not tuned recommendations for power networks. Sampled stress is
smaller when drawn distances better match graph shortest paths. Stability
summarizes how much layouts vary after alignment. Use the same scoring samples
and definitions across candidates.

```{r candidate-view, echo=FALSE}
layouts <- lapply(cmp$layouts,function(z) z[['1']])
names(layouts) <- paste('GRIP:',names(layouts))
documentation.view(layouts,g$edges,
  description='494_bus: three GRIP presets, seed 1 in every view. Rigid alignment; no resizing.')
```

The figure always shows seed 1; the table summarizes all three seeds. Rotate
and switch presets to inspect branches and crowded regions. Neither the
matrix nor these layouts provide a known geographical reference.

## Select settings without hiding variability

`cmp$runs` retains the individual runs. A good average should be considered
alongside variation among seeds and any unsuccessful attempts.

```{r runs}
knitr::kable(cmp$runs[,intersect(c('candidate','seed','status','sampled.stress'),
                               names(cmp$runs))],digits=3)
```

Use `params.from.summary()` to recover a candidate's settings. If further
improvement matters, try a small neighborhood of those settings and retain the
initial comparison as a separate result. A composite score ranks candidates
within its chosen criteria and weights; it is not a universal quality grade.

## Choose an appropriate score

| What matters? | Function | What the score means |
|---|---|---|
| Overall graph-distance agreement | `score.layout()` | Sampled drawn distances versus graph shortest paths, with fitted scale |
| Edge, route, and endpoint errors separately | `score.gmds()` | Distinct residuals with an explicit scale policy |
| Locations of corresponding vertices | `score.coordinates()` | Vertex displacement after the requested alignment |
| Shape of an already aligned triangular surface | `score.surface()` | Area-weighted surface distance; does not align the surfaces |

Score the original 3D coordinates, not their 2D projection. An improvement in
edge lengths can accompany worse distances between distant vertices. Likewise,
a close surface overlay does not establish recovery of corresponding vertices.
See the [score definitions](function-guide.html#choose-a-score) for units,
alignment choices, and the differences between these quantities.

For graphs with meaningful labels, examine those labels alongside the geometry.
The [HMP/U01 case study](hmp-u01-case-study.html) illustrates this with biological
community labels. Label separation alone does not establish accurate geometry.

## Keep computation proportional to the question

Full MDS requires dense distances. Use `estimate.preparation()` before a large
preparation; its matrix-size estimate is a lower bound, not peak memory.
Start with a few seeds and candidates. Use [traces](grip-trace-and-diagnostics.html)
to investigate a fold or collapse before increasing a search indiscriminately.

The [historical engine comparison](grip-vs-alternatives.html) preserves the
published-in-this-package grip 0.2.0-era results and their original settings.
Those results do not benchmark the current metric-MDS backend.
