---
title: "Tracing and Diagnosing Layouts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracing and Diagnosing Layouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.8,
  fig.height = 5.2
)
```

## Why trace a layout?

Most graph-layout functions return only the final coordinate matrix. That is
enough for plotting, but it hides the most useful diagnostic information:

- how the multiscale hierarchy opened,
- whether the geometry stabilized gradually or abruptly,
- whether a parameter setting improved edge-length regularity or only moved the
  final picture around,
- and whether a weighted solve is actually getting closer to the intended
  metric geometry.

The `trace.grip()` API makes those questions visible through the same metric
choice as `grip()`:

- `trace.grip(metric = "hop")` for combinatorial layouts,
- `trace.grip(metric = "edge_length")` for edge-length geometry-aware layouts.

```{r}
library(grip)
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

## A first combinatorial trace

The example below traces a planar recursive triangle. Its nested holes make
the multiscale construction visible. Using
`trace = "level"` keeps the vignette readable by recording the coarse
initialization, the start of each new level, and the final layout.

```{r}
mesh.edges <- edges.sierpinski.triangle(4)

mesh.trace <- trace.grip(
  mesh.edges,
  n = max(mesh.edges),
  dim = 2,
  preset = "carpet",
  rounds = 48,
  final.rounds = 96,
  trace = "level",
  diagnostics = "light",
  seed = 1
)

knitr::kable(mesh.trace$meta)
knitr::kable(mesh.trace$diagnostics[, c(
  "frame",
  "edge.length.cv",
  "median.edge.length",
  "sampled.nonedge.sep.ratio"
)], digits = 3)
```

```{r trace-view, echo=FALSE}
documentation.trace(mesh.trace,mesh.edges)
```

Move the slider to inspect recorded stages of the construction. Bounds stay
fixed, and each stage shows only its active vertices and edges. This is a
planar graph, so a 2D view makes its recursive holes easier to inspect.


The two most useful tables are:

- `meta`, which explains where each frame came from,
- `diagnostics`, which tracks simple quality signals frame by frame.

## Level traces versus round traces

When you need more detail, `trace = "round"` records intermediate refinement
rounds within each active level.

```{r}
mesh.trace.round <- trace.grip(
  mesh.edges,
  n = max(mesh.edges),
  dim = 2,
  preset = "carpet",
  rounds = 6,
  final.rounds = 8,
  trace = "round",
  trace.every = 4,
  diagnostics = "light",
  seed = 1
)

data.frame(
  trace_mode = c("level", "round"),
  n.frames = c(length(mesh.trace$frames), length(mesh.trace.round$frames))
)
```

Use:

- `trace = "level"` for most reporting and teaching,
- `trace = "round"` when you are diagnosing a tuning issue.

## Read changes over the trace

```{r diagnostic-curve, echo=FALSE}
plot(mesh.trace$diagnostics$frame,mesh.trace$diagnostics$edge.length.cv,
  type='b',pch=16,col='#1769AA',xlab='Recorded frame',ylab='Edge-length variation')
```

Early frames contain only the active coarse vertices. Read changes in the
score together with the stage and active graph; adjacent frames may not score
identical vertex sets. Level traces show insertion stages; round traces show
more of the refinement within each stage. They are different sampling schedules.

For a weighted graph, use `trace.grip(metric='edge_length', ...)` with its
lengths. The [weighted guide](weighted-grip-intro.html) explains this metric.
Detailed construction animations and their reproduction commands are described
in the [asset guide](https://github.com/pgajer/grip/blob/main/tools/README.md#readme-and-showcase-assets).

## Trace objects are analysis objects

A trace object is more than a list of pictures. It can support:

- frame-by-frame diagnostics,
- manual inspection of where a solve destabilizes,
- teaching demonstrations of the multiscale hierarchy,
- and regression checks for future algorithm changes.

The most useful components are:

- `final`: the final coordinate matrix,
- `frames`: the stored coordinate frames,
- `meta`: frame provenance,
- `diagnostics`: per-frame metrics,
- `lgkk.polish`: optional post-layout LGKK result when enabled as an advanced
  experimental refinement step.

## Practical guidance

- Start with `trace = "level"` and `diagnostics = "light"`.
- Use `trace = "round"` only when you need finer-grained diagnostics.
- Supply `target.coords` when a meaningful geometric target exists.
- Prefer weighted traces for weighted graph families; they are much more
  interpretable than tracing the combinatorial solver on a graph with
  nontrivial edge lengths.

## Where to go next

For an API map, see [Finding your way around grip](function-guide.html).
For reproducible bundles and reference scoring, see
[Graph examples](synthetic-graph-families.html).
These guides and the interactive-explorer walkthrough are installed vignettes.



- `Getting Started with grip` gives the short overview of the main APIs.
- `Weighted Graph Layouts with grip` focuses on weighted solving and scoring.
- `Choosing Layouts for Real Data` shows how trace fits into a real-data
  selection workflow.
