---
title: "Interactive Layout Exploration with gripui"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Layout Exploration with gripui}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5.2
)

options(rgl.useNULL = TRUE)

gripui_available <- all(vapply(
  c("shiny", "bslib", "DT", "htmltools", "rgl"),
  requireNamespace,
  logical(1L),
  quietly = TRUE
))

if (gripui_available && requireNamespace("rgl", quietly = TRUE)) {
  rgl::setupKnitr(autoprint = TRUE)
}
```

## Overview

The `gripui` module provides a Shiny-based explorer for comparing graph
layouts produced by `grip.compare.layouts()` or assembled from on-disk
artifacts. The interface combines a scatter-plot "landscape" of layout quality
scores, a filterable catalog table, and an interactive 2D or 3D viewer for the
selected layout.

The public API consists of six functions:

| Function | Purpose |
|:--|:--|
| `gripui_project()` | Build a project from an in-memory graph and a layouts data frame |
| `gripui_project_from_compare()` | Convert `grip.compare.layouts()` output directly |
| `gripui_project_from_dir()` | Assemble a project from saved artifacts on disk |
| `gripui_validate_project()` | Check a project object for structural correctness |
| `gripui_app()` | Build the Shiny app object (for embedding or testing) |
| `run_gripui()` | Launch the Shiny app in a browser |

All of these require the optional packages listed in `Suggests` (`shiny`,
`bslib`, `DT`, `htmltools`, `rgl`). If any are missing, the functions stop
with a clear message listing the missing packages.

## Quick start: from `grip.compare.layouts()`

The most common entry point is to run a layout comparison and then explore the
results interactively. The example below uses a small mesh graph so the
vignette builds quickly, but the same pattern works for graphs of any size.

```{r compare-example}
library(grip)

edges <- edges.mesh(5, 5)
n <- 25L

cmp <- grip.compare.layouts(
  edges = edges,
  n = n,
  dim = 2,
  candidates = list(
    default = list(),
    tuned = list(
      rounds = 40L,
      final_rounds = 80L,
      num_init = 4L,
      num_nbrs = 6L,
      r = 0.05,
      s = 4.5
    )
  ),
  seeds = 1:3,
  return.layouts = TRUE
)

knitr::kable(
  cmp$summary[, c("candidate", "sampled.stress.mean",
                   "edge.length.cv.mean", "score.composite")],
  digits = 4
)
```

With the comparison result in hand, `gripui_project_from_compare()` wraps it
into a project object that the Shiny app understands. Passing a graph enables
edge drawing and vertex-data coloring in the viewer.

```{r project-from-compare, eval=gripui_available}
adj <- vector("list", n)
for (i in seq_len(nrow(edges))) {
  adj[[edges[i, 1L]]] <- c(adj[[edges[i, 1L]]], edges[i, 2L])
  adj[[edges[i, 2L]]] <- c(adj[[edges[i, 2L]]], edges[i, 1L])
}
graph <- list(adj_list = lapply(adj, function(x) sort(unique(as.integer(x)))))

project <- gripui_project_from_compare(
  cmp,
  graph = graph,
  title = "5x5 mesh comparison"
)
project$meta$title
nrow(project$layouts)
```

To launch the app you would call `run_gripui(project)`. Since vignettes
cannot start a live Shiny session, the rest of this document illustrates the
building blocks without calling `run_gripui()`.

## Building a project by hand

When layouts come from sources other than `grip.compare.layouts()` — for
example from a custom pipeline or an external optimizer — you can build the
project manually with `gripui_project()`.

The `layouts` argument is a data frame with at least `candidate`, `stage`,
`seed`, and `status` columns. Additional columns such as `coords_path`,
`html_path`, and any numeric quality scores are picked up automatically.

```{r manual-project, eval=gripui_available}
layouts_df <- data.frame(
  candidate = c("run_a", "run_b"),
  stage     = c("layout", "layout"),
  seed      = c(1L, 2L),
  status    = c("ok", "ok"),
  my_score  = c(0.42, 0.37),
  stringsAsFactors = FALSE
)

manual_project <- gripui_project(
  graph = NULL,
  layouts = layouts_df,
  title = "Manual example"
)
manual_project$meta$title
names(manual_project$layouts)
```

The constructor normalizes column names, generates unique `layout_id` values,
and marks each row as `interactive`, `artifact-only`, or `summary-only`
depending on whether in-memory coordinates, on-disk coordinate files, or
neither are available.

## Validation

`gripui_validate_project()` checks that the required columns are present, that
`layout_id` values are unique, and that graph objects are structurally sound.
It is called automatically by the constructor, but you can also call it
explicitly after modifying a project.

```{r validate, eval=gripui_available}
gripui_validate_project(project)
```

If the graph is a named list of sub-graphs (for multi-stage projects), the
validator also checks that every `graph_id` referenced in the layouts table
resolves to a key in the graph list.

## Loading a project from disk

For larger projects where layout artifacts have been saved to a directory,
`gripui_project_from_dir()` assembles the project by reading catalog TSVs,
resolving relative paths, and attaching artifact manifests. This function
currently recognizes two directory layouts:

1. **Normalized bundle format** — a `catalog/layout_catalog.tsv` file with an
   optional `graph/graph.rds` and `metadata/notes.md`.
2. **HMP/U01 compatibility format** — stage-specific run tables
   (`coarse_stage/coarse_stage_runs.tsv`, `full_stage1/full_stage1_runs.tsv`,
   etc.) with optional visual manifests.

```r
# Not run — requires a directory of saved outputs
project <- gripui_project_from_dir(
  root = "/path/to/saved/outputs",
  title = "My HMP project"
)
run_gripui(project)
```

## The gripui Shiny interface

When launched, the app presents four panels:

1. **Sidebar** — stage filters, candidate search, axis and color pickers, and
   edge-display toggle.
2. **Landscape scatter plot** — plots any two numeric columns from the catalog
   against each other, colored by a third variable. Clicking a point selects
   that layout.
3. **Viewer** — an interactive 3D `rgl` widget or a 2D base-graphics plot
   depending on the coordinate dimensionality, with optional edge overlay and
   vertex coloring by any column in `vertex_data`.
4. **Catalog table** — a searchable, sortable `DT` datatable. Clicking a row
   selects that layout.

The landscape and the table stay synchronized: selecting a layout in either
panel highlights it in both.

## Graph objects

The `graph` argument accepted by the project constructors and the app is a
list that may contain any combination of:

| Field | Type | Purpose |
|:--|:--|:--|
| `adj_list` | list of integer vectors | Adjacency list for edge drawing |
| `weight_list` | list of numeric vectors | Edge weights (same structure as `adj_list`) |
| `vertex_data` | data frame | Per-vertex metadata for coloring |
| `graph_info` | list | Graph-level metadata (currently informational) |

For multi-stage projects where different stages use different graphs (e.g., a
coarsened graph and the full graph), pass a named list of such graph objects.
The `graph_id` column in the layouts table is then used to look up the correct
sub-graph for each row.

## Vertex coloring

The viewer offers a "Viewer color" selector. The special value `"plain"` draws
all vertices in a single color. Any column name from `vertex_data` enables
data-driven coloring:

- **Numeric columns** map to a sequential blue palette via
  `grDevices::colorRampPalette()`. Non-finite values are drawn in gray.
- **Categorical columns** map to `grDevices::hcl.colors()` with the "Dark 3"
  palette. `NA` values are drawn in gray.

## Tips for CRAN-compatible usage

Because `gripui` depends on Shiny-ecosystem packages that are listed in
`Suggests`, all the gripui functions guard their availability at runtime. If
you write code that calls `gripui_app()` or `run_gripui()` in a context
where those packages might not be installed (e.g., in a package that re-exports
gripui functionality), wrap the call in a
`if (requireNamespace("shiny", quietly = TRUE))` check or use `@examplesIf`
in roxygen blocks, as the grip package itself does.
