---
title: "Getting Started with grip"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with grip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.5,
  fig.height = 5.2
)
```

## What this vignette covers

`grip` has a small set of entry points that cover most package use:

- `grip.layout()` for ordinary unweighted graphs,
- `grip.compare.layouts()` when you want to compare a short list of candidate
  settings,
- `grip.layout.legacy()` for backward-compatible reproduction,
- `grip.layout.weighted()` when edge lengths carry geometry you want to keep.

This vignette is the shortest path through the default workflow. It shows how
to:

- compute a first unweighted layout,
- score it,
- compare a few plausible candidates,
- and decide when to switch to the specialist guides.

For weighted layouts, real-data search, tracing, and interactive exploration,
the later guides go deeper.

```{r}
library(grip)
```

The helper below keeps side-by-side comparisons compact.

```{r}
plot.layout.pair <- function(coords.left,
                             coords.right,
                             edges,
                             titles,
                             projection = NULL,
                             vertex.cols = c("black", "black"),
                             edge.col = "gray82") {
  op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3, 1.2), bg = "white")
  on.exit(par(op), add = TRUE)

  grip.plot(
    coords.left, edges,
    projection = projection,
    main = titles[[1]],
    vertex.col = vertex.cols[[1]],
    edge.col = edge.col
  )
  grip.plot(
    coords.right, edges,
    projection = projection,
    main = titles[[2]],
    vertex.col = vertex.cols[[2]],
    edge.col = edge.col
  )
}
```

## A first unweighted layout

For an ordinary unweighted graph, `grip.layout()` is the default starting
point. Here is a small mesh in 2D.

```{r}
mesh.edges <- edges.mesh(5, 5)
mesh.coords <- grip.layout(
  mesh.edges,
  n = 25,
  dim = 2,
  preset = "mesh",
  seed = 1
)

mesh.score <- grip.score.layout(mesh.coords, edges = mesh.edges, n = 25)
knitr::kable(mesh.score[, c(
  "sampled.stress",
  "edge.length.cv",
  "sampled.nonedge.sep.ratio"
)], digits = 3)
```

```{r fig.width=5.2, fig.height=4.6}
grip.plot(
  mesh.coords,
  mesh.edges,
  main = "grip.layout() on a 5x5 mesh",
  pch = 16,
  cex = 0.65,
  edge.col = "gray82"
)
```

For small and medium unweighted graphs, that is often all you need:

- choose `dim = 2` or `dim = 3`,
- optionally start from a preset,
- and score or compare a few candidate settings if the graph is important.

## Compare a few plausible candidates

If the first picture matters, it is usually better to compare a short candidate
list than to tune blindly. `grip.compare.layouts()` runs several seeds and
summarizes the results in a score table.

```{r}
mesh.cmp <- grip.compare.layouts(
  edges = mesh.edges,
  n = 25,
  dim = 2,
  candidates = c("default", "mesh", "tree"),
  seeds = 1:2,
  sample.size.stress = 500L,
  sample.size.nonedge = 1000L,
  edge.crossings = "never"
)

knitr::kable(mesh.cmp$summary[, c(
  "candidate",
  "sampled.stress.mean",
  "edge.length.cv.mean",
  "sampled.nonedge.sep.ratio.mean",
  "score.composite"
)], digits = 3)
```

That same pattern scales to real graphs:

- shortlist a few plausible presets,
- run several seeds,
- inspect the summary table before choosing a favorite picture,
- then search locally only if the graph is important enough to justify it.

## Current GRIP versus legacy GRIP

The package still keeps the historical entry point through
`grip.layout.legacy()`. That is useful for backward-compatible reproduction,
but the recommended path for new work is `grip.layout()`.

```{r}
mesh.current <- grip.layout(
  mesh.edges,
  n = 25,
  dim = 2,
  preset = "mesh",
  seed = 2
)

mesh.legacy <- grip.layout.legacy(
  mesh.edges,
  n = 25,
  dim = 2,
  seed = 2
)
```

```{r fig.width=9.4, fig.height=4.2}
plot.layout.pair(
  mesh.legacy,
  mesh.current,
  edges = mesh.edges,
  titles = c("legacy", "current"),
  vertex.cols = c("black", "#1F3B73")
)
```

The practical distinction is:

- use `grip.layout.legacy()` when you need the historical local-force behavior,
- use `grip.layout()` for the current package default,
- and treat the legacy path as a reproduction tool rather than a new default.

## When should you switch workflows?

Start with `grip.layout()` when the graph is fundamentally unweighted and you
mainly care about its combinatorial structure.

Switch to the other guides when the task changes:

- use `grip.layout.weighted()` when edge lengths encode geometry you care
  about,
- use `grip.compare.layouts()` on real graphs when you want a disciplined
  shortlist rather than a single run,
- use `grip.layout.trace()` or `grip.layout.trace.weighted()` when you want to
  inspect how a solve evolves,
- use `run_gripui()` or `run_gripui_family()` in an interactive R session when
  you want app-based exploration.

## Where to go next

- `Weighted Graph Layouts with grip` covers weighted solving, geodesic scoring,
  and 2D-versus-3D decisions.
- `Choosing Layouts for Real Data` focuses on candidate shortlisting, local
  search, and real-data evaluation.
- `Tracing and Diagnosing Layouts` covers trace objects and per-frame
  diagnostics.
- `Interactive Exploration with gripui` is a website article about the
  package's Shiny tools.
- `Synthetic Graph Families and Geometries` is a website article about the
  benchmark and geometry library.
