---
title: "Interactive exploration with gripui"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Interactive exploration with gripui}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE,comment='#>',fig.width=9,fig.height=5,out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

The project explorer lets you select saved layouts, inspect their scores, and
rotate the selected graph. The graph-family explorer instead constructs
synthetic examples from parameters. Both are optional Shiny applications.

## Prepare an application graph

This walkthrough uses the same **dwt_307** graph as Getting Started. Two presets
and two seeds provide a small collection to browse.

```{r compare-example}
data(zheng.graphs)
g <- zheng.graphs[['dwt_307']]
cmp <- compare.layouts(edges=g$edges,n=g$n,dim=3,
  candidates=c('default','tree'),seeds=1:2,return.layouts=TRUE,
  edge.crossings='never')
adj <- vector('list',g$n)
for(i in seq_len(nrow(g$edges))) {
  a <- g$edges[i,1];b <- g$edges[i,2]
  adj[[a]] <- c(adj[[a]],b);adj[[b]] <- c(adj[[b]],a)
}
project <- gripui.project.from.compare(cmp,
  graph=list(adj_list=adj),title='dwt_307: presets and seeds')
gripui.validate.project(project)
```

## Launch and explore

```{r launch, eval=FALSE}
run.gripui(project)
```

The app requires **shiny**, **bslib**, **DT**, **htmltools**, and **rgl**, all
listed as optional dependencies of grip. Launching it is an interactive step;
building this vignette does not start a server.

| Step | What to do | What to inspect |
|---|---|---|
| 1. Choose a result | Select a row in the layout catalog | Candidate, seed, status, and available coordinates |
| 2. Inspect the drawing | Rotate the graph and compare another row | Branches, folds, and crowded regions |
| 3. Read the scores | Compare the score landscape and table | Distance agreement alongside visual differences |
| 4. Check stability | Compare seeds within a candidate | Whether the apparent structure persists |

The figure below previews two layouts from this project. It is an ivue figure,
not a screenshot of the Shiny application. The app provides the catalog and
score-selection workflow described above.

```{r project-view, echo=FALSE}
documentation.view(list('Default: seed 1'=cmp$layouts$default[['1']],
  'Tree: seed 1'=cmp$layouts$tree[['1']]),g$edges,
  description='Saved project layouts: default and tree GRIP presets. Rigid alignment; no resizing.')
```

```{r project-table}
knitr::kable(project$layouts[,c('layout_id','candidate','seed','status','availability')])
```

## Explore graph families

For a new synthetic construction rather than saved layout results, launch:

```{r families, eval=FALSE}
run.gripui.family()
```

Choose a graph family, alter its construction parameters, and inspect its
geometry. The [graph gallery](synthetic-graph-families.html) introduces the
families and their ordinary R recipes; the function overview lists all
[interactive exploration functions](function-guide.html#interactive-exploration).

## Import existing results

Use `gripui.project()` to assemble results from another source, or
`gripui.project.from.dir()` to read an existing output directory. Keep vertex
ordering consistent between graph data and every coordinate matrix.

```{r from-directory, eval=FALSE}
project <- gripui.project.from.dir(root='/path/to/saved/outputs',
  title='My saved search')
gripui.validate.project(project)
run.gripui(project)
```

The function help documents the project fields and directory structure.
For layout selection principles, see [Choosing and evaluating](grip-real-data.html).
