---
title: "Selecting Layouts on the Coarsened HMP/U01 Graph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting Layouts on the Coarsened HMP/U01 Graph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.8
)
```

## Why this graph is useful

The package now ships a coarsened real-world graph derived from the giant
connected component of the HMP+U01 16S amplicon iKNN graph. This is a useful
layout-selection example because it is large enough to behave like a real
analysis graph, but still small enough to bundle with the package and to inspect
reproducibly.

The bundled object is `hmp.u01.gc.coarse`. It contains:

- a weighted undirected graph with `1828` coarse vertices,
- aggregated labels including `cst` and `subcst`,
- graph-level provenance describing the upstream `>=1%` relative-abundance +
  PCA branch and the coarsening path from `6474` to `1828` vertices.

```{r}
library(grip)

data(hmp.u01.gc.coarse)

graph.info <- data.frame(
  quantity = c(
    "source dataset",
    "representation",
    "selected k",
    "original giant-component vertices",
    "coarsened vertices",
    "weighted undirected edges"
  ),
  value = c(
    hmp.u01.gc.coarse$graph_info$source_dataset,
    hmp.u01.gc.coarse$graph_info$representation,
    hmp.u01.gc.coarse$graph_info$selected_k,
    hmp.u01.gc.coarse$graph_info$original_vertices,
    hmp.u01.gc.coarse$graph_info$coarse_vertices,
    hmp.u01.gc.coarse$graph_info$edge_count
  ),
  stringsAsFactors = FALSE
)

knitr::kable(graph.info)
```

```{r}
hmp.coarsening.rounds <- hmp.u01.gc.coarse$graph_info$coarsening_rounds
knitr::kable(hmp.coarsening.rounds)
```

The raw bundled provenance files are also available under:

```{r}
system.file("extdata", "hmp_u01_gc_coarse", package = "grip")
```

## A lightweight real-data workflow

For a graph like this there is no canonical embedding, so layout choice is a
model-selection problem rather than a reconstruction problem. In practice the
workflow is:

1. shortlist a few plausible presets,
2. compare them in 3D,
3. run a small local search around the strongest region, and
4. inspect intrinsic and label-aware metrics together.

The full HMP/U01 search is too expensive to rebuild inside a CRAN vignette, so
this vignette uses precomputed comparison results bundled with the package. That
keeps the vignette fast while still showing an actual real-data workflow.

```{r}
read.hmp.vignette.results <- function() {
  path <- system.file(
    "extdata", "hmp_u01_gc_coarse", "vignette_results.rds",
    package = "grip"
  )
  if (!nzchar(path)) {
    stop("could not locate bundled vignette results")
  }
  readRDS(path)
}

hmp.results <- read.hmp.vignette.results()
```

## Projecting 3D layouts for static figures

The layouts below are true 3D layouts. For a vignette page we need a static
projection, so each panel shows a fixed 3D view rather than a 2D layout.

```{r}
project.coords <- function(coords, azimuth = 35, elevation = 22) {
  az <- azimuth * pi / 180
  el <- elevation * pi / 180
  rz <- matrix(c(
    cos(az), -sin(az), 0,
    sin(az),  cos(az), 0,
    0,        0,       1
  ), nrow = 3, byrow = TRUE)
  rx <- matrix(c(
    1, 0, 0,
    0, cos(el), -sin(el),
    0, sin(el),  cos(el)
  ), nrow = 3, byrow = TRUE)
  rotated <- coords %*% rz %*% rx
  rotated[, 1:2, drop = FALSE]
}

edge.matrix.from.adj <- function(adj.list) {
  edges <- list()
  idx <- 0L
  for (u in seq_along(adj.list)) {
    nbrs <- adj.list[[u]]
    nbrs <- nbrs[nbrs > u]
    if (!length(nbrs)) next
    for (v in nbrs) {
      idx <- idx + 1L
      edges[[idx]] <- c(u, v)
    }
  }
  do.call(rbind, edges)
}

plot.projected.graph <- function(coords,
                                 edges,
                                 main = "",
                                 vertex.col = "black",
                                 edge.col = "gray85",
                                 azimuth = 35,
                                 elevation = 22,
                                 pch = 16,
                                 cex = 0.55) {
  xy <- project.coords(coords, azimuth = azimuth, elevation = elevation)
  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- 0.08 * diff(xlim)
  ypad <- 0.08 * diff(ylim)
  if (!is.finite(xpad) || xpad == 0) xpad <- 0.2
  if (!is.finite(ypad) || ypad == 0) ypad <- 0.2

  plot(
    xy[, 1], xy[, 2],
    type = "n",
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    xlim = xlim + c(-xpad, xpad),
    ylim = ylim + c(-ypad, ypad),
    main = main
  )
  apply(edges, 1, function(e) {
    graphics::segments(
      xy[e[1], 1], xy[e[1], 2],
      xy[e[2], 1], xy[e[2], 2],
      col = edge.col
    )
  })
  points(xy[, 1], xy[, 2], pch = pch, cex = cex, col = vertex.col)
}

hmp.edges <- edge.matrix.from.adj(hmp.u01.gc.coarse$adj_list)
hmp.cst <- hmp.u01.gc.coarse$vertex_data$cst
cst.levels <- sort(unique(hmp.cst))
cst.cols <- setNames(grDevices::hcl.colors(length(cst.levels), "Dark 3"), cst.levels)
hmp.cst.col <- cst.cols[hmp.cst]
```

## Step 1: shortlist a few presets

The bundled preset comparison compared three candidates:

- `default`
- `tree`
- `torus`

The metrics below were computed with `grip.compare.layouts()` using the bundled
graph and CST labels. Lower `sampled.stress.mean` and lower
`score.composite` are better. Higher `cluster.separation.mean` is better.

```{r}
preset.summary <- hmp.results$preset_summary
preset.keep <- c(
  "candidate",
  "sampled.stress.mean",
  "edge.length.cv.mean",
  "sampled.nonedge.sep.ratio.mean",
  "cluster.separation.mean",
  "stability.procrustes.mean",
  "score.composite"
)
knitr::kable(preset.summary[, preset.keep])
```

```{r fig.width=11, fig.height=4.4}
op <- par(mfrow = c(1, 3), mar = c(1.2, 1.2, 3, 1.2))
on.exit(par(op), add = TRUE)

plot.projected.graph(
  hmp.results$layouts$preset$default,
  hmp.edges,
  main = "default",
  vertex.col = hmp.cst.col
)
plot.projected.graph(
  hmp.results$layouts$preset$tree,
  hmp.edges,
  main = "tree",
  vertex.col = hmp.cst.col
)
plot.projected.graph(
  hmp.results$layouts$preset$torus,
  hmp.edges,
  main = "torus",
  vertex.col = hmp.cst.col
)
```

On this graph, `tree` and `torus` both improve on `default`. That makes sense:
the graph has a compact core with several arm-like extensions, so a purely
generic layout is not the strongest prior.

## Step 2: search a small neighborhood around the strongest region

The next step is not to search the whole parameter space blindly. Instead, use a
small local grid around the strongest preset family. For this graph the search
was centered on 3D barycenter layouts with stronger refinement and moderate
repulsion. The bundled search varies only:

- `final.rounds` in `{224, 288}`
- `repulsion.factor` in `{0.75, 1.25}`

```{r}
local.summary <- hmp.results$local_search_summary
local.keep <- c(
  "candidate",
  "rounds",
  "final.rounds",
  "num.init",
  "r",
  "s",
  "repulsion.factor",
  "sampled.stress.mean",
  "cluster.separation.mean",
  "score.composite"
)
knitr::kable(head(local.summary[, local.keep], 8))
```

```{r fig.width=9, fig.height=4.5}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.2, 1))
on.exit(par(op), add = TRUE)

plot(
  local.summary$repulsion.factor,
  local.summary$sampled.stress.mean,
  pch = 19,
  col = grDevices::hcl.colors(nrow(local.summary), "Blues 3"),
  xlab = "repulsion.factor",
  ylab = "sampled.stress.mean",
  main = "search candidates: stress"
)
text(
  local.summary$repulsion.factor,
  local.summary$sampled.stress.mean,
  labels = local.summary$candidate,
  pos = 4,
  cex = 0.7
)

plot(
  local.summary$repulsion.factor,
  local.summary$cluster.separation.mean,
  pch = 19,
  col = grDevices::hcl.colors(nrow(local.summary), "Greens 3"),
  xlab = "repulsion.factor",
  ylab = "cluster.separation.mean",
  main = "search candidates: CST separation"
)
text(
  local.summary$repulsion.factor,
  local.summary$cluster.separation.mean,
  labels = local.summary$candidate,
  pos = 4,
  cex = 0.7
)
```

The important lesson is that stronger repulsion does not automatically improve a
real graph. On this example, the best local candidates sit in a moderate
repulsion range rather than at the maximum tested value.

```{r fig.width=11, fig.height=4.4}
top.local <- hmp.results$top_local_candidates
op <- par(mfrow = c(1, 3), mar = c(1.2, 1.2, 3, 1.2))
on.exit(par(op), add = TRUE)

for (nm in top.local) {
  plot.projected.graph(
    hmp.results$layouts$local[[nm]],
    hmp.edges,
    main = nm,
    vertex.col = hmp.cst.col
  )
}
```

## What the metrics mean in plain language

- `sampled.stress.mean`: lower is better. It measures how well Euclidean
  distances in the plot track graph distances.
- `edge.length.cv.mean`: lower is better. It tells you whether neighboring
  vertices are drawn with reasonably consistent edge lengths.
- `sampled.nonedge.sep.ratio.mean`: higher is better. It penalizes clutter and
  near-collisions.
- `cluster.separation.mean`: higher is better when you have meaningful external
  labels such as CST or subCST.
- `stability.procrustes.mean`: lower is better. It tells you how much layouts
  change across seeds.
- `score.composite`: lower is better. It is a weighted summary of the metrics
  above and is useful as a ranking aid, not as the only decision rule.

For a graph like this, the choice is not about a single metric. The practical
goal is to find a layout that is faithful enough globally, keeps clutter under
control, and separates the biological labels in a way that remains visually
interpretable.

## How to rerun or expand the search

The vignette uses precomputed HMP/U01 results to stay lightweight, but the
comparison workflow is fully reproducible. A direct rerun looks like this:

```{r eval=FALSE}
cmp <- grip.compare.layouts(
  adj_list = hmp.u01.gc.coarse$adj_list,
  weight_list = hmp.u01.gc.coarse$weight_list,
  n = length(hmp.u01.gc.coarse$adj_list),
  dim = 3,
  candidates = c("default", "tree", "torus"),
  clusters = hmp.u01.gc.coarse$vertex_data$cst,
  seeds = 1,
  sample.size.stress = 1200,
  sample.size.nonedge = 3000
)
cmp$summary
```

And a small local search around the stronger region looks like this:

```{r eval=FALSE}
search.cmp <- grip.compare.layouts(
  adj_list = hmp.u01.gc.coarse$adj_list,
  weight_list = hmp.u01.gc.coarse$weight_list,
  n = length(hmp.u01.gc.coarse$adj_list),
  dim = 3,
  search = list(
    candidate.prefix = "hmp.local",
    placement = "barycenter",
    rounds = 192L,
    final_rounds = c(224L, 288L),
    num_init = 30L,
    num_nbrs = 10L,
    r = 0.10,
    s = 2.0,
    repulsion_factor = c(0.75, 1.25)
  ),
  clusters = hmp.u01.gc.coarse$vertex_data$cst,
  seeds = 1,
  sample.size.stress = 1200,
  sample.size.nonedge = 3000
)
search.cmp$summary
```

## Takeaway

This coarsened HMP/U01 graph is a good example of how `grip` should be used on
real data:

- start with a few plausible presets in 3D,
- use external labels when they are scientifically meaningful,
- search locally around the strongest region rather than sweeping blindly,
- and inspect both metrics and projected views before deciding on a final
  layout.

The bundled graph is intentionally realistic enough to show this workflow while
still being small enough to ship with the package.
