---
title: "Getting Started with grip"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with grip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.5,
  fig.height = 5.2
)
```

## What this vignette covers

`grip` now supports three closely related layout workflows:

- `grip.layout()` for the current quality-first multiscale GRIP engine,
- `grip.layout.legacy()` for the historical GRIP behavior,
- `grip.layout.weighted()` for geometry-aware layouts of weighted graphs.

This vignette is the shortest path through that structure. It shows:

- a first unweighted layout,
- the difference between the current and legacy entry points,
- a first weighted layout where edge lengths matter,
- and a compact trace example.

For deeper workflows, see the real-data, weighted-layouts, and interactive
vignettes.

```{r}
library(grip)
```

The helper functions below keep the figures compact.

```{r}
plot.layout.pair <- function(coords.left,
                             coords.right,
                             edges,
                             titles,
                             projection = NULL,
                             vertex.cols = c("black", "black"),
                             edge.col = "gray82") {
  op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3, 1.2), bg = "white")
  on.exit(par(op), add = TRUE)

  grip.plot(
    coords.left, edges,
    projection = projection,
    main = titles[[1]],
    vertex.col = vertex.cols[[1]],
    edge.col = edge.col
  )
  grip.plot(
    coords.right, edges,
    projection = projection,
    main = titles[[2]],
    vertex.col = vertex.cols[[2]],
    edge.col = edge.col
  )
}

plot.layout.triptych <- function(coords.list,
                                 edges,
                                 titles,
                                 projection = NULL,
                                 vertex.cols = rep("black", length(coords.list)),
                                 edge.col = "gray82") {
  op <- par(
    mfrow = c(1, length(coords.list)),
    mar = c(1.2, 1.2, 3, 1.2),
    bg = "white"
  )
  on.exit(par(op), add = TRUE)

  for (i in seq_along(coords.list)) {
    grip.plot(
      coords.list[[i]], edges,
      projection = projection,
      main = titles[[i]],
      vertex.col = vertex.cols[[i]],
      edge.col = edge.col
    )
  }
}

plot.trace.frame <- function(coords, edges, main = "") {
  active <- stats::complete.cases(coords[, 1:2, drop = FALSE])
  xy <- coords[active, 1:2, drop = FALSE]

  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- 0.08 * diff(xlim)
  ypad <- 0.08 * diff(ylim)
  if (!is.finite(xpad) || xpad == 0) xpad <- 0.2
  if (!is.finite(ypad) || ypad == 0) ypad <- 0.2

  plot(
    xy[, 1], xy[, 2],
    type = "n",
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    xlim = xlim + c(-xpad, xpad),
    ylim = ylim + c(-ypad, ypad),
    main = main
  )

  active.edges <- edges[active[edges[, 1]] & active[edges[, 2]], , drop = FALSE]
  if (nrow(active.edges) > 0) {
    apply(active.edges, 1, function(e) {
      graphics::segments(
        coords[e[1], 1], coords[e[1], 2],
        coords[e[2], 1], coords[e[2], 2],
        col = "gray82"
      )
    })
  }

  points(xy[, 1], xy[, 2], pch = 16, cex = 0.55, col = "black")
}
```

## A first unweighted layout

For an ordinary unweighted graph, `grip.layout()` is the default starting
point. Here is a small mesh in 2D.

```{r}
mesh.edges <- edges.mesh(5, 5)
mesh.coords <- grip.layout(
  mesh.edges,
  n = 25,
  dim = 2,
  preset = "mesh",
  seed = 1
)

mesh.score <- grip.score.layout(mesh.coords, edges = mesh.edges, n = 25)
knitr::kable(mesh.score[, c(
  "sampled.stress",
  "edge.length.cv",
  "sampled.nonedge.sep.ratio"
)], digits = 3)
```

```{r fig.width=5.2, fig.height=4.6}
grip.plot(
  mesh.coords,
  mesh.edges,
  main = "Current GRIP on a 5x5 mesh",
  pch = 16,
  cex = 0.65,
  edge.col = "gray82"
)
```

For small and medium unweighted graphs, that is often all you need:

- choose `dim = 2` or `dim = 3`,
- optionally start from a preset,
- and score or compare a few candidate settings if the graph is important.

## Current GRIP versus legacy GRIP

The package still keeps the historical entry point through
`grip.layout.legacy()`. That is useful for backward-compatible reproduction,
but the main package path is the current `grip.layout()` engine.

```{r}
mesh.current <- grip.layout(
  mesh.edges,
  n = 25,
  dim = 2,
  preset = "mesh",
  seed = 2
)

mesh.legacy <- grip.layout.legacy(
  mesh.edges,
  n = 25,
  dim = 2,
  seed = 2
)
```

```{r fig.width=9.4, fig.height=4.2}
plot.layout.pair(
  mesh.legacy,
  mesh.current,
  edges = mesh.edges,
  titles = c("legacy", "current"),
  vertex.cols = c("black", "#1F3B73")
)
```

The practical distinction is:

- use `grip.layout.legacy()` when you need the historical local-force behavior,
- use `grip.layout()` for the current package default,
- use `grip.layout.trace()` when you want to inspect how a layout forms.

## When edge lengths matter: weighted GRIP

Some graphs carry real geometry in their edge lengths. In those cases, the
combinatorial graph alone is not the whole story. The weighted sister API keeps
the same multiscale spirit while making the hierarchy and refinement steps
length-aware.

The example below starts from a rectangular mesh whose edge lengths are induced
by a curved 3D saddle surface.

```{r}
weighted.mesh <- mesh.surface.graph(
  5, 5,
  surface = "saddle",
  amplitude = 0.9
)

coords.combinatorial <- grip.layout(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  dim = 3,
  preset = "mesh",
  seed = 3
)

coords.weighted <- grip.layout.weighted(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  seed = 3
)

gkk.prepared <- grip.prepare.geodesic.kk(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights
)

gkk.summary <- do.call(
  rbind,
  list(
    cbind(
      method = "Combinatorial GRIP",
      grip.score.geodesic.kk(
        coords.combinatorial,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rmse",
        "gkk.mean.abs.path.error",
        "gkk.mean.rel.path.error"
      )]
    ),
    cbind(
      method = "Weighted GRIP",
      grip.score.geodesic.kk(
        coords.weighted,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rmse",
        "gkk.mean.abs.path.error",
        "gkk.mean.rel.path.error"
      )]
    )
  )
)

knitr::kable(gkk.summary, digits = 3)
```

```{r fig.width=12.6, fig.height=4.2}
plot.layout.triptych(
  list(
    weighted.mesh$coords_surface,
    coords.combinatorial,
    coords.weighted
  ),
  edges = weighted.mesh$edges,
  titles = c("Target geometry", "Combinatorial GRIP", "Weighted GRIP"),
  projection = "ortho",
  vertex.cols = c("#555555", "black", "#1F3B73")
)
```

This is the basic rule of thumb:

- if the graph is fundamentally unweighted, start with `grip.layout()`,
- if the edge lengths encode geometry you care about, start with
  `grip.layout.weighted()`,
- and keep 3D in mind for weighted geometric families, because many of them are
  intrinsically awkward in 2D.

## Tracing a layout

`grip.layout.trace()` records snapshots of the current multiscale solve. That
is useful for debugging, teaching, and understanding why a parameter setting is
working or failing.

```{r}
mesh.trace <- grip.layout.trace(
  mesh.edges,
  n = 25,
  dim = 2,
  preset = "mesh",
  rounds = 12,
  final_rounds = 16,
  trace = "level",
  diagnostics = "light",
  seed = 4
)

knitr::kable(mesh.trace$meta)
knitr::kable(mesh.trace$diagnostics[, c(
  "frame",
  "edge.length.cv",
  "sampled.nonedge.sep.ratio"
)], digits = 3)
```

```{r fig.width=10.5, fig.height=3.8}
sel <- c(1L, max(2L, floor(length(mesh.trace$frames) / 2)), length(mesh.trace$frames))

op <- par(mfrow = c(1, 3), mar = c(1.2, 1.2, 3, 1.2), bg = "white")
on.exit(par(op), add = TRUE)

for (idx in sel) {
  plot.trace.frame(
    mesh.trace$frames[[idx]],
    mesh.edges,
    main = paste(
      mesh.trace$meta$phase[[idx]],
      "(frame", idx, ")"
    )
  )
}
```

When you need the same idea for weighted graphs, use
`grip.layout.trace.weighted()`.

## Where to go next

- `Weighted Graph Layouts with grip` covers the weighted sister API in more
  detail.
- `Choosing Layouts for Real Data` focuses on layout search and scoring.
- `Interactive Exploration with gripui` covers the package's Shiny tools.
- `Synthetic Graph Families and Geometries` explains the growing benchmark and
  geometry library.
