---
title: "GRIP Layout Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GRIP Layout Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5.2
)

if (requireNamespace("rgl", quietly = TRUE)) {
  options(rgl.useNULL = TRUE)
  rgl::setupKnitr(autoprint = TRUE)
}
```

## Why this vignette is different

This vignette is a synthetic gallery rather than a minimal API reference. The
goal is to show the graph families where the current package has been tuned and
validated during development:

- `preset = "mesh"` for rectangular lattice graphs,
- `preset = "carpet"` for Sierpinski-carpet-like graphs,
- `preset = "tree"` for symmetric force-directed tree layouts,
- `preset = "torus"` for 3D torus and cylinder layouts,
- explicit tuned parameters for a 3D Sierpinski tetrahedron example.

For real-data workflows and layout scoring, see the separate vignettes
`Choosing Layouts for Real Data` and `Selecting Layouts on the Coarsened
HMP/U01 Graph`.

```{r}
library(grip)
```

## Warm-up: even a cycle benefits from tuning

The core interface is still simple: pass an edge list and the number of
vertices to `grip.layout()`. But this is also a nice place to show why the
comparison tools matter. On an 18-cycle, the untuned default can be visibly far
from a regular polygon, while a tiny local search around a more suitable
initial placement almost solves the problem.

```{r}
edges <- edges.cycle(18)

cycle.cmp <- grip.compare.layouts(
  edges = edges,
  n = 18,
  dim = 2,
  candidates = list(
    default = list(),
    circle = list(placement = "circle"),
    circle_refined = list(
      placement = "circle",
      rounds = 64L,
      final_rounds = 128L,
      num_init = 6L,
      num_nbrs = 8L,
      r = 0.05,
      s = 4.5,
      repulsion_factor = 1.25
    ),
    bary_refined = list(
      placement = "barycenter",
      rounds = 64L,
      final_rounds = 128L,
      num_init = 6L,
      num_nbrs = 8L,
      r = 0.05,
      s = 4.5,
      repulsion_factor = 1.25
    )
  ),
  seeds = 1:3,
  return.layouts = TRUE
)

knitr::kable(cycle.cmp$summary[, c(
  "candidate",
  "sampled.stress.mean",
  "edge.length.cv.mean",
  "sampled.nonedge.sep.ratio.mean",
  "score.composite"
)])
```

The composite score ranks the candidates from best to worst. Lower stress and
lower edge-length CV are better; higher non-edge separation ratio is better.
Even on this small graph, the refined candidates noticeably outperform the
default.

```{r fig.width=10.5, fig.height=4.2}
cycle.best <- cycle.cmp$summary$candidate[[1L]]
cycle.2d.default <- cycle.cmp$layouts$default[[1L]]
cycle.2d.best <- cycle.cmp$layouts[[cycle.best]][[1L]]

op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3.2, 1.2))
on.exit(par(op), add = TRUE)

grip.plot(cycle.2d.default, edges, main = "Cycle default", pch = 16, cex = 0.7)
grip.plot(cycle.2d.best, edges, main = paste("Cycle tuned:", cycle.best), pch = 16, cex = 0.7)
```

## A small preset map

The table below summarizes the presets currently shipped with the package and
the graph families they were tuned for.

| Family | Good starting point | What it tends to help with |
|:--|:--|:--|
| Rectangular grid or lattice | `preset = "mesh"` | More regular spacing and cleaner boundaries |
| Fractal carpet-like graph | `preset = "carpet"` | Better separation of holes and recursive cells |
| Tree-like graph | `preset = "tree"` | More symmetric branching structure |
| 3D torus or cylinder-like graph | `preset = "torus"` | Cleaner tubular or toroidal geometry |
| 3D tetrahedral Sierpinski graph | Explicit tuned parameters | Better simplex-like separation in 3D |

## Plot helpers for 3D layouts

The 3D examples below are shown as `rglwidget` widgets when the optional `rgl`
package is available. That makes the vignette more faithful to the actual 3D
geometry: you can rotate the layout rather than relying on a single static
projection. The two helper functions defined here are local to this vignette;
for 2D plotting the package exports `grip.plot()`.

```{r}
make.rglwidget.graph <- function(coords,
                                 edges,
                                 theta = 35,
                                 phi = 22,
                                 zoom = 0.8,
                                 point.col = "#1F3B73",
                                 edge.col = "gray80",
                                 point.size = 5,
                                 width = 800,
                                 height = 520) {
  if (!requireNamespace("rgl", quietly = TRUE)) {
    return("Install the optional 'rgl' package to see the interactive 3D widget here.")
  }

  rgl::open3d(useNULL = TRUE)
  rgl::bg3d(color = "white")

  apply(edges, 1, function(e) {
    rgl::segments3d(
      coords[e, 1],
      coords[e, 2],
      coords[e, 3],
      color = edge.col
    )
  })

  rgl::points3d(
    coords[, 1],
    coords[, 2],
    coords[, 3],
    color = point.col,
    size = point.size
  )

  rgl::view3d(theta = theta, phi = phi, zoom = zoom)
  widget <- rgl::rglwidget(width = width, height = height)
  rgl::close3d()
  widget
}

plot.trace.frame <- function(coords, edges, main = "", pch = 16, cex = 0.55) {
  active <- stats::complete.cases(coords[, 1:2, drop = FALSE])
  xy <- coords[active, 1:2, drop = FALSE]

  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- 0.08 * diff(xlim)
  ypad <- 0.08 * diff(ylim)
  if (!is.finite(xpad) || xpad == 0) xpad <- 0.2
  if (!is.finite(ypad) || ypad == 0) ypad <- 0.2

  plot(
    xy[, 1], xy[, 2],
    type = "n",
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    xlim = xlim + c(-xpad, xpad),
    ylim = ylim + c(-ypad, ypad),
    main = main
  )

  active.edges <- edges[active[edges[, 1]] & active[edges[, 2]], , drop = FALSE]
  if (nrow(active.edges) > 0) {
    apply(active.edges, 1, function(e) {
      graphics::segments(
        coords[e[1], 1], coords[e[1], 2],
        coords[e[2], 1], coords[e[2], 2],
        col = "gray80"
      )
    })
  }

  points(xy[, 1], xy[, 2], pch = pch, cex = cex, col = "black")
}
```

## Tuned 2D families

Each subsection below generates a default layout and a preset layout for the
same graph, plotted side by side so the difference is easy to spot.

### Mesh

The mesh preset was tuned on regular lattice graphs. On a `10 x 10` mesh it
produces visibly more uniform spacing than the generic defaults.

```{r fig.width=10.5, fig.height=4.2}
edges <- edges.mesh(10, 10)

mesh.2d.default <- grip.layout(edges, n = 100, dim = 2, seed = 11)
mesh.2d.preset <- grip.layout(edges, n = 100, dim = 2, preset = "mesh", seed = 11)

op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3.2, 1.2))
on.exit(par(op), add = TRUE)

grip.plot(mesh.2d.default, edges, main = "Mesh default", pch = 16, cex = 0.45)
grip.plot(mesh.2d.preset, edges, main = "Mesh preset", pch = 16, cex = 0.45)
```

### Sierpinski carpet

The carpet preset was tuned on level 3 and 4 Sierpinski carpets. This family is
a good stress test because it combines a dense recursive interior with large
holes that should remain visually open.

```{r fig.width=10.5, fig.height=4.2}
edges <- edges.sierpinski.carpet(3)
n <- max(edges)

carpet.2d.default <- grip.layout(edges, n = n, dim = 2, seed = 12)
carpet.2d.preset <- grip.layout(edges, n = n, dim = 2, preset = "carpet", seed = 12)

op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3.2, 1.2))
on.exit(par(op), add = TRUE)

grip.plot(carpet.2d.default, edges, main = "Carpet default", pch = 16, cex = 0.28)
grip.plot(carpet.2d.preset, edges, main = "Carpet preset", pch = 16, cex = 0.28)
```

### Binary tree

The tree preset was tuned to encourage more symmetric force-directed tree
layouts. It is not a dendrogram layout; it is still a GRIP force-directed
layout, just with a more tree-friendly parameter profile.

```{r fig.width=10.5, fig.height=4.2}
edges <- edges.kary.tree(2, 6)
n <- max(edges)

tree.2d.default <- grip.layout(edges, n = n, dim = 2, seed = 13)
tree.2d.preset <- grip.layout(edges, n = n, dim = 2, preset = "tree", seed = 13)

op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3.2, 1.2))
on.exit(par(op), add = TRUE)

grip.plot(tree.2d.default, edges, main = "Tree default", pch = 16, cex = 0.45)
grip.plot(tree.2d.preset, edges, main = "Tree preset", pch = 16, cex = 0.45)
```

## Tuned 3D families

The same default-versus-tuned comparison extends to three dimensions. Rotate the
interactive widgets to appreciate how the tuned parameters recover the expected
geometry.

### Cylinder

The torus preset was validated on torus graphs, but it also works well on
cylinders. This makes a good interactive example because the tubular geometry
is easy to inspect by rotation.

```{r}
edges <- edges.cylinder(8, 12)
n <- max(edges)

cylinder.3d.default <- grip.layout(edges, n = n, dim = 3, seed = 21)
cylinder.3d.preset <- grip.layout(edges, n = n, dim = 3, preset = "torus", seed = 21)
```

The default layout tends to flatten or twist the cylinder, while the torus
preset recovers the expected tubular shape.

Default cylinder layout:

```{r}
make.rglwidget.graph(cylinder.3d.default, edges)
```

Tuned cylinder layout:

```{r}
make.rglwidget.graph(cylinder.3d.preset, edges)
```

### Sierpinski tetrahedron

We do not currently expose a tetrahedron preset, but development tuning found a
strong 3D parameter region for this family. The explicit parameter set below is
the current best validated tetrahedron example in the package workflow.

```{r}
edges <- edges.sierpinski.tetrahedron(4)
n <- max(edges)

tetrahedron.3d.default <- grip.layout(edges, n = n, dim = 3, seed = 22)
tetrahedron.3d.tuned <- grip.layout(
  edges, n = n, dim = 3,
  placement = "barycenter",
  rounds = 128,
  final_rounds = 192,
  num_init = 12,
  num_nbrs = 16,
  r = 0.07,
  s = 9.0,
  repulsion_factor = 1.5,
  seed = 22
)
```

Compare the default, which often collapses the four tetrahedral faces into a
flatter shape, with the tuned version that opens up the simplex structure.

Default tetrahedron layout:

```{r}
make.rglwidget.graph(tetrahedron.3d.default, edges)
```

Tuned tetrahedron layout:

```{r}
make.rglwidget.graph(tetrahedron.3d.tuned, edges)
```

## Tracing a layout

`grip.layout.trace()` records how the multiscale GRIP layout evolves. The trace
is especially useful when you want to understand how new vertices enter the
layout and how the later refinement stages clean up the geometry.

```{r}
edges <- edges.mesh(6, 6)
mesh.2d.trace <- grip.layout.trace(
  edges,
  n = max(edges),
  dim = 2,
  preset = "mesh",
  seed = 31,
  trace = "level"
)

head(mesh.2d.trace$meta)
```

The four panels below show the mesh layout at four stages of the multiscale
process. The coarsest level places only a handful of representative vertices;
successive refinement stages interpolate the remaining vertices and improve
spacing until the full grid emerges.

```{r fig.width=10.5, fig.height=8}
frame.idx <- unique(c(
  1L,
  max(2L, floor(length(mesh.2d.trace$frames) / 3)),
  max(3L, floor(2 * length(mesh.2d.trace$frames) / 3)),
  length(mesh.2d.trace$frames)
))

frame.labels <- c(
  "Coarsest start",
  "Early refinement",
  "Late refinement",
  "Final layout"
)

op <- par(mfrow = c(2, 2), mar = c(1.2, 1.2, 3.0, 1.2))
on.exit(par(op), add = TRUE)

for (i in seq_along(frame.idx)) {
  plot.trace.frame(
    mesh.2d.trace$frames[[frame.idx[[i]]]],
    # In early trace frames, only active vertices are shown.
    edges,
    main = frame.labels[[i]],
    cex = 0.45
  )
}
```

## Where to go next

This vignette shows the synthetic families where the current package has the
strongest tuned examples. The scoring metrics used throughout --- sampled
stress, edge-length CV, and non-edge separation ratio --- are computed by
`grip.score.layout()` and aggregated by `grip.compare.layouts()`. For
real-data selection workflows:

- use `grip.compare.layouts()` and `grip.score.layout()`,
- compare defaults, presets, and small local searches in 3D,
- then inspect the real-data vignettes for worked examples.
