---
title: "Historical layout-engine comparison: grip 0.2.0"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Historical layout-engine comparison: grip 0.2.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  fig.width = 10,
  fig.height = 4.5
)
```

## What is compared

This appendix preserves the external-engine benchmark run with grip 0.2.0 on
23 August 2026. It is historical evidence, not a benchmark of the current MDS
backend. The [current selection guide](grip-real-data.html) explains how to
compare layouts; [Metric MDS](metric-mds-backends.html) compares the current
SGD and SMACOF implementations on their separately recorded examples.

## Shared scoring idea

`score.layout()` can score a realized layout from any source. That remains
one of the most useful parts of the package, because it lets you compare
candidate layouts even when they come from different packages.

```{r score-helper, eval=FALSE}
score_layout <- function(coords, edges, n,
                         clusters = NULL,
                         edge.crossings = "auto",
                         sample.size.stress = 2000L,
                         sample.size.nonedge = 5000L) {
  score.layout(
    coords = coords,
    edges = edges,
    n = n,
    clusters = clusters,
    sample.size.stress = sample.size.stress,
    sample.size.nonedge = sample.size.nonedge,
    stress.seed = 42L,
    nonedge.seed = 42L,
    edge.crossings = edge.crossings
  )
}
```

For weighted geometry-aware graphs, the package also provides retained-path
diagnostics that answer a different question from sampled chord stress:

- `prepare.geodesic.kk()`
- `score.geodesic.kk()`

Those GKK/LGKK-style tools are advanced public experimental utilities rather
than the default onboarding path, and they are used in the second half of this
article for the weighted comparison track.

```{r}
plot_layout <- function(coords,
                        edges,
                        title,
                        vertex.col = "#1F3B73",
                        edge.col = "gray80",
                        cex = 0.55) {
  if (ncol(coords) == 2L) {
    plot(
      coords[, 1], coords[, 2],
      asp = 1,
      pch = 16,
      cex = cex,
      col = vertex.col,
      xlab = "",
      ylab = "",
      axes = FALSE,
      main = title
    )
    if (!is.null(edges) && nrow(edges) > 0) {
      segments(
        coords[edges[, 1], 1], coords[edges[, 1], 2],
        coords[edges[, 2], 1], coords[edges[, 2], 2],
        col = edge.col
      )
    }
  } else {
    plot.layout(
      coords,
      edges,
      projection = "ortho",
      azimuth = 35,
      elevation = 22,
      vertex.col = vertex.col,
      edge.col = edge.col,
      main = title
    )
  }
}
```

## Track 1: external-package comparisons on unweighted graphs

For this track the comparison remains package-to-package:

- `igraph`
- `graphlayouts`
- `grip`

To keep the article lightweight, the cross-package results are precomputed and
bundled as an RDS file. The reproduction script is still:

- `inst/scripts/precompute-vs-alternatives.R`

```{r load-results}
res.path <- system.file(
  "extdata", "vs_alternatives", "benchmark_results.rds",
  package = "grip"
)
if (!nzchar(res.path)) {
  stop(
    "Precomputed benchmark results not found. ",
    "Run inst/scripts/precompute-vs-alternatives.R first."
  )
}
res <- readRDS(res.path)
```

### Provenance of the saved comparison

The external results were generated on **23 August 2026 with grip 0.2.0**.
They are frozen results, not reruns of the development version serving this page.
The [provenance record](https://github.com/pgajer/grip/blob/main/inst/extdata/vs_alternatives/BENCHMARK_PROVENANCE.md)
and [reproduction script](https://github.com/pgajer/grip/blob/main/inst/scripts/precompute-vs-alternatives.R)
record the settings and timing boundaries. The artifact's software versions are:

```{r external-versions}
knitr::kable(data.frame(
  Package = names(res$benchmark_metadata$package_versions),
  Version = unname(res$benchmark_metadata$package_versions)
), row.names = FALSE)
```

The karate graph has **34 vertices and 78 edges**; the 12 by 12 mesh has
**144 vertices and 264 edges**. Each table below scores a saved layout with
hop-distance stress, scoring seeds 42, 2,000 requested stress pairs, and up to
5,000 sampled nonedges. Exact crossings are enabled for these small graphs.
The stress score fits a common scale; it is not the raw stress objective of
`metric.mds()`. Preset/tuned rows are labeled separately from ordinary defaults.
Keep the results where another method scores better when interpreting the comparison.

### Benchmark 1: Zachary karate club

This is still a useful small-graph sanity check. The graph is small enough that
single-scale methods can compete directly.

```{r karate-table}
karate.display <- res$karate$scores[, c(
  "method",
  "sampled.stress",
  "edge.length.cv",
  "edge.crossings",
  "sampled.nonedge.sep.ratio",
  "cluster.separation"
)]

knitr::kable(
  karate.display,
  digits = 4,
  row.names = FALSE,
  caption = "Karate club: saved grip 0.2.0 benchmark; one seeded layout per method."
)
```

```{r karate-plot, fig.width=12, fig.height=7.5}
k <- res$karate
par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")
plot_layout(k$layouts$fr,            k$edges, "FR (igraph)", cex = 0.6)
plot_layout(k$layouts$kk,            k$edges, "KK (igraph)", cex = 0.6)
plot_layout(k$layouts$drl,           k$edges, "DrL (igraph)", cex = 0.6)
plot_layout(k$layouts$stress,        k$edges, "Stress (graphlayouts)", cex = 0.6)
plot_layout(k$layouts$grip.default,  k$edges, "grip default", cex = 0.6)
plot_layout(k$layouts$grip.tuned,    k$edges, "grip tuned", cex = 0.6)
```

On graphs this small, `grip` does not have a special structural advantage.
That is an honest result and a useful reminder: multiscale methods are not
automatically best on every graph.

### Benchmark 2: 12x12 mesh

Meshes are a good structured topological benchmark because a strong layout
should recover an orderly grid with uniform edge lengths and few or no
crossings.

```{r mesh-table}
mesh.display <- res$mesh$scores[, c(
  "method",
  "sampled.stress",
  "edge.length.cv",
  "edge.crossings",
  "sampled.nonedge.sep.ratio"
)]

knitr::kable(
  mesh.display,
  digits = 4,
  row.names = FALSE,
  caption = "12x12 mesh: saved grip 0.2.0 benchmark; one seeded layout per method."
)
```

```{r mesh-plot, fig.width=12, fig.height=7.5}
m <- res$mesh
par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")
plot_layout(m$layouts$fr,            m$edges, "FR (igraph)", cex = 0.5)
plot_layout(m$layouts$kk,            m$edges, "KK (igraph)", cex = 0.5)
plot_layout(m$layouts$drl,           m$edges, "DrL (igraph)", cex = 0.5)
plot_layout(m$layouts$stress,        m$edges, "Stress (graphlayouts)", cex = 0.5)
plot_layout(m$layouts$grip.default,  m$edges, "grip default", cex = 0.5)
plot_layout(m$layouts$grip.mesh,     m$edges, "grip mesh preset", cex = 0.5)
```

This benchmark shows a different pattern from the karate graph: on a canonical
structured family like a mesh, the saved grip 0.2.0 default result
shows a clean uncrossed layout. The saved `mesh` preset result has slightly
lower sampled stress. These are particular seeded results, not an estimate of
variation across layout seeds. At the same time, stress and KK remain very strong on a regular
lattice.

### Larger graph: saved HMP timing and quality

The same artifact includes an unweighted symmetric-neighbor graph of HMP
microbiome samples: **4,391 vertices and 9,067 edges**. PCA-space edge lengths
were not supplied to any layout engine. The recorded machine was an Apple
M4 Max with 64 GiB memory, macOS/Darwin 25.6.0, and R 4.6.1.

```{r hmp-saved-results}
knitr::kable(res$hmp$timing, digits = 3, row.names = FALSE,
  caption = "HMP graph: elapsed seconds for saved grip 0.2.0-era runs.")
knitr::kable(res$hmp$scores[, c("method", "sampled.stress", "edge.length.cv")],
  digits = 4, row.names = FALSE,
  caption = "HMP graph: hop-distance quality of the first timed layout per method.")
```

The **five timing repeats reset seed 1 before each layout call**. Their medians
and interquartile ranges measure timing variation for that seeded setup, not
variation across independent stochastic layouts. Timing includes only the layout
call; graph construction, garbage collection, and scoring are excluded. Scores
use the first repeat, seeds 42, 2,000 stress pairs, and up to 5,000 nonedges;
exact crossings are disabled for this graph. These hardware-specific times and
single-seed quality results do not establish general speed or accuracy rankings.

### What the unweighted track tells us

The saved external-package benchmark has a limited scope:

- it compares engines on ordinary unweighted graphs,
- it uses a shared scoring function on the same graph for each comparison,
- but it does not answer the geometry-aware weighted-layout question.

That second question needs a different benchmark track.

## Reproducing the external benchmark bundle

To run the comparison recipe again with your installed package version, write
to a new output file. Reproducing the archived numerical results additionally
requires the software versions recorded above:

```{r reproduce, eval=FALSE}
Sys.setenv(GRIP_VS_ALTERNATIVES_OUTPUT = "benchmark_results.rds")
source(system.file(
  "scripts", "precompute-vs-alternatives.R",
  package = "grip"
))
```

## Session info for the bundled external results

```{r session-info-precomputed}
res$session_info
```
