---
title: "Getting Started with grip"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with grip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE, comment='#>', fig.width=9, fig.height=5,
                     out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

Start with a graph whose structure you want to explore. This example uses
**dwt_307**, a 307-vertex structural graph from the
[SuiteSparse Matrix Collection](https://sparse.tamu.edu/HB/dwt_307), supplied
with grip. Its edges describe connectivity; the drawing has no known spatial
reference that the layout must recover.

## Load a graph and draw it

```{r first-layout}
data(zheng.graphs)
g <- zheng.graphs[['dwt_307']]
coords <- grip(edges=g$edges, n=g$n, dim=3, seed=1)
```

Each row of `g$edges` contains two vertex indices, numbered from 1 to `g$n`.
Each row of `coords` gives the position of the corresponding vertex.
For an unweighted graph, `grip()` uses shortest-path hop counts to guide its
multiscale layout. Here all edges have unit length.

```{r first-view, echo=FALSE}
documentation.view(list(GRIP=coords),g$edges,
  description='dwt_307: 307 vertices. GRIP with seed 1. Drag to rotate; scroll to zoom.')
```

Rotate the graph to inspect its structure. All vertices and edges are shown.
For a quick view in an interactive R session, use:

```{r plot-recipe, eval=FALSE}
plot.layout(coords, edges=g$edges)
```

## Assess the drawing

```{r score}
quality <- score.layout(coords, edges=g$edges, n=g$n,
                        edge.crossings='never')
knitr::kable(quality[,c('sampled.stress','edge.length.cv',
                       'sampled.nonedge.sep.ratio')],digits=3)
```

Sampled stress compares drawn distances with graph shortest paths after fitting
one common scale; smaller values mean closer agreement. Edge-length variation
shows how evenly the unit edges are drawn. Nonedge separation describes how
far apart unconnected vertices lie relative to edges. These measures describe
different aspects of a drawing; no single number captures readability.

## A metric MDS alternative

`metric.mds()` directly fits Euclidean distances to graph shortest-path
distances. Its default SGD backend adjusts pairs of vertices in shuffled order,
following [Zheng, Pawar and Goodman (2018)](https://doi.org/10.1109/TVCG.2018.2859997)
with the learning-rate schedule of
[Hangan, Kobourov and Miller (2026)](https://arxiv.org/abs/2605.00641).

```{r metric-mds-default}
mds <- suppressWarnings(metric.mds(
  edges=g$edges, n=g$n, dim=3, init='random',
  max.iter=80, seed=1, diagnostics=FALSE))
mds$metadata$termination
```

The 80-pass allowance is a short illustrative setting. Schedule completion
reports `iteration_limit`; it does not establish convergence. The figure
compares this fit with the GRIP layout above, allowing rotation, reflection,
and translation, and matching the overall size for display. GRIP and MDS use
different coordinate scales; one uniform multiplier per layout makes their
shapes comparable. This does not change the fitted coordinates or scores.

```{r methods-view, echo=FALSE}
documentation.view(list('metric-MDS'=mds$coords,GRIP=coords),g$edges,common.size=TRUE,
  description='dwt_307: metric-MDS and GRIP. Aligned at a common overall size for display. Seed 1.')
```

Select a layout below the figure, or show both. For optimizer and weighting
choices, continue to [Metric MDS](metric-mds-backends.html). Larger graphs may
benefit from [sparse metric MDS](sparse-metric-mds.html).

## Choose the next step

| Your next question | Vignette |
|---|---|
| What functions are available? | [Finding your way around grip](function-guide.html) |
| What other graphs can I draw? | [Graph gallery](synthetic-graph-families.html) |
| Which settings or seeds should I use? | [Choosing and evaluating a layout](grip-real-data.html) |
| My edges represent distances | [Layouts with meaningful edge lengths](weighted-grip-intro.html) |
| How did the drawing develop? | [Tracing and diagnosing layouts](grip-trace-and-diagnostics.html) |
| How do I explore saved results? | [Interactive exploration with gripui](gripui-explorer.html) |

Edge values that describe connection strength are not automatically suitable
as distances. In particular, the bundled SuiteSparse examples use unit traversal
lengths and retain their original matrix values separately. The graph gallery
provides source attribution and conversion details.
