---
title: "GRIP: Graph dRawing with Intelligent Placement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GRIP: Graph dRawing with Intelligent Placement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Visualizing large graphs is a fundamental problem in network analysis,
bioinformatics, social-network research, and many other fields. *Force-directed*
methods are among the most popular approaches: they model the graph as a
physical system where edges act as springs pulling connected vertices together,
while all vertex pairs exert repulsive forces to prevent overlap. The layout is
obtained by minimizing the total energy of this system. Classical algorithms
such as Kamada--Kawai (1989) and Fruchterman--Reingold (1991) produce
aesthetically pleasing drawings, but their running time grows quickly with
graph size---Kamada--Kawai, for instance, is $O(n^3)$ per iteration---making
them impractical for graphs with more than a few thousand vertices.

GRIP (Graph dRawing with Intelligent Placement) addresses this scalability
bottleneck. The algorithm was introduced in two companion papers:

1. Gajer, P., Goodrich, M.T., and Kobourov, S.G. (2004). A multi-dimensional
   approach to force-directed layouts of large graphs. *Computational Geometry:
   Theory and Applications*, 29(1):3--18.
   (Preliminary version in *Proc. 8th Symposium on Graph Drawing*, 2000.)

2. Gajer, P. and Kobourov, S.G. (2002). GRIP: Graph dRawing with Intelligent
   Placement. *Journal of Graph Algorithms and Applications*, 6(3):203--224.

The key observation behind GRIP is that force-directed refinement converges much
faster when vertices start close to their final positions. GRIP therefore
combines three ideas into a single multi-scale pipeline:

1. **Maximal Independent Set Filtration (MISF).** The vertex set is decomposed
   into a nested sequence $V = V_0 \supset V_1 \supset \cdots \supset V_k$,
   where each $V_i$ is a maximal independent set of the subgraph induced by
   $V_{i-1}$ with the additional constraint that the graph distance between any
   two vertices in $V_i$ is at least $2^{i-1}+1$. The coarsest level $V_k$
   contains only a small number of vertices that are spread far apart in the
   graph.

2. **Intelligent placement.** At each level, newly introduced vertices are
   placed at an initial position computed from the already-positioned neighbors
   (barycenter of nearby vertices, or along a circle in 2D). Because these
   neighbors are already well separated, the initial position is typically close
   to optimal.

3. **Fast force-directed refinement.** Each vertex is refined using only a
   small, local neighborhood of graph-distance neighbors (controlled by the
   `num_nbrs` parameter). This makes the per-vertex cost constant on average,
   yielding near-linear overall running time. Two engine variants are provided:
   `mish_v5` uses Kamada--Kawai-style spring forces, while `mish_v6` adds
   Fruchterman--Reingold-style repulsive forces for improved spacing.

The result is an algorithm that draws graphs with tens of thousands of
vertices in seconds, using negligible memory, while producing layouts that are
competitive in quality with the much slower classical methods.

## Quick start

```{r}
library(grip)
```

## Parameter overview

| Parameter      | Meaning |
|:---------------|:--------|
| `dim`          | Layout dimension (2 or 3). |
| `engine`       | Force engine: `"mish_v5"` (spring-only) or `"mish_v6"` (spring + repulsive). |
| `placement`    | Initial placement strategy: `"barycenter"` (default) or `"circle"` (2D only). |
| `rounds`       | Refinement rounds at coarse levels. |
| `final_rounds` | Refinement rounds at the finest level. |
| `num_init`     | Number of initial vertices at the coarsest filtration level. |
| `num_nbrs`     | Number of graph-distance neighbors used during refinement. |
| `r`, `s`       | Local temperature schedule parameters (adaptive cooling). |
| `seed`         | RNG seed for reproducibility. |

The examples below use `rounds = 25` and `final_rounds = 25` for good layout
quality. The remaining parameters (`num_init`, `num_nbrs`) are kept small
since the example graphs are small; for larger graphs the defaults
(`num_init = 36`, `num_nbrs = 10`) generally work well.

## Helper functions

We define small helpers to generate edge lists for several standard graph
families.

```{r}
edges_path <- function(n) {
  if (n < 2) return(matrix(integer(), ncol = 2))
  cbind(seq_len(n - 1L), seq_len(n - 1L) + 1L)
}

edges_cycle <- function(n) {
  if (n < 2) return(matrix(integer(), ncol = 2))
  rbind(edges_path(n), c(n, 1L))
}

edges_mesh <- function(h, w = h) {
  stopifnot(h >= 1, w >= 1)
  idx <- function(i, j) (i - 1L) * w + j
  edges <- list()
  for (i in seq_len(h)) {
    for (j in seq_len(w)) {
      v <- idx(i, j)
      if (i < h) edges[[length(edges) + 1L]] <- c(v, idx(i + 1L, j))
      if (j < w) edges[[length(edges) + 1L]] <- c(v, idx(i, j + 1L))
    }
  }
  do.call(rbind, edges)
}

edges_cylinder <- function(h, w = h) {
  stopifnot(h >= 1, w >= 1)
  idx <- function(i, j) (i - 1L) * w + j
  edges <- list()
  for (i in seq_len(h)) {
    for (j in seq_len(w)) {
      v <- idx(i, j)
      if (i < h) edges[[length(edges) + 1L]] <- c(v, idx(i + 1L, j))
      edges[[length(edges) + 1L]] <- c(v, idx(i, (j %% w) + 1L))
    }
  }
  do.call(rbind, edges)
}

edges_torus <- function(h, w = h) {
  stopifnot(h >= 1, w >= 1)
  idx <- function(i, j) (i - 1L) * w + j
  edges <- list()
  for (i in seq_len(h)) {
    for (j in seq_len(w)) {
      v <- idx(i, j)
      edges[[length(edges) + 1L]] <- c(v, idx((i %% h) + 1L, j))
      edges[[length(edges) + 1L]] <- c(v, idx(i, (j %% w) + 1L))
    }
  }
  do.call(rbind, edges)
}

edges_sierpinski <- function(level = 2) {
  stopifnot(level >= 0)

  merge_nodes <- function(edges, from, to) {
    edges[edges == from] <- to
    edges
  }

  build <- function(k) {
    if (k == 0) {
      edges <- rbind(c(1L, 2L), c(2L, 3L), c(3L, 1L))
      return(list(edges = edges, corners = c(1L, 2L, 3L), n = 3L))
    }
    left <- build(k - 1)
    right <- build(k - 1)
    top <- build(k - 1)

    off1 <- left$n
    off2 <- left$n + right$n

    edges <- rbind(left$edges,
                   right$edges + off1,
                   top$edges + off2)

    L <- left$corners
    R <- right$corners + off1
    T <- top$corners + off2

    edges <- merge_nodes(edges, R[1], L[2])
    edges <- merge_nodes(edges, T[1], L[3])
    edges <- merge_nodes(edges, T[2], R[3])

    ids <- sort(unique(c(edges)))
    map <- seq_along(ids)
    names(map) <- ids
    edges <- cbind(map[as.character(edges[, 1])],
                   map[as.character(edges[, 2])])
    edges <- t(apply(edges, 1, sort))
    edges <- unique(edges)

    corners <- c(map[as.character(L[1])],
                 map[as.character(R[2])],
                 map[as.character(T[3])])

    list(edges = edges, corners = corners, n = length(ids))
  }

  build(level)$edges
}
```

## 2D layout examples

The examples below lay out several common graph families in two dimensions.
The `"circle"` placement strategy, which arranges initial vertices along a
circle before refinement, is available only in 2D; the default `"barycenter"`
strategy works in both 2D and 3D.

### Path graph

A simple path on 12 vertices. The algorithm should recover a (roughly) linear
arrangement.

```{r}
edges <- edges_path(12)
coords <- grip.layout(edges, n = 12, dim = 2,
                      engine = "mish_v5",
                      placement = "barycenter",
                      rounds = 25, final_rounds = 25,
                      num_init = 5, num_nbrs = 6,
                      seed = 1)
grip.plot(coords, edges, main = "Path (2D)", pch = 16, cex = 0.6)
```

### Cycle graph

A cycle on 16 vertices. A good layout should approximate a regular polygon.

```{r}
edges <- edges_cycle(16)
coords <- grip.layout(edges, n = 16, dim = 2,
                      engine = "mish_v5",
                      placement = "barycenter",
                      rounds = 25, final_rounds = 25,
                      num_init = 5, num_nbrs = 6,
                      seed = 2)
grip.plot(coords, edges, main = "Cycle (2D)", pch = 16, cex = 0.6)
```

### Mesh (grid) graph

A 5 x 5 grid graph. This uses the `mish_v6` engine, which adds repulsive
forces and can produce more uniform spacing on regular lattices.

```{r}
edges <- edges_mesh(5, 5)
coords <- grip.layout(edges, n = 25, dim = 2,
                      engine = "mish_v6",
                      placement = "barycenter",
                      rounds = 25, final_rounds = 25,
                      num_init = 6, num_nbrs = 8,
                      seed = 3)
grip.plot(coords, edges, main = "Mesh (2D)", pch = 16, cex = 0.6)
```

### Sierpinski triangle

The level-2 Sierpinski triangle graph has a fractal structure that is a
useful stress test for layout algorithms. Here we use the `"circle"`
placement strategy.

```{r}
edges <- edges_sierpinski(2)
n <- max(edges)
coords <- grip.layout(edges, n = n, dim = 2,
                      engine = "mish_v5",
                      placement = "circle",
                      rounds = 25, final_rounds = 25,
                      num_init = 5, num_nbrs = 7,
                      seed = 4)
grip.plot(coords, edges, main = "Sierpinski (2D, circle placement)", pch = 16, cex = 0.6)
```

## 3D layout examples

Setting `dim = 3` produces a layout in three-dimensional space. The output is a
numeric matrix with three columns. GRIP was designed from the outset as a
multi-dimensional algorithm---the original papers showed that working in higher
dimensions and then projecting can sometimes yield smoother, more symmetric
drawings.

For interactive 3D viewing, the **rgl** package can be used (see the optional
plotting snippet at the end of this section).

### 3D mesh

```{r}
edges <- edges_mesh(4, 4)
coords <- grip.layout(edges, n = 16, dim = 3,
                      engine = "mish_v5",
                      placement = "barycenter",
                      rounds = 25, final_rounds = 25,
                      num_init = 5, num_nbrs = 7,
                      seed = 5)
head(coords)
```

## Adjacency list and weighted edges

In addition to the edge-list interface, `grip.layout()` accepts an adjacency
list with an optional parallel list of edge weights. This is convenient when
your graph is already stored in adjacency-list form. Edge weights are
interpreted as desired edge lengths: a larger weight means the algorithm will
try to place the two endpoints farther apart.

`weight_list` is optional; omit it to treat all edges as having unit length.

```{r}
adj_list <- list(c(2), c(1, 3), c(2, 4), c(3))
weight_list <- list(c(1.0), c(1.0, 2.0), c(2.0, 1.5), c(1.5))
coords <- grip.layout(adj_list = adj_list,
                      weight_list = weight_list,
                      n = 4,
                      dim = 2,
                      engine = "mish_v5",
                      placement = "barycenter",
                      rounds = 25, final_rounds = 25,
                      num_init = 3, num_nbrs = 3,
                      seed = 12)
grip.plot(coords, main = "Adjacency list input (2D)")
```

### 3D cylinder

A cylinder graph wraps the horizontal axis of a grid into a cycle, producing
a surface that should curve into a tube-like shape in 3D.

```{r}
edges <- edges_cylinder(4, 6)
coords <- grip.layout(edges, n = 24, dim = 3,
                      engine = "mish_v5",
                      placement = "barycenter",
                      rounds = 25, final_rounds = 25,
                      num_init = 6, num_nbrs = 8,
                      seed = 6)
head(coords)
```

### 3D torus

A torus wraps both axes of a grid into cycles. In 3D, the algorithm should
recover the characteristic doughnut shape.

```{r}
edges <- edges_torus(4, 4)
coords <- grip.layout(edges, n = 16, dim = 3,
                      engine = "mish_v6",
                      placement = "barycenter",
                      rounds = 25, final_rounds = 25,
                      num_init = 5, num_nbrs = 7,
                      seed = 7)
head(coords)
```

```{r, eval=FALSE}
# Optional 3D plotting (requires rgl)
if (requireNamespace("rgl", quietly = TRUE)) {
  rgl::plot3d(coords[, 1], coords[, 2], coords[, 3],
              size = 4, col = "steelblue")
}
```

## References

1. Gajer, P., Goodrich, M.T., and Kobourov, S.G. (2004). A multi-dimensional
   approach to force-directed layouts of large graphs. *Computational Geometry:
   Theory and Applications*, 29(1):3--18.
   [doi:10.1016/j.comgeo.2004.03.014](https://doi.org/10.1016/j.comgeo.2004.03.014).

2. Gajer, P. and Kobourov, S.G. (2002). GRIP: Graph dRawing with Intelligent
   Placement. *Journal of Graph Algorithms and Applications*, 6(3):203--224.
   [doi:10.7155/jgaa.00052](https://doi.org/10.7155/jgaa.00052).

3. Kamada, T. and Kawai, S. (1989). An algorithm for drawing general
   undirected graphs. *Information Processing Letters*, 31(1):7--15.

4. Fruchterman, T.M.J. and Reingold, E.M. (1991). Graph drawing by
   force-directed placement. *Software: Practice and Experience*,
   21(11):1129--1164.
