---
title: "Tracing and Diagnosing Layouts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracing and Diagnosing Layouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.8,
  fig.height = 5.2
)
```

## Why trace a layout?

Most graph-layout functions return only the final coordinate matrix. That is
enough for plotting, but it hides the most useful debugging information:

- how the multiscale hierarchy opened,
- whether the geometry stabilized gradually or abruptly,
- whether a parameter setting improved edge-length regularity or only moved the
  final picture around,
- and whether a weighted solve is actually getting closer to the intended
  metric geometry.

The trace APIs in `grip` make those questions visible:

- `grip.layout.trace()` for combinatorial layouts,
- `grip.layout.trace.weighted()` for weighted geometry-aware layouts.

```{r}
library(grip)
```

```{r}
plot.trace.frame <- function(coords, edges, main = "", vertex.col = "black") {
  active <- stats::complete.cases(coords[, 1:2, drop = FALSE])
  xy <- coords[active, 1:2, drop = FALSE]

  xlim <- range(xy[, 1])
  ylim <- range(xy[, 2])
  xpad <- 0.08 * diff(xlim)
  ypad <- 0.08 * diff(ylim)
  if (!is.finite(xpad) || xpad == 0) xpad <- 0.2
  if (!is.finite(ypad) || ypad == 0) ypad <- 0.2

  plot(
    xy[, 1], xy[, 2],
    type = "n",
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    xlim = xlim + c(-xpad, xpad),
    ylim = ylim + c(-ypad, ypad),
    main = main
  )

  active.edges <- edges[active[edges[, 1]] & active[edges[, 2]], , drop = FALSE]
  if (nrow(active.edges) > 0) {
    apply(active.edges, 1, function(e) {
      graphics::segments(
        coords[e[1], 1], coords[e[1], 2],
        coords[e[2], 1], coords[e[2], 2],
        col = "gray82"
      )
    })
  }

  points(xy[, 1], xy[, 2], pch = 16, cex = 0.55, col = vertex.col)
}

pick.trace.frames <- function(trace.obj) {
  n <- length(trace.obj$frames)
  unique(c(1L, max(2L, floor((n + 1L) / 2L)), n))
}
```

## A first combinatorial trace

The example below traces a 2D layout of a small mesh. Using
`trace = "level"` keeps the vignette readable by recording the coarse
initialization, the start of each new level, and the final layout.

```{r}
mesh.edges <- edges.mesh(5, 5)

mesh.trace <- grip.layout.trace(
  mesh.edges,
  n = 25,
  dim = 2,
  preset = "mesh",
  rounds = 12,
  final_rounds = 16,
  trace = "level",
  diagnostics = "light",
  seed = 1
)

knitr::kable(mesh.trace$meta)
knitr::kable(mesh.trace$diagnostics[, c(
  "frame",
  "edge.length.cv",
  "median.edge.length",
  "sampled.nonedge.sep.ratio"
)], digits = 3)
```

```{r fig.width=10.5, fig.height=3.8}
sel <- pick.trace.frames(mesh.trace)

op <- par(mfrow = c(1, length(sel)), mar = c(1.2, 1.2, 3, 1.2), bg = "white")
on.exit(par(op), add = TRUE)

for (idx in sel) {
  plot.trace.frame(
    mesh.trace$frames[[idx]],
    mesh.edges,
    main = paste(mesh.trace$meta$phase[[idx]], "(frame", idx, ")")
  )
}
```

The two most useful tables are:

- `meta`, which explains where each frame came from,
- `diagnostics`, which tracks simple quality signals frame by frame.

## Level traces versus round traces

When you need more detail, `trace = "round"` records intermediate refinement
rounds within each active level.

```{r}
mesh.trace.round <- grip.layout.trace(
  mesh.edges,
  n = 25,
  dim = 2,
  preset = "mesh",
  rounds = 6,
  final_rounds = 8,
  trace = "round",
  trace.every = 4,
  diagnostics = "light",
  seed = 1
)

data.frame(
  trace_mode = c("level", "round"),
  n.frames = c(length(mesh.trace$frames), length(mesh.trace.round$frames))
)
```

Use:

- `trace = "level"` for most reporting and teaching,
- `trace = "round"` when you are diagnosing a tuning issue.

## Weighted traces

Weighted traces answer a slightly different question. Instead of only asking
whether the layout becomes visually cleaner, we can also ask whether it moves
toward a target geometry.

Here the graph is a mesh whose edge lengths come from a curved 3D saddle
surface.

```{r}
weighted.mesh <- mesh.surface.graph(
  4, 4,
  surface = "saddle",
  amplitude = 0.8
)

weighted.trace <- grip.layout.trace.weighted(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  rounds = 8,
  final_rounds = 12,
  trace = "level",
  diagnostics = "light",
  target_coords = weighted.mesh$coords_surface,
  seed = 2
)

knitr::kable(weighted.trace$meta)
knitr::kable(weighted.trace$diagnostics[, c(
  "frame",
  "edge.length.cv",
  "sampled.nonedge.sep.ratio",
  "procrustes.rmse"
)], digits = 3)
```

```{r fig.width=10.8, fig.height=3.8}
sel.w <- pick.trace.frames(weighted.trace)

op <- par(mfrow = c(1, length(sel.w)), mar = c(1.2, 1.2, 3, 1.2), bg = "white")
on.exit(par(op), add = TRUE)

for (idx in sel.w) {
  frame.xy <- grip.project.3d(weighted.trace$frames[[idx]], azimuth = 35, elevation = 20)
  plot.trace.frame(
    frame.xy,
    weighted.mesh$edges,
    main = paste(weighted.trace$meta$phase[[idx]], "(frame", idx, ")"),
    vertex.col = "#1F3B73"
  )
}
```

The important extra column is `procrustes.rmse`. When a meaningful target
geometry exists, it gives a direct per-frame measure of whether the layout is
getting closer to that target up to rigid motion and scale.

## Trace objects are analysis objects

A trace object is more than a list of pictures. It can support:

- frame-by-frame diagnostics,
- manual inspection of where a solve destabilizes,
- teaching demonstrations of the multiscale hierarchy,
- and regression checks for future algorithm changes.

The most useful components are:

- `final`: the final coordinate matrix,
- `frames`: the stored coordinate frames,
- `meta`: frame provenance,
- `diagnostics`: per-frame metrics,
- `lgkk.polish`: optional post-layout LGKK result when enabled.

## Practical guidance

- Start with `trace = "level"` and `diagnostics = "light"`.
- Use `trace = "round"` only when you need finer-grained debugging.
- Supply `target_coords` when a meaningful geometric target exists.
- Prefer weighted traces for weighted graph families; they are much more
  interpretable than tracing the combinatorial solver on a graph with
  nontrivial edge lengths.

## Where to go next

- `Getting Started with grip` gives the short overview of the main APIs.
- `Weighted Graph Layouts with grip` focuses on weighted solving and scoring.
- `Choosing Layouts for Real Data` shows how trace fits into a real-data
  selection workflow.

