---
title: "Higher-dimensional metric MDS for 3D visualization"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Higher-dimensional metric MDS for 3D visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.rglWebGL { max-width: 100%; }
.grip-layout-selector { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE,comment='#>')
source(system.file('scripts','high-dimensional-mds.R',package='grip'),local=TRUE)
bundle <- readRDS(system.file('extdata','high-dimensional-mds','examples.rds',package='grip'))
```

Can fitting distances in more than three dimensions give a better **final 3D
drawing**? This vignette compares direct 3D metric MDS with a higher-dimensional
fit followed by principal component analysis (PCA), with or without a further
3D metric-MDS fit. Higher-dimensional fitting provides another initialization
strategy; it does not guarantee a better drawing or a faster calculation.

The small recipe below runs with grip's public interfaces. The interactive
examples and repeated-start tables use a compact subset of a saved experiment,
so rebuilding this vignette does not repeat the larger fits. Every figure is
an interactive **ivue** 3D view. Its **Layout** dropdown selects the method and
intermediate dimension while retaining the camera. Drag to rotate and scroll
to zoom. If ivue is unavailable, explanatory text replaces the viewer and the
numerical tables remain; no static projection is substituted.

This guide uses the development metric-MDS interface, introduced after grip
0.2.0. For backend and weighting choices, see [Metric MDS](metric-mds-backends.html).

## Three routes to a 3D drawing

1. **Direct 3D:** minimize metric-MDS stress with three coordinates per vertex.
2. **Higher-dimensional MDS + PCA:** fit, for example, ten coordinates per
   vertex; center them and retain the first three principal components.
3. **Higher-dimensional MDS + PCA + 3D refinement:** use those three PCA
   coordinates as the starting configuration for another metric-MDS fit.
   Every vertex can then move freely in 3D.

PCA chooses a three-dimensional linear subspace that retains coordinate
variance. It does not minimize error against the original distance targets.
Free 3D refinement optimizes those original targets again; the result need
not remain a linear projection of the higher-dimensional coordinates. This
is different from the optional `edge.kk()` stage in the
[synthetic graph gallery](synthetic-graph-families.html#optional-edge-length-comparison),
which fits directly connected vertices rather than all finite pairs.

An optimized higher-dimensional fit can have low stress while its PCA
projection has much higher stress. Evaluate the actual three-column result,
not just the higher-dimensional fit. Also compare a **classical-MDS start
followed by 3D metric MDS**: a useful initialization need not require a
higher-dimensional iterative fit.

## A runnable 10D example

We use the bundled 200-point paraboloid with Euclidean edge lengths on a
symmetric six-nearest-neighbor graph. Its targets are shortest-path distances.
Preparing the graph once lets every route use exactly the same target matrix.

```{r recipe-input}
library(grip)
bundle <- readRDS(system.file('extdata','high-dimensional-mds','examples.rds',package='grip'))
example <- bundle$cases[['paraboloid_17_6']]
prepared <- prepare.graph.geodesic.mds(
  edges=example$edges, n=example$n, edge.weights=example$weights)
D <- prepared$distance_matrix
stopifnot(isTRUE(all.equal(unname(D),unname(example$distances),tolerance=1e-10)))
```

The following is a short usage demonstration, with 20 SGD passes per fitted
stage and one start. It is not a convergence study or an equal-time benchmark.
The 10D route performs an additional fitting stage. To expose the initialization,
we supply random matrices explicitly; the direct start uses the first three
columns of the same draw. Internal scale adjustment differs between dimensions.

```{r recipe-fit}
set.seed(17)
start10 <- matrix(rnorm(example$n * 10),example$n,10)
fit <- function(dimension, initial) {
  # Record schedule termination below; suppress its expected warning here.
  withCallingHandlers(
    metric.mds(prepared=prepared, dim=dimension, init=initial,
      backend='sgd', approximation='full', pair.weights='uniform',
      n.init=1, max.iter=20, seed=17, diagnostics=FALSE,
      sgd.control=list(checkpoint.every=1L)),
    warning=function(w) {
      if (grepl('iteration_limit',conditionMessage(w),fixed=TRUE))
        invokeRestart('muffleWarning')
    })
}
direct <- fit(3,start10[,1:3])
high <- fit(10,start10)
pca3 <- prcomp(high$coords,center=TRUE,scale.=FALSE,rank.=3)$x[,1:3,drop=FALSE]
refined <- fit(3,pca3)
classical.start <- fit(3,'classical')
```

Use `scale.=FALSE`: rescaling the ten coordinate axes to unit variance would
change their Euclidean geometry before projection. `project.3d()` is a drawing
helper, not PCA; it uses the first three coordinate columns.

We score the returned coordinates without an additional fitted scale:

\[
S_{\mathrm{norm}}(Z)=
\frac{\sum_{i<j:\ d_{ij}<\infty}(\lVert z_i-z_j\rVert-d_{ij})^2}
     {\sum_{i<j:\ d_{ij}<\infty}d_{ij}^{2}}.
\]

Lower is better; zero is an exact realization. This dimensionless **normalized
stress** is the square of the **normalized error** reported in the
[backend comparison](metric-mds-backends.html). The metric-MDS solver itself
fits an overall scale when returning coordinates. Here we evaluate those
coordinates as returned, and the PCA coordinates as projected, without another
rescaling during evaluation.

```{r recipe-score}
normalized.stress <- function(z,D) {
  target <- as.vector(as.dist(D))
  keep <- is.finite(target)
  sum((as.vector(dist(z))[keep]-target[keep])^2)/sum(target[keep]^2)
}
layouts <- list('Direct 3D'=direct$coords, '10D before projection'=high$coords,
  '10D + PCA'=pca3, '10D + PCA + 3D refinement'=refined$coords,
  'Classical + 3D refinement'=classical.start$coords)
knitr::kable(data.frame(Route=names(layouts),
  Dimensions=vapply(layouts,ncol,0L),
  `Normalized stress`=vapply(layouts,normalized.stress,0,D=D),
  check.names=FALSE),digits=7,row.names=FALSE)
```

```{r recipe-termination, echo=FALSE}
knitr::kable(data.frame(Stage=c('Direct 3D','10D','3D refinement','Classical-start 3D'),
  Termination=vapply(list(direct,high,refined,classical.start),function(f) f$metadata$termination,'')),
  row.names=FALSE)
```

`iteration_limit` means that the prescribed schedule ended, not that an optimum
was reached. These fresh demonstration fits are separate from the saved results
below; their starts and implementation wrapper are not identical.

## Saved examples and comparison conditions

The saved study used full, uniformly weighted SGD-MDS on nine 200-point clouds
(plane, paraboloid, saddle; three data seeds), each with five distance targets,
plus six SuiteSparse graphs. It retained dimensions 4–20 and optimization
starts 17, 29, and 43. Here we show four target datasets and dimensions
4, 6, 8, 10, 15, and 20. The examples were selected **after examining the study**
to teach contrasting outcomes; they are not an unbiased success-rate sample.
Every displayed fit uses start 17, without choosing its appearance or score.
Tables retain all three starts; classical MDS is deterministic and appears once.

Synthetic fits allowed 20 passes per stage; SuiteSparse fits allowed four.
The PCA-refined route therefore includes both the higher-dimensional stage and
a second 3D stage. **Route seconds** include the preceding fitting stages,
their initialization and projection, but exclude shared target preparation and
subsequent evaluation/rendering. Equal pass counts across dimensions do not
mean equal time or arithmetic work.

To compare cost, the study also continued direct 3D SGD and retained its best
scored checkpoint within each refined route's charged time allowance. The
dimension tables compare against that stronger direct baseline. The allowance
includes some archival overhead absent from direct continuation, favoring the
direct method. Times describe these short recorded runs on one machine, not
isolated solver speed or performance on an idle machine. Neither route is
claimed to have converged.

The viewers show the saved initial direct stage, whose coordinates exist.
The time-matched direct checkpoints have saved scores but not saved coordinates;
their scores appear only in tables. A displayed direct drawing must therefore
not be mistaken for the time-matched comparator.

All views preserve coordinate scale. Display alignment permits translation,
rotation and reflection using corresponding vertex identities. It does not
deform the drawing or affect scores. Gray reference points, where present,
show the generating cloud; similarity to that cloud is a different question
from fitting graph-path distances. Blue denotes direct/classical fits, orange
PCA projections, and green refined fits. Use individual selections before
overlaying all layouts.

## One paraboloid, two distance targets

The 200 points have data seed 17 and satisfy \(z=x^2+y^2\), with parameters
sampled uniformly from the unit disk. This samples the parameter domain rather
than surface area uniformly.

### Ambient Euclidean distances

These are straight-line distances between the generating points. An exact 3D
realization exists. Additional fitting dimensions are not needed to represent
the target; this case is a useful check on projection and optimization error.
No edges are drawn in this view.

```{r ambient-view, echo=FALSE}
highdim.view(bundle$cases[['paraboloid_17_ambient']])
```

```{r ambient-scores, echo=FALSE}
highdim.table(highdim.routes(bundle$cases[['paraboloid_17_ambient']]))
```

### Shortest paths on the six-neighbor graph

The same points define a symmetric nearest-neighbor graph with Euclidean edge
lengths. The graph construction permits minimum-spanning-tree repair between
components. These finite graph-path distances differ from ambient distances
and from exact geodesics on the continuous paraboloid. A best-fitting 3D
configuration need not reproduce the generating surface.

```{r graph-view, echo=FALSE}
highdim.view(bundle$cases[['paraboloid_17_6']])
```

```{r graph-scores, echo=FALSE}
highdim.table(highdim.routes(bundle$cases[['paraboloid_17_6']]))
```

Compare the higher-dimensional stress with the PCA-only row, then with free
3D refinement. The first change measures the cost of projection; the second
shows how much a subsequent 3D fit recovers. The following table changes the
intermediate dimension and uses the time-matched direct baseline.

```{r graph-dimensions, echo=FALSE}
highdim.table(highdim.dimensions(bundle$cases[['paraboloid_17_6']]))
```

**Paired gain** is \(100(S_{\mathrm{direct}}-S_{\mathrm{refined}})/S_{\mathrm{direct}}\)
for each start, comparing the refined result with its time-matched direct
checkpoint. Positive values favor refinement. The mean is the mean of these
three percentages, and the range shows variation across starts. Near zero
stress, relative percentages can exaggerate negligible absolute differences;
inspect the stress values as well.

## SuiteSparse: an improvement and a counterexample

These are unweighted graph conversions of HB/illc1033 and HB/lock_700 from the
SuiteSparse Matrix Collection, not embeddings of their original numeric matrix
values. HB/illc1033 uses distinct row and column vertices with an edge for each
nonzero matrix entry. HB/lock_700 uses symmetrized off-diagonal support. Every
edge has unit traversal length. The package's `suitesparse.graph()` catalogue
retains the original source attribution and conversion metadata.

Disconnected components were fitted separately, and pairs in different
components are excluded from stress. All vertices are retained. HB/lock_700's
nine isolates are placed in a separate display row; their positions relative
to the main component have no distance-based interpretation. For legibility,
each dense graph view draws at most 2,500 edges, selected at evenly spaced
indices in the saved edge order. Fitting and scoring use the full graph.

### HB/illc1033

```{r illc-view, echo=FALSE}
highdim.view(bundle$cases[['HB__illc1033']])
```

```{r illc-scores, echo=FALSE}
highdim.table(highdim.routes(bundle$cases[['HB__illc1033']]))
highdim.table(highdim.dimensions(bundle$cases[['HB__illc1033']]))
```

### HB/lock_700

```{r lock-view, echo=FALSE}
highdim.view(bundle$cases[['HB__lock_700']])
```

```{r lock-scores, echo=FALSE}
highdim.table(highdim.routes(bundle$cases[['HB__lock_700']]))
highdim.table(highdim.dimensions(bundle$cases[['HB__lock_700']]))
```

```{r contrast, echo=FALSE, results='asis'}
a <- highdim.dimensions(bundle$cases[['HB__illc1033']])
b <- highdim.dimensions(bundle$cases[['HB__lock_700']])
cat(sprintf('\nAt intermediate dimension 10, mean paired time-matched gain is **%.2f%%** for HB/illc1033 and **%.2f%%** for HB/lock_700. These examples show why an additional fitting stage should be evaluated at its total cost, rather than compared only with one short direct stage.\n',
  a$`Mean paired gain (%)`[a$Dimension==10],b$`Mean paired gain (%)`[b$Dimension==10]))
```

## Choosing a workflow

Start with direct 3D MDS and a classical-initialized 3D control. Try several
random starts before attributing a difference to dimension. If higher-dimensional
fitting is promising, compare PCA-only and refined results at a few dimensions
and include the total route cost. Ten is an example, not a recommended optimum;
the original study's preselected summary dimension was eight.

Across all six study graphs, refinement at dimension eight lowered mean
time-matched stress by more than 1% on three graphs. The short schedules and
small number of starts limit that finding. Choosing the best dimension after
examining these tables is an exploratory choice, not independent validation.
Stress also does not determine visual usefulness: inspect neighborhoods,
overlap and recognizable structure in the interactive views.

Full MDS requires dense all-pairs distances and quadratic work/storage. Larger
problems may need [sparse MDS](sparse-metric-mds.html). Sparse fitting has a
different pair set and objective; its proxy score cannot be substituted for
the full uniform stress used here. Higher-dimensional sparse fitting is
supported, but its performance is not tested by these saved comparisons.

## Reproduction and saved evidence

`inst/extdata/high-dimensional-mds/examples.rds` contains the four graph
fixtures, two small synthetic distance matrices, the displayed 3D coordinates,
and selected three-start scores/timings. Its source-hash table records the
original study files. The saved study used a timing-instrumented SGD wrapper
with component handling; the short public-API recipe above does not reproduce
its exact starts, checkpoints or timings.

From a source checkout, the maintainer can extract this subset from the retained
study archive, verifying its hashes and independently rescoring every included
3D layout, with:

```{bash extract, eval=FALSE}
Rscript data-raw/build-high-dimensional-mds.R /path/to/high_dimensional_mds_to_3d
make high-dimensional-vignette
```

The extraction runs no optimization. Rebuilding the vignette runs only the
small public-API demonstration; the saved tables and 3D views require no
external archive, server, or gflowui installation. The bundle's provenance
document describes selection, timing and attribution. See also the
[function guide](function-guide.html#dimensions-result-objects-and-methods)
for dimension support and drawing conventions.
