---
title: "A biological network: HMP/U01"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{A biological network: HMP/U01}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE,comment='#>',fig.width=9,fig.height=5,out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

This graph summarizes similarities among HMP+U01 16S amplicon samples.
The supplied object records a representation based on taxa at at least 1%
relative abundance followed by principal component analysis, and a
nearest-neighbor construction with k=3. Coarsening reduces the original giant
component to 1,828 vertices and 4,656 weighted edges.

The biological labels are **community state types (CSTs)** and finer subtypes.
Use them to interpret groups in the drawing; separated colors do not by
themselves establish that the geometry is accurate.

## Inspect the graph

```{r graph}
data(hmp.u01.gc.coarse)
g <- hmp.u01.gc.coarse
knitr::kable(data.frame(Quantity=c('Original giant-component vertices',
  'Coarsened vertices','Undirected edges','Selected k'),
  Value=unlist(g$graph_info[c('original_vertices','coarse_vertices','edge_count','selected_k')],use.names=FALSE)))
```

The original giant component has 6,474 vertices. Seven other components were
discarded in the recorded preparation; no claim is made here about their
arrangement. This coarsened weighted HMP/U01 object differs from the unweighted
HMP graph in the historical external-engine benchmark.

## Interpret a saved layout

The saved search used grip 0.2.0, seed 1, and the default, tree, and torus presets.
It supplied lengths to the hop-based GRIP workflow: adjacent attractions use
lengths, while standard hierarchy searches use hops. **These are historical
hop-based fits, not new `metric='edge_length'` results.** The displayed
configuration is the default preset, selected for this walkthrough without
ranking it by appearance.

```{r saved, include=FALSE}
results <- readRDS(system.file('extdata','hmp_u01_gc_coarse','vignette_results.rds',package='grip'))
edges <- do.call(rbind,lapply(seq_along(g$adj_list),function(i) {
  j <- g$adj_list[[i]];j <- j[j>i];cbind(rep(i,length(j)),j)
}))
```

```{r hmp-view, echo=FALSE}
documentation.groups(results$layouts$preset$default,edges,g$vertex_data$cst,
  description='HMP/U01: saved default-preset layout, seed 1. Colors show community state types. Drag to rotate; scroll to zoom.')
```

Rotation helps distinguish actual branches from overlaps in projection.
The coordinates are shown in their saved units, without alignment to a putative
biological reference configuration.

## Compare the recorded candidates

```{r preset-scores}
knitr::kable(results$preset_summary[,c('candidate','sampled.stress.mean',
  'edge.length.cv.mean','cluster.separation.mean','score.composite')],digits=3)
```

Stress describes graph-distance agreement; cluster separation describes the
CST labels. Their priorities may differ. This saved search has one seed per
candidate, so it does not measure variation over random starts.

The recorded local search varied final refinement rounds and repulsion near
the selected settings. Its full summaries and individual outcomes are retained
in `results$local_search_summary` and `results$local_search_runs`.

```{r local-scores}
knitr::kable(results$local_search_summary[,c('candidate','final.rounds',
  'repulsion.factor','sampled.stress.mean','cluster.separation.mean')],digits=3)
```

## Fit the weighted graph today

Use the explicit edge-length mode for a new weighted-geodesic layout. This
recipe is separate from the saved historical results above.

```{r new-weighted, eval=FALSE}
coords <- grip(adj.list=g$adj_list,weight.list=g$weight_list,
  n=length(g$adj_list),dim=3,metric='edge_length',seed=1)
score.layout(coords,adj.list=g$adj_list,weight.list=g$weight_list,
  n=length(g$adj_list),clusters=g$vertex_data$cst,edge.crossings='never')
```

See [weighted layouts](weighted-grip-intro.html) for the metric semantics and
[Choosing and evaluating](grip-real-data.html) for repeated seeds and candidate
selection.

## Data and coarsening record

The object contains `adj_list`, parallel `weight_list`, `vertex_data`, and
`graph_info`. The vertex metadata includes aggregated CST and subtype labels.
The recorded coarsening rounds are:

```{r coarsening}
knitr::kable(g$graph_info$coarsening_rounds)
```

Saved results reside in `extdata/hmp_u01_gc_coarse/vignette_results.rds` in the
installed package. They include candidate settings, scores, fitted coordinates,
and the software session. To regenerate that historical workflow into a new
file, explicitly set the output path before sourcing its script:

```{r reproduce, eval=FALSE}
Sys.setenv(GRIP_HMP_RESULTS_OUTPUT='fresh-hmp-results.rds')
source(system.file('scripts','precompute-hmp-results.R',package='grip'))
```

That script reproduces its documented hop-based workflow; it does not rerun
the new weighted recipe above. Keep any new runs separate from the saved evidence.
