---
title: "Metric MDS: SGD, SMACOF, and stress weighting"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Metric MDS: SGD, SMACOF, and stress weighting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE,comment='#>',fig.width=9,fig.height=5,out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

`metric.mds()` fits Euclidean coordinates to graph shortest-path distances.
The default `backend='sgd'` moves pairs of vertices in shuffled order, following
[Zheng, Pawar and Goodman (2018)](https://doi.org/10.1109/TVCG.2018.2859997),
with the learning-rate schedule of
[Hangan, Kobourov and Miller (2026)](https://arxiv.org/abs/2605.00641).
`backend='smacof'` updates the full configuration using the optional
[smacof package](https://doi.org/10.18637/jss.v102.i10). Its majorization steps
are designed to reduce total stress; individual SGD pair adjustments need not.

This guide illustrates the differences using existing saved synthetic examples.
For a broader comparison of the methods, see Hangan, Kobourov and Miller (2026).
The [complete experiment](metric-mds-results.html) retains all 420 original
fits, the separate 60-fit weighting comparison, failed outcomes, numerical
results, and reproduction instructions.

## Fit a graph with either backend

```{r recipe, eval=FALSE}
data(zheng.graphs)
g <- zheng.graphs[['dwt_307']]
sgd <- metric.mds(edges=g$edges,n=g$n,dim=3,seed=1,
                  init='random',max.iter=100,diagnostics=FALSE)
smacof.fit <- metric.mds(edges=g$edges,n=g$n,dim=3,seed=1,
                  init='random',max.iter=100,diagnostics=FALSE,
                  backend='smacof',eps=1e-8)
```

This is a usage recipe, not a new benchmark on dwt_307. Both backends support
uniform and inverse-squared stress weights. The default full mode uses uniform
weights. An SGD pass and a SMACOF iteration are different amounts of work;
compare achieved error and elapsed time, not iteration counts alone. SGD's
schedule completion reports `iteration_limit`, which does not certify
convergence. Its schedule defaults remain provisional calibration choices.

## Two distance targets on the same saddle

The saved experiment uses 64- and 144-point clouds, five paired random starts,
and separate allowances of 10, 30, or 100 passes/iterations. Both methods receive
the same starting coordinates within each pair. Here we show the 144-point
saddle at seed 1 and allowance 100, with uniform stress weights. These displayed
runs were selected in advance, not chosen for their appearance.

**Ambient distances** are straight-line distances between generating points;
an exact 3D realization exists. **Local graph paths** sum edge lengths along
a finite grid with horizontal, vertical, and diagonal edges. They are not
exact geodesics on the continuous saddle, and their best 3D drawing need not
reproduce the generating surface.

```{r data, include=FALSE}
bundle <- readRDS(system.file('extdata','metric-mds-comparison','benchmark.rds',package='grip'))
```

```{r selected-saddles, results='asis', echo=FALSE}
for(id in c('saddle-144','saddle_graph-144')) {
  case <- bundle$cases[[id]]; fits <- comparison_representative(bundle,id)
  cat('\n\n### ',case$target,'\n\n',sep='')
  child <- c(sprintf('```{r selected-%s, echo=FALSE}',id),
    'documentation.full.view(case,fits)','```')
  cat(knitr::knit_child(text=child,envir=environment(),quiet=TRUE),sep='\n')
  tab <- do.call(rbind,lapply(fits,function(f) f$row[,c('backend','error','seconds','iterations','termination')]))
  cat(knitr::kable(tab,digits=4,row.names=FALSE,
    col.names=c('Backend','Normalized error','Fit seconds','Passes/iterations','Termination')),sep='\n')
}
```

Gray shows the generating reference; blue is SGD and orange is SMACOF.
The ambient case uses matched-vertex alignment. The local-graph case uses
surface-shape alignment: translation, rotation, and reflection match nearby
surface locations without requiring vertex labels to agree. Neither alignment
resizes or deforms the layouts. Scores use the original returned coordinates.

**Normalized error** is the square root of raw stress divided by the sum of
squared target distances; lower is better. The figures below summarize both
cloud sizes across five starts, with median errors and fitting times and
interquartile ranges. Each point is a separate fit restarted with its own
iteration allowance, not a checkpoint along one run. Fitting times exclude
shared graph preparation, alignment, and rendering.

```{r tradeoff, echo=FALSE}
comparison_tradeoff(bundle,c('saddle_graph-64','saddle_graph-144'))
```

## The effect of stress weighting

Uniform weighting minimizes squared errors in distance units. The paper's
inverse-squared weighting minimizes squared **relative** distance errors:

\[
\sum_{i<j} \frac{(\lVert z_i-z_j\rVert-d_{ij})^2}{d_{ij}^2}.
\]

An error of 0.2 on a target distance of 1 then counts as much as an error of
2 on a target of 10. Edge lengths determine the target distances;
`pair.weights` determines their influence on the objective.

```{r weighting-recipe, eval=FALSE}
fit <- metric.mds(edges=g$edges,n=g$n,dim=3,
  pair.weights='inverse_squared',backend='sgd',seed=1)
```

The following comparison uses the saved 64-vertex saddle graph, seed 1,
and allowance 100 for each backend and weighting. All four layouts use
surface-shape alignment without resizing. Select a method and weighting below
the figure, or show all four.

```{r weighting, echo=FALSE}
w <- readRDS(system.file('extdata','metric-mds-weighting','benchmark.rds',package='grip'))
case <- w$cases[['saddle_graph-64']]
fits <- comparison_weighting_fits(w,case$id)
documentation.full.view(case,fits,weighting=TRUE)
```

```{r weighting-table, echo=FALSE}
knitr::kable(do.call(rbind,lapply(fits,function(f)
  f$row[,intersect(c('backend','pair_weights','uniform_error','relative_error','seconds'),names(f$row))])),digits=4)
```

Compare errors within the same column: uniform and relative error express
different priorities. The [complete results](metric-mds-results.html#the-effect-of-stress-weighting)
include the five-start summaries, paraboloid and helix examples, and the
surface-registration checks. For larger graphs, continue to
[Sparse SGD](sparse-metric-mds.html).
