---
title: "Sparse SGD for larger graphs"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Sparse SGD for larger graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE,comment='#>',fig.width=9,fig.height=5,out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

Full metric MDS fits all vertex pairs. Sparse SGD keeps graph edges and
approximates longer-range influences using representative vertices called
**pivots**. It uses the sparse model of
[Ortmann, Klimenta and Brandes (2017)](https://doi.org/10.7155/jgaa.00440)
and the SGD updates of
[Zheng, Pawar and Goodman (2018)](https://doi.org/10.1109/TVCG.2018.2859997).

With `n` vertices, `m` edges, and `h` pivots, storage and work per pass are
roughly proportional to `m + n*h`, avoiding an all-pairs distance matrix.
More pivots cost more; improvement in every individual run is not guaranteed.

## Use sparse SGD

```{r sparse-recipe, eval=FALSE}
data(zheng.graphs)
g <- zheng.graphs[['dwt_307']]
fit <- metric.mds(edges=g$edges,n=g$n,edge.weights=g$edge_weights,
  dim=3,approximation='sparse',backend='sgd',seed=1,
  sparse.control=list(n.pivots=32),max.iter=100)
plot.layout(fit$coords,edges=g$edges)
```

Full MDS remains the default. Sparse mode currently supports SGD only, with
inverse-squared stress weights. Its defaults are random initialization, up to
200 pivots, and 30 epochs: provisional calibration choices. The 32 pivots and
100 epochs above are explicit illustrative choices. Supply connected graphs
with positive lengths; disconnected graphs are rejected without adding edges.

Sparse mode accepts supplied starting coordinates but does not accept classical
initialization, uniform weights, multiple starts, prepared objects, or dense
diagnostics. These restrictions avoid an unexpected all-pairs calculation.
Its `sparse_proxy_stress` describes the fitted sparse constraints, not full stress.
When a supplied edge is longer than an alternative route, sparse and full MDS
can use different targets for that edge. That distinction does not arise for
the unit and Euclidean chord lengths used in this experiment.

## Quality, time, and memory on existing examples

The saved experiment compares full SGD, sparse SGD at several pivot counts,
and weighted GRIP. Each setting uses seeds 1–3. SGD uses 100 epochs with the
same random initial coordinates for paired runs; GRIP uses its own initialization.
Full MDS retains its best checkpoint and fits scale; sparse SGD returns its
terminal iterate without fitting scale. These are comparisons of the supplied
methods, not equal work per epoch.

The structural graph below connects this comparison to the
[graph gallery](synthetic-graph-families.html). The larger saddle illustrates
approximation at 4,096 vertices. Full SGD was not attempted above 512 vertices
in this bounded experiment; that cutoff is not a measured capacity limit.

```{r sparse-data, include=FALSE}
source(system.file('scripts','sparse-mds-comparison.R',package='grip'),local=TRUE)
bundle <- readRDS(system.file('extdata','sparse-mds-comparison','benchmark.rds',package='grip'))
summary <- sparse_comparison_summary(bundle)
method_label <- function(x) ifelse(x=='full','Full SGD',ifelse(x=='grip','Weighted GRIP',
  paste('Sparse SGD:',sub('sparse','',x),'pivots')))
```

Errors are independently evaluated against graph shortest paths: all pairs for
dwt_307, and all distances from 32 independently sampled source vertices for
the larger saddle. **Original error** uses returned coordinates.
**Scale-adjusted error** additionally fits one positive scalar, a diagnostic
added after inspecting the original run; no layouts were refitted.

**Total seconds** includes graph preparation and fitting. **Peak MiB** includes
the entire isolated R process, including package loading. Tables report medians
over returned runs and show both the returned and attempted counts.

```{r selected-sparse, results='asis', echo=FALSE}
for(id in c('dwt_307','saddle-4096')) {
  case <- bundle$cases[[id]]
  cat('\n\n### ',case$label,' (',case$n,' vertices)\n\n',sep='')
  med <- summary[summary$case==id,];med$method <- method_label(med$method)
  cat(knitr::kable(med[,c('method','returned','attempted','relative_error',
    'profiled_relative_error','seconds','peak_mib')],digits=3,row.names=FALSE,
    col.names=c('Method','Returned','Attempted','Original error','Scale-adjusted error','Total seconds','Peak MiB')),sep='\n')
  selected <- Filter(function(f) f$row$case==id && f$row$seed==1,bundle$fits)
  fits <- setNames(selected,vapply(selected,function(f) f$row$method,''))
  child <- c(sprintf('```{r sparse-view-%s, echo=FALSE}',id),
    'sparse_comparison_view(case,fits,saved.display=TRUE)','```')
  cat(knitr::knit_child(text=child,envir=environment(),quiet=TRUE),sep='\n')
  cat('\n\nSeed-1 display scale multipliers: ',paste(vapply(fits,function(f)
    paste(method_label(f$row$method),format(f$row$relative_scale,digits=3)),''),collapse='; '),'.\n\n',sep='')
}
```

The figures show seed 1 with each method's independently fitted scale multiplier.
The saddle uses surface-shape alignment after scaling; the structural graph
uses matched-vertex alignment to full SGD, which is a comparison reference,
not a known true shape. The selector preserves the camera and scale.

## Complete results and reproduction

The [experiment appendix](sparse-mds-results.html) preserves every example,
including the smaller saddle, paraboloid, and helix, all outcomes, numerical
protocol, resource limits, and reproduction commands. A returned fit does not
establish convergence. Compare independently evaluated error rather than
sparse proxy values from different pivot counts.
