---
title: "Graph gallery"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Graph gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE,comment='#>',fig.width=9,fig.height=5,out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

Explore a structural mesh, a branching power network, a recursive carpet,
and a porous volume. Every figure compares **metric-MDS** with **metric-MDS +
edge-KK** using the same graph and lengths. The remaining nine examples,
including dwt_307, Les Miserables, curved surfaces, and the intrinsic tree,
are in the [extended gallery](graph-gallery-more.html).

The layouts use SGD, grip's default metric-MDS backend. They use uniform stress
weights; they do not reproduce the inverse-squared weighting in the Zheng et al.
paper. See [Metric MDS](metric-mds-backends.html) for that comparison.

```{r gallery-data, include=FALSE}
source(system.file('scripts','graph-examples.R',package='grip'),local=TRUE)
bundle <- readRDS(system.file('extdata','graph-examples','layouts.rds',package='grip'))
have.ivue <- documentation.interactive()
```

## Reading the gallery

**Blue is metric MDS; orange is MDS followed by edge-KK.** Gray, when present,
shows the coordinates used to construct the graph. Each visible layout includes
all its edges. Switching layouts preserves the camera and scale. Each layout is
**Procrustes-aligned** to the generating coordinates (or to metric MDS when no
generating coordinates are available). Corresponding vertices are matched by
their graph indices; translation and rotation/reflection minimize their summed
squared separations. No resizing is used to make the layouts agree.
Generating coordinates show where a geometric graph came from, but need not
realize its graph shortest-path distances as straight-line separations.

Each graph has an interactive 3D view made with
[ivue](https://github.com/pgajer/ivue). Use the visible layout selector to choose
**metric-MDS**, **metric-MDS + edge-KK**, or **both**. Gray generating coordinates
remain visible where available. Drag to rotate and scroll to zoom.
Without a compatible ivue installation, a static view shows the same
metric-MDS coordinates and alignment as the initial interactive view.

Each table reports two errors. **Graph-distance error** measures the fit to
all shortest-path distances; **edge-length error** measures the fit to supplied
edge lengths. Both are root-mean-square errors divided by the root-mean-square
target length, so zero is exact agreement and lower is better. Scores use the
returned coordinates without fitting an extra scale. Each stage's elapsed time
is also listed; preparation is excluded, and the refinement time is additional
to the MDS time. These are single-workstation timings, not a speed comparison.

```{r selected, include=FALSE}
gallery.ids <- c('dwt_1005','1138_bus','carpet','porous')
```

```{r gallery-pages, results='asis', echo=FALSE}
cat(knitr::knit_child(system.file('scripts','graph-gallery.Rmd',package='grip'),
  envir=environment(),quiet=TRUE),sep='\n')
```

## Draw a graph in two steps

This small example creates a weighted tree, draws it in 3D, and refines its
edge lengths. The gallery uses this same recipe for every graph.

```{r small-example}
g <- kary.tree.weighted.graph(k = 2, depth = 5, depth.decay = .85,
                            branch.spread = .2, normalize = 'none')
prepared <- prepare.graph.geodesic.mds(
  edges = g$edges, n = g$n, edge.weights = g$edge_weights)
mds <- suppressWarnings(metric.mds(
  prepared = prepared, dim = 3, init = 'random',
  n.init = 3, max.iter = 80, seed = 2026, diagnostics = FALSE))
refined <- edge.kk(
  coords = mds$coords, prepared = prepared, dim = 3,
  stiffness.method = 'uniform', density.mix.schedule = 1,
  scale.mode = 'identity', max.iter = 100,
  return.trace = FALSE, diagnostics = FALSE)
```

MDS tries to fit shortest-path distances between **all pairs of vertices**.
Edge-KK adjusts the drawing to fit **directly connected vertices**. Here every
edge has equal influence during refinement, and `scale.mode = "identity"`
keeps its target length fixed. An improvement in edge lengths can come with a
worse fit to distances between more widely separated vertices.

We use three random starts for MDS, retaining the one with the lowest stress,
and allow 80 passes per start. Reaching that limit can produce an
`iteration_limit` warning; the small example suppresses it, while the saved
gallery records it. Refinement allows 100 iterations. These are illustration
settings, not settings chosen to make either stage win. All fits use seed 2026.

For the six new datasets, replace the graph-construction line with:

```{r supplied-data}
data(zheng.graphs)
names(zheng.graphs)
g <- zheng.graphs[['lesmis']]
head(g$vertex_data)
```

Each dataset supplies `n`, `edges`, `edge_weights`, `vertex_data`, and
`graph_info`. Vertex indices follow the original matrix rows. The six graphs
are connected; no vertices were dropped or connecting edges added. We use
unit edge lengths and preserve the original matrix values separately.
For example, Les Miserables coappearance counts describe connection strength,
so they are not passed to the layout functions as distances.


## More graph constructors

| Need | Public entry points | Output |
|---|---|---|
| Topology alone | `edges.path()`, `edges.mesh()`, `edges.kary.tree()`, the other primitive `edges.*()` generators | Two-column edge matrix; no bundled reference coordinates or lengths. |
| A finite perforated grid | `keep.periodic.holes()`, `keep.asymmetric.notches()`, `keep.slit.channels()`, `keep.staggered.windows()`, then `occupied.mesh.surface.graph()` | A keep matrix followed by a complete weighted graph bundle. |
| A recursive pattern | Square `mask.*()` helpers, cubic `mask.cube.*()` arrays, or triangle/tetrahedron masks, then the matching `recursive.*.surface.graph()` | A mask followed by a recursively generated weighted bundle. |
| A regular reference surface | `mesh.surface.graph()`, `cylinder.surface.graph()`, `sphere.surface.graph()`, `torus.surface.graph()` | Edges, lengths, 3D reference coordinates, and parameters. |
| Irregular surfaces or volumes | The `irregular.*.surface.graph()` and `irregular.*.solid.graph()` families | Deterministic geometric constructions; “irregular” does not imply random sampling. |
| Triangular-lattice or polyhedral topology | The three `triangulated.*.surface.graph()` families | Graph edges and reference coordinates; the public bundles do not expose triangular faces. |
| Randomly sampled rectangle | `sampled.rectangle.surface.graph()` or plural `sampled.rectangle.surface.graphs()` | One graph or a sequence on a shared sampled point set. |
| Intrinsic lengths without an embedding | `kary.tree.weighted.graph()` | Tree edges, lengths, depths, parents, and edge metadata. |

The construction recipes in the [extended gallery](graph-gallery-more.html#construction-examples) illustrate additional families. Use complete public bundles; standalone embeddings
and specialized edge builders used inside them are internal functions.
A name ending in `surface.graph` is not a promise of a triangle mesh:
occupied-cube families, for example, connect occupied cells and can include
interior vertices.



## Constructing other graph families

Use a constructor when you need controlled topology or generating coordinates.
The [extended gallery](graph-gallery-more.html#construction-examples) includes
executable recipes and 3D views of a folded irregular annulus and a recursive
Vicsek pattern. Change one feature at a time when studying its effect.
The complete constructor catalog is in
[Finding your way around grip](function-guide.html#synthetic-graphs-complete-surface-and-solid-bundles).

## Repeat the gallery

The small tree example runs when this vignette is built. The gallery uses
saved layouts so rebuilding the document does not repeat the larger fits.
To draw any graph yourself, use its construction line above followed by the
two-step example. Keep `dim = 3` and the same edge lengths in both steps.

The full recipe is installed as `scripts/graph-examples.R`. Instructions for
rebuilding the saved layouts and images are in the
[reproduction guide](https://github.com/pgajer/grip/blob/main/tools/README.md#graph-examples-gallery).
The saved file retains the settings, software version, all MDS starts, warnings,
and any unsuccessful fits. This gallery uses dense shortest-path distances;
check graph size before increasing the construction sizes substantially.

## Data sources

The six graph datasets are derived from the
[SuiteSparse Matrix Collection](https://sparse.tamu.edu/), distributed under
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Original matrices,
author credits, character labels, and descriptions of the conversion are
included in `extdata/zheng-graphs/`. See `help('zheng.graphs')` for details.

- Zheng, J. X., Pawar, S. and Goodman, D. F. M. (2018).
  *Graph Drawing by Stochastic Gradient Descent*.
  [doi:10.1109/TVCG.2018.2859997](https://doi.org/10.1109/TVCG.2018.2859997).
- Davis, T. A. and Hu, Y. (2011). *The University of Florida Sparse Matrix Collection*.
  [doi:10.1145/2049662.2049663](https://doi.org/10.1145/2049662.2049663).
- Knuth, D. E. (1993). *The Stanford GraphBase: A Platform for Combinatorial Computing*.
  Addison-Wesley. Source of the Les Miserables coappearance data.
