---
title: "Synthetic graph families and layout examples"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Synthetic graph families and layout examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1000px; }
td, th { vertical-align: top; }
td code { white-space: normal; overflow-wrap: anywhere; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     fig.width = 6, fig.height = 4.5)
```

Choose a synthetic family to make a layout question reproducible. A tree
isolates branching and edge-length effects; a masked grid adds holes and
bottlenecks; a surface bundle supplies reference coordinates. These examples
use small graphs to demonstrate the workflow, not to rank layout methods.
For the complete function catalog and score definitions, see
[Finding your way around grip](function-guide.html).

## Choose the kind of input you need

| Need | Public entry points | Output |
|---|---|---|
| Topology alone | `edges.path()`, `edges.mesh()`, `edges.kary.tree()`, the other primitive `edges.*()` generators | Two-column edge matrix; no bundled reference coordinates or lengths. |
| A finite perforated grid | `keep.periodic.holes()`, `keep.asymmetric.notches()`, `keep.slit.channels()`, `keep.staggered.windows()`, then `occupied.mesh.surface.graph()` | A keep matrix followed by a complete weighted graph bundle. |
| A recursive pattern | Square `mask.*()` helpers, cubic `mask.cube.*()` arrays, or triangle/tetrahedron masks, then the matching `recursive.*.surface.graph()` | A mask followed by a recursively generated weighted bundle. |
| A regular reference surface | `mesh.surface.graph()`, `cylinder.surface.graph()`, `sphere.surface.graph()`, `torus.surface.graph()` | Edges, lengths, 3D reference coordinates, and parameters. |
| Irregular surfaces or volumes | The `irregular.*.surface.graph()` and `irregular.*.solid.graph()` families | Deterministic geometric constructions; “irregular” does not imply random sampling. |
| Triangular-lattice or polyhedral topology | The three `triangulated.*.surface.graph()` families | Graph edges and reference coordinates; the public bundles do not expose triangular faces. |
| Randomly sampled rectangle | `sampled.rectangle.surface.graph()` or plural `sampled.rectangle.surface.graphs()` | One graph or a sequence on a shared sampled point set. |
| Intrinsic lengths without an embedding | `kary.tree.weighted.graph()` | Tree edges, lengths, depths, parents, and edge metadata. |

The [website gallery](https://pgajer.github.io/grip/articles/gallery.html)
illustrates more families. Use complete public bundles; standalone embeddings
and specialized edge builders used inside them are internal functions.
A name ending in `surface.graph` is not a promise of a triangle mesh:
occupied-cube families, for example, connect occupied cells and can include
interior vertices.

## Read a bundle before using it

```{r bundle-structure}
library(grip)
mesh <- mesh.surface.graph(3, 4, surface = "saddle", normalize = "none")
names(mesh)
c(vertices = mesh$n, edges = nrow(mesh$edges),
  reference_dimensions = ncol(mesh$coords_surface))
head(mesh$edges)
```

| Field | Meaning and ordering |
|---|---|
| `edges` | Integer matrix with two columns, one undirected edge per row; endpoints are 1-based vertex indices. |
| `n` | Total number of vertices. Use this instead of `max(edges)` so isolated vertices are retained. |
| `edge_weights` | Positive lengths parallel to edge rows. These are traversal lengths, not strengths. |
| `coords_surface` | When supplied, an `n` by 3 reference-coordinate matrix. Row `i` is vertex `i`, in the same order as the edge endpoints. Tree bundles omit it. |
| `coords_param` | When supplied, coordinates before the geometric deformation. Often `n` by 2, but polyhedra and volumetric families can use three columns. These are not automatically layout targets. |
| `weight_scale` | Divisor used to normalize raw lengths: supplied length equals raw length divided by `weight_scale`. The reference coordinates remain in their original units. |
| `normalize`, `family`, `label`, and construction controls | Record the selected normalization and family-specific settings. Exact fields differ by constructor. |

Most geometric families induce lengths from the Euclidean distances between
reference endpoints. `kary.tree.weighted.graph()` assigns lengths intrinsically.
Sampled rectangles also return `raw_edge_weights` (3D endpoint distances) and
`iknn_witness_edge_weights` (intersection-neighbor construction diagnostics).
Even with `graph_space = "param"`, the final sampled-rectangle lengths use
3D endpoint distances; `graph_space` changes graph construction.

Do not assume a shared schema beyond the documented constructor fields.
The plural sampled-rectangle function returns a container with `$graphs`,
`$k`, `$k_statistics`, and shared coordinate matrices; pass one member of
`$graphs` to the layout workflow. It sorts and deduplicates the supplied
neighborhood sizes. No public bundle used here supplies `faces` or `triangles`.
`score.surface()` needs explicit triangle connectivity supplied separately;
it cannot infer a surface from these edge lists.

Ordering is part of reproducibility. The rectangular mesh enumerates each
row from left to right. Mask matrices use display orientation (top row first,
left column first); occupied-cell coordinates and graph indices share that
order. Sphere bundles put the north pole first, then latitude rings from
north to south, then the south pole. Trees put the root first and enumerate
children by parent and child slot. For other families, retain the constructor's
returned order rather than reconstructing it from a drawing. Reordering rows
requires remapping edges and every reference and metadata field together.

## Example 1: a tree with intrinsic edge lengths

Use a primitive path when you need only topology:

```{r primitive-path}
path.edges <- edges.path(7)
path.coords <- grip(path.edges, n = 7, dim = 2, preset = "tree", seed = 21)
stopifnot(identical(dim(path.edges), c(6L, 2L)))
```

A complete weighted tree separates depth-dependent tapering from child-slot
asymmetry. Here a binary tree has three levels below the root (15 vertices).
It has no generating reference coordinates; a graph-distance diagnostic is
therefore the appropriate first score.

```{r weighted-tree}
tree <- kary.tree.weighted.graph(
  k = 2, depth = 3, depth_decay = 0.8,
  branch_spread = 0.25, normalize = "none"
)
knitr::kable(head(tree$edge_table), digits = 3)
tree.coords <- grip(tree$edges, n = tree$n,
                    edge_weights = tree$edge_weights,
                    metric = "edge_length", length_normalization = "none",
                    dim = 2, preset = "tree", seed = 21)
plot.layout(tree.coords, edges = tree$edges,
            vertex.col = tree$vertex_depth + 1L, pch = 16,
            main = "Intrinsic lengths on a binary tree")
tree.score <- score.layout(tree.coords, edges = tree$edges, n = tree$n,
                            edge_weights = tree$edge_weights,
                            sample.size.stress = 150, stress.seed = 21,
                            edge.crossings = "never")
knitr::kable(tree.score[, c("sampled.stress", "edge.length.cv")], digits = 3)
```

`parent` has one entry per vertex and uses 0 for the root's missing parent;
`vertex_depth` starts at 0. `edge_table` follows the edge rows and records
parent, child, both depths, child slot, raw length, and normalized length.
Unequal drawn edges are intentional here, so a low edge-length coefficient of
variation is not the goal. Weighted sampled stress evaluates chords against
tree-path distances after fitting scale. Branching tree metrics cannot in
general be preserved exactly as Euclidean separations in two dimensions.

## Example 2: a recursive graph with holes

A keep mask describes which cells survive each subdivision. The border of a
3 by 3 mask removes the center at each step. At level 2 there are 64 retained
cells. This example uses a flat lift (`amplitude = 0`) and omits lengths when
fitting and scoring, so both operations use hop counts.

```{r masked-grid}
keep <- mask.border(3)
keep
carpet <- recursive.mask.grid.surface.graph(
  mask = keep, level = 2, surface = "saddle", amplitude = 0
)
stopifnot(carpet$n == 64L, nrow(carpet$coords_param) == carpet$n)
carpet.coords <- grip(carpet$edges, n = carpet$n,
                      metric = "hop", dim = 2, preset = "carpet", seed = 22)
```

```{r masked-grid-plot, fig.width=8, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(3, 3, 2, 1))
plot.layout(carpet$coords_param, edges = carpet$edges, pch = 16, cex = 0.5,
            main = "Generating grid")
plot.layout(carpet.coords, edges = carpet$edges, pch = 16, cex = 0.5,
            main = "Hop-metric drawing")
par(op)
```

```{r masked-grid-score}
carpet.score <- score.layout(carpet.coords, edges = carpet$edges, n = carpet$n,
                              sample.size.stress = 300, stress.seed = 22,
                              edge.crossings = "never")
carpet.score$sampled.stress
```

The layout preserves the graph's vertex and edge identities, but it need not
reproduce the original square arrangement. Holes and narrow connections make
this a useful example for [tracing](grip-trace-and-diagnostics.html).
A different keep pattern can disconnect a graph or isolate cells; `grip()`
packs components by default. Such packing is not evidence of distances
between disconnected components.

Square and cubic recursive grids require `level >= 1`. Triangle and tetrahedron
recursions and triangulated polyhedra also allow level 0. Triangle masks have
four named child slots; tetrahedron masks have four corner slots. A cubic
keep array is not interchangeable with either. Increasing depth multiplies
the retained cells, so inspect `n` before launching a layout or all-pairs score.

## Example 3: graph fidelity versus reference recovery

An octahedron subdivided once gives 18 vertices and a three-dimensional
reference configuration. The two fits below keep topology, vertex order,
supplied lengths, seed, dimension, preset, and iteration controls fixed.
Both receive the same **unnormalized lengths**. Hop mode uses those lengths
for adjacent-edge forces; edge-length mode additionally uses them for graph
neighborhoods and hierarchy construction. Its solver normalization is set to
`"none"` to retain the same input length scale.

```{r surface-fits}
surface <- triangulated.polyhedron.surface.graph(
  base = "octahedron", level = 1, surface = "inflated", normalize = "none"
)
common <- list(edges = surface$edges, n = surface$n,
               edge_weights = surface$edge_weights, dim = 3,
               preset = "mesh", rounds = 40, final_rounds = 60, seed = 23)
fits <- list(
  hop = do.call(grip, c(common, list(metric = "hop"))),
  edge_length = do.call(grip, c(common, list(
    metric = "edge_length", length_normalization = "none")))
)
stopifnot(surface$n == 18L,
          all(vapply(fits, function(x) all(is.finite(x)), logical(1))))
reference.scores <- lapply(fits, score.coordinates,
                           reference = surface$coords_surface,
                           alignment = "similarity")
summary <- do.call(rbind, lapply(names(fits), function(method) {
  graph.score <- score.layout(
    fits[[method]], edges = surface$edges, n = surface$n,
    edge_weights = surface$edge_weights, sample.size.stress = 200,
    stress.seed = 23, edge.crossings = "never"
  )
  data.frame(method = method,
             weighted_chord_stress = graph.score$sampled.stress,
             reference_relative_rmse = reference.scores[[method]]$relative_rmse,
             alignment_scale = reference.scores[[method]]$scale)
}))
knitr::kable(summary, digits = 3)
```

Both drawings are scored against the **same weighted graph distances**, with
the same pair sampling settings. The coordinate score separately uses the
same reference and vertex correspondence, fitting translation, an orthogonal
transformation (including reflection), and uniform scale. Relative coordinate
RMSE is root mean squared vertex displacement divided by the reference's RMS
radius. Similarity alignment removes overall size differences, so it does
not establish recovery of absolute length scale. Use `alignment = "rigid"`
when scale must count as error.

```{r surface-projections, fig.width=9, fig.height=3.6}
op <- par(mfrow = c(1, 3), mar = c(1, 1, 2, 1))
plot.layout(surface$coords_surface, edges = surface$edges,
            projection = "ortho", main = "Reference")
for (method in names(fits)) {
  plot.layout(reference.scores[[method]]$coords, edges = surface$edges,
              projection = "ortho",
              main = if (method == "hop") "Hop metric" else "Edge-length metric",
              vertex.col = "#215A87")
}
par(op)
```

These are static orthographic views of 3D coordinates, not new 2D fits.
Scores above use all three columns. The generating coordinates provide one
reference: graph symmetries, bending, missing constraints, and the choice of
objective can permit other drawings. Matching graph-path distances, matching
endpoint separations, and recovering this reference are distinct goals.
Here the rounded scores agree closely, showing that the two metric choices
can give similar results on a symmetric graph. One graph and one seed do not
establish general superiority of either mode.

## Reproducibility and practical limits

Primitive generators and the regular, recursive, and irregular families are
deterministic for fixed construction arguments. The sampled-rectangle family
uses random sampling: a supplied integer `seed` fixes that sample and restores
the caller's R random-number state; `seed = NULL` uses and advances the current
stream. A plural call samples once for all requested neighborhood sizes.
Graph sampling and layout fitting have separate seeds.

```{r sampled-reproducibility}
sample.one <- sampled.rectangle.surface.graph(n = 20, k = 4, seed = 24)
sample.two <- sampled.rectangle.surface.graph(n = 20, k = 4, seed = 24)
stopifnot(identical(sample.one$edges, sample.two$edges),
          identical(sample.one$coords_surface, sample.two$coords_surface))
c(vertices = sample.one$n, edges = nrow(sample.one$edges))
```

For `grip()`, pass an explicit `seed` (the default is 6); `NULL` uses the
current time rather than the caller's R random stream. Record the package
version, graph arguments, vertex order, graph and solver normalization, layout
controls, and scoring seeds. Fixed seeds aid repeatability within an
implementation; they are not promises of identical floating-point results
across future versions and platforms.

The surface and solid bundles here return three reference columns, while
`grip()` fits either 2D or 3D. Higher-dimensional layout workflows are described
in the [function guide](function-guide.html); changing layout dimension does
not create a higher-dimensional reference bundle. Sphere and wrapped families
have minimum grid-size restrictions; intrinsic trees can have depth 0, with
one vertex and no edges. Use constructor help for parameter bounds.

No example here needs Shiny, rgl, an interactive device, or a browser. For
larger families, first inspect the graph size and use modest sampled scores;
full shortest-path matrices and route caches can dominate memory. Optional
[interactive exploration](https://pgajer.github.io/grip/articles/gripui-explorer.html)
and the [gallery](https://pgajer.github.io/grip/articles/gallery.html) are
website follow-ups. The installed [weighted-layout guide](weighted-grip-intro.html)
and [real-data guide](grip-real-data.html) develop the corresponding workflows.
