---
title: "Layouts with meaningful edge lengths"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Layouts with meaningful edge lengths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1050px; }
body:has(nav.navbar) { max-width: none; }
.ivue-legend { max-width: 100%; }
.rglWebGL { max-width: 100%; }
th, td { vertical-align: top; }
pre { overflow-x: auto; }
</style>

```{r setup, include=FALSE}
library(grip)
knitr::opts_chunk$set(collapse=TRUE, comment='#>', fig.width=9, fig.height=5,
                     out.width='100%')
source(system.file('scripts','vignette-views.R',package='grip'),local=TRUE)
```

An edge length specifies a desired traversal cost or separation. It differs
from a connection strength, and from the weight assigned to a distance error
inside an optimizer. This guide demonstrates the metric choice on a torus and
then shows a tree whose lengths are defined without a generating surface.

## The same lengths, two graph metrics

A torus has two periodic directions. The constructor supplies 3D generating
coordinates, edges, and Euclidean lengths between adjacent surface points.
These lengths define paths on a finite graph, not exact continuous-surface
geodesics.

```{r torus-fits}
g <- torus.surface.graph(12,8,normalize='none')
hop <- grip(edges=g$edges,n=g$n,edge.weights=g$edge_weights,
  dim=3,metric='hop',seed=1)
weighted <- grip(edges=g$edges,n=g$n,edge.weights=g$edge_weights,
  dim=3,metric='edge_length',length.normalization='none',seed=1)
```

Both fits receive the **same edge lengths**, topology, dimension, and seed.
In hop mode, those lengths affect adjacent-edge attractions, while the standard
hierarchy and neighborhoods use hop counts. In edge-length mode, sums of the
lengths guide the multiscale graph searches. Omitting lengths from the hop call
would instead give a purely unweighted baseline, which is a different comparison.

```{r torus-view, echo=FALSE}
documentation.view(list('GRIP: hop'=hop,'GRIP: edge lengths'=weighted),
  g$edges,reference=g$coords_surface,common.size=TRUE,
  description='Torus: identical input lengths in both fits. Gray: generating coordinates. Aligned at a common overall size for display.')
```

The gray torus shows how the graph was constructed. Graph shortest-path
distances need not be realizable as straight-line distances on that surface.
The views use matched-vertex rotation/reflection and translation, with a uniform display multiplier to match each fit to the
reference's root-mean-square radius. This separates shape from overall size;
the scores below use the original coordinates. A close overlay and low graph-distance stress answer different questions.

```{r weighted-scores}
fits <- list(Hop=hop,'Edge lengths'=weighted)
scores <- do.call(rbind,lapply(fits,function(z)
  score.layout(z,edges=g$edges,n=g$n,edge.weights=g$edge_weights,
    sample.size.stress=2000L,edge.crossings='never')))
knitr::kable(scores[,c('sampled.stress','edge.length.cv',
                     'sampled.nonedge.sep.ratio')],digits=3)
```

Both layouts are evaluated against the same weighted graph distances.
Unequal drawn edges can be appropriate here: low edge-length variation is not
itself the objective for a graph with unequal target lengths.

## Lengths without a reference surface

This binary tree has shorter edges at greater depth and unequal sibling
branches. Its lengths are intrinsic inputs; there is no generating 3D shape
to recover.

```{r tree}
tree <- kary.tree.weighted.graph(k=2,depth=5,depth.decay=.85,
  branch.spread=.2,normalize='none')
tree.coords <- grip(edges=tree$edges,n=tree$n,edge.weights=tree$edge_weights,
  dim=3,metric='edge_length',length.normalization='none',seed=1)
```

```{r tree-view, echo=FALSE}
documentation.view(list('Weighted GRIP'=tree.coords),tree$edges,
  description='An intrinsic weighted tree: shorter branches at greater depth. No generating reference.')
```

## Units, normalization, and stress weighting

`length.normalization='none'` retains the numerical input lengths in the
weighted GRIP engine; the default median normalization changes their common
scale. It does not make the final coordinates an exact realization of every
target length. Check the returned layout with a score appropriate to the task.

For metric MDS, `edge.weights` determines the shortest-path targets, while
`pair.weights` determines how errors at different distances influence the fit.
The [uniform versus inverse-squared comparison](metric-mds-backends.html#the-effect-of-stress-weighting)
illustrates this distinction. The two parameters serve different purposes.

Use [Choosing and evaluating a layout](grip-real-data.html) to compare settings,
[the graph gallery](synthetic-graph-families.html) for other surfaces and
recursive constructions, and [HMP/U01](hmp-u01-case-study.html) for a biological
graph with distance-valued edges. Specialized path-based refinements and their
scores remain documented in the [function overview](function-guide.html#experimental-refinement-and-current-initializers).
