---
title: "Weighted Graph Layouts with grip"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weighted Graph Layouts with grip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.8,
  fig.height = 5.3
)
```

## Why weighted layouts deserve their own vignette

Classical GRIP is fundamentally combinatorial: it sees the graph topology and
uses graph-distance neighborhoods, but it does not treat edge lengths as a
first-class geometric signal throughout the multiscale hierarchy.

That is exactly why `grip` now includes a weighted sister API:

- `weighted.grip()`
- `globalrep.weighted.grip()`
- `trace.weighted.grip()`
- `build.weighted.misf()`

Use these when edge lengths represent geometry you care about rather than just
optional metadata.

The package decision rule is:

- use `grip()` for ordinary unweighted or topology-first graphs,
- use `weighted.grip()` when edge lengths carry geometry you want to
  preserve,
- add GKK/LGKK only after you already have weighted candidate layouts and need
  advanced experimental geodesic-aware scoring or polish.

```{r}
library(grip)
```

```{r}
plot.layout.triptych <- function(coords.list,
                                 edges,
                                 titles,
                                 projection = NULL,
                                 vertex.cols = rep("black", length(coords.list)),
                                 edge.col = "gray82") {
  op <- par(
    mfrow = c(1, length(coords.list)),
    mar = c(1.2, 1.2, 3, 1.2),
    bg = "white"
  )
  on.exit(par(op), add = TRUE)

  for (i in seq_along(coords.list)) {
    plot.layout(
      coords.list[[i]], edges,
      projection = projection,
      main = titles[[i]],
      vertex.col = vertex.cols[[i]],
      edge.col = edge.col
    )
  }
}
```

## A first weighted surface example

The helper below creates a plain mesh topology whose edge lengths are induced
by a curved 3D surface. The topology stays simple, but the intended metric is
no longer the flat grid metric.

```{r}
surface.mesh <- mesh.surface.graph(
  5, 5,
  surface = "saddle",
  amplitude = 0.9
)

coords.unweighted <- grip(
  surface.mesh$edges,
  n = surface.mesh$n,
  dim = 3,
  preset = "mesh",
  seed = 1
)

coords.weighted <- weighted.grip(
  surface.mesh$edges,
  n = surface.mesh$n,
  edge_weights = surface.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  seed = 1
)

gkk.prepared <- prepare.geodesic.kk(
  surface.mesh$edges,
  n = surface.mesh$n,
  edge_weights = surface.mesh$edge_weights
)

surface.summary <- do.call(
  rbind,
  list(
    cbind(
      method = "Combinatorial GRIP",
      score.geodesic.kk(
        coords.unweighted,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rmse",
        "gkk.mean.abs.path.error",
        "gkk.mean.rel.path.error"
      )]
    ),
    cbind(
      method = "Weighted GRIP",
      score.geodesic.kk(
        coords.weighted,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rmse",
        "gkk.mean.abs.path.error",
        "gkk.mean.rel.path.error"
      )]
    )
  )
)

knitr::kable(surface.summary, digits = 3)
```

```{r fig.width=12.8, fig.height=4.2}
plot.layout.triptych(
  list(
    surface.mesh$coords_surface,
    coords.unweighted,
    coords.weighted
  ),
  edges = surface.mesh$edges,
  titles = c("Target geometry", "Combinatorial GRIP", "Weighted GRIP"),
  projection = "ortho",
  vertex.cols = c("#666666", "black", "#1F3B73")
)
```

The important pattern is not that one method always wins on every graph. It is
that weighted GRIP is solving a different problem: it tries to respect the
graph's edge-length geometry, not only its combinatorial adjacency structure.

## 2D versus 3D on the same weighted graph

For many weighted geometric families, 3D is the more informative target space.
The graph metric can be difficult or impossible to represent faithfully in 2D
without substantial distortion.

```{r}
coords.weighted.2d <- weighted.grip(
  surface.mesh$edges,
  n = surface.mesh$n,
  edge_weights = surface.mesh$edge_weights,
  dim = 2,
  preset = "mesh",
  seed = 2
)

coords.weighted.3d <- weighted.grip(
  surface.mesh$edges,
  n = surface.mesh$n,
  edge_weights = surface.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  seed = 2
)

dim.summary <- do.call(
  rbind,
  list(
    cbind(
      dim = "2D",
      score.geodesic.kk(
        coords.weighted.2d,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rmse",
        "gkk.mean.abs.path.error",
        "gkk.mean.rel.path.error"
      )]
    ),
    cbind(
      dim = "3D",
      score.geodesic.kk(
        coords.weighted.3d,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rmse",
        "gkk.mean.abs.path.error",
        "gkk.mean.rel.path.error"
      )]
    )
  )
)

knitr::kable(dim.summary, digits = 3)
```

```{r fig.width=10.2, fig.height=4.3}
op <- par(mfrow = c(1, 2), mar = c(1.2, 1.2, 3, 1.2), bg = "white")
on.exit(par(op), add = TRUE)

plot.layout(
  coords.weighted.2d,
  surface.mesh$edges,
  main = "Weighted GRIP in 2D",
  vertex.col = "black",
  edge.col = "gray82"
)

plot.layout(
  coords.weighted.3d,
  surface.mesh$edges,
  projection = "ortho",
  main = "Weighted GRIP in 3D",
  vertex.col = "#1F3B73",
  edge.col = "gray82"
)
```

This is why the weighted benchmark work in `grip` treats 3D as the primary
track and 2D as an informative limitation track.

## Weighted presets

The weighted API keeps explicit presets tuned for the major weighted-family
classes currently shipped with the package.

| Family class | Good weighted preset | Typical use |
|:--|:--|:--|
| Lifted mesh surfaces | `preset = "mesh"` | Rectangular weighted surfaces |
| Cylindrical grids | `preset = "cylinder"` | Open wrapped surfaces |
| Toroidal grids | `preset = "torus"` | Closed wrapped surfaces |
| Near-spherical surfaces | `preset = "sphere"` | Closed surface families |
| Irregular manifolds and porous families | `preset = "irregular"` | Non-lattice weighted manifolds |
| Intrinsic weighted trees | `preset = "tree"` | Edge-length-driven tree geometry |
| Recursive carpet-like lattices | `preset = "carpet"` | Recursive hole-rich weighted grids |

These presets are starting points, not declarations that the graph belongs to a
single correct family.

## Intrinsic weighted trees

Weighted families do not need to come from ambient surfaces. They can also be
intrinsically weighted. The example below keeps the topology of a binary tree
but assigns edge lengths by depth and branch position.

```{r}
tree.graph <- kary.tree.weighted.graph(
  k = 2,
  depth = 4,
  depth_rule = "geometric",
  depth_decay = 0.82,
  branch_rule = "linear",
  branch_spread = 0.25
)

tree.coords <- weighted.grip(
  tree.graph$edges,
  n = tree.graph$n,
  edge_weights = tree.graph$edge_weights,
  dim = 2,
  preset = "tree",
  seed = 3
)

knitr::kable(
  head(tree.graph$edge_table[, c(
    "parent",
    "child",
    "child_depth",
    "branch_index",
    "edge_weight"
  )]),
  digits = 3
)
```

```{r fig.width=5.4, fig.height=4.6}
plot.layout(
  tree.coords,
  tree.graph$edges,
  main = "Intrinsic weighted tree",
  vertex.col = "#1F3B73",
  edge.col = "gray80",
  pch = 16,
  cex = 0.55
)
```

This kind of example is useful because the geometry lives in the edge lengths
themselves rather than in a chosen 3D embedding.

## Trace and advanced geodesic hooks

The weighted API also supports:

- `trace.weighted.grip()` for traced weighted solves,
- `prepare.geodesic.kk()` / `score.geodesic.kk()` for full GKK
  evaluation,
- `prepare.landmark.geodesic.kk()` /
  `score.landmark.geodesic.kk()` for sparse LGKK evaluation,
- `lgkk_polish_rounds` for post-layout landmark geodesic KK refinement,
- `lgkk_multiscale_rounds` and the stage-specific LGKK controls for in-core
  multiscale refinement.

These GKK/LGKK tools are public, but they are not the default entry path. For
most weighted problems, start with `weighted.grip()` and use the
geodesic tools only when you need a stronger metric comparison or an
experimental polish step.

Here is the smallest traced weighted example pattern:

```{r eval=FALSE}
surface.trace <- trace.weighted.grip(
  surface.mesh$edges,
  n = surface.mesh$n,
  edge_weights = surface.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  trace = "level",
  diagnostics = "light",
  seed = 1
)

head(surface.trace$meta)
head(surface.trace$diagnostics)
```

For small or medium weighted graphs where geodesic fidelity matters strongly,
it is often worth comparing:

- weighted GRIP alone,
- weighted GRIP + experimental core LGKK,
- weighted GRIP + experimental polish LGKK,
- and the KK->GKK or KK->LGKK baselines.

## Where to go next

- `Getting Started with grip` shows how the weighted API fits into the broader
  package.
- `Choosing Layouts for Real Data` focuses on candidate search and scoring.
- `Tracing and Diagnosing Layouts` covers the trace APIs in more detail.
- `Synthetic Graph Families and Geometries` explains the benchmark-family
  library that supports these weighted workflows.
