---
title: "Supplement S3: Controlled examples and the sampled-saddle experiment"
author: "Pawel Gajer and Jacques Ravel"
date: ""
output:
  bookdown::pdf_document2:
    keep_tex: true
    citation_package: natbib
    latex_engine: pdflatex
    toc: false
fontsize: 10pt
papersize: a4
geometry: margin=22mm
bibliography: S3-controlled-examples.bib
header-includes:
  - \usepackage{amsmath,array,booktabs,float,needspace}
  - \usepackage{xurl}
  - \providecommand{\CRANpkg}[1]{\textbf{#1}}
  - \providecommand{\pkg}[1]{\textbf{#1}}
  - \renewcommand{\thefigure}{S3.\arabic{figure}}
  - \renewcommand{\thetable}{S3.\arabic{table}}
  - \renewcommand{\thepage}{S3--\arabic{page}}
  - \setlength{\emergencystretch}{2em}
---

```{r setup-s3, include=FALSE}
# Load at knit time, when knit_root_dir is the manuscript directory.
do.call(knitr::read_chunk, list("reproducibility/scripts/edge-kk-workflow.R"))
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
  fig.align="center", out.width="100%", fig.pos="H",
  fig.path="build/supplement/S3_files/figure-latex/")
stopifnot(requireNamespace("grip", quietly=TRUE),
          requireNamespace("dgraphs", quietly=TRUE),
          requireNamespace("igraph", quietly=TRUE),
          requireNamespace("png", quietly=TRUE))
library(grip)
source("reproducibility/scripts/plot-two-fidelity.R")
pilot <- readRDS("reproducibility/precomputed/two-fidelity-saddle.rds")
benchmark.results <- readRDS("reproducibility/precomputed/vs_alternatives/benchmark_results.rds")
paper_table_caption <- function(title, abbreviations) paste(title, "Abbreviations:", abbreviations)
s3_kable <- function(x, caption, position="H", ...) {
  knitr::kable(x, caption=gsub("%", "\\%", caption, fixed=TRUE),
    position=position, ...)
}
accessible_kable <- function(x, caption, alt=NULL, row.names=FALSE, align=NULL,
  col.names=NULL, digits=3, position="H", group_after=integer(), ...) {
  if (is.null(col.names)) col.names <- names(x)
  s3_kable(x, caption=caption, row.names=row.names, align=align,
    col.names=col.names, digits=digits, format="latex", booktabs=TRUE,
    position=position, linesep="")
}
```

# Sampled-saddle protocol and numerical checks

This supplement documents the five-sample experiment in the manuscript and
retains additional controlled examples. The sampled-saddle study is a
fixed-budget illustration of the two fidelity stages. It does not establish
a general ranking of graph constructors or embedding methods.

## Sampling and graph construction

Each cloud contains 1,000 independent observations on
$(x,y,0.8(x^2-y^2))$, with $(x,y)\in[-1,1]^2$, without measurement noise.
Rejection sampling gives uniform surface area: a uniform-square proposal is
accepted with probability $\sqrt{1+2.56(x^2+y^2)}/\sqrt{6.12}$.
R seeds are `2211000 + replicate`, with Mersenne-Twister, Inversion normals,
and Rejection sampling explicitly selected. Reference-source seeds are
`3211000 + replicate`; 128 vertices are chosen without replacement, independently
of graph construction and layout scores.

Graphs use `dgraphs::create.sknn.graph()` with exact neighbors, symmetric union,
Euclidean edge lengths in the original 3D observations, no pruning, and exact
component-MST repair. The number and lengths of added bridges are saved.
Graph shortest paths for calibration are computed with `igraph::distances()`.
The initial integer sweep `3:20` had boundary minima, as did its extension to
`3:40`; the pilot therefore expanded to `3:80` for every cloud. This adaptive
range finding is part of the pilot, not a predeclared confirmatory study.

## Numerical reference for smooth-surface distances

The reference domain is the **bounded saddle patch**, not the entire unbounded
surface. Every original observation is included as a vertex, together with a
regular parameter grid. Delaunay triangulation is performed in `(x,y)`, then
vertices are lifted to the saddle. Endpoints are not snapped to nearby grid
vertices. Paths can cross triangle interiors and cannot leave the patch.

The mesh solver is `pygeodesic` 0.1.11, an interface to Kirsanov's implementation
of the exact polyhedral geodesic algorithm [@pygeodesic]. "Exact" concerns
the triangulated surface in floating-point arithmetic; it does not make these
distances exact on the smooth saddle.

The primary reference uses an 81-by-81 grid plus the 1,000 observations.
Distances from 128 random source vertices to all observations cover 119,744
distinct unordered pairs, after removing self-pairs and duplicates. All five
clouds also use a 41-by-41 grid with the same 128 sources and a 161-by-161 grid
with 16 of those sources. The most resolution-sensitive cloud receives a further
161-by-161 check on all 128 sources. These comparisons assess numerical
sensitivity rather than provide a certified uniform error bound.

Independent controls solve the smooth-surface geodesic boundary-value equation
for 128 source-target pairs per cloud. For $f(x,y)=0.8(x^2-y^2)$, a geodesic in
parameter coordinates $u(t)=(x(t),y(t))$ with an affine parameter satisfies
\[
u''=-\frac{\nabla f(u)\,[u'^{\mathsf T}\nabla^2 f(u)\,u']}
               {1+\|\nabla f(u)\|_2^2},
\qquad
\nabla f(u)=(1.6x,-1.6y),\quad
\nabla^2 f(u)=\operatorname{diag}(1.6,-1.6).
\]
The SciPy boundary-value solver uses tolerance $10^{-8}$ and at most
10,000 collocation nodes [@scipy-bvp]. Curve length is integrated
on 1,001 evaluation points. Each check requires successful termination and
checks that the evaluated curve stays inside the parameter square. These are
independent numerical controls, not an interval-arithmetic proof. A separate
flat-surface test compares the polyhedral solver with exact Euclidean lengths.

## The two stages and the scale convention

For each cloud, choose the graph minimizing the **unprofiled** surface-to-graph
error on the reference pairs:

\[
E_{XG}=\left\{\frac{\sum_{(i,j)\in\mathcal P}(d_{ij}^G-d_{ij}^X)^2}
                         {\sum_{(i,j)\in\mathcal P}(d_{ij}^X)^2}\right\}^{1/2},
\]
where $\mathcal P$ is the set of reference pairs.

Graph and surface distances retain their common physical units. The full
error-versus-k curve, fitted-scale diagnostic, signed relative bias, fraction
of graph distances shorter than the reference, and 32/64/128-source sensitivity
are retained. The minimizing integer is an **oracle choice against a numerical
reference**. Shallow minima and reference/source-subset sensitivity prevent
interpreting that integer as a uniquely established population optimum.

On the selected graph, all three configurations have three coordinate columns:

1. Original saddle observations (known edge/path-fidelity control).
2. `grip::metric.mds(dim=3)`, classical scaling of graph distances.
3. That same MDS result followed by `grip::edge.kk(dim=3)`.

The primary refinement uses the package's density-based stiffness continuation
with mixing coefficients `0, .25, .5, .75, 1`, at most 200 iterations per stage,
profiled scale, and zero edge-length stabilizer. A separate sensitivity run
continues its result for at most 1,000 additional iterations at mixing
coefficient 1. The additional run never replaces the primary comparison.
There is no direct optimization of the fixed-path GMDS objective in this pilot.

`grip::prepare.geodesic.kk(tie_mode="single")` retains one input route for each
of all 499,500 unordered vertex pairs. All three candidates use these same
routes. Their embedded lengths are sums of Euclidean edge lengths, without
recomputing shortest paths after embedding.

For embedded lengths $y_p$ and graph lengths $g_p$, the fitted scale
$a=\sum_p y_p g_p/\sum_p g_p^2$ gives
\[
\operatorname{RelRMSE}=
\left\{\frac{\sum_p(y_p-a g_p)^2}{\sum_p(a g_p)^2}\right\}^{1/2}.
\]
Edges and retained paths are fitted separately. MDS Stress-1 uses endpoint
chords for $y_p$ and replaces the denominator by $\sum_p y_p^2$.
An independent C++
route sum and a direct R route sample are checked against `score.gmds()`.
The package's near-tie tolerance can create tiny discrepancies against
igraph's strict shortest-path distances; these are checked and recorded.

For an additional **end-to-end** comparison against $D_X$, convert each
configuration into input units using its edge scale (estimated only from
graph edges), then compare its fixed-path lengths with $D_X$ without fitting
another scale against the surface. A Procrustes error measures coordinate
shape discrepancy, distinct from all distance-fidelity statistics.

The original 3D coordinates preserve every Euclidean edge and retained graph
path, including MST bridges. This does not make chord stress zero or establish
surface-geodesic accuracy. Edge/path fidelity also need not identify the
original coordinate shape uniquely.


## Per-sample results and numerical reference checks

All 390 graph cases and all five primary layout comparisons completed; none
was excluded. All 25 primary continuation stages reached their iteration cap.
The surface-to-graph minima and layout results are recorded separately below.
All table entries headed by an error statistic are percentages.

```{r primary-results}
mds <- pilot$scores[pilot$scores$method == "Metric MDS", ]
refined <- pilot$scores[pilot$scores$method == "MDS + edge-KK", ]
original <- pilot$scores[pilot$scores$method == "Original saddle", ]
stopifnot(identical(mds$replicate, refined$replicate))
s3_kable(data.frame(Cloud=mds$replicate, k=mds$k,
  "X-G"=100*mds$xg_error, "MDS path"=100*mds$path_rel,
  "Refined path"=100*refined$path_rel, "MDS Stress-1"=100*mds$stress1,
  "Refined Stress-1"=100*refined$stress1, check.names=FALSE),
  digits=3, booktabs=TRUE, caption="Selected-graph and layout errors for all five samples.",
  linesep="")
```

```{r edge-shape-results}
s3_kable(data.frame(Cloud=mds$replicate, "MDS edge"=100*mds$edge_rel,
  "Refined edge"=100*refined$edge_rel, "MDS shape"=100*mds$procrustes,
  "Refined shape"=100*refined$procrustes, check.names=FALSE),
  digits=3, booktabs=TRUE, caption="Edge relative RMSE and normalized Procrustes shape discrepancy (%).",
  linesep="")
```

The shape statistic centers both configurations and minimizes coordinate RMSE
over a common scale and orthogonal transformation, normalized by the
root-mean-square radius of the generating coordinates. It is separate from
the edge, path, and chord statistics. Its increase in this experiment does
not establish that shape recovery is impossible: the original coordinates
provide an exact zero-edge/path-loss realization.

The smooth boundary-value checks converged for all 640 examined pairs, with
evaluated curves inside the bounded patch. Relative RMSE between these
numerical controls and the primary mesh reference was 0.002935%--0.003141%
across samples. Refining the parameter grid from 81 to 161 points per axis
changed reference distances by 0.002167%--0.002417% on the 16-source subsets.
The further 128-source fine-grid check for sample 5 changed the reference by
0.002266% and retained $k=73$. A planar control agreed with Euclidean
distances to a maximum absolute error of $1.33\times10^{-15}$.
These checks assess numerical sensitivity; they are not certified error bounds.

```{r reference-results}
v <- pilot$validation
s3_kable(data.frame(Cloud=v$replicate,
  "Mesh vs smooth (%)"=100*v$bvp_relative_difference,
  "Primary vs fine (%)"=100*v$finer_relative_difference, check.names=FALSE),
  digits=6, booktabs=TRUE, caption="Independent smooth-geodesic and mesh-refinement checks.",
  linesep="")
```

## Connectivity and the identity-embedding control

At $k=3$, all five graphs needed component-MST repair. Their surface-to-graph
errors ranged from 62.2% to 94.7%, even though every Euclidean graph edge and
every retained graph path is preserved at the generating coordinates.
Figure \@ref(fig:s3-identity) uses the same sample and endpoints on a sparse
repaired graph and its calibrated graph.

```{r s3-identity, fig.width=10.2, fig.height=4.6, fig.cap="The same original coordinates preserve two different graphs exactly. Blue highlights a graph shortest path between the same endpoints; dashed brown marks MST bridges. Both configurations have zero fixed-path graph-to-embedding loss, despite their markedly different surface-to-graph errors. The displayed sample is the one selected for the manuscript by median minimum surface-to-graph error."}
plot_pilot_controls(pilot)
```

## Additional iterations and end-to-end distances

A separate continuation of each primary result used at most 1,000 additional
iterations at the final stiffness mixture. All five runs reached this limit
and further reduced the fixed-path error. They do not replace the primary
results. The table reports both achieved values.

```{r extra-results}
a <- pilot$audit[match(refined$replicate, pilot$audit$replicate), ]
s3_kable(data.frame(Cloud=refined$replicate,
  "Primary path"=100*refined$path_rel, "Additional path"=100*a$path_rel,
  "Primary edge"=100*refined$edge_rel, "Additional edge"=100*a$edge_rel,
  check.names=FALSE), digits=4, booktabs=TRUE,
  caption="Sensitivity to 1,000 additional edge-KK iterations; relative RMSE (%).", linesep="")
```

An end-to-end comparison returns each layout to input units using only its
fitted graph-edge scale. It then compares embedded fixed-path lengths directly
with the same surface reference, without fitting another scale against that
reference. Median relative RMSE was 0.211% for the input graph, 4.553% for
metric-MDS paths, and 0.253% for refined paths.

```{r s3-end-to-end, fig.width=8.5, fig.height=4.2, fig.cap="Surface-reference errors before and after embedding. Each line represents one sample. The input graph and both embedded configurations use the same reference pairs. The layout-to-input conversion uses only graph-edge information, not surface distances. This comparison supplements the two separate fidelity assessments."}
plot_pilot_end_to_end(pilot)
```

## Reproduction and supplied inputs

The supplementary archive includes the compact publication input
\path{reproducibility/precomputed/two-fidelity-saddle.rds}, containing the
representative coordinates and graphs, the calibration curves, all primary
and additional-budget scores, fit statuses, and validation summaries.
\path{reproducibility/scripts/plot-two-fidelity.R} draws the manuscript and
supplementary figures without rerunning the experiment.

The full protocol, pinned Python dependencies, sampling and graph-construction
scripts, reference solver, fitting code, and independent scoring checks are
in \path{reproducibility/experiments/two-fidelity-pilot/}.
Its README gives the regeneration order; \path{pilot-summary/} retains
per-sample numerical results, environment records, timing boundaries, and
input checksums. Larger intermediate distance matrices are regenerated rather
than distributed. After validation, \path{export-paper-data.R} reconstructs
the compact publication input.

The reference computation used Python 3.12, NumPy 2.5.2, SciPy 1.18.1, and
pygeodesic 0.1.11. The layout computation used grip 0.2.0, dgraphs 0.2.0,
igraph 2.3.3, Rcpp 1.1.2, and R development version 4.7.0
(2026-06-24 r90190). The environment records identify the particular runs;
regeneration in another numerical environment can differ slightly.

\clearpage

# Reference-coordinate and surface agreement

The following diagnostics evaluate the saved five-cloud configurations; they
do not refit them. For aligned coordinates $\widetilde Z$, coordinate RMSE is
$[n^{-1}\sum_i\|\widetilde z_i-x_i\|^2]^{1/2}$. Dividing by
$[n^{-1}\sum_i\|x_i-\bar x\|^2]^{1/2}$ gives the relative coordinate error.
Rigid alignment fits translation and rotation/reflection without rescaling;
similarity alignment also fits one common scale. The latter reproduces the
normalized Procrustes discrepancy already reported for this experiment.

To compare surfaces without imposing point correspondence, the Delaunay
triangulation of the original $(x,y)$ coordinates is carried unchanged to
each aligned configuration, forming a piecewise-planar surface $S_Z$.
Subdividing each original triangle twice and lifting its vertices to the
analytic saddle gives a reference $S_X$ over the same convex-hull footprint.
The symmetric surface RMS distance is
$$
\left[\frac12\left\{
\frac{1}{|S_Z|}\int_{S_Z}d(u,S_X)^2\,dA(u)+
\frac{1}{|S_X|}\int_{S_X}d(v,S_Z)^2\,dA(v)
\right\}\right]^{1/2},
$$
where $d(u,S)$ is the shortest Euclidean distance to the triangular surface
and $|S|$ its area. Each integral is estimated with 8,000 independent
area-uniform points, using exact point-to-triangle distances. Seeds are
`1901 + replicate`. This is not a geodesic loss or enclosed volume. It
depends on the surface mesh and its footprint, and does not certify the
absence of folds or self-intersections.

```{r reference-summary}
reference <- readRDS("reproducibility/precomputed/saddle-reference-diagnostics.rds")
rs <- subset(reference$scores,sample_size==8000)
summary <- aggregate(cbind(coordinate_relative_rmse,surface_rms)~method+alignment,rs,median)
summary$coordinate_relative_rmse <- 100*summary$coordinate_relative_rmse
summary$method <- c("metric-MDS + edge-KK","metric-MDS","Original saddle")[
  match(summary$method,c("MDS + edge-KK","Metric MDS","Original saddle"))]
s3_kable(summary,booktabs=TRUE,digits=3,row.names=FALSE,
 col.names=c("Configuration","Alignment","Coordinate error (%)","Surface RMS"),
 caption="Median reference discrepancies across the five original clouds. Coordinate error is normalized by reference RMS radius; surface RMS is in coordinate units.")
```

Refinement improves the surface-distance summary while worsening
corresponding-coordinate agreement. These measurements distinguish proximity
of surfaces from recovery of individual observations. The original-coordinate
surface score is nonzero because a finite triangulation replaces curved patches
by planar faces; this baseline must not be attributed to embedding error.
Long boundary triangles can contribute appreciably even for dense point clouds.

```{r reference-sampling,include=FALSE}
low <- subset(reference$scores,sample_size==2000)
key <- function(d) paste(d$replicate,d$method,d$alignment)
low <- low[match(key(rs),key(low)),]
sampling.change <- abs(rs$surface_rms-low$surface_rms)/rs$surface_rms
```

Increasing the sample count from 2,000 to 8,000 per direction changes surface
RMS by a median `r sprintf("%.2f",100*median(sampling.change))`% and at most
`r sprintf("%.2f",100*max(sampling.change))`% across the evaluated configurations
and alignments. This checks integration variability, not reference-mesh bias.

```{r refinement-overlay,fig.width=9,fig.height=3.5,fig.cap="Two views of the same refinement comparison for cloud 5. Left: triangulated metric-MDS (blue) and metric-MDS + edge-KK (orange), independently similarity-aligned to the original observations. Right: segments joining 50 systematically selected corresponding vertices. These are displacements between final aligned configurations, not optimization trajectories or a decomposition into tangential and normal movement. Interactive versions are provided in Supplementary Figure S4.3A--B."}
par(mfrow=c(1,2),mar=rep(0,4))
for(nm in c("overlay","displacement")) {
  im <- png::readPNG(paste0("reproducibility/figures/saddle/",nm,".png"))
  plot.new(); plot.window(xlim=c(0,4),ylim=c(0,3),asp=1,xaxs="i",yaxs="i")
  rasterImage(im,0,0,4,3)
}
```

## Separate neighborhood-size illustration

A separate cloud of 500 area-uniform observations (seed 1, the same saddle and
parameter domain) illustrates how these diagnostics vary over $k=3,\ldots,20$.
Each graph is fitted in three dimensions with metric-MDS and then five edge-KK
continuation stages with a maximum of 200 iterations each. This is not a sixth
pilot replicate: its sample size and neighborhood range differ. The curves describe within-cloud
sensitivity, not uncertainty across independent samples.

```{r single-reference,include=FALSE}
source("reproducibility/experiments/two-fidelity-pilot/plot-saddle-reference.R")
single <- read.csv("reproducibility/precomputed/single-saddle-reference.csv")
single.figure <- "build/supplement/single-saddle-reference.png"
plot.saddle.reference(single,single.figure,columns=2L)
```
```{r single-reference-figure,fig.cap="Reference recovery and graph fidelity in a separate 500-observation cloud. The same original parameter-plane triangulation is used for all configurations. Rigid and similarity alignments distinguish size-sensitive errors from shape comparisons; the nonzero original-surface baseline reflects triangulation. Neighborhood sizes are not independent replicates."}
knitr::include_graphics(normalizePath(single.figure),rel_path=FALSE)
```

The new scoring code uses grip development commit `b72f61d`, including
`score.coordinates()` and `score.surface()`; the saved pilot fits remain those
from grip 0.2.0. Supplement S4 uses ivue commit `872f9d4`. The scripts
`export-reference-diagnostics.R` and `run-saddle-reference.R` reproduce the
five-cloud extension and separate 500-point illustration, respectively.
Their CSV inputs and compact coordinates accompany the archive.

\clearpage

# Minimal executable circle example

A variable-density unit circle makes the two fidelity stages observable because
its intrinsic arc-distance matrix $D_X$ is known. The example draws 60
unevenly spaced points, constructs symmetric $k$-nearest-neighbor graphs with
the companion \CRANpkg{dgraphs} package, and compares $D_X$ with each graph
distance matrix $D_G$. For every graph, \CRANpkg{grip} then compares a weighted
GRIP layout and metric MDS using the same graph-to-embedding diagnostics.

```{r paired-fidelity-example, echo=TRUE}
set.seed(271)
n.circle <- 60L
circle.parameter <- sort(stats::runif(n.circle)^1.65)
circle.angle <- 2 * pi * circle.parameter
circle.data <- cbind(cos(circle.angle), sin(circle.angle))
circle.arc <- outer(circle.angle, circle.angle, function(a, b) abs(a - b))
D.circle <- pmin(circle.arc, 2 * pi - circle.arc)

paired.rows <- lapply(c(5L, 7L, 9L), function(k) {
  graph <- dgraphs::create.sknn.graph(
    circle.data,
    k = k,
    neighbor.method = "exact",
    connect.components = TRUE
  )
  stopifnot(graph$n_components_before == 1L, nrow(graph$mst_edge_matrix) == 0L)
  D.graph <- dgraphs::graph.geodesic.distances(graph)
  graph.loss <- dgraphs::isometry.rel.rms.error(D.graph, D.circle)
  prepared <- prepare.graph.geodesic.mds(
    adj_list = graph$adj_list,
    weight_list = graph$weight_list
  )

  layouts <- list(
    "Weighted GRIP" = grip(
      adj_list = graph$adj_list,
      weight_list = graph$weight_list,
      n = n.circle,
      dim = 2,
      metric = "edge_length",
      seed = 19
    ),
    "Metric MDS" = metric.mds(prepared = prepared, dim = 2)$coords
  )

  do.call(rbind, lapply(names(layouts), function(method) {
    score <- score.gmds(layouts[[method]], prepared = prepared)
    data.frame(
      k = k,
      `X-G Rel` = graph.loss,
      Embedding = method,
      `G-Z Path` = score$gmds.stress,
      `G-Z Chord` = score$metric.chord.stress,
      check.names = FALSE
    )
  }))
})
paired.fidelity <- do.call(rbind, paired.rows)
```

```{r circle-summary}
accessible_kable(
  paired.fidelity,
  digits = 3,
  row.names = FALSE,
  align = c("r", "r", "l", "r", "r"),
  group_after = c(2L, 4L),
  caption = paper_table_caption(
    paste(
      "Separate data-to-graph and graph-to-embedding losses for one",
      "controlled variable-density circle realization. All three graphs",
      "are connected without adding edges between components."
    ),
    paste(
      "X-G Rel = target-normalized data-to-graph relative RMSE;",
      "G-Z Path = fixed-path relative RMSE; G-Z Chord = chord relative RMSE."
    )
  ),
  alt = paste(
    "Six rows report separate data-to-graph and graph-to-embedding losses for",
    "three symmetric nearest-neighbor graphs, each embedded with weighted GRIP",
    "and metric MDS."
  ),
  position = "H"
)
```

This example illustrates the two fidelity assessments, rather than comparing
the performance of graph constructors or embedding methods. In this realization,
the $k=5$ graph is connected but fails
to close the sparsely sampled part of the circle, producing a large
data-to-graph error. Metric MDS represents that supplied graph with small
graph-to-embedding loss, but the low value cannot recover the circular relationship
already lost during graph construction. The $k=7$ and $k=9$ graphs approximate $D_X$ much more
closely, while the graph-to-embedding columns still distinguish how their
geometry is represented. The two losses are therefore retained as separate,
conditional diagnostics rather than added or multiplied. A substantive study
would repeat this calculation over a prespecified calibration family and
multiple realizations, as discussed in the main manuscript.


\clearpage

# Fixed weighted graph: graph-geodesic assessment

An edge-length metric matters when a graph's topology is simple but its geometry
is not. Here a regular $10 \times 10$ grid of $(x,y)$ values in $[-1,1]^2$ is
lifted to the saddle surface $z=0.8(x^2-y^2)$. These generating coordinates form
the target saddle in Figure \@ref(fig:weighted-figure). The graph joins
horizontal and vertical grid neighbors, assigning each of its 180 edges the
Euclidean distance between its 3D endpoints, divided by the median edge length.
Its topology is unchanged, but the resulting graph metric differs from that
of the flat grid. Distances in $D_G$ follow these mesh edges; they are not
continuous-surface geodesic distances.

The comparison includes combinatorial and weighted GRIP, metric MDS, and weighted
Kamada--Kawai from \CRANpkg{igraph}, together with edge-KK refinement of both the
metric-MDS and weighted-GRIP configurations. Six LGKK refinement rounds provide a
second refinement of the same weighted-GRIP coordinates. All layouts are
three-dimensional. Both edge-KK runs use the default density-stiffness
continuation schedule with at most 50 iterations per stage; stochastic calls
use seed 1. The graph and one selected shortest path for each of its 4,950
unordered vertex pairs are fixed before scoring. `score.gmds()` measures edge
and fixed-path relative RMSE, while `score.geodesic.kk()` supplies the weighted
path diagnostics. This comparison concerns graph-to-embedding fidelity, not
how well the graph represents an underlying data set.

The coordinates and scores used in this comparison are precomputed. The
supplementary archive provides these results, a $15 \times 15$ comparison under
the same settings, and the scripts used to generate both.

```{r weighted-summary}
weighted.saddle <- benchmark.results$weighted_saddle
weighted.resolutions <- benchmark.results$weighted_saddle_resolutions

weighted.score.cols <- c(
  "edge.rel.rmse",
  "gmds.stress",
  "gkk.weighted.rel.rmse",
  "gkk.mean.rel.path.error"
)

weighted.summary <- weighted.saddle$scores[, c("method", weighted.score.cols)]
# Keep stored method identifiers unchanged; use the shorter label in the table.
weighted.summary$method <- sub(" + LGKK polish", " + LGKK",
                               weighted.summary$method, fixed = TRUE)
# Keep small, nonzero errors visible rather than rounding them to exact zero.
weighted.summary[weighted.score.cols] <- lapply(
  weighted.summary[weighted.score.cols], formatC, format = "g", digits = 3
)

accessible_kable(
  weighted.summary,
  col.names = c("Method", "Edge Rel", "GMDS Rel", "GKK Rel", "Path Rel"),
  align = c("l", rep("r", 4)),
  caption = paper_table_caption(
    paste("Three-dimensional weighted saddle-surface layouts, scored on the",
          "same 180 edges and 4,950 fixed paths with separately profiled scales."),
    paste("Edge Rel and GMDS Rel = unweighted edge and fixed-path relative RMSE;",
          "GKK Rel = weighted relative RMSE; Path Rel = mean relative path error",
          "at the gKK scale.")
  ),
  alt = paste(
    "Seven rows compare combinatorial and weighted GRIP, metric MDS,",
    "edge-KK refinement of the latter two, LGKK-refined weighted GRIP,",
    "and weighted igraph Kamada-Kawai. Both edge-KK refinements reduce",
    "edge and fixed-path relative RMSE below one millionth."
  ),
  digits = 3
)
```

Both edge-KK refinements reduce edge and fixed-path relative RMSE below
$10^{-6}$ in this example. The unrefined metric-MDS and weighted-GRIP rows show
the change from each initialization; combinatorial GRIP remains as a
topology-only baseline. These are in-sample optimization results, with different
iteration budgets for edge-KK and LGKK, not a comparison of computational
efficiency or independent validation of the graph geometry. Under the same
refinement budget, the $15 \times 15$ comparison also reduces both errors, but
its residuals remain between $5\times10^{-6}$ and $2\times10^{-5}$; the near-zero
results should not be interpreted as a size-independent convergence guarantee.

\Needspace{5\baselineskip}

The target saddle is scored at its generating 3D coordinates, without an MDS
fit. Its Stress-1 value of 0.138 compares their endpoint chords with $D_G$,
using all 4,950 pairs and a least-squares fitted common scale. Its fixed-path
loss is zero after profiling out the edge-normalization scale, because those
coordinates realize every input edge length up to that scale. The nonzero
chord stress measures the mismatch between graph paths and endpoint distances
in this configuration. It is neither the smallest attainable stress over all
embeddings nor a measure of surface curvature alone: $D_G$ also reflects the
restriction to horizontal and vertical mesh edges.

Figure \@ref(fig:weighted-figure) also exposes a limitation of fixed-path fidelity.
Preserving the weighted graph paths does not uniquely determine the target's
extrinsic curvature: a comparatively flat realization can have low path loss.
The panels are similarity-aligned to the target and use one projection and one
coordinate scale, so this difference is not a viewing-axis artifact.

```{r weighted-figure, fig.cap="Three-dimensional layouts of a 10 by 10 weighted saddle mesh after similarity alignment to the generating coordinates on $z=0.8(x^2-y^2)$, displayed with a common orthographic projection and scale. The target panel shows those generating coordinates, not an MDS fit. In each row, the right-hand layout starts from the coordinates in the middle panel. The annotations report MDS Stress-1 and fixed-path GMDS relative RMSE from the full 3D coordinates, using all 4,950 vertex pairs, the same retained input paths, and separately fitted least-squares scales. Stress-1 is configuration-normalized; path relative RMSE is input-normalized, so the two magnitudes are not directly interchangeable. Values below $10^{-12}$ are shown as an upper bound. Low fixed-path loss need not recover the target's extrinsic curvature.", fig.alt="Six consistently projected panels show three-dimensional layouts of a 100-vertex saddle mesh, each annotated with MDS Stress-1 and fixed-path GMDS relative RMSE. The middle and right columns compare metric MDS and weighted GRIP before and after edge-KK. The left column contains the generating saddle coordinates and weighted igraph Kamada-Kawai. The target has near-zero path loss but nonzero chord stress; the refined weighted-GRIP layout also has near-zero path loss while remaining visibly flatter than the target.", fig.width=11.2, fig.height=7.0, fig.pos="H"}
source("reproducibility/scripts/plot-weighted-saddle.R")
plot_weighted_saddle(weighted.saddle, weighted.resolutions$display_limits)
```


\clearpage

# Minimal edge-KK refinement example

This minimal edge-KK example starts from a distorted
$4\times4$ mesh and applies eight refinement iterations. The code and figure
below use `reproducibility/scripts/edge-kk-workflow.R`.
The panels are centered and scaled only for display, consistently with the
profiled edge diagnostic; scoring uses the original coordinates.

```{r geodesic-workflow, echo=TRUE, results="hide"}
```

```{r edge-kk-workflow, fig.cap=sprintf("Illustrative output of the edge-KK workflow. Eight refinement iterations reduce profiled edge relative RMSE from %.3f to %.3f by equalizing the embedded mesh-edge lengths. The edge-only objective does not select a unique shape, so this deterministic example demonstrates the workflow rather than a general performance comparison.", before$edge.rel.rmse, after$edge.rel.rmse), fig.alt="Two panels show the 16-vertex 4-by-4 mesh before and after edge-KK refinement. The initial coordinates form a compressed, curved grid; refinement produces a more regular grid with nearly equal embedded edge lengths. Each panel reports profiled edge relative RMSE.", fig.width=7.5, fig.height=3.4, fig.pos="H"}
```

\clearpage

# Family-specific presets

The graph generators and presets support reproducible examples and tests for
unintended changes in results. Table \@ref(tab:preset-benchmark) compares the
default settings with family-specific presets on two controlled graph families,
each run across ten seeds.

```{r preset-benchmark}
mesh.edges <- edges.mesh(12, 12)
mesh.compare <- compare.layouts(
  edges = mesh.edges,
  n = 144,
  dim = 2,
  candidates = c("default", "mesh"),
  seeds = 1:10,
  edge.crossings = "always"
)

carpet.edges <- edges.sierpinski.carpet(3)
carpet.n <- max(carpet.edges)
carpet.compare <- compare.layouts(
  edges = carpet.edges,
  n = carpet.n,
  dim = 2,
  candidates = c("default", "carpet"),
  seeds = 1:10,
  edge.crossings = "never"
)

summary.cols <- c(
  "candidate", "n.runs", "elapsed.sec.mean", "sampled.stress.mean",
  "edge.length.cv.mean", "stability.procrustes.mean"
)

preset.benchmark <- rbind(
  transform(mesh.compare$summary[, summary.cols], graph = "12x12 mesh"),
  transform(
    carpet.compare$summary[, summary.cols],
    graph = "Sierpinski carpet, level 3"
  )
)

preset.benchmark <- preset.benchmark[, c(
  "graph", "candidate", "n.runs", "elapsed.sec.mean",
  "sampled.stress.mean", "edge.length.cv.mean",
  "stability.procrustes.mean"
)]
preset.benchmark$candidate <- factor(
  preset.benchmark$candidate,
  levels = c("default", "mesh", "carpet")
)
preset.benchmark <- preset.benchmark[
  order(preset.benchmark$graph, preset.benchmark$candidate),
]

accessible_kable(
  preset.benchmark,
  col.names = c("Graph", "Cand", "Runs", "Sec", "Stress", "EdgeCV", "PStab"),
  align = c("l", "l", rep("r", 5)),
  caption = paper_table_caption(
    "Preset benchmark on two structured graph families, each averaged over ten seeds.",
    "Runs = number of seeds; Sec = mean elapsed seconds; Stress = sampled stress; EdgeCV = edge-length coefficient of variation; PStab = Procrustes stability."
  ),
  alt = paste(
    "Four rows compare default and family-specific presets on a mesh and",
    "Sierpinski carpet using runtime, stress, edge variation, and stability",
    "metrics."
  ),
  digits = 3
)
```

For these controlled families, the package-supplied presets reduce sampled
stress and edge-length variation and produce more reproducible layouts across
seeds. This is family-specific evidence for the supplied defaults, not a claim
that these settings are preferable for arbitrary data-derived graphs. Figure
\@ref(fig:carpet-figure) shows the corresponding carpet layouts.

```{r carpet-figure, fig.cap="Level-3 Sierpinski carpet laid out with the default settings (left) and the carpet preset (right). The preset yields the lower-stress and more stable solution reported in the preceding benchmark table.", fig.alt="Two panels showing a Sierpinski carpet graph. The default layout is more distorted, while the preset layout better preserves the recursive holes.", fig.width=10, fig.height=4, fig.pos="H"}
carpet.default.coords <- grip(carpet.edges, n = carpet.n, dim = 2, seed = 12)
carpet.preset.coords <- grip(
  carpet.edges, n = carpet.n, dim = 2, preset = "carpet", seed = 12
)

op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
plot.layout(
  carpet.default.coords, carpet.edges,
  vertex.col = "black", edge.col = "gray82", pch = 16, cex = 0.3,
  main = "Default", axes = FALSE, xlab = "", ylab = ""
)
plot.layout(
  carpet.preset.coords, carpet.edges,
  vertex.col = "black", edge.col = "gray82", pch = 16, cex = 0.3,
  main = "Carpet preset", axes = FALSE, xlab = "", ylab = ""
)
par(op)
```

\clearpage

# Method families and principal functions

Table \@ref(tab:family-scope) lists the multiscale layout and GMDS method
families, their purposes, and their principal functions.

```{r family-scope, results="asis"}
family.scope <- data.frame(
  Family = c(
    "Multiscale GRIP",
    "Full gKK",
    "LGKK",
    "MISF gKK",
    "Edge KK",
    "Repulsive terms"
  ),
  Purpose = c(
    "Layouts and intermediate stages.",
    "All-pairs scoring and refinement.",
    "Landmark scoring and refinement.",
    "Multiscale scoring and refinement.",
    "Edge scoring and refinement.",
    "Dispersion and edge regularization."
  ),
  `Principal functions` = c(
    "grip(), trace.grip()",
    "prepare.geodesic.kk(), score.geodesic.kk(), geodesic.kk()",
    "prepare.landmark.geodesic.kk(), score.landmark.geodesic.kk(), landmark.geodesic.kk()",
    "prepare.misf.geodesic.kk(), score.misf.geodesic.kk(), misf.geodesic.kk()",
    "prepare.edge.kk(), edge.kk()",
    "repulsive.state(), repulsive.stage(), edge.repulsive.state(), edge.repulsive.stage()"
  ),
  check.names = FALSE
)
family.scope[["Principal functions"]] <- vapply(
  strsplit(family.scope[["Principal functions"]], ", ", fixed = TRUE),
  function(x) paste(paste0("\\texttt{", x, "}"), collapse = "\\newline "),
  character(1)
)
family.table <- knitr::kable(
  family.scope, format = "latex", booktabs = TRUE, escape = FALSE,
  row.names = FALSE, align = c("l", "l", "l"), position = "H",
  caption = "Method families, purposes, and principal functions in grip. Multiscale GRIP provides the standard layout methods; the remaining families belong to the experimental GMDS functionality."
)
# Fixed-width columns keep each function name intact and wrap the prose.
stopifnot(grepl("\\begin{tabular}[t]{lll}", family.table, fixed = TRUE))
cat(sub(
  "\\begin{tabular}[t]{lll}",
  paste0("\\begin{tabular}[t]{@{}",
    ">{\\raggedright\\arraybackslash}p{0.17\\linewidth}",
    ">{\\raggedright\\arraybackslash}p{0.27\\linewidth}",
    ">{\\raggedright\\arraybackslash}p{0.50\\linewidth}@{}}"),
  family.table, fixed = TRUE
))
```

# References {-}
