---
title: "Supplement S4: Interactive saddle configurations"
author: "Pawel Gajer and Jacques Ravel"
output:
  html_document:
    self_contained: true
    theme: null
    mathjax: null
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE)
source("reproducibility/scripts/saddle-widgets.R")
pilot <- readRDS("reproducibility/precomputed/two-fidelity-saddle.rds")
reference <- readRDS("reproducibility/precomputed/saddle-reference-diagnostics.rds")
views <- saddle_widgets(pilot,reference)
dir.create("build/saddle-widgets",recursive=TRUE,showWarnings=FALSE)
for(nm in names(views)) htmlwidgets::saveWidget(views[[nm]],
  paste0("build/saddle-widgets/",nm,".html"),selfcontained=TRUE)
```

<style>
body {font:17px/1.5 system-ui,sans-serif;max-width:1100px;margin:32px auto;padding:0 20px;color:#25313b}
h1,h2 {line-height:1.2} .rglWebGL {max-width:100%} a {color:#2166ac}
h2,h3 {scroll-margin-top:20px}
</style>

Cloud 5 of the five-cloud pilot: **1,000 observations, k = 73**.
These are the saved configurations used in Figure 7, not new fits.
Drag a scene to rotate it; use the scroll wheel to zoom. All scenes start with
the same orthographic camera and bounding box. Reload to restore the initial views.

[Figure S4.1: triangulated views](#fig-s4-1) ·
[Figure S4.2: reference overlays](#fig-s4-2) ·
[Figure S4.3: refinement comparison](#fig-s4-3)

<span id="triangulated-views"></span>

## Figure S4.1. Triangulated configurations {#fig-s4-1}

Panels [A](#fig-s4-1a), [B](#fig-s4-1b), and [C](#fig-s4-1c) correspond to
main-text Figure 7C, D, and E, respectively.

The Delaunay triangles are constructed once in the original (x,y) parameter
plane and carried to each configuration with vertex correspondence unchanged.
They are a visualization mesh, **not the calibrated skNN graph**. Colors encode
original x, from blue (−1) to brown (+1). Blue lines mark the same retained graph
route and brown lines its endpoint chord. Each configuration is independently
similarity-aligned to the observations (translation, rotation/reflection, and
one uniform scale); this display alignment does not change the graph scores.

### Figure S4.1A. Triangulated original saddle {#fig-s4-1a}
```{r}
views$mesh1
```
### Figure S4.1B. Triangulated metric-MDS configuration {#fig-s4-1b}
```{r}
views$mesh2
```
### Figure S4.1C. Triangulated metric-MDS + edge-KK configuration {#fig-s4-1c}
```{r}
views$mesh3
```

<span id="reference-overlays"></span>

## Figure S4.2. Generating-surface overlays {#fig-s4-2}

Panels [A](#fig-s4-2a), [B](#fig-s4-2b), and [C](#fig-s4-2c) show the original
observations, metric-MDS, and metric-MDS + edge-KK against the same reference.

The translucent blue reference is z = 0.8(x² − y²), over [−1,1]².
It remains fixed while the observed or embedded points change. Surface scoring
instead uses the matched sampled convex hull, as explained in Supplement S3;
these full-square display overlays do not define the numerical score.

### Figure S4.2A. Original observations and generating surface {#fig-s4-2a}
```{r}
views$reference1
```
### Figure S4.2B. metric-MDS and generating surface {#fig-s4-2b}
```{r}
views$reference2
```
### Figure S4.2C. metric-MDS + edge-KK and generating surface {#fig-s4-2c}
```{r}
views$reference3
```

<span id="refinement-comparison"></span>

## Figure S4.3. Refinement comparison {#fig-s4-3}

Panel [A](#fig-s4-3a) overlays the two meshes; panel [B](#fig-s4-3b) shows
corresponding-vertex displacements. These are interactive versions of the
two views in Figure S3.3. Panel [C](#fig-s4-3c) adds the generating saddle
to the two-mesh comparison in panel A.

Blue: metric-MDS. Orange: metric-MDS + edge-KK. Both use the same triangles and
are independently similarity-aligned to the reference observations. Overlapping
meshes can obscure differences; resemblance alone does not establish tangential
movement or recovery of the generating surface.

### Figure S4.3A. metric-MDS and metric-MDS + edge-KK mesh overlay {#fig-s4-3a}

```{r}
views$overlay
```

### Figure S4.3B. Corresponding-vertex displacements {#fig-s4-3b}

This view joins corresponding vertices for a systematic subset of 50 of
the 1,000 observations. Segments show displacement between the two aligned
configurations, not optimizer trajectories or surface-normal displacements.

```{r}
views$displacement
```

### Figure S4.3C. Both fitted configurations and the generating saddle {#fig-s4-3c}

The light-gray translucent surface is the analytic saddle
z = 0.8(x² − y²), over [−1,1]², not another fitted mesh. The blue metric-MDS
and orange metric-MDS + edge-KK configurations retain the independent
similarity alignments, triangulation, colors, and initial camera of panel A.
Only the fixed reference surface has been added. As in Figure S4.2, its
full-square display footprint differs from the matched convex-hull footprint
used for surface scoring.

```{r}
views$overlay_reference
```

## Reproducibility

Fits: grip 0.2.0. Additional reference diagnostics: grip development commit
`b72f61d`. Visualization: ivue commit `872f9d4`. Data, source scripts, and the
reference-score definitions accompany Supplements S3 and S4. This self-contained
HTML needs a WebGL-capable browser but no network connection.
