---
title: "Exploring Default-Only Benchmark Candidates for grip vs Alternatives"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  fig.width = 12,
  fig.height = 7.5
)
```

## Why this review document exists

This is a review-only exploration document for revising the public article
`Comparing grip with Alternative Layout Engines`.

The main goals are:

- use default settings only for all external layout engines and for `grip`,
- look at a broader synthetic-family panel than karate plus mesh,
- and identify two or three strong candidate benchmarks for the final vignette.

This document intentionally does **not** use presets.

```{r}
library(grip)
library(igraph)
library(graphlayouts)
```

## Shared helpers

```{r}
as_coord_matrix <- function(layout_result) {
  matrix(as.numeric(layout_result), ncol = ncol(layout_result))
}

score_layout <- function(coords, edges, n,
                         edge.crossings = "auto",
                         sample.size.stress = 2000L,
                         sample.size.nonedge = 5000L) {
  grip.score.layout(
    coords = coords,
    edges = edges,
    n = n,
    sample.size.stress = sample.size.stress,
    sample.size.nonedge = sample.size.nonedge,
    stress.seed = 42L,
    nonedge.seed = 42L,
    edge.crossings = edge.crossings
  )
}

plot_layout <- function(coords,
                        edges,
                        title,
                        vertex.col = "#1F3B73",
                        edge.col = "gray80",
                        cex = 0.5) {
  plot(
    coords[, 1], coords[, 2],
    asp = 1,
    pch = 16,
    cex = cex,
    col = vertex.col,
    xlab = "",
    ylab = "",
    axes = FALSE,
    main = title
  )
  if (!is.null(edges) && nrow(edges) > 0) {
    segments(
      coords[edges[, 1], 1], coords[edges[, 1], 2],
      coords[edges[, 2], 1], coords[edges[, 2], 2],
      col = edge.col
    )
  }
}

review_rank <- function(score_table) {
  rank_sum <-
    rank(score_table$sampled.stress, ties.method = "min") +
    rank(score_table$edge.length.cv, ties.method = "min") +
    rank(score_table$edge.crossings, ties.method = "min") +
    rank(-score_table$sampled.nonedge.sep.ratio, ties.method = "min")
  rank(rank_sum, ties.method = "min")
}

run_case <- function(case) {
  built <- case$build()
  edges <- built$edges
  n <- built$n
  ig <- igraph::graph_from_edgelist(edges, directed = FALSE)

  set.seed(1); coords.fr <- as_coord_matrix(layout_with_fr(ig))
  set.seed(1); coords.kk <- as_coord_matrix(layout_with_kk(ig))
  set.seed(1); coords.drl <- as_coord_matrix(layout_with_drl(ig))
  set.seed(1); coords.stress <- as_coord_matrix(layout_with_stress(ig))
  coords.grip <- grip.layout(edges, n = n, dim = 2, seed = 1)

  crossing.mode <- if (n <= 300L) "always" else "never"

  scores <- rbind(
    cbind(
      method = "FR (igraph)",
      score_layout(coords.fr, edges, n, edge.crossings = crossing.mode)
    ),
    cbind(
      method = "KK (igraph)",
      score_layout(coords.kk, edges, n, edge.crossings = crossing.mode)
    ),
    cbind(
      method = "DrL (igraph)",
      score_layout(coords.drl, edges, n, edge.crossings = crossing.mode)
    ),
    cbind(
      method = "Stress (graphlayouts)",
      score_layout(coords.stress, edges, n, edge.crossings = crossing.mode)
    ),
    cbind(
      method = "grip default",
      score_layout(coords.grip, edges, n, edge.crossings = crossing.mode)
    )
  )

  scores <- as.data.frame(scores, stringsAsFactors = FALSE)
  num.cols <- setdiff(names(scores), "method")
  for (nm in num.cols) {
    scores[[nm]] <- as.numeric(scores[[nm]])
  }
  scores$review.rank <- review_rank(scores)

  list(
    meta = data.frame(
      case_id = case$id,
      label = case$label,
      family = case$family,
      why = case$why,
      n = n,
      m = nrow(edges),
      stringsAsFactors = FALSE
    ),
    edges = edges,
    layouts = list(
      fr = coords.fr,
      kk = coords.kk,
      drl = coords.drl,
      stress = coords.stress,
      grip.default = coords.grip
    ),
    scores = scores
  )
}

plot_case_panel <- function(case.result) {
  old.par <- par(no.readonly = TRUE)
  on.exit(par(old.par), add = TRUE)
  par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")
  plot_layout(case.result$layouts$fr, case.result$edges, "FR (igraph)")
  plot_layout(case.result$layouts$kk, case.result$edges, "KK (igraph)")
  plot_layout(case.result$layouts$drl, case.result$edges, "DrL (igraph)")
  plot_layout(case.result$layouts$stress, case.result$edges, "Stress (graphlayouts)")
  plot_layout(case.result$layouts$grip.default, case.result$edges, "grip default")

  plot.new()
  title("Case note", line = 0.5)
  lines <- c(
    sprintf("Family: %s", case.result$meta$family[[1L]]),
    sprintf("n = %d, m = %d", case.result$meta$n[[1L]], case.result$meta$m[[1L]]),
    "",
    strwrap(case.result$meta$why[[1L]], width = 28)
  )
  text(
    x = 0.03,
    y = 0.95,
    labels = paste(lines, collapse = "\n"),
    adj = c(0, 1),
    cex = 0.95
  )
}
```

## Candidate case panel

The panel below is intentionally broad. It spans trees, wrapped lattices,
fractals, triangulated manifolds, irregular manifolds, and a porous 3D-derived
graph, all viewed through a 2D default-layout comparison.

```{r}
case.catalog <- list(
  list(
    id = "kary_tree",
    label = "3-ary tree (depth 4)",
    family = "tree",
    why = "Acyclic branching benchmark. Useful for seeing whether default layouts keep hierarchy legible without tuning.",
    build = function() {
      edges <- edges.kary.tree(k = 3, depth = 4)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "cylinder",
    label = "Cylinder 8 x 16",
    family = "wrapped lattice",
    why = "A mesh-like graph with one periodic direction and boundary effects in the other.",
    build = function() {
      edges <- edges.cylinder(h = 8, w = 16)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "torus",
    label = "Torus 8 x 16",
    family = "wrapped lattice",
    why = "Doubly periodic surface graph. A strong candidate if we want a non-boundary surface instead of a plain mesh.",
    build = function() {
      edges <- edges.torus(h = 8, w = 16)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "sphere",
    label = "Sphere 8 x 16",
    family = "surface graph",
    why = "Spherical adjacency introduces poles and non-grid local structure while staying quite interpretable.",
    build = function() {
      edges <- edges.sphere(h = 8, w = 16)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "sierpinski_triangle",
    label = "Sierpinski triangle (level 4)",
    family = "recursive fractal",
    why = "Sparse self-similar triangular benchmark with strong hierarchy and obvious holes.",
    build = function() {
      edges <- edges.sierpinski.triangle(level = 4)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "sierpinski_carpet",
    label = "Sierpinski carpet (level 2)",
    family = "recursive fractal",
    why = "Square fractal with holes and bottlenecks; much more interesting than a plain grid but still easy to read.",
    build = function() {
      edges <- edges.sierpinski.carpet(level = 2)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "vicsek",
    label = "Vicsek cross fractal (level 3)",
    family = "recursive fractal",
    why = "Cross-shaped recursive benchmark with strong symmetry and narrow connectors.",
    build = function() {
      edges <- edges.vicsek(level = 3)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "triangulated_annulus",
    label = "Triangulated annulus (resolution 8)",
    family = "triangulated manifold",
    why = "Boundary-rich triangulated surface without rectangular-grid bias.",
    build = function() {
      edges <- edges.triangulated.annulus(resolution = 8)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "irregular_annulus",
    label = "Irregular annulus",
    family = "irregular manifold",
    why = "Point-sampled annulus with irregular valence and no lattice regularity.",
    build = function() {
      edges <- edges.irregular.annulus(rings = 6, outer_count = 24)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "irregular_torus",
    label = "Irregular torus",
    family = "irregular manifold",
    why = "Point-sampled torus with irregular counts per ring, useful if we want a less lattice-biased torus benchmark.",
    build = function() {
      edges <- edges.irregular.torus(major_rings = 8, tube_count = 16)
      list(edges = edges, n = max(edges))
    }
  ),
  list(
    id = "porous_cube",
    label = "Porous cube channel network",
    family = "porous 3D-derived graph",
    why = "A non-surface topology coming from a cubical channel network. Interesting if we want a structurally rich graph that is not a classical surface.",
    build = function() {
      edges <- edges.cube.channel.network(level = 1, side = 5)
      list(edges = edges, n = max(edges))
    }
  )
)

results <- lapply(case.catalog, run_case)
names(results) <- vapply(case.catalog, `[[`, character(1), "id")

case.overview <- do.call(
  rbind,
  lapply(results, function(x) {
    grip.row <- x$scores[x$scores$method == "grip default", , drop = FALSE]
    best.row <- x$scores[which.min(x$scores$review.rank), , drop = FALSE]
    cbind(
      x$meta,
      grip_default_review_rank = grip.row$review.rank,
      grip_default_crossings = grip.row$edge.crossings,
      grip_default_stress = grip.row$sampled.stress,
      best_method = best.row$method
    )
  })
)

knitr::kable(case.overview, digits = 3)
```

## Suggested shortlist for manual review

This table is not meant to replace visual judgment. It is just a first-pass
screen showing where the default `grip` layout is structurally competitive under
the simple review rank used here.

```{r}
shortlist <- case.overview[
  order(case.overview$grip_default_review_rank,
        case.overview$grip_default_crossings,
        case.overview$grip_default_stress),
  c(
    "label",
    "family",
    "grip_default_review_rank",
    "grip_default_crossings",
    "grip_default_stress",
    "best_method"
  )
]

knitr::kable(shortlist, digits = 3)
```

## Case-by-case review

```{r candidate-review, results='asis', fig.width=12, fig.height=7.5}
for (nm in names(results)) {
  case.result <- results[[nm]]
  cat("\n\n### ", case.result$meta$label[[1L]], "\n\n", sep = "")
  cat(
    case.result$meta$why[[1L]],
    "\n\n",
    sep = ""
  )
  print(
    knitr::kable(
      case.result$scores[
        ,
        c(
          "method",
          "sampled.stress",
          "edge.length.cv",
          "edge.crossings",
          "sampled.nonedge.sep.ratio",
          "review.rank"
        )
      ],
      digits = 3,
      row.names = FALSE
    )
  )
  plot_case_panel(case.result)
}
```

## Takeaway for the public vignette

This exploration document is intended to help us decide which two or three
synthetic examples belong in the public comparison article.

The editorial rules for that next revision should be:

- keep the public comparison track default-only,
- use synthetic examples where `grip` default is at least credible,
- avoid benchmarks whose main lesson is just that a preset exists,
- and keep the weighted geometry-aware story separate from the external-package
  default comparison.

## Session info

```{r}
sessionInfo()
```
