---
title: "Comparing grip with igraph and graphlayouts"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  fig.width = 10,
  fig.height = 4.5
)
```

## Introduction

This article provides a reproducible, side-by-side comparison of graph layouts
from three R packages --- **grip**, **igraph**, and **graphlayouts** --- on the
same benchmark graphs, evaluated using the same quality metrics. The goal is not
to declare a winner but to show where each tool does well and to illustrate how
`grip.score.layout()` can score *any* layout, regardless of which package
produced it. grip is complementary to igraph and graphlayouts, not a
replacement.

## Setup

```{r libraries, message=FALSE}
library(grip)
library(igraph)
library(graphlayouts)
```

We define a helper that takes a coordinate matrix (from any source) and an edge
list, then calls `grip.score.layout()` to return a standardized metrics row.
This keeps the comparison code DRY and ensures every layout is scored
identically.

```{r score-helper}
score_layout <- function(coords, edges, n,
                         clusters = NULL,
                         edge.crossings = "auto",
                         sample.size.stress = 2000L,
                         sample.size.nonedge = 5000L) {
  grip.score.layout(
    coords            = coords,
    edges             = edges,
    n                 = n,
    clusters          = clusters,
    sample.size.stress  = sample.size.stress,
    sample.size.nonedge = sample.size.nonedge,
    stress.seed       = 42L,
    nonedge.seed      = 42L,
    edge.crossings    = edge.crossings
  )
}
```

A second helper converts an igraph layout (which is already an n×2 or n×3
matrix) into a plain numeric matrix, since some igraph methods return objects
with extra attributes.

```{r normalize-helper}
as_coord_matrix <- function(layout_result) {
  matrix(as.numeric(layout_result), ncol = ncol(layout_result))
}
```

Finally, a plotting helper for consistent base-R panels across all methods.

```{r plot-helper}
plot_layout <- function(coords, edges, title,
                        vertex.col = "#1F3B73",
                        edge.col = "gray80",
                        cex = 0.5) {
  plot(coords[, 1], coords[, 2],
       asp = 1, pch = 16, cex = cex,
       col = vertex.col, xlab = "", ylab = "",
       axes = FALSE, main = title)
  if (!is.null(edges) && nrow(edges) > 0) {
    segments(coords[edges[, 1], 1], coords[edges[, 1], 2],
             coords[edges[, 2], 1], coords[edges[, 2], 2],
             col = edge.col)
  }
}
```

## Benchmark 1: Zachary karate club (34 vertices, 78 edges)

The Zachary karate club is a small, well-studied social network with known
faction membership. It is small enough that all methods should perform well.

```{r karate-setup}
karate.edges <- as.matrix(utils::read.csv(
  system.file("extdata", "karate-club-edges.csv", package = "grip"),
  stringsAsFactors = FALSE
))
karate.n <- max(karate.edges)
karate.m <- nrow(karate.edges)

karate.labels <- utils::read.csv(
  system.file("extdata", "karate-club-membership.csv", package = "grip"),
  stringsAsFactors = FALSE
)
karate.labels <- karate.labels[order(karate.labels$vertex), ]
karate.clubs  <- karate.labels$club

cat(sprintf("Vertices: %d, Edges: %d\n", karate.n, karate.m))
```

Build an igraph object for the alternative layout functions.

```{r karate-igraph}
karate.ig <- graph_from_edgelist(karate.edges, directed = FALSE)
```

### Compute layouts

```{r karate-layouts}
set.seed(1)
karate.fr  <- as_coord_matrix(layout_with_fr(karate.ig))
set.seed(1)
karate.kk  <- as_coord_matrix(layout_with_kk(karate.ig))
set.seed(1)
karate.drl <- as_coord_matrix(layout_with_drl(karate.ig))
set.seed(1)
karate.stress <- as_coord_matrix(layout_with_stress(karate.ig))

karate.grip.default <- grip.layout(karate.edges, n = karate.n, dim = 2, seed = 1)
karate.grip.tree    <- grip.layout(karate.edges, n = karate.n, dim = 2,
                                    preset = "tree", seed = 1)
```

### Score all layouts

```{r karate-score}
karate.results <- rbind(
  cbind(method = "FR (igraph)",
        score_layout(karate.fr, karate.edges, karate.n,
                     clusters = karate.clubs, edge.crossings = "always")),
  cbind(method = "KK (igraph)",
        score_layout(karate.kk, karate.edges, karate.n,
                     clusters = karate.clubs, edge.crossings = "always")),
  cbind(method = "DrL (igraph)",
        score_layout(karate.drl, karate.edges, karate.n,
                     clusters = karate.clubs, edge.crossings = "always")),
  cbind(method = "Stress (graphlayouts)",
        score_layout(karate.stress, karate.edges, karate.n,
                     clusters = karate.clubs, edge.crossings = "always")),
  cbind(method = "grip default",
        score_layout(karate.grip.default, karate.edges, karate.n,
                     clusters = karate.clubs, edge.crossings = "always")),
  cbind(method = "grip tree preset",
        score_layout(karate.grip.tree, karate.edges, karate.n,
                     clusters = karate.clubs, edge.crossings = "always"))
)

karate.display <- karate.results[, c("method", "sampled.stress",
                                      "edge.length.cv", "edge.crossings",
                                      "sampled.nonedge.sep.ratio",
                                      "cluster.separation")]
knitr::kable(karate.display, digits = 4, row.names = FALSE,
             caption = "Karate club: quality metrics (lower stress/CV/crossings is better; higher separation is better)")
```

### Visual comparison

```{r karate-plot, fig.width=12, fig.height=7.5}
par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")

plot_layout(karate.fr,           karate.edges, "FR (igraph)",             cex = 0.6)
plot_layout(karate.kk,           karate.edges, "KK (igraph)",             cex = 0.6)
plot_layout(karate.drl,          karate.edges, "DrL (igraph)",            cex = 0.6)
plot_layout(karate.stress,       karate.edges, "Stress (graphlayouts)",   cex = 0.6)
plot_layout(karate.grip.default, karate.edges, "grip default",            cex = 0.6)
plot_layout(karate.grip.tree,    karate.edges, "grip tree preset",        cex = 0.6)
```

On a graph this small, all methods produce readable layouts. Stress
minimization (from graphlayouts) and Kamada-Kawai typically achieve low stress
values, which is expected given that they directly optimize a stress objective.
grip's metrics help quantify differences that are difficult to see by eye.

## Benchmark 2: 12×12 rectangular mesh (144 vertices, 264 edges)

A regular lattice is a structural benchmark: the ideal layout recovers the grid
exactly, with uniform edge lengths and zero crossings.

```{r mesh-setup}
mesh.edges <- edges.mesh(12, 12)
mesh.n <- 12 * 12
mesh.m <- nrow(mesh.edges)

cat(sprintf("Vertices: %d, Edges: %d\n", mesh.n, mesh.m))
```

```{r mesh-igraph}
mesh.ig <- graph_from_edgelist(mesh.edges, directed = FALSE)
```

### Compute layouts

```{r mesh-layouts}
set.seed(1)
mesh.fr  <- as_coord_matrix(layout_with_fr(mesh.ig))
set.seed(1)
mesh.kk  <- as_coord_matrix(layout_with_kk(mesh.ig))
set.seed(1)
mesh.drl <- as_coord_matrix(layout_with_drl(mesh.ig))
set.seed(1)
mesh.stress <- as_coord_matrix(layout_with_stress(mesh.ig))

mesh.grip.default <- grip.layout(mesh.edges, n = mesh.n, dim = 2, seed = 1)
mesh.grip.mesh    <- grip.layout(mesh.edges, n = mesh.n, dim = 2,
                                  preset = "mesh", seed = 1)
```

### Score all layouts

```{r mesh-score}
mesh.results <- rbind(
  cbind(method = "FR (igraph)",
        score_layout(mesh.fr, mesh.edges, mesh.n, edge.crossings = "always")),
  cbind(method = "KK (igraph)",
        score_layout(mesh.kk, mesh.edges, mesh.n, edge.crossings = "always")),
  cbind(method = "DrL (igraph)",
        score_layout(mesh.drl, mesh.edges, mesh.n, edge.crossings = "always")),
  cbind(method = "Stress (graphlayouts)",
        score_layout(mesh.stress, mesh.edges, mesh.n, edge.crossings = "always")),
  cbind(method = "grip default",
        score_layout(mesh.grip.default, mesh.edges, mesh.n, edge.crossings = "always")),
  cbind(method = "grip mesh preset",
        score_layout(mesh.grip.mesh, mesh.edges, mesh.n, edge.crossings = "always"))
)

mesh.display <- mesh.results[, c("method", "sampled.stress",
                                  "edge.length.cv", "edge.crossings",
                                  "sampled.nonedge.sep.ratio")]
knitr::kable(mesh.display, digits = 4, row.names = FALSE,
             caption = "12x12 mesh: quality metrics")
```

### Visual comparison

```{r mesh-plot, fig.width=12, fig.height=7.5}
par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")

plot_layout(mesh.fr,           mesh.edges, "FR (igraph)",           cex = 0.5)
plot_layout(mesh.kk,           mesh.edges, "KK (igraph)",           cex = 0.5)
plot_layout(mesh.drl,          mesh.edges, "DrL (igraph)",          cex = 0.5)
plot_layout(mesh.stress,       mesh.edges, "Stress (graphlayouts)", cex = 0.5)
plot_layout(mesh.grip.default, mesh.edges, "grip default",          cex = 0.5)
plot_layout(mesh.grip.mesh,    mesh.edges, "grip mesh preset",      cex = 0.5)
```

On lattice graphs, stress minimization and Kamada-Kawai both tend to recover
near-perfect grids. grip's mesh preset is specifically tuned for this family and
should produce very uniform edge lengths (low CV). Fruchterman-Reingold and DrL
often introduce slight barrel distortion on regular meshes.

## Benchmark 3: Level-4 Sierpinski carpet (~4,000+ vertices)

This fractal graph is large enough that algorithm scalability starts to matter.
Edge-crossing computation is skipped (it would be extremely slow at this size).

```{r carpet-setup}
carpet.edges <- edges.sierpinski.carpet(4)
carpet.n <- max(carpet.edges)
carpet.m <- nrow(carpet.edges)

cat(sprintf("Vertices: %d, Edges: %d\n", carpet.n, carpet.m))
```

```{r carpet-igraph}
carpet.ig <- graph_from_edgelist(carpet.edges, directed = FALSE)
```

### Compute layouts with timing

```{r carpet-layouts}
set.seed(1)
carpet.time.fr <- system.time(
  carpet.fr <- as_coord_matrix(layout_with_fr(carpet.ig))
)
set.seed(1)
carpet.time.kk <- system.time(
  carpet.kk <- as_coord_matrix(layout_with_kk(carpet.ig))
)
set.seed(1)
carpet.time.drl <- system.time(
  carpet.drl <- as_coord_matrix(layout_with_drl(carpet.ig))
)
set.seed(1)
carpet.time.stress <- system.time(
  carpet.stress <- as_coord_matrix(layout_with_stress(carpet.ig))
)

carpet.time.grip.default <- system.time(
  carpet.grip.default <- grip.layout(carpet.edges, n = carpet.n, dim = 2, seed = 1)
)
carpet.time.grip.carpet <- system.time(
  carpet.grip.carpet <- grip.layout(carpet.edges, n = carpet.n, dim = 2,
                                     preset = "carpet", seed = 1)
)
```

### Score all layouts

```{r carpet-score}
carpet.results <- rbind(
  cbind(method = "FR (igraph)",
        score_layout(carpet.fr, carpet.edges, carpet.n,
                     edge.crossings = "never")),
  cbind(method = "KK (igraph)",
        score_layout(carpet.kk, carpet.edges, carpet.n,
                     edge.crossings = "never")),
  cbind(method = "DrL (igraph)",
        score_layout(carpet.drl, carpet.edges, carpet.n,
                     edge.crossings = "never")),
  cbind(method = "Stress (graphlayouts)",
        score_layout(carpet.stress, carpet.edges, carpet.n,
                     edge.crossings = "never")),
  cbind(method = "grip default",
        score_layout(carpet.grip.default, carpet.edges, carpet.n,
                     edge.crossings = "never")),
  cbind(method = "grip carpet preset",
        score_layout(carpet.grip.carpet, carpet.edges, carpet.n,
                     edge.crossings = "never"))
)

carpet.display <- carpet.results[, c("method", "sampled.stress",
                                      "edge.length.cv",
                                      "sampled.nonedge.sep.ratio")]
knitr::kable(carpet.display, digits = 4, row.names = FALSE,
             caption = "Level-4 Sierpinski carpet: quality metrics (edge crossings skipped)")
```

### Timing

Timing depends on hardware, so treat these numbers as relative rather than
absolute. All benchmarks were run on the same machine in a single R session.

```{r carpet-timing}
carpet.timing <- data.frame(
  method = c("FR (igraph)", "KK (igraph)", "DrL (igraph)",
             "Stress (graphlayouts)", "grip default", "grip carpet preset"),
  elapsed.sec = c(carpet.time.fr[["elapsed"]],
                  carpet.time.kk[["elapsed"]],
                  carpet.time.drl[["elapsed"]],
                  carpet.time.stress[["elapsed"]],
                  carpet.time.grip.default[["elapsed"]],
                  carpet.time.grip.carpet[["elapsed"]])
)
knitr::kable(carpet.timing, digits = 2, row.names = FALSE,
             caption = "Level-4 Sierpinski carpet: elapsed time (seconds)")
```

### Visual comparison

```{r carpet-plot, fig.width=12, fig.height=7.5}
par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")

plot_layout(carpet.fr,           carpet.edges, "FR (igraph)",           cex = 0.3)
plot_layout(carpet.kk,           carpet.edges, "KK (igraph)",           cex = 0.3)
plot_layout(carpet.drl,          carpet.edges, "DrL (igraph)",          cex = 0.3)
plot_layout(carpet.stress,       carpet.edges, "Stress (graphlayouts)", cex = 0.3)
plot_layout(carpet.grip.default, carpet.edges, "grip default",          cex = 0.3)
plot_layout(carpet.grip.carpet,  carpet.edges, "grip carpet preset",    cex = 0.3)
```

The Sierpinski carpet has a self-similar hole structure that is a strong test of
layout fidelity. grip's carpet preset is specifically tuned for this fractal
family and should preserve the recursive holes more clearly. On a graph of this
size, Kamada-Kawai can be slow (its time complexity is O(n²)), while DrL and
grip's multiscale approach are designed to handle larger graphs efficiently.

## Benchmark 4: HMP microbial network (1,828 vertices)

This real-world co-occurrence network from the Human Microbiome Project has
community labels (CST), making it a good test of cluster separation. It is also
weighted, though we use unweighted versions for igraph/graphlayouts layouts to
keep the comparison straightforward. For grip, we pass the full weighted graph.

```{r hmp-setup}
data(hmp.u01.gc.coarse)

hmp.adj   <- hmp.u01.gc.coarse$adj_list
hmp.wt    <- hmp.u01.gc.coarse$weight_list
hmp.n     <- length(hmp.adj)
hmp.cst   <- hmp.u01.gc.coarse$vertex_data$cst

## Build an edge matrix from the adjacency list
edge.matrix.from.adj <- function(adj.list) {
  edges <- list()
  idx <- 0L
  for (u in seq_along(adj.list)) {
    nbrs <- adj.list[[u]]
    nbrs <- nbrs[nbrs > u]
    if (!length(nbrs)) next
    for (v in nbrs) {
      idx <- idx + 1L
      edges[[idx]] <- c(u, v)
    }
  }
  do.call(rbind, edges)
}

hmp.edges <- edge.matrix.from.adj(hmp.adj)
hmp.m     <- nrow(hmp.edges)

cat(sprintf("Vertices: %d, Edges: %d\n", hmp.n, hmp.m))
```

```{r hmp-igraph}
hmp.ig <- graph_from_edgelist(hmp.edges, directed = FALSE)
```

### Compute layouts with timing

Note: Kamada-Kawai can be very slow on a graph of this size. If it takes more
than a few minutes on your machine, consider skipping it.

```{r hmp-layouts}
set.seed(1)
hmp.time.fr <- system.time(
  hmp.fr <- as_coord_matrix(layout_with_fr(hmp.ig))
)
set.seed(1)
hmp.time.drl <- system.time(
  hmp.drl <- as_coord_matrix(layout_with_drl(hmp.ig))
)
set.seed(1)
hmp.time.stress <- system.time(
  hmp.stress <- as_coord_matrix(layout_with_stress(hmp.ig))
)

hmp.time.grip.default <- system.time(
  hmp.grip.default <- grip.layout(
    adj_list = hmp.adj, weight_list = hmp.wt, n = hmp.n,
    dim = 2, seed = 1
  )
)
```

We skip `layout_with_kk()` for this benchmark because its O(n²) per-iteration
cost makes it impractical for a graph with 1,828 vertices and dense local
connectivity. All other methods complete in a reasonable time.

### Score all layouts

```{r hmp-score}
hmp.results <- rbind(
  cbind(method = "FR (igraph)",
        score_layout(hmp.fr, hmp.edges, hmp.n,
                     clusters = hmp.cst, edge.crossings = "never")),
  cbind(method = "DrL (igraph)",
        score_layout(hmp.drl, hmp.edges, hmp.n,
                     clusters = hmp.cst, edge.crossings = "never")),
  cbind(method = "Stress (graphlayouts)",
        score_layout(hmp.stress, hmp.edges, hmp.n,
                     clusters = hmp.cst, edge.crossings = "never")),
  cbind(method = "grip default",
        score_layout(hmp.grip.default, hmp.edges, hmp.n,
                     clusters = hmp.cst, edge.crossings = "never"))
)

hmp.display <- hmp.results[, c("method", "sampled.stress",
                                "edge.length.cv",
                                "sampled.nonedge.sep.ratio",
                                "cluster.separation")]
knitr::kable(hmp.display, digits = 4, row.names = FALSE,
             caption = "HMP microbial network: quality metrics (edge crossings skipped)")
```

### Timing

```{r hmp-timing}
hmp.timing <- data.frame(
  method = c("FR (igraph)", "DrL (igraph)",
             "Stress (graphlayouts)", "grip default"),
  elapsed.sec = c(hmp.time.fr[["elapsed"]],
                  hmp.time.drl[["elapsed"]],
                  hmp.time.stress[["elapsed"]],
                  hmp.time.grip.default[["elapsed"]])
)
knitr::kable(hmp.timing, digits = 2, row.names = FALSE,
             caption = "HMP microbial network: elapsed time (seconds)")
```

### Visual comparison

```{r hmp-plot, fig.width=10, fig.height=9}
cst.levels <- sort(unique(hmp.cst))
cst.cols   <- setNames(grDevices::hcl.colors(length(cst.levels), "Dark 3"),
                        cst.levels)
hmp.cst.col <- cst.cols[hmp.cst]

par(mfrow = c(2, 2), mar = c(1, 1, 3, 1), bg = "white")

plot_layout(hmp.fr,           hmp.edges, "FR (igraph)",
            vertex.col = hmp.cst.col, cex = 0.4)
plot_layout(hmp.drl,          hmp.edges, "DrL (igraph)",
            vertex.col = hmp.cst.col, cex = 0.4)
plot_layout(hmp.stress,       hmp.edges, "Stress (graphlayouts)",
            vertex.col = hmp.cst.col, cex = 0.4)
plot_layout(hmp.grip.default, hmp.edges, "grip default",
            vertex.col = hmp.cst.col, cex = 0.4)

legend("bottomleft", legend = cst.levels, col = cst.cols, pch = 16,
       bty = "n", cex = 0.7, title = "CST")
```

On a real biological network, cluster separation is the metric of primary
interest. grip accepts the full weighted graph directly, which allows it to
exploit edge-weight information during layout. The other methods are applied to
the unweighted projection here, which may put them at a slight disadvantage on
this particular graph.

## Discussion

Across these four benchmarks, several patterns emerge.

On small, well-connected graphs like the karate club, all methods produce
comparable results. The differences are small enough that the choice of layout
algorithm matters less than the choice of visual encoding. This is expected:
when the graph fits comfortably in memory and has low diameter, most
force-directed and stress-based methods converge to similar solutions.

On structured graphs with a known ideal layout (the mesh, the Sierpinski
carpet), grip's presets tend to produce lower stress and more uniform edge
lengths. This is because the presets encode domain knowledge about the graph
family --- tuned repulsion factors, iteration counts, and neighborhood sizes
that help the multiscale solver find the right geometry. graphlayouts'
`layout_with_stress()` also performs well on the mesh, since stress minimization
is a natural fit for lattice graphs. Fruchterman-Reingold and DrL, which were
designed for different use cases, sometimes introduce distortion on these
regular structures.

On the larger HMP network, scalability differences become apparent.
Kamada-Kawai's O(n²) cost makes it impractical. DrL is fast but can sacrifice
layout quality for speed. graphlayouts' stress minimization strikes a good
balance between quality and performance on moderate-sized graphs. grip's
multiscale approach is designed for this regime and can additionally exploit edge
weights when available.

The key takeaway is not that grip is universally better, but that
`grip.score.layout()` provides a principled way to compare layouts from any
source. Rather than choosing a layout algorithm by eye, you can compute stress,
edge-length uniformity, non-edge separation, and cluster separation for
candidate layouts and make an informed decision. The scoring function works with
any coordinate matrix and edge list --- it does not require that the layout was
produced by grip.

grip is best thought of as a complement to igraph and graphlayouts: install it
alongside your existing graph tools, use its presets when your graph matches a
known family, and use its scoring functions to evaluate and compare layouts
regardless of origin.

## Session info

```{r session-info}
sessionInfo()
```
