---
title: "Interactive Exploration with gripui"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5.2
)

options(rgl.useNULL = TRUE)

gripui_available <- all(vapply(
  c("shiny", "bslib", "DT", "htmltools", "rgl"),
  requireNamespace,
  logical(1L),
  quietly = TRUE
))
```

## Supported interactive workflows

For the first public release, `grip` documents two supported Shiny-based
exploration workflows:

1. the original **project explorer**, launched with `run_gripui()`, for
   browsing realized layout catalogs;
2. the **graph-family geometry explorer**, launched with `run_gripui_family()`,
   for interactively generating and viewing synthetic graph families.

They solve different problems:

- use `run_gripui()` when you already have layout results and want to compare,
  filter, and inspect them;
- use `run_gripui_family()` when you want to browse the package's synthetic
  graph-geometry library and its parameter space.

```{r}
library(grip)
```

## Optional dependencies

These two documented apps rely on packages listed in `Suggests`:

- `shiny`
- `bslib`
- `DT`
- `htmltools`
- `rgl`

If any of these are missing, the app constructors stop with a clear message.

## Part 1: the project explorer

The project explorer is built around a normalized `gripui_project` object. The
usual workflow is:

1. generate layouts,
2. wrap them into a project,
3. launch the app.

### Building a project from `compare.layouts()`

The shortest path is `gripui_project_from_compare()`.

```{r compare-example}
edges <- edges.mesh(5, 5)
n <- 25L

cmp <- compare.layouts(
  edges = edges,
  n = n,
  dim = 2,
  candidates = list(
    default = list(),
    tuned = list(
      preset = "mesh",
      rounds = 48L,
      final_rounds = 96L
    )
  ),
  seeds = 1:2,
  return.layouts = TRUE
)

knitr::kable(
  cmp$summary[, c("candidate", "sampled.stress.mean",
                  "edge.length.cv.mean", "score.composite")],
  digits = 4
)
```

```{r project-from-compare}
adj <- vector("list", n)
for (i in seq_len(nrow(edges))) {
  adj[[edges[i, 1L]]] <- c(adj[[edges[i, 1L]]], edges[i, 2L])
  adj[[edges[i, 2L]]] <- c(adj[[edges[i, 2L]]], edges[i, 1L])
}
graph <- list(adj_list = lapply(adj, function(x) sort(unique(as.integer(x)))))

project <- gripui_project_from_compare(
  cmp,
  graph = graph,
  title = "5x5 mesh comparison"
)

project$meta$title
knitr::kable(project$layouts[, c(
  "layout_id", "candidate", "stage", "seed", "status", "availability"
)])
```

### Building a project by hand

When the layouts come from another source, use `gripui_project()` directly.

```{r manual-project}
layouts_df <- data.frame(
  candidate = c("run_a", "run_b"),
  stage = c("layout", "layout"),
  seed = c(1L, 2L),
  status = c("ok", "ok"),
  my_score = c(0.42, 0.37),
  stringsAsFactors = FALSE
)

manual_project <- gripui_project(
  graph = NULL,
  layouts = layouts_df,
  title = "Manual example"
)

manual_project$meta$title
```

### Validation and directory loading

Two helpers round out the project workflow:

- `gripui_validate_project()` checks structure and consistency,
- `gripui_project_from_dir()` assembles a project from saved artifacts on disk.

```{r validate-project}
gripui_validate_project(project)
```

```r
# Not run: requires a saved output directory
project <- gripui_project_from_dir(
  root = "/path/to/saved/outputs",
  title = "My saved search"
)
run_gripui(project)
```

### Launching the project explorer

`gripui_app()` builds the app object, while `run_gripui()` launches it.

```{r app-objects, eval=gripui_available}
project.app <- gripui_app(project)
inherits(project.app, "shiny.appobj")
```

```r
run_gripui(project)
```

The project explorer is the right tool when you want:

- a catalog table of realized layouts,
- a score landscape,
- a synchronized 2D or 3D viewer,
- and artifact-aware browsing of saved search results.

That applies to both unweighted and weighted layout catalogs once you have
already generated the candidate runs.

## Part 2: the graph-family geometry explorer

The second app is geometry-first. Instead of browsing a catalog of realized
layouts, it lets you interactively generate the synthetic graph families that
support the package's testing and benchmark workflows.

Its main entry points are:

- `gripui_graph_family_catalog()`
- `gripui_family_app()`
- `run_gripui_family()`

### The family catalog

The catalog is a registry of family descriptors used by the geometry explorer.
Each descriptor records the family name, category, presets, parameter controls,
builder call, and source references.

```{r family-catalog}
catalog <- gripui_graph_family_catalog()

length(catalog)
head(names(catalog))
catalog$mesh$function_name
```

### What the family explorer is for

The family explorer is useful when you want to:

- browse the synthetic family collection,
- change family parameters without writing code,
- inspect 2D and 3D embeddings interactively,
- compare several family presets side by side,
- and recover the corresponding R call for a chosen example.

This app is particularly helpful when designing benchmark suites or reviewing
which graph families belong in a paper, test, or demo.

### Launching the family explorer

```{r family-app, eval=gripui_available}
family.app <- gripui_family_app()
inherits(family.app, "shiny.appobj")
```

```r
run_gripui_family()
```

The geometry explorer includes:

- an `Explore` mode for one family at a time,
- a `Compare` mode for side-by-side family or preset panels,
- a dynamic parameter sidebar that adapts to the current family,
- interactive 3D viewing through `rgl`,
- and graph/weight/geometry summaries for the generated example.

## Which app should you use?

Use `run_gripui()` when:

- you already have layout results,
- you want to compare candidates or saved searches,
- you care about score tables and layout catalogs.

Use `run_gripui_family()` when:

- you want to inspect the synthetic family collection,
- you are choosing benchmark families,
- you want to understand or demo the package's geometry generators,
- you need a quick interactive way to recover family-building calls.

## Practical notes

- The apps are optional-package features, so they are best launched in an
  interactive R session rather than inside a scripted report.
- `run_gripui_family()` is self-contained because it builds its own synthetic
  examples from the family catalog.
- `run_gripui()` is best thought of as a viewer for a prepared `gripui_project`.

## Where to go next

- `Getting Started with grip` introduces the main layout APIs.
- `Synthetic Graph Families and Geometries` describes the family collection
  that powers the geometry explorer.
- `Choosing Layouts for Real Data` shows how the project explorer fits into a
  real-data layout-selection workflow.
