---
title: "Comparing grip with Alternative Layout Engines"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  fig.width = 10,
  fig.height = 4.5
)
```

## Why this comparison article was rewritten

`grip` is no longer just an R implementation of the classical GRIP algorithm.
The package now contains:

- a quality-first GRIP workflow for ordinary graphs,
- weighted geometry-aware sister APIs,
- and geodesic scoring and polishing utilities.

That means there are now **two different comparison questions**:

1. how the current `grip` engine compares with other R layout engines on
   ordinary unweighted graphs,
2. how geometry-aware weighted `grip` methods behave on weighted graphs, where
   plain topology is not the whole problem.

This article treats those as separate tracks.

```{r}
library(grip)
```

## Shared scoring idea

`grip.score.layout()` can score a realized layout from any source. That remains
one of the most useful parts of the package, because it lets you compare
candidate layouts even when they come from different packages.

```{r score-helper, eval=FALSE}
score_layout <- function(coords, edges, n,
                         clusters = NULL,
                         edge.crossings = "auto",
                         sample.size.stress = 2000L,
                         sample.size.nonedge = 5000L) {
  grip.score.layout(
    coords = coords,
    edges = edges,
    n = n,
    clusters = clusters,
    sample.size.stress = sample.size.stress,
    sample.size.nonedge = sample.size.nonedge,
    stress.seed = 42L,
    nonedge.seed = 42L,
    edge.crossings = edge.crossings
  )
}
```

For weighted geometry-aware graphs, the package also provides a stronger metric:

- `grip.prepare.geodesic.kk()`
- `grip.score.geodesic.kk()`

Those GKK/LGKK-style tools are advanced public experimental utilities rather
than the default onboarding path, and they are used in the second half of this
article for the weighted comparison track.

```{r}
plot_layout <- function(coords,
                        edges,
                        title,
                        vertex.col = "#1F3B73",
                        edge.col = "gray80",
                        cex = 0.55) {
  if (ncol(coords) == 2L) {
    plot(
      coords[, 1], coords[, 2],
      asp = 1,
      pch = 16,
      cex = cex,
      col = vertex.col,
      xlab = "",
      ylab = "",
      axes = FALSE,
      main = title
    )
    if (!is.null(edges) && nrow(edges) > 0) {
      segments(
        coords[edges[, 1], 1], coords[edges[, 1], 2],
        coords[edges[, 2], 1], coords[edges[, 2], 2],
        col = edge.col
      )
    }
  } else {
    grip.plot(
      coords,
      edges,
      projection = "ortho",
      azimuth = 35,
      elevation = 22,
      vertex.col = vertex.col,
      edge.col = edge.col,
      main = title
    )
  }
}
```

## Track 1: external-package comparisons on unweighted graphs

For this track the comparison remains package-to-package:

- `igraph`
- `graphlayouts`
- `grip`

To keep the article lightweight, the cross-package results are precomputed and
bundled as an RDS file. The reproduction script is still:

- `inst/scripts/precompute-vs-alternatives.R`

```{r load-results}
res.path <- system.file(
  "extdata", "vs_alternatives", "benchmark_results.rds",
  package = "grip"
)
if (!nzchar(res.path)) {
  stop(
    "Precomputed benchmark results not found. ",
    "Run inst/scripts/precompute-vs-alternatives.R first."
  )
}
res <- readRDS(res.path)
```

### Benchmark 1: Zachary karate club

This is still a useful small-graph sanity check. The graph is small enough that
single-scale methods can compete directly.

```{r karate-table}
karate.display <- res$karate$scores[, c(
  "method",
  "sampled.stress",
  "edge.length.cv",
  "edge.crossings",
  "sampled.nonedge.sep.ratio",
  "cluster.separation"
)]

knitr::kable(
  karate.display,
  digits = 4,
  row.names = FALSE,
  caption = "Karate club: unweighted comparison metrics."
)
```

```{r karate-plot, fig.width=12, fig.height=7.5}
k <- res$karate
par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")
plot_layout(k$layouts$fr,            k$edges, "FR (igraph)", cex = 0.6)
plot_layout(k$layouts$kk,            k$edges, "KK (igraph)", cex = 0.6)
plot_layout(k$layouts$drl,           k$edges, "DrL (igraph)", cex = 0.6)
plot_layout(k$layouts$stress,        k$edges, "Stress (graphlayouts)", cex = 0.6)
plot_layout(k$layouts$grip.default,  k$edges, "grip default", cex = 0.6)
plot_layout(k$layouts$grip.tuned,    k$edges, "grip tuned", cex = 0.6)
```

On graphs this small, `grip` does not have a special structural advantage.
That is an honest result and a useful reminder: multiscale methods are not
automatically best on every graph.

### Benchmark 2: 12x12 mesh

Meshes are a good structured topological benchmark because a strong layout
should recover an orderly grid with uniform edge lengths and few or no
crossings.

```{r mesh-table}
mesh.display <- res$mesh$scores[, c(
  "method",
  "sampled.stress",
  "edge.length.cv",
  "edge.crossings",
  "sampled.nonedge.sep.ratio"
)]

knitr::kable(
  mesh.display,
  digits = 4,
  row.names = FALSE,
  caption = "12x12 mesh: unweighted comparison metrics."
)
```

```{r mesh-plot, fig.width=12, fig.height=7.5}
m <- res$mesh
par(mfrow = c(2, 3), mar = c(1, 1, 3, 1), bg = "white")
plot_layout(m$layouts$fr,            m$edges, "FR (igraph)", cex = 0.5)
plot_layout(m$layouts$kk,            m$edges, "KK (igraph)", cex = 0.5)
plot_layout(m$layouts$drl,           m$edges, "DrL (igraph)", cex = 0.5)
plot_layout(m$layouts$stress,        m$edges, "Stress (graphlayouts)", cex = 0.5)
plot_layout(m$layouts$grip.default,  m$edges, "grip default", cex = 0.5)
plot_layout(m$layouts$grip.mesh,     m$edges, "grip mesh preset", cex = 0.5)
```

This benchmark shows a different pattern from the karate graph: on a canonical
structured family like a mesh, the current `grip` default is already robust and
produces a clean uncrossed layout. The `mesh` preset still helps a bit, but it
now acts as a modest family-specific refinement rather than a rescue from a bad
default. At the same time, stress and KK remain very strong on a regular
lattice.

### What the unweighted track tells us

The external-package benchmark is still useful, but its scope is now clear:

- it compares engines on ordinary unweighted graphs,
- it is a fair way to evaluate layout quality with a shared scoring function,
- but it does not answer the geometry-aware weighted-layout question.

That second question needs a different benchmark track.

## Track 2: geometry-aware comparisons on weighted graphs

For weighted graphs with meaningful edge lengths, a pure topology-only
comparison is no longer enough. The weighted track in `grip` is built around:

- `grip.layout.weighted()`
- optional experimental LGKK-based refinement on the legacy weighted path
- and full geodesic-KK scoring.

The small example below is built and scored live in the vignette. It uses a
plain mesh topology whose edge lengths are induced by a curved saddle surface.

```{r weighted-case}
weighted.mesh <- mesh.surface.graph(
  5, 5,
  surface = "saddle",
  amplitude = 0.8
)

coords.combinatorial <- grip.layout(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  dim = 3,
  preset = "mesh",
  seed = 1
)

coords.weighted <- grip.layout.weighted(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  seed = 1
)

coords.weighted.lgkk <- grip.layout.weighted.legacy(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights,
  dim = 3,
  preset = "mesh",
  lgkk_polish_rounds = 6L,
  seed = 1
)

gkk.prepared <- grip.prepare.geodesic.kk(
  weighted.mesh$edges,
  n = weighted.mesh$n,
  edge_weights = weighted.mesh$edge_weights
)

weighted.summary <- do.call(
  rbind,
  list(
    cbind(
      method = "Combinatorial GRIP",
      grip.score.geodesic.kk(
        coords.combinatorial,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rel.rmse",
        "gkk.weighted.rmse",
        "gkk.mean.rel.path.error"
      )]
    ),
    cbind(
      method = "Weighted GRIP",
      grip.score.geodesic.kk(
        coords.weighted,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rel.rmse",
        "gkk.weighted.rmse",
        "gkk.mean.rel.path.error"
      )]
    ),
    cbind(
      method = "Weighted GRIP + LGKK polish",
      grip.score.geodesic.kk(
        coords.weighted.lgkk,
        prepared = gkk.prepared
      )[, c(
        "gkk.weighted.rel.rmse",
        "gkk.weighted.rmse",
        "gkk.mean.rel.path.error"
      )]
    )
  )
)

knitr::kable(
  weighted.summary,
  digits = 4,
  row.names = FALSE,
  caption = "Weighted mesh surface: geodesic-KK comparison metrics."
)
```

```{r weighted-plot, fig.width=11.2, fig.height=8}
par(mfrow = c(2, 2), mar = c(1, 1, 3, 1), bg = "white")
plot_layout(weighted.mesh$coords_surface, weighted.mesh$edges, "Target geometry")
plot_layout(coords.combinatorial, weighted.mesh$edges, "Combinatorial GRIP")
plot_layout(coords.weighted, weighted.mesh$edges, "Weighted GRIP")
plot_layout(coords.weighted.lgkk, weighted.mesh$edges, "Weighted GRIP + LGKK polish")
```

This weighted section is not a package-versus-package benchmark in the old
sense. Instead, it answers the more relevant question for modern `grip`:

- what happens when the graph metric matters?

On this example the progression is exactly what the current package design aims
for:

- combinatorial GRIP gives a topology-respecting layout,
- weighted GRIP improves geodesic fidelity,
- weighted GRIP plus experimental LGKK polish improves it further.

## So what should you compare with what?

For ordinary unweighted graphs:

- compare `grip` with `igraph` and `graphlayouts`,
- use `grip.score.layout()` to evaluate all of them on the same graph,
- and expect the answer to depend on graph family and graph size.

For weighted geometric graphs:

- start with `grip.layout.weighted()`,
- use 3D as the primary layout space when the geometry demands it,
- and evaluate candidate layouts with geodesic-aware criteria such as
  `grip.score.geodesic.kk()`.

That is the main conceptual change in the package since the original version of
this article was written.

## Reproducing the external benchmark bundle

To regenerate the precomputed external-package results:

```{r reproduce, eval=FALSE}
source(system.file("scripts", "precompute-vs-alternatives.R", package = "grip"))
```

The weighted section of this article is generated directly inside the vignette,
so it always reflects the current package implementation.

## Session info for the bundled external results

```{r session-info-precomputed}
res$session_info
```
