---
title: "HMP/U01 Weighted Graph Case Study"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.6
)
```

## What this article is for

The main real-data workflow now lives in:

- `Choosing Layouts for Real Data`

This companion keeps the HMP/U01-specific material that is still useful on its
own:

- the structure of the bundled `hmp.u01.gc.coarse` object,
- the coarsening provenance,
- and the location of the bundled precomputed search artifacts.

```{r}
library(grip)

data(hmp.u01.gc.coarse)
```

## The bundled graph object

`hmp.u01.gc.coarse` is a weighted, coarsened real-world graph derived from the
giant connected component of the HMP+U01 16S amplicon iKNN graph.

```{r}
graph.info <- data.frame(
  quantity = c(
    "source dataset",
    "representation",
    "selected k",
    "original giant-component vertices",
    "coarsened vertices",
    "weighted undirected edges"
  ),
  value = c(
    hmp.u01.gc.coarse$graph_info$source_dataset,
    hmp.u01.gc.coarse$graph_info$representation,
    hmp.u01.gc.coarse$graph_info$selected_k,
    hmp.u01.gc.coarse$graph_info$original_vertices,
    hmp.u01.gc.coarse$graph_info$coarse_vertices,
    hmp.u01.gc.coarse$graph_info$edge_count
  ),
  stringsAsFactors = FALSE
)

knitr::kable(graph.info)
```

The object contains four top-level components:

```{r}
names(hmp.u01.gc.coarse)
```

- `adj_list`
- `weight_list`
- `vertex_data`
- `graph_info`

The `vertex_data` frame includes aggregated biological labels such as `cst` and
`subcst`:

```{r}
names(hmp.u01.gc.coarse$vertex_data)
```

## Coarsening provenance

The coarsening history records how the graph was reduced at each round,
including the number of retained vertices and the matching ratio.

```{r}
knitr::kable(hmp.u01.gc.coarse$graph_info$coarsening_rounds)
```

This provenance is useful when the graph is used as a benchmark or manuscript
example, because it makes the bundled object reproducible as a derived graph
rather than just a static artifact.

## Bundled search artifacts

The package ships precomputed HMP/U01 search results used by the main real-data
vignette. The artifact directory is:

```{r}
system.file("extdata", "hmp_u01_gc_coarse", package = "grip")
```

The bundled results file used in the real-data vignette is:

```{r}
path <- system.file(
  "extdata", "hmp_u01_gc_coarse", "vignette_results.rds",
  package = "grip"
)
path
```

It contains:

```{r}
hmp.results <- readRDS(path)
names(hmp.results)
```

These precomputed artifacts keep the full HMP/U01 workflow available without
making regular vignette builds excessively heavy.

## Where the workflow now lives

For the actual layout-selection workflow on this graph, including:

- preset shortlisting,
- local search,
- metric tables,
- and CST-colored projected layouts,

see:

- `Choosing Layouts for Real Data`

That vignette now contains the HMP/U01 case study in a form that sits directly
alongside the smaller unweighted and weighted real-data examples.
